### 2023-02-02 scripts/0_whole_game.qmd
removed extra files

---
title: "whole_game"
format: html
editor: visual
---

## Setup

```{r setup}
library(tidyverse)
library(terra)
library(sf)
library(bcmaps)
library(sgsR)
library(exactextractr)

#### setup ----
terraOptions(
  memfrac = 0.75,
  tempdir = "E:\\temp",
  todisk = T,
  progress = 100
)
options(scipen = 999)

source(here::here("scripts", "get_keys.R"))
```

Needed things

```{r loads}
vlce <- rast("G:/mosaiced/VLCE2.0/LC_Class_HMM_2015_v20_v20.dat")
bcb <- bcmaps::bc_bound() %>%
  vect()

bcb_rast <- bcb %>%
  rasterize(y = vlce) %>%
  crop(bcb)

vlce <- crop(vlce, bcb_rast) %>%
  mask(bcb_rast)
```

Run the focal cell analysis. Only runs if not done yet via if statement

```{r run-focal}
focal_loc <- "G:/mosaiced/focal"

if (length(list.files(focal_loc)) != 11) {
  print("running focal analysis on vlce and 10 structure variables")
  rmarkdown::render(here::here("scripts", "1_bc_focal_processing.qmd"))
} else {
  print("focal analysis already done")
}
```

Run the sampling across each BEC zone

```{r run-sample}
sample_loc <- here::here("data", "samples")

list.files(sample_loc, pattern = ".shp$")

rmarkdown::render(here::here("scripts", "2_sampling.qmd"))
```

```{r}
print("hello")
```

### 2023-02-02 scripts/mosaic_working.R

@@ -1,769 +1,0 @@
-library(tidyverse)
-library(terra)
-library(bcmaps)
-library(sgsR)
-library(sf)
-library(ggrepel)
-
-#### setup ----
-terraOptions(
-  memfrac = 0.75,
-  tempdir = "E:\\temp",
-  todisk = T,
-  progress = 100
-)
-options(scipen = 999)
-
-source(here::here("scripts", "get_keys.R"))
-merged_loc <- here::here("G://", "mosaiced")
-
-dirs <- list.dirs(merged_loc, recursive = F)
-
-#### data rasters ----
-vlce <- str_subset(dirs, "VLCE2.0") %>%
-  list.files(pattern = ".dat$", full.names = T) %>%
-  rast()
-
-bcb <- bc_bound() %>%
-  vect()
-
-bcb_rast <- bcb %>%
-  rasterize(y = vlce) %>%
-  crop(bcb)
-
-vlce <- crop(vlce, bcb_rast, mask = T)
-
-all_rasts_path <- here::here("data", "rasts", "all_rasts.dat")
-if (!file.exists(all_rasts_path)) {
-  bec_rast <- bec() %>%
-    select(ZONE) %>%
-    vect() %>%
-    rasterize(y = vlce, field = "ZONE") %>%
-    crop(vlce) %>%
-    mask(bcb_rast)
-  
-  # adjust rasters to suit my needs
-  
-  # generate reclassification matrix to select only forests AND retain
-  # their land cover class number
-  vlce_rcl <- keys$vlce %>%
-    mutate(class_name = ifelse(forest == "Forest", class_val, NA)) %>%
-    select(class_val, class_name) %>%
-    as.matrix(ncol = 2)
-  
-  forests <- classify(vlce, vlce_rcl)
-  
-  struct_names <- str_subset(dirs, "structure") %>%
-    list.dirs(full.names = F) %>%
-    .[-1] # removes first index which is the base folder
-  
-  structure <- str_subset(dirs, "structure") %>%
-    list.files(pattern = ".dat$",
-               full.names = T,
-               recursive = T) %>%
-    map(rast)
-  
-  names(structure) <- struct_names
-  
-  structure <- rast(structure)  %>%
-    crop(vlce) %>%
-    mask(bcb_rast)
-  
-  dhi_names <- str_subset(dirs, "DHI") %>%
-    list.files(pattern = ".tif$") %>%
-    str_extract("\\w{3}DHI")
-  
-  dhi <- str_subset(dirs, "DHI") %>%
[…]

### 2023-02-02 scripts/rasterize_bec_zones.R

@@ -1,33 +1,0 @@
-library(tidyverse)
-library(terra)
-library(sf)
-
-terraOptions(memfrac = 0.75,
-             tempdir = "E:\\temp",
-             todisk = T,
-             progress = 100)
-
-vect <- read_sf(here::here("E:\\Sync\\Masters\\analysis\\data\\shapefiles\\bc_ppa_bec_hres.shp")) %>%
-  select(zone, subzone, NAME_E, IUCN_CA, geometry)
-
-
-pa_subzone_rast <- vect %>%
-  mutate(protected = !is.na(IUCN_CA),
-         rasterize = paste0(ifelse(protected, "p", "np"), "-", zone, "_", subzone)) %>%
-  vect() %>%
-  rasterize(y = rast("G:\\Merged\\harvest_mask\\harvest_mask.tif"), field = "rasterize") %>%
-  writeRaster("G:\\Merged\\bec_singles\\pa_subzones.dat",
-              filetype = "envi")
-
-subzone_rast <- vect %>%
-  mutate(rasterize = paste0(zone, "_", subzone)) %>%
-  vect() %>%
-  rasterize(y = rast("G:\\Merged\\harvest_mask\\harvest_mask.tif"), field = "rasterize") %>%
-  writeRaster("G:\\Merged\\bec_singles\\subzones.dat",
-              filetype = "envi")
-
-zone_rast <- vect %>%
-  vect() %>%
-  rasterize(y = rast("G:\\Merged\\harvest_mask\\harvest_mask.tif"), field = "zone")  %>%
-  writeRaster("G:\\Merged\\bec_singles\\zones.dat",
-              filetype = "envi")

### 2023-02-02 scripts/semPowerN.R

@@ -1,8 +1,0 @@
-library(tidyverse)
-library(semPower)
-
-models <-  "cumDHI ~ percentage_first_returns_above_2m + loreys_height + total_biomass
-  percentage_first_returns_above_2m ~ loreys_height
-  total_biomass ~ loreys_height"
-
-n <- semPower(type = "a-priori", effect = 0.05, effect.measure = "RMSEA", alpha = 0.5, power = 0.8, df = semPower.getDf(models))$requiredN

### 2023-02-02 scripts/1_bc_focal_processing.qmd
changed to whole game approach rather than various random scripts. uses python for focal processing. implemented log %>% scale for all variables. prior to log %>% scale, DHI vars are min-max scaled from 0-1.

---
title: "focals"
format: html
editor: visual
---

Make extended tilings

```{r make-grid}
grid_ncol <- 2
grid_nrow <- 2

# make tiles to grab samples from
grid <-
  rast(ncol = grid_ncol,
       nrow = grid_ncol,
       ext = ext(bcb_rast))

values(grid) <- 1:(grid_ncol*grid_nrow)

tiles_focal_loc <- here::here("scratch", "tiles_focal")

dir.create(tiles_focal_loc, showWarnings = F)

focal_files <- list.files(tiles_focal_loc,
                        pattern = "^b.*\\.dat$",
                        full.names = T)

max_tile <- str_extract(focal_files, "[0-9]{1,2}.dat$") %>%
  str_extract("[0-9]{1,2}") %>%
  as.numeric() %>%
  max()

if (max_tile != (grid_ncol * grid_nrow)) {
  file.remove(list.files(focal_files, full.names = T))

  focal_files <- terra::makeTiles(
    x = bcb_rast,
    y = grid,
    here::here(tiles_focal_loc, "tile_.dat"),
    na.rm = TRUE,
    overwrite = TRUE,
    filetype = "ENVI",
    datatype = "INT1U"
  )
}

extended_focal <- focal_files %>%
  map(rast) %>%
  map(extend, y = 1)
```

```{r forest-rcl}
forest_rcl <- keys$vlce %>%
  mutate(class_name = ifelse(forest == "Forest", class_val, NA)) %>%
  select(class_val, class_name) %>%
  as.matrix(ncol = 2)

forests <- classify(vlce, forest_rcl)
```

Function to allow mapping on the inverted variable order

```{r function}
rev_crop <- function(x, y) {
  crop(y, x)
}
```

Extended tiles (vlce)

```{r tile-extend}
forests_focal_tile_loc <- here::here(tiles_focal_loc, "forests", "input")
dir.create(forests_focal_tile_loc, showWarnings = F, recursive = T)

forest_save_names <- here::here(forests_focal_tile_loc,
                                paste0("focal-forests-", 1:length(extended_focal), ".dat"))

forests_extend <- extended_focal %>%
  map(rev_crop, y = forests) %>%
  map2(.x = ., .y = forest_save_names, .f = writeRaster,
       filetype = "ENVI", overwrite = T)
```

Make clean structure directories

```{r dir-creation}
structure_focal_tile_loc <- here::here(tiles_focal_loc, "structure", "input")
dir.create(structure_focal_tile_loc, showWarnings = F, recursive = T)

struct_loc <- here::here("G://", "mosaiced", "structure")

struct_vars <- list.dirs(struct_loc, full.names = F) %>%
  stringi::stri_remove_empty()

structure_subdirs <- here::here(structure_focal_tile_loc, struct_vars)

map(structure_subdirs, dir.create, showWarnings = F)
```

Make extended tiles for each structure variable

```{r structure-extend}
for(var in struct_vars) {
  print(var)

  r <- here::here(struct_loc, var, paste0(var, "_2015.dat")) %>%
  rast() %>%
  crop(bcb_rast) %>%
  mask(bcb_rast)

struct_save_names <- here::here(structure_focal_tile_loc,
                                var,
                                paste0("focal-", var, "-", 1:length(extended_focal), ".dat"))

struct_extend <- extended_focal %>%
  map(rev_crop, y = r) %>%
  map2(.x = ., .y = struct_save_names, .f = writeRaster,
       filetype = "ENVI", overwrite = T
  )
}
```

```{r reticulate}
# this does the focal cell processing for all of BC based on the tiles 
# created earlier in this script
library(reticulate)
use_python("C:/Users/evanmuis/Miniconda3/envs/geospatial/python.exe")

source_python(here::here("scripts", "1a_focal_tiles.py"))

rm(stack)
rm(nptile)
rm(open_tile)
rm(valid_locs)
```

```{r mosaic}
# mosaics the focal tiles together

folders <- "E:/Sync/Masters/analysis_sem/scratch/tiles_focal" %>%
  list.dirs() %>%
  str_subset("output")

for_fold <- str_subset(folders, "forests")

struct_fold <- folders %>%
  str_subset("structure") %>%
  str_subset("output$", negate = T)

iter_folds = c(for_fold, struct_fold)

focal_save <- here::here("G:/", "mosaiced", "focal")

# only do the 3 currently used

# iter_folds <- c(iter_folds[1], iter_folds[8:9])

mosaic_rcl <- matrix(c(0, 1, NA, 1), ncol = 2)

for (folder in iter_folds) {
  print(folder)
  var = folder %>%
    list.files() %>%
    .[1] %>% # first file
    str_split("-") %>%
    .[[1]] %>% # unlist
    .[2] # second item (variable)

  # lol what the fuck is this evan nobody can read it
  # take first file w/o full names (since naming conventions the same)
  # make it not a list
  # take the second item in that vector (should always be the variable)

  save_name = here::here(focal_save, paste0("focal-", var, ".tif"))
  print(save_name)

  mosaiced <- list.files(folder, full.names = T, pattern = ".dat$") %>%
    map(rast) %>%
    map(trim, padding = 0, value = 0) %>%
    sprc() %>%
    mosaic(fun = "min") %>%
    extend(bcb_rast)

  mosaiced_rc <- classify(mosaiced, mosaic_rcl)

  #mosaiced <- mosaiced == 1

  # if (var != "forests") {
  #   save_name = here::here(focal_save, "structure", var, paste0("focal-", var, ".tif"))
  # }

  plot(mosaiced_rc, col = "green")

  dir.create(dirname(save_name), showWarnings = F)

  writeRaster(mosaiced_rc, save_name, overwrite = T)

}
```

### 2023-02-02 scripts/3_sem_visualization.qmd
changed to whole game approach rather than various random scripts. uses python for focal processing. implemented log %>% scale for all variables. prior to log %>% scale, DHI vars are min-max scaled from 0-1.

---
title: "sem"
format: html
editor: visual
---

```{r}
library(lavaan)

theme_set(theme_light())
```

extract datasets

```{r}
samples <- here::here("data", "samples", "bec") %>%
  list.files(full.names = T, pattern = ".shp$") %>%
  map(read_sf) %>%
  bind_rows()

samples %>%
  st_drop_geometry() %>%
  count(zone) %>%
  ggplot(aes(
    x = n,
    y = fct_reorder(zone, n),
    fill = zone
  )) +
  geom_col() +
  ggsci::scale_fill_igv() +
  theme_bw() +
  labs(x = "n",
       y = NULL,
       fill = NULL)
```

```{r}
all_rasts_loc <- here::here("data", "rasts", "all_rasts.dat")

if (!file.exists(all_rasts_loc)) {
  struct_locs <- here::here("G://", "mosaiced", "structure")
  struct_rasts <- list.files(
    struct_locs,
    recursive = T,
    pattern = ".dat$",
    full.names = T
  ) %>%
    map(rast) %>%
    rast()

  struct_names <- str_split(sources(struct_rasts), pattern = "/") %>%
    lapply("[[", 4) %>% # get fourth index from names
    unlist()

  names(struct_rasts) <- struct_names

  struct_rasts <- struct_rasts %>%
    crop(bcb_rast, mask = T)

  dhi_rasts <-
    list.files(here::here("G://", "mosaiced", "DHI"),
               pattern = "2010s.tif$",
               full.names = T) %>%
    map(rast) %>%
    rast()

  dhi_names <- str_split(sources(dhi_rasts), pattern = "/") %>%
    lapply("[[", 4) %>% # get fourth index from names
    unlist() %>%
    str_split(pattern = "[[:punct:]]") %>%
    lapply("[[", 2) %>% # get second index from split names
    unlist()

  names(dhi_rasts) <- dhi_names

  dhi_rasts <- dhi_rasts %>%
    crop(bcb_rast, mask = T)

  dhi_mm <- minmax(dhi_rasts)

  dhi_rasts_mm <- (dhi_rasts - dhi_mm[1,]) / (dhi_mm[2,] - dhi_mm[1,])

  all_rasts <- c(vlce, struct_rasts, dhi_rasts_mm)

  writeRaster(all_rasts, all_rasts_loc, overwrite = T, filetype = "envi",
              datatype = "FLT4S")
}

all_rasts <- rast(all_rasts_loc)
```

```{r}
extracted_samples <- extract(all_rasts, samples, bind = T)
```

)

```{r}
invalid_zones <- c("CMA", "IMA", "BAFA", "BG")
# alpine (CMA, IMA, BAFA) & non-forest dominant (BG)

extracted_df <- extracted_samples %>% 
  st_as_sf() %>%
  st_drop_geometry() %>%
  filter(!(loreys_height < elev_mean | elev_p95 < elev_mean)) %>%
  pivot_longer(basal_area:total_biomass, names_to = "variable") %>%
  left_join(keys$continuous) %>%
  mutate(value = value / divide_by) %>%
  select(zone, category, ends_with("DHI"), variable, value) %>%
  pivot_wider(names_from = "variable", values_from = "value") %>%
  unchop(everything()) %>% # somehow does the fix i want
  filter(elev_cv <= 1) %>% # remove erroneous values
  drop_na() # remove anything w/ NA

# groups by zone and all other structure variables 
# takes the mean of each DHI dataset
# effectively removing duplicates, but takes the mean instead of the first value
extracted_df_b <- extracted_df %>%
  group_by(across(c(zone, category, basal_area:total_biomass))) %>%
  summarize(across(ends_with("DHI"), mean)) %>%
  ungroup()
```

```{r}
# downsampling procedure - allows for comparisons between models if sample size is the same (AIC and BIC are based on sample)

downsample_df <- extracted_df_b %>%
  count(zone) %>%
  filter(!(zone %in% invalid_zones))

downsample_no <- downsample_df %>%
  pull(n) %>%
  min()
```

```{r}
set.seed(69420)

clean_df <- extracted_df_b %>%
  mutate(
    overstory_per = percentage_first_returns_above_mean,
    understory_per = percentage_first_returns_above_2m - percentage_first_returns_above_mean,
    total_cover = percentage_first_returns_above_2m,
    overstory_range = loreys_height - elev_mean,
    understory_range = elev_mean - 2
  ) %>%
  relocate(zone,
           category,
           ends_with("DHI")) %>%
  filter(!(zone %in% invalid_zones)) %>%
  group_by(zone) %>%
  slice_sample(n = downsample_no) %>%
  select(!starts_with("perc"))
```

```{r}
mylog <- function(data) {
  if (any(data <= 0)) {
    return(log1p(data))
  } else {
    return(log(data))
  }
}

normal <- function(data) {
  data %>%
    mylog() %>%
    scale() %>% 
    as.numeric()
}

log_normal_df <- clean_df %>%
  group_by(zone) %>%
  mutate(across(CumDHI:understory_range, .fns = ~ normal(.x)))

log_normal_df %>%
  summarize(across(CumDHI:understory_range, .fns = ~ shapiro.test(.x)$p.value))
```

```{r}
log_splits <- log_normal_df %>%
  group_by(zone) %>%
  group_split()
```

sample each bec zone down to the lowest number of samples

```{r}
set.seed(69420)

zone_dfs <- extracted_df %>%
  filter(!(zone %in% invalid_zones)) %>%
  group_by(zone) %>%
  group_split() %>%
  map(slice_sample, n = downsample_no)
```

```{r}
library(Hmisc)
library(corrplot)
corr <- log_normal_df %>% 
  ungroup() %>%
  select(CumDHI:total_biomass) %>%
  as.matrix() %>%
  rcorr()

corrplot(corr$r,
         type = "upper",
         tl.col = "black",
         tl.srt = 45,
         p.mat = corr$P, 
         sig.level = 0.01)
```

```{r}
sample_names <- names(extracted_samples)

dhi_vars <-
  sample_names[(length(sample_names) - 2):length(sample_names)]

v_models <-
  paste0(
    dhi_vars,
    " ~ loreys_height + total_biomass + elev_stddev
elev_stddev ~ loreys_height
total_biomass ~ loreys_height"
  )

h_models <-
  paste0(
    dhi_vars,
    " ~ basal_area + percentage_first_returns_above_2m + percentage_first_returns_above_mean
percentage_first_returns_above_mean ~ basal_area
percentage_first_returns_above_2m ~ basal_area"
  )

models <- c(v_models, h_models)
```

```{r some-weird-map-lol}
bec_zones <- bec() %>%
  mutate(area = st_area(.)) %>%
  group_by(ZONE) %>%
  dplyr::summarize(area = sum(area)) %>%
  st_intersection(bc_bound())

bec_counts_sf <- bec_zones %>%
  left_join(bec_counts, by = c("ZONE" = "zone")) %>%
  select(ZONE, n) %>%
  mutate(ZONE = ifelse(!(ZONE %in% invalid_zones), ZONE, "Not Included"))

ggplot(bec_counts_sf) +
  geom_sf(aes(fill = ZONE), colour = "#00000000") +
  ggsci::scale_fill_igv() +
  scale_alpha_manual(values = c(0.5, 1)) +
  theme_void()
```

```{r zonal-models}
sem_extract <- function(split_df, model) {
  zone <- split_df %>%
    pull(zone) %>%
    unique()

  var <- str_split(model, pattern = " ")[[1]][1]
  fit <- sem(model = model, data = split_df)

  fits <-
    fitmeasures(fit, c("chisq", "df", "pvalue", "cfi", "rmsea", "aic", "r2", "srmr"))
  pars <- parameterestimates(fit) %>%
    tibble() %>%
    mutate(zone = zone, var = var, model = model)

  fits <- tibble(fit = names(fits), value = as.numeric(fits)) %>%
    pivot_wider(names_from = fit, values_from = value) %>%
    mutate(zone = zone,
           var = var, model = model)

  interpreted <- fit %>%
    effectsize::interpret() %>%
    mutate(zone = zone, var = var, model = model)

  return(list(
    # fitmeasures = fits,
    parameters = pars,
    interpreted = interpreted,
    fits = fits
  ))
}

crossed <- crossing(zone_dfs, models)

dfs <- crossed %>% pull(zone_dfs)
mods <- crossed %>% pull(models)

outs <- map2(.x = dfs, .y = mods, .f = sem_extract)

params <- map_df(outs, function(x) {
    x$parameters
  })

# fitmeasures <- map_df(outs, function(x) {
#   x$fitmeasures
# })

fitmeasures <- map_df(outs, function(x) {
  x$interpreted
})

fits <- map_df(outs, function(x) {
  x$fits
})
```

```{r}
vlce_samples <- here::here("data", "samples", "vlce") %>%
  list.files(pattern = ".shp", full.names = T) %>%
  map(read_sf) %>%
  bind_rows()

extracted_vlce <- extract(all_rasts, vlce_samples)
```

```{r}
extracted_vlce %>%
  st_drop_geometry() %>%
  count(category)
```

```{r}
fitmeasures %>% 
  group_by(zone) %>%
  dplyr::summarize(across(cfi:aic, mean))

fitmeasures %>%
  mutate(v_h = ifelse(model %in% h_models, "h", "v")) %>%
  group_by(model) %>%
  dplyr::summarize(across(c(cfi:aic, chisq), mean))
```

```{r}
params %>%
  filter(op == "~") %>%
  group_by(model, lhs, rhs) %>%
  dplyr::summarize(per_sig = (pvalue > 0.05) / n()) %>%
  view()
```

```{r}
test_mod <- "CumDHI ~ loreys_height + percentage_first_returns_above_2m + percentage_first_returns_above_mean
percentage_first_returns_above_2m ~ total_biomass
percentage_first_returns_above_mean ~ percentage_first_returns_above_2m + total_biomass
total_biomass ~ loreys_height"

data <- dfs[[19]] %>%
  mutate(across(basal_area:total_biomass, scale),
         across(basal_area:total_biomass, as.numeric))

(egg <- sem(test_mod, data = data))

varTable(egg)
fitmeasures(egg)
parameterestimates(egg)

graph_sem(egg)
```

```{r}
vlce_samples <- here::here("data", "samples", "vlce") %>%
  list.files(full.names = T, pattern = ".shp$") %>%
  map(read_sf) %>%
  bind_rows()

vlce_samples
```

### 2023-02-02 scripts/1a_focal_tiles.py

@@ -1,0 +1,81 @@
+import xarray as xr
+import rioxarray
+import os
+import time
+import itertools
+import numpy as np
+
+# goal of this script is to run focal analyses to speed up my time in sampling so i can iterate quicker and run more models with various restrictions on the sampling regime
+
+tiles_focal = "E:/Sync/Masters/analysis_sem/scratch/tiles_focal"
+
+forest_in = os.path.join(tiles_focal, "forests", "input")
+
+if not os.path.isdir(forest_in.replace("input", "output")):
+    os.mkdir(forest_in.replace("input", "output"))
+
+forest_tiles = [os.path.join(forest_in, x) for x in os.listdir(forest_in) if x.endswith(".dat")]
+
+for tile in forest_tiles:
+    print(tile)
+    open_tile = xr.open_dataset(tile, engine = "rasterio").to_array()
+
+    nptile = open_tile.data[0, 0, ...]
+
+    stack = np.empty((nptile.shape[0]+2, nptile.shape[1]+2, 9))
+
+    for index, (xshift, yshift) in enumerate(itertools.product([-1, 0, 1], [-1, 0, 1])):
+        # print(xshift, yshift)
+        stack[xshift+1:nptile.shape[0]+xshift+1, yshift+1:nptile.shape[1]+yshift+1, index] = nptile
+
+    valid_locs = np.ptp(stack, axis=-1) == 0
+    np.count_nonzero(valid_locs)
+
+    valid_locs.shape
+
+    open_tile.data[0, 0, ...] = valid_locs[1:-1, 1:-1]
+
+    savename = tile.replace("input", "output")
+    
+    open_tile.squeeze().rio.to_raster(savename, driver = "envi")
+
+struct_in = os.path.join(tiles_focal, "structure", "input")
+
+if not os.path.isdir(struct_in.replace("input", "output")):
+    os.mkdir(struct_in.replace("input", "output"))
+
+struct_folds = os.listdir(struct_in)
+
+for fold in struct_folds:
+    struct_tiles = [os.path.join(struct_in, fold, x) for x in os.listdir(os.path.join(struct_in, fold)) if x.endswith(".dat")]
+
+    for tile in struct_tiles:
+        print(tile)
+        open_tile = xr.open_dataset(tile, engine = "rasterio").to_array()
+
+        nptile = open_tile.data[0, 0, ...]
+
[…]

### 2023-02-02 scripts/manuscript/manuscript.qmd
added manuscript subfolder

---
title: Disentangling linkages between satellite derived forest structure and productivity across a range of forest types and ecosystems.
author:
  - name: Evan R. Muise
    email: evan.muise@student.ubc.ca
    affiliations: 
        - id: ubc
          name: University of British Columbia
          department: Forest Resources Management
          address: 2424 Main Mall
          city: Vancouver, BC, Canada
          postal-code: V6T 1Z4
    attributes:
        corresponding: true
  - name: Nicholas C. Coops
    email: nicholas.coops@ubc.ca
    affiliations: 
        - id: ubc
          name: University of British Columbia
          department: Forest Resources Management
          address: 2424 Main Mall
          city: Vancouver, BC, Canada
          postal-code: V6T 1Z4
  - name: Txomin Hermosilla
    email: txomin.hermosillagomez@NRCan-RNCan.gc.ca
    affiliations: 
        - id: cfs
          name: Natural Resources Canada
          department: Canada Forest Service (Pacific Forestry Centre)
          address: 506 Burnside Rd W
          city: Victoria, BC, Canada
          postal-code: V8Z 1M5
  - name: A. Cole Burton
    email: cole.burton@ubc.ca
    affiliations: 
        - id: ubc
          name: University of British Columbia
          department: Forest Resources Management
          address: 2424 Main Mall
          city: Vancouver, BC, Canada
          postal-code: V6T 1Z4
  - name: Margaret E. Andrew
    email: M.Andrew@murdoch.edu.au
    affiliations: 
        - id: murd
          name: Murdoch University
          department: Environmental and Conservation Sciences and Harry Butler Institute
          city: Murdoch, WA, Australia
          postal-code: 6150
  - name: Stephen S. Ban
    email: Stephen.Ban@gov.bc.ca
    affiliations: 
        - id: bcp
          name: Ministry of Environment and Climate Change Strategy
          department: BC Parks
          address: 525 Superior Street
          city: Victoria, BC, Canada
          postal-code: V8V 1T7
abstract: |
  insert abstract here
keywords: 
  - structural equation modelling
  - remote sensing
  - landsat
  - forest structure
  - forest productivity
date: last-modified
bibliography: bibliography.bib
csl: remote-sensing-of-environment.csl
format:
  elsevier-pdf:
    keep-tex: true
    keep-md: true
    journal:
      name: Remote Sensing of Environment
      formatting: review
      model: 3p
      cite-style: authoryear
      layout: twocolumn
  docx: 
    reference-doc: my-styles.docx
    keep-md: true
  pdf: 
    pdf-engine: xelatex
editor: 
  markdown: 
    wrap: 72
---

# Introduction

Biodiversity is the summation of variation in biological life, across
genes, species, communities, and ecosystems. Currently, biodiversity is
in decline, facing extinction rates above the background extinction rate
[@thomas2004; @urban2015], and homogenization of communities at various
scales [@mcgill2015]. In response, the global biodiversity community is
making efforts to assess and halt the degradation of biodiversity on
earth. The Group for Earth Observation Biodiversity Observation Network
has developed the Essential Biodiversity Variables [EBVs, @pereira2013],
designed as an analog to the Essential Climate Variables framework
[@bojinski2014]. EBVs are designed to be global in scope, relevant to
biodiversity information, feasible to utilize, and complementary to one
another [@skidmore2021].

There are six EBV classes, each corresponding to a different facet of
biodiversity, including, genetic composition, species populations,
species traits, community composition, ecosystem structure, and
ecosystem function [@pereira2013]. @fernández2020 divides the six
classes into two approaches, with one focusing on species biodiversity,
and the other focusing on ecosystem diversity. Remote sensing has proven
to be capable of measure five of the six classes, with the exception
being genetic composition, which requires in-situ observation and
samples [@skidmore2021]. Notably, while remote sensing can provide
information on the remaining two species EBV classes, this information
is typically acquired with an *ad-hoc* approach. This collects high
resolution, small spatial extent information, rather than the global or
regional scales required for biodiversity assessment. Community
composition falls into a similar dilemma, requiring species population
information which necessitates high resolution spatial data.

The remaining two classes, ecosystem structure and function, are
incredibly well suited to be examined at global or regional scales using
mid-resolution satellite imagery, such as that provided by the Landsat
series of satellites. Advances in satellite remote sensing processing
have allowed 3d forest structure data to be imputed across wide spatial
scales [@matasci2018; @coops2021] using data fusion approaches involving
collected lidar data and optical/radar data. Other advances in image
compositing have allowed yearly summaries of vegetation productivity to
be calculated at regional to global scales, summarizing the yearly
energy totals, minimums, and variations [@radeloff2019]. These datasets
correspond quite well with the EBV classes ecosystem structure (forest
structural diversity metrics), and ecosystem function (forest
productivity metrics).

Forest structural diversity has been linked to biodiversity at various
scales [@guo2017; @bergen2009; @gao2014]. Increased structural
complexity is hypothesized to create additional niches, leading to
increased species diversity [@bergen2009]. The relationship between
forest structure and biodiversity is commonly assessed using avian
species diversity metrics [@macarthur1961; @goetz2007], however, other
clades (and habitats), have been used [@davies2014; @nelson2005]. Many
metrics derived from lidar remote sensing have been used as local
indicators of biodiversity, including canopy cover, canopy height,
vertical profiles, and aboveground biomass, while other 2nd order
derived metrics such as canopy texture, height class distribution,
edges, and patch metrics have been used to examine habitat and
biodiversity at landscape scales [@bergen2009].

The dynamic habitat indices (DHIs) are indicators of productivity
calculated by summarizing vegetation indices over the course of one (or
multiple) years [@radeloff2019]. These indices have been related to
multiple facets of biodiversity at a range of scales, including species
occurrence and abundance [@razenkova2020], alpha [@radeloff2019] and
beta diversity [@andrew2012]. Hypotheses behind the biodiversity
productivity relationships have been established, including the
species-energy hypothesis, the environmental stress hypothesis, and the
environmental stability hypothesis [@coops2019]. The cumulative DHI
calculates the total amount of energy available in a given pixel over
the course of a year. Cumulative DHI is strongly linked to the available
energy hypothesis, which suggests that with greater available energy
species richness will increase [@wright1983]. The minimum DHI, which
calculates the lowest productivity over the course of a year can be
matched to the environmental stress hypothesis, which proposes that
higher levels of minimum available energy will lead to higher species
richness [@currie2004]. Finally, the variation DHI, which calculates the
coefficient of variance in a vegetation index through the course of a
year, corresponds to the environmental stability hypothesis which states
that lower energy variation throughout a year will lead to increased
species richness [@williams2008].

Linkages between forest structure and productivity (namely, vegetation
indices) have been examined for nearly 20 years [@huete2002,
@knyazikhin1998; @myneni1994]. While there is significant theoretical
and empirical evidence for their relationship at single time points
(within a single image) [@myneni1994], various relationship directions
and shapes have been found between forest structure and productivity
metrics [@ali2019]. These relationships, their shapes, and their
strengths have been attributed to multiple possible hypotheses and can
vary based on environmental conditions [@ali2019]. The relationship
between forest structural diversity metrics and annual productivity
summaries has yet to be fully examined, including the DHIs.

Some vegetation structure metrics are simpler, and more accurate, to
calculate than others [@coops2021]. These basic metrics, such as canopy
height and canopy cover,can then be used to estimate additional
structure metrics, such as basal area or total biomass. Canopy height
and cover are commonly used as an indicator of vertical and horizontal
variation, respectively. Recently, more attention has been paid to
internal structural complexity metrics, which can be more difficult and
time consuming to generate [@coops2021; @ma2022] .

In this study, we seek to untangle the relationship between two EBVs:
forest structure diversity metrics and yearly summaries of forest
productivity. To accomplish this, we assess this relationship using path
analysis to assess the direct and indirect (as mediated by more complex
forest structural diversity metrics) effects of commonly collected
forest structural metrics (canopy height and canopy cover) on yearly
productivity summaries. Doing so will in turn assess their utility as
complementary EBVs. We ran this analysis separately across four forested
land covers and the forested ecosystems of British Columbia, Canada.

# Methods

## Study Area

British Columbia is the westernmost province of Canada, and is home to a
variety of terrestrial ecosystems [@pojar1987]. Approximately 64% of the
province is forested [@bcministryofforests2003]. There is a large amount
of ecosystem variation in the province, with large climate and
topographic gradients. The Biogeoclimatic Ecosystem Classification (BEC)
system identifies 16 zones based on the dominant tree species and the
ecosystems general climate. These zones can be further split into
subzones, variants, and phases based on microclimate, precipitation, and
topography.

## Data

## Sampling

## Analysis

# Results

# Discussion

# Conclusion

\newpage

# References {.unnumbered}

### 2023-02-02 scripts/manuscript/manuscript.tex
added manuscript subfolder

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  authoryear,
  review,
  3p,
  twocolumn]{elsarticle}

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\usepackage{float}
\makeatletter
\let\oldlt\longtable
\let\endoldlt\endlongtable
\def\longtable{\@ifnextchar[\longtable@i \longtable@ii}
\def\longtable@i[#1]{\begin{figure}[H]
\onecolumn
\begin{minipage}{0.5\textwidth}
\oldlt[#1]
}
\def\longtable@ii{\begin{figure}[H]
\onecolumn
\begin{minipage}{0.5\textwidth}
\oldlt
}
\def\endlongtable{\endoldlt
\end{minipage}
\twocolumn
\end{figure}}
\makeatother
\journal{Remote Sensing of Environment}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Disentangling linkages between satellite derived forest structure and productivity across a range of forest types and ecosystems.},
  pdfauthor={Evan R. Muise; Nicholas C. Coops; Txomin Hermosilla; A. Cole Burton; Margaret E. Andrew; Stephen S. Ban},
  pdfkeywords={structural equation modelling, remote
sensing, landsat, forest structure, forest productivity},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\setlength{\parindent}{6pt}
\begin{document}

\begin{frontmatter}
\title{Disentangling linkages between satellite derived forest structure
and productivity across a range of forest types and ecosystems.}
\author[1]{Evan R. Muise%
\corref{cor1}%
}
 \ead{evan.muise@student.ubc.ca} 
\author[1]{Nicholas C. Coops%
%
}
 \ead{nicholas.coops@ubc.ca} 
\author[2]{Txomin Hermosilla%
%
}
 \ead{txomin.hermosillagomez@NRCan-RNCan.gc.ca} 
\author[1]{A. Cole Burton%
%
}
 \ead{cole.burton@ubc.ca} 
\author[3]{Margaret E. Andrew%
%
}
 \ead{M.Andrew@murdoch.edu.au} 
\author[4]{Stephen S. Ban%
%
}
 \ead{Stephen.Ban@gov.bc.ca}

\affiliation[1]{organization={University of British Columbia, Forest
Resources Management},addressline={2424 Main Mall},city={Vancouver, BC,
Canada},postcode={V6T 1Z4},postcodesep={}}
\affiliation[2]{organization={Natural Resources Canada, Canada Forest
Service (Pacific Forestry Centre)},addressline={506 Burnside Rd
W},city={Victoria, BC, Canada},postcode={V8Z 1M5},postcodesep={}}
\affiliation[3]{organization={Murdoch University, Environmental and
Conservation Sciences and Harry Butler Institute},city={Murdoch, WA,
Australia},postcode={6150},postcodesep={}}
\affiliation[4]{organization={Ministry of Environment and Climate Change
Strategy, BC Parks},addressline={525 Superior Street},city={Victoria,
BC, Canada},postcode={V8V 1T7},postcodesep={}}

\cortext[cor1]{Corresponding author}

\begin{abstract}
This is the abstract. Lorem ipsum dolor sit amet, consectetur adipiscing
elit. Vestibulum augue turpis, dictum non malesuada a, volutpat eget
velit. Nam placerat turpis purus, eu tristique ex tincidunt et. Mauris
sed augue eget turpis ultrices tincidunt. Sed et mi in leo porta
egestas. Aliquam non laoreet velit. Nunc quis ex vitae eros aliquet
auctor nec ac libero. Duis laoreet sapien eu mi luctus, in bibendum leo
molestie. Sed hendrerit diam diam, ac dapibus nisl volutpat vitae.
Aliquam bibendum varius libero, eu efficitur justo rutrum at. Sed at
tempus elit.
\end{abstract}

\begin{keyword}
    structural equation modelling \sep remote
sensing \sep landsat \sep forest structure \sep 
    forest productivity
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, sharp corners, interior hidden, frame hidden, enhanced, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Biodiversity is the summation of variation in biological life, across
genes, species, communities, and ecosystems. Currently, biodiversity is
in decline, facing extinction rates above the background extinction rate
\citep{thomas2004, urban2015}, and homogenization of communities at
various scales \citep{mcgill2015}. In response, the global biodiversity
community is making efforts to assess and halt the degradation of
biodiversity on earth. The Group for Earth Observation Biodiversity
Observation Network has developed the Essential Biodiversity Variables
\citep[EBVs,][]{pereira2013}, designed as an analog to the Essential
Climate Variables framework \citep{bojinski2014}. EBVs are designed to
be global in scope, relevant to biodiversity information, feasible to
utilize, and complementary to one another \citep{skidmore2021}.

There are six EBV classes, each corresponding to a different facet of
biodiversity, including, genetic composition, species populations,
species traits, community composition, ecosystem structure, and
ecosystem function \citep{pereira2013}. \citet{fernández2020} divides
the six classes into two approaches, with one focusing on species
biodiversity, and the other focusing on ecosystem diversity. Remote
sensing has proven to be capable of measure five of the six classes,
with the exception being genetic composition, which requires in-situ
observation and samples \citep{skidmore2021}. Notably, while remote
sensing can provide information on the remaining two species EBV
classes, this information is typically acquired with an \emph{ad-hoc}
approach. This collects high resolution, small spatial extent
information, rather than the global or regional scales required for
biodiversity assessment. Community composition falls into a similar
dilemma, requiring species population information which necessitates
high resolution spatial data.

The remaining two classes, ecosystem structure and function, are
incredibly well suited to be examined at global or regional scales using
mid-resolution satellite imagery, such as that provided by the Landsat
series of satellites. Advances in satellite remote sensing processing
have allowed 3d forest structure data to be imputed across wide spatial
scales \citep{matasci2018, coops2021} using data fusion approaches
involving collected lidar data and optical/radar data. Other advances in
image compositing have allowed yearly summaries of vegetation
productivity to be calculated at regional to global scales, summarizing
the yearly energy totals, minimums, and variations \citep{radeloff2019}.
These datasets correspond quite well with the EBV classes ecosystem
structure (forest structural diversity metrics), and ecosystem function
(forest productivity metrics).

Forest structural diversity has been linked to biodiversity at various
scales \citep{guo2017, bergen2009, gao2014}. Increased structural
complexity is hypothesized to create additional niches, leading to
increased species diversity \citep{bergen2009}. The relationship between
forest structure and biodiversity is commonly assessed using avian
species diversity metrics \citep{macarthur1961, goetz2007}, however,
other clades (and habitats), have been used
\citep{davies2014, nelson2005}. Many metrics derived from lidar remote
sensing have been used as local indicators of biodiversity, including
canopy cover, canopy height, vertical profiles, and aboveground biomass,
while other 2nd order derived metrics such as canopy texture, height
class distribution, edges, and patch metrics have been used to examine
habitat and biodiversity at landscape scales \citep{bergen2009}.

The dynamic habitat indices (DHIs) are indicators of productivity
calculated by summarizing vegetation indices over the course of one (or
multiple) years \citep{radeloff2019}. These indices have been related to
multiple facets of biodiversity at a range of scales, including species
occurrence and abundance \citep{razenkova2020}, alpha
\citep{radeloff2019} and beta diversity \citep{andrew2012}. Hypotheses
behind the biodiversity productivity relationships have been
established, including the species-energy hypothesis, the environmental
stress hypothesis, and the environmental stability hypothesis
\citep{coops2019}. The cumulative DHI calculates the total amount of
energy available in a given pixel over the course of a year. Cumulative
DHI is strongly linked to the available energy hypothesis, which
suggests that with greater available energy species richness will
increase \citep{wright1983}. The minimum DHI, which calculates the
lowest productivity over the course of a year can be matched to the
environmental stress hypothesis, which proposes that higher levels of
minimum available energy will lead to higher species richness
\citep{currie2004}. Finally, the variation DHI, which calculates the
coefficient of variance in a vegetation index through the course of a
year, corresponds to the environmental stability hypothesis which states
that lower energy variation throughout a year will lead to increased
species richness \citep{williams2008}.

Linkages between forest structure and productivity (namely, vegetation
indices) have been examined for nearly 20 years \citetext{\citealp[
\citet{knyazikhin1998}]{huete2002}; \citealp{myneni1994}}. While there
is significant theoretical and empirical evidence for their relationship
at single time points (within a single image) \citep{myneni1994},
various relationship directions and shapes have been found between
forest structure and productivity metrics \citep{ali2019}. These
relationships, their shapes, and their strengths have been attributed to
multiple possible hypotheses and can vary based on environmental
conditions \citep{ali2019}. The relationship between forest structural
diversity metrics and annual productivity summaries has yet to be fully
examined, including the DHIs.

Some vegetation structure metrics are simpler, and more accurate, to
calculate than others \citep{coops2021}. These basic metrics, such as
canopy height and canopy cover,can then be used to estimate additional
structure metrics, such as basal area or total biomass. Canopy height
and cover are commonly used as an indicator of vertical and horizontal
variation, respectively. Recently, more attention has been paid to
internal structural complexity metrics, which can be more difficult and
time consuming to generate \citep{coops2021, ma2022} .

In this study, we seek to untangle the relationship between two EBVs:
forest structure diversity metrics and yearly summaries of forest
productivity. To accomplish this, we assess this relationship using path
analysis to assess the direct and indirect (as mediated by more complex
forest structural diversity metrics) effects of commonly collected
forest structural metrics (canopy height and canopy cover) on yearly
productivity summaries. Doing so will in turn assess their utility as
complementary EBVs. We ran this analysis separately across four forested
land covers and the forested ecosystems of British Columbia, Canada.

\hypertarget{methods}{%
\section{Methods}\label{methods}}

\hypertarget{study-area}{%
\subsection{Study Area}\label{study-area}}

British Columbia is the westernmost province of Canada, and is home to a
variety of terrestrial ecosystems \citep{pojar1987}. Approximately 64\%
of the province is forested \citep{bcministryofforests2003}. There is a
large amount of ecosystem variation in the province, with large climate
and topographic gradients. The Biogeoclimatic Ecosystem Classification
(BEC) system identifies 16 zones based on the dominant tree species and
the ecosystems general climate. These zones can be further split into
subzones, variants, and phases based on microclimate, precipitation, and
topography.

\hypertarget{data}{%
\subsection{Data}\label{data}}

\hypertarget{sampling}{%
\subsection{Sampling}\label{sampling}}

\hypertarget{analysis}{%
\subsection{Analysis}\label{analysis}}

\hypertarget{results}{%
\section{Results}\label{results}}

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

\newpage

\renewcommand\refname{References}
  \bibliography{bibliography.bib}

\end{document}

### 2023-03-20 scripts/0_whole_game.qmd
first revisions to ncc 2023-03-17

@@
 library(sgsR)
-library(exactextractr)
+library(cols4all)
+# library(exactextractr)
 
@@
 ```{r loads}
-vlce <- rast("G:/mosaiced/VLCE2.0/LC_Class_HMM_2015_v20_v20.dat")
+vlce <- rast("F:/mosaiced/VLCE2.0/LC_Class_HMM_2015_v20_v20.dat")
 bcb <- bcmaps::bc_bound() %>%
@@
 ```{r run-focal}
-focal_loc <- "G:/mosaiced/focal"
+focal_loc <- "F:/mosaiced/focal"
 

### 2023-03-20 scripts/1_bc_focal_processing.qmd
first revisions to ncc 2023-03-17

@@
 
-struct_loc <- here::here("G://", "mosaiced", "structure")
+struct_loc <- here::here("F://", "mosaiced", "structure")
 
@@
 library(reticulate)
-use_python("C:/Users/evanmuis/Miniconda3/envs/geospatial/python.exe")
+use_python("C:/Users/evanmuis/AppData/Local/miniconda3/envs/geospatial/python.exe")
 
@@
 
-focal_save <- here::here("G:/", "mosaiced", "focal")
+focal_save <- here::here("F:/", "mosaiced", "focal")
 

### 2023-03-20 scripts/3_sem_visualization.qmd
first revisions to ncc 2023-03-17

@@
 library(lavaan)
+library(lavaanPlot)
+library(semPlot)
 
 theme_set(theme_light())
+```
+
+```{r}
+mylog <- function(data) {
+  if (any(data <= 0)) {
+    return(log1p(data))
+  } else {
+    return(log(data))
+  }
+}
+
+# normalization from grace et al 2016
+# first does mylog (does natural logarithm. if 0s present, does log(one + x))
+# then scales to z scores
+# results it more normal looking data
+normal <- function(data) {
+  data %>%
+    mylog() %>%
+    scale() %>%
+    as.numeric()
+}
 ```
@@
 ```{r}
-samples <- here::here("data", "samples", "bec") %>%
+bec_samples <- here::here("data", "samples", "bec") %>%
   list.files(full.names = T, pattern = ".shp$") %>%
@@
 
-samples %>%
-  st_drop_geometry() %>%
-  count(zone) %>%
-  ggplot(aes(
-    x = n,
-    y = fct_reorder(zone, n),
-    fill = zone
-  )) +
-  geom_col() +
-  ggsci::scale_fill_igv() +
-  theme_bw() +
-  labs(x = "n",
-       y = NULL,
-       fill = NULL)
+vlce_samples <- here::here("data", "samples", "vlce") %>%
+  list.files(full.names = T, pattern = ".shp$") %>%
+  map(read_sf) %>%
+  bind_rows()
 ```
@@
 ```{r}
-all_rasts_loc <- here::here("data", "rasts", "all_rasts.dat")
+
+all_rasts_loc <- here::here("data", "rasts", "all_rasts_bec.dat")
 
 if (!file.exists(all_rasts_loc)) {
-  struct_locs <- here::here("G://", "mosaiced", "structure")
+  # bec_rast <- bec() %>%
+  #   rasterize(y = vlce, field = "ZONE") %>%
+  #   crop(bcb_rast) %>%
+  #   mask(bcb_rast)
+  # 
+  # names(bec_rast) <- "bec"
+
+  struct_locs <- here::here("F://", "mosaiced", "structure")
   struct_rasts <- list.files(
@@
   dhi_rasts <-
-    list.files(here::here("G://", "mosaiced", "DHI"),
+    list.files(here::here("F://", "mosaiced", "DHI"),
                pattern = "2010s.tif$",
@@
 
-  dhi_mm <- minmax(dhi_rasts)
+  # dhi_mm <- minmax(dhi_rasts)
+  #
+  # dhi_rasts_mm <- (dhi_rasts - dhi_mm[1,]) / (dhi_mm[2,] - dhi_mm[1,])
 
-  dhi_rasts_mm <- (dhi_rasts - dhi_mm[1,]) / (dhi_mm[2,] - dhi_mm[1,])
+  all_rasts <- c(vlce, struct_rasts, dhi_rasts) %>%
+    mask(bcb_rast)
 
-  all_rasts <- c(vlce, struct_rasts, dhi_rasts_mm)
-
-  writeRaster(all_rasts, all_rasts_loc, overwrite = T, filetype = "envi",
-              datatype = "FLT4S")
+  writeRaster(
+    all_rasts,
+    all_rasts_loc,
+    overwrite = T,
+    filetype = "envi",
+    datatype = "FLT4S"
+  )
 }
@@
 all_rasts <- rast(all_rasts_loc)
+
+bec_rast <- bec() %>%
+  rasterize(y = vlce, field = "ZONE") %>%
+  crop(bcb_rast) %>%
+  mask(bcb_rast)
+
+names(bec_rast) <- "bec"
+all_rasts <- c(bec_rast, vlce, all_rasts)
 ```
@@
 ```{r}
-extracted_samples <- extract(all_rasts, samples, bind = T)
+bec_sample_loc <- here::here("data", "samples", "bec_extracted.shp")
+vlce_sample_loc <- here::here("data", "samples", "vlce_extracted.shp")
+
+if (!file.exists(bec_sample_loc)) {
+  bec_samples_e <- extract(all_rasts, bec_samples, bind = T)
+  writeVector(bec_samples_e, bec_sample_loc, overwrite = T)
+}
+if (!file.exists(vlce_sample_loc)) {
+  vlce_samples_e <- extract(all_rasts, vlce_samples, bind = T)
+  writeVector(vlce_samples_e, vlce_sample_loc, overwrite = T)
+}
+
+bec_samples_e <- vect(bec_sample_loc)
+vlce_samples_e <- vect(vlce_sample_loc)
 ```
-
-)
 
@@
 
-extracted_df <- extracted_samples %>% 
+bec_df <- bec_samples_e %>% 
   st_as_sf() %>%
   st_drop_geometry() %>%
+  select(bec:VarDHI) %>%
+  drop_na() %>%
+  tibble() %>%
+  filter(!(bec %in% invalid_zones))
+
+names_manual <- c(
+  "bec",
+  "category",
+  "basal_area",
+  "elev_cv",
+  "elev_mean",
+  "elev_p95",
+  "elev_stddev",
+  "gross_stem_volume",
+  "loreys_height",
+  "percentage_first_returns_above_2m",
+  "percentage_first_returns_above_mean",
+  "total_biomass",
+  "CumDHI",
+  "MinDHI",
+  "VarDHI"
+)
+
+# names(bec_df) <- names(all_rasts)
+names(bec_df) <- names_manual
+
+```
+
+```{r}
+
+bec_df <- bec_df %>%
+  mutate(strata = bec) %>%
+  relocate(strata)
+
+vlce_df <- vlce_samples %>%
+  st_as_sf() %>%
+  st_drop_geometry() %>%
+  select(bec:VarDHI)
+
+# names(bec_df) <- names(all_rasts)
+names(vlce_df) <- names_manual
+
+vlce_df <- vlce_df %>%
+  mutate(strata = category) %>%
+  relocate(strata)
+
+strata_df <- bind_rows(bec_df, vlce_df) %>%
   filter(!(loreys_height < elev_mean | elev_p95 < elev_mean)) %>%
@@
   mutate(value = value / divide_by) %>%
-  select(zone, category, ends_with("DHI"), variable, value) %>%
+  select(strata, bec, category, ends_with("DHI"), variable, value) %>%
   pivot_wider(names_from = "variable", values_from = "value") %>%
@@
 # effectively removing duplicates, but takes the mean instead of the first value
-extracted_df_b <- extracted_df %>%
-  group_by(across(c(zone, category, basal_area:total_biomass))) %>%
-  summarize(across(ends_with("DHI"), mean)) %>%
-  ungroup()
+strata_df_b <- strata_df %>%
+  group_by(across(c(strata, bec, category, basal_area:total_biomass))) %>%
+  dplyr::summarize(across(ends_with("DHI"), mean)) %>%
+  ungroup() %>%
+  filter(!(bec %in% invalid_zones))
 ```
@@
 ```{r}
-# downsampling procedure - allows for comparisons between models if sample size is the same (AIC and BIC are based on sample)
 
-downsample_df <- extracted_df_b %>%
-  count(zone) %>%
-  filter(!(zone %in% invalid_zones))
+# see number of pixels
+n_dups <- strata_df %>%
+  group_by(across(c(
+    strata, bec, category, basal_area:total_biomass
+  ))) %>%
+  summarize(n = n())
 
-downsample_no <- downsample_df %>%
+egg <- n_dups %>%
+  filter(!(bec %in% invalid_zones)) %>%
+  group_by(strata, duplicated = n != 1) %>%
+  summarize(duplicate_pixels = sum(n)) %>%
+  pivot_wider(names_from = duplicated, 
+              values_from = duplicate_pixels) %>%
+  mutate(total = `TRUE` + `FALSE`,
+         per_dup = `TRUE` / total,
+         strata_name = glue::glue("{strata} (n = {total})"),
+         label = glue::glue(" n[dup]: {total}"))
+
+n_dups %>% 
+  ggplot(aes(x = n, fill = strata)) + 
+  geom_histogram() + 
+  facet_wrap(~strata, scales = "free")
+
+egg %>%
+  ggplot(aes(x = fct_reorder(strata_name, per_dup), 
+             y = per_dup)) +
+
+  geom_col() +
+  geom_text(aes(label = label), hjust = 0,
+            parse = T) + 
+  lims(y = c(0, 1)) +
+  labs(x = NULL,
+         y = "Proportion of Duplicates") +
+  coord_flip() +
+  theme(panel.grid = element_blank())
+
+n_dups %>%
+  ggplot(aes(x = n)) +
+  geom_histogram() +
+  labs(x = "n",
+       y = NULL,
+       title = "Number of duplicated structure pixels") +
+  theme_light()
+
+```
+
+```{r}
+# downsampling procedure - allows for comparisons between models if 
+# sample size is the same (AIC and BIC are based on sample)
+
+downsample_no <- strata_df_b %>%
+  count(strata) %>%
   pull(n) %>%
@@
 set.seed(69420)
+# downsample the vlce with equivalent sampling across becs
+# if possible, if not it takes the maximum available samples and adjusts to 
+# account
 
-clean_df <- extracted_df_b %>%
-  mutate(
-    overstory_per = percentage_first_returns_above_mean,
-    understory_per = percentage_first_returns_above_2m - percentage_first_returns_above_mean,
-    total_cover = percentage_first_returns_above_2m,
-    overstory_range = loreys_height - elev_mean,
-    understory_range = elev_mean - 2
-  ) %>%
-  relocate(zone,
-           category,
-           ends_with("DHI")) %>%
-  filter(!(zone %in% invalid_zones)) %>%
-  group_by(zone) %>%
-  slice_sample(n = downsample_no) %>%
-  select(!starts_with("perc"))
+downsample_prop <- function(strat_tib) {
+  # this function samples equal to the downsample number
+  # across each bec zone, for each strata
+  # if a zone does not have enough to be an equal number of samples
+  # it takes all of those samples, and adjusts other zones to meet
+  # the total number of desired samples (downsample number)
+  bec_counts <- strat_tib %>%
+    count(bec)
+
+  total_bec <- bec_counts %>%
+    pull(bec) %>%
+    length()
+
+  bec_sub <- bec_counts %>%
+    filter(n <= ceiling(downsample_no / 12)) %>%
+    pull(bec)
+
+  df_sub <- strat_tib %>%
+    filter(bec %in% bec_sub)
+
+  n_sub <- nrow(df_sub)
+
+  n_rest <- downsample_no - n_sub
+
+  df_over <- strat_tib %>%
+    filter(!(bec %in% bec_sub)) %>%
+    group_by(bec) %>%
+    slice_sample(n = ceiling(n_rest / (total_bec - length(bec_sub)))) %>%
+    ungroup() %>%
+    slice_sample(n = n_rest)
+
+  bind_rows(df_over, df_sub)
+}
+
+vlce_ds <- strata_df_b %>%
+  filter(strata == category) %>%
+  group_by(strata) %>%
+  group_split() %>%
+  map(downsample_prop) %>%
+  bind_rows()
+
+# downsample bec zones. does not matter which forest type they come from
+
+bec_ds <- strata_df_b %>%
+  filter(strata != category) %>%
+  group_by(strata) %>%
+  slice_sample(n = downsample_no)
+
+strata_ds <- bind_rows(vlce_ds, bec_ds)
+
+meg_dir <- here::here("data", "samples", "meg")
+
+dir.create(meg_dir, showWarnings = F)
+write_csv(strata_ds, here::here(meg_dir, "downsampled.csv"))
+write_csv(strata_df_b, here::here(meg_dir, "full.csv"))
 ```
@@
 ```{r}
-mylog <- function(data) {
-  if (any(data <= 0)) {
-    return(log1p(data))
-  } else {
-    return(log(data))
-  }
-}
+clean_df <- strata_ds %>%
+  mutate(
+    percentage_first_returns_above_mean = percentage_first_returns_above_mean,
+    understory_per = percentage_first_returns_above_2m - percentage_first_returns_above_mean,
+    percentage_first_returns_above_2m = percentage_first_returns_above_2m,
+    overstory_range = loreys_height - elev_mean,
+    understory_range = elev_mean - 2,
+    over_biomass = total_biomass * (percentage_first_returns_above_mean / percentage_first_returns_above_2m),
+    under_biomass = total_biomass * (understory_per / percentage_first_returns_above_2m)
+  ) %>%
+  relocate(strata,
+           bec,
+           category,
+           ends_with("DHI"))
 
-normal <- function(data) {
-  data %>%
-    mylog() %>%
-    scale() %>% 
-    as.numeric()
-}
+# applies normalization by strata of analysis (i.e. bec zone or forest type)
 
-log_normal_df <- clean_df %>%
-  group_by(zone) %>%
-  mutate(across(CumDHI:understory_range, .fns = ~ normal(.x)))
+normal_df <- clean_df %>%
+  group_by(strata) %>%
+  mutate(across(CumDHI:under_biomass, .fns = ~ normal(.x)))
 
-log_normal_df %>%
-  summarize(across(CumDHI:understory_range, .fns = ~ shapiro.test(.x)$p.value))
+write_csv(normal_df, here::here(meg_dir, "downsampled_normal.csv"))
 ```
@@
 ```{r}
-log_splits <- log_normal_df %>%
-  group_by(zone) %>%
+dhi_vars <- c("CumDHI",
+              "VarDHI",
+              "MinDHI")
+models <- c(
+  paste0(
+    dhi_vars,
+    " ~ elev_p95 + total_biomass + elev_cv
+    elev_cv ~ elev_p95
+    total_biomass ~ elev_p95"
+  ),
+  paste0(
+    dhi_vars,
+    " ~ percentage_first_returns_above_2m + basal_area + percentage_first_returns_above_mean
+    percentage_first_returns_above_mean ~ percentage_first_returns_above_2m
+    basal_area ~ percentage_first_returns_above_2m"
+  )
+)
+
+splits <- normal_df %>%
+  group_by(strata) %>%
   group_split()
@@
 
-sample each bec zone down to the lowest number of samples
+```{r}
+lavaans <- crossing(data = splits, models = models) %>%
+  mutate(outs = map2(.x = models, .y = data, .f = sem))
+
+fitmeasures_list <- c("chisq", "df", "pvalue", "cfi", "rmsea", "aic", "srmr")
+
+myfits <- function(lavaan) {
+  try(fitmeasures(lavaan, fit.measures = fitmeasures_list), silent = T)
+}
+
+tib_out <- lavaans %>%
+  mutate(var = word(models, start = 1),
+         strata = data %>% map(pull, var = strata) %>% map(unique) %>% unlist(),
+         fits = outs %>% map(myfits),
+         pars = outs %>% map(parameterestimates) %>% map(tibble)#,
+         # interpreted = outs %>% map(effectsize::interpret)
+         ) %>%
+  relocate(data, models)
+
+non_converge <- tib_out %>%
+  filter(fits %>% map(class) == "try-error")
+
+converge <- tib_out %>%
+  # remove any models that don't converge
+  filter(!(fits %>% map(class) == "try-error"))
+
+converge_fm <- converge %>%
+  select(models, var, strata, fits, pars) %>%
+  mutate(fits = fits %>% map(as.numeric) %>%
+           map(tibble) %>%
+           map(mutate, measures = fitmeasures_list) %>%
+           map(rename, value = `<dbl>`) %>%
+           map(pivot_wider, names_from = measures, values_from = value)) %>%
+  unnest(fits) %>%
+  mutate(mod_follows = str_remove(models, ".*DHI"))
+```
+
+```{r}
+global_fits <- converge_fm %>%
+  mutate(global_sig = pvalue > 0.05,
+         mod_follows = str_remove(models, ".*DHI")) %>%
+  group_by(mod_follows) %>%
+  dplyr::summarize(per_sig = (sum(global_sig) / n()),
+                   count = n()) %>%
+  arrange(desc(per_sig))
+
+global_fits
+
+# fan et al. (2016) says chi ^ 2 saturates w/ samples size and as such isn't
+# the be all end all
+# need to also report: SRMR, CFI, and RMSEA
+# aic can be used to compare models
+# CFI > 0.95
+# RMSEA < 0.06
+# SRMR < 0.09
+```
+
+```{r}
+vlce_colours <- keys$vlce %>%
+  filter(forest == "Forest") %>%
+  select(strata = name_clean, colour = ntems_colour)
+
+# bc government bec colours.
+# i prefer the igv palette
+bec_colours <- bcmaps::bec_colors() %>%
+  tibble(strata = names(bcmaps::bec_colours()), colour = .)
+
+bec_colours <- keys$bec %>%
+  mutate(colour = ggsci::pal_igv("default")(16)) %>%
+  select(strata = zone, colour) %>% 
+  filter(!(strata %in% invalid_zones))
+
+strata_colours <- bind_rows(vlce_colours, bec_colours) %>%
+  arrange(strata)
+
+mod_names <- c("Vertical Structure",
+               "Horizontal Structure")
+
+selected_models <- models %>%
+  str_remove(".*DHI") %>%
+  unique()
+
+names_df <- tibble(mod_follows = selected_models,
+                   mod_names,
+                   order = 1:2)
+
+converge_fm %>%
+  mutate(mod_follows = str_remove(models, ".*DHI")) %>%
+  left_join(names_df) %>%
+  ggplot(aes(x = pvalue, y = chisq, col = strata)) +
+  geom_point() +
+  geom_vline(xintercept = 0.05,
+             col = "red",
+             lty = "dashed") +
+  annotate(
+    "Text",
+    label = "Global Fit Accepted",
+    x = 0.051,
+    y = 100,
+    angle = 270,
+    col = "red"
+  ) +
+  facet_wrap(~ fct_reorder(mod_names, order),
+             ncol = 1) +
+  scale_colour_manual(values = strata_colours %>% pull(colour) %>% unname()) +
+  theme_bw() +
+  labs(x = "P-value",
+       y = (expression(chi^2)),
+       colour = NULL) +
+  theme(panel.grid = element_blank())
+```
+
+```{r}
+par_df <- converge_fm %>%
+  group_by(strata) %>%
+  select(models, var, strata, pars, mod_follows) %>%
+  unnest(cols = pars) %>%
+  filter(op == "~")
+
+split_dfs <- par_df %>%
+  group_by(strata) %>%
+  group_split()
+
+fit_df <- converge_fm %>%
+  select(models, var, strata, mod_follows, pvalue:srmr)
+
+coords_df <-
+  tibble(
+    pos = c(
+      "elev_p95",
+      "elev_cv",
+      "total_biomass",
+      "percentage_first_returns_above_2m",
+      "basal_area",
+      "percentage_first_returns_above_mean",
+      "dhi",
+      "dhi"
+    ),
+    x = c(0, 2, 2, 0, 2, 2, 4, 4),
+    y = c(0, 1,-1, 0, 1,-1, 0, 0),
+    mod_follows = c(
+      selected_models[1],
+      selected_models[1],
+      selected_models[1],
+      selected_models[2],
+      selected_models[2],
+      selected_models[2],
+      selected_models[1],
+      selected_models[2]
+    )
+  )
+
+arrows_df <-
+  tibble(
+    rhs = c(
+      "elev_p95",
+      "elev_cv",
+      "total_biomass",
+      "elev_p95",
+      "elev_p95"
+    ),
+    lhs = c(
+      "dhi",
+      "dhi",
+      "dhi" ,
+      "elev_cv",
+      "total_biomass"
+    )
+  ) %>%
+  left_join(coords_df, by = c(lhs = "pos")) %>%
+  rename(xend = x,
+         yend = y) %>%
+  left_join(coords_df, by = c(rhs = "pos")) %>%
+  rename(xstart = x,
+         ystart = y) %>%
+  mutate(
+    xend = xend * 0.975,
+    yend = yend * 0.9,
+    yend = case_when(ystart == 1 ~ 0.15,
+                     ystart == -1 ~ -0.15,
+                     T ~ yend),
+    ystart = ystart * 0.9
+  )
+
+arrows_df2 <-
+  tibble(
+    rhs = c(
+      "percentage_first_returns_above_2m",
+      "basal_area",
+      "percentage_first_returns_above_mean",
+      "percentage_first_returns_above_2m",
+      "percentage_first_returns_above_2m"
+    ),
+    lhs = c(
+      "dhi",
+      "dhi",
+      "dhi",
+      "basal_area",
+      "percentage_first_returns_above_mean"
+    )
+  ) %>%
+  left_join(coords_df, by = c(lhs = "pos")) %>%
+  rename(xend = x,
+         yend = y) %>%
+  left_join(coords_df, by = c(rhs = "pos")) %>%
+  rename(xstart = x,
+         ystart = y) %>%
+  mutate(
+    xend = xend * 0.975,
+    yend = yend * 0.9,
+    yend = case_when(ystart == 1 ~ 0.15,
+                     ystart == -1 ~ -0.15,
+                     T ~ yend),
+    ystart = ystart * 0.9
+  )
+
+arrows_both <- bind_rows(arrows_df, arrows_df2) %>%
+  rename(mod_follows = mod_follows.x) %>%
+  select(-mod_follows.y)
+
+label_df <- coords_df %>% 
+  crossing(var = c("CumDHI", "VarDHI", "MinDHI")) %>% 
+  mutate(pos = case_when(pos == "dhi" ~ var, T ~ pos)) %>%
+  mutate(x = case_when(pos == "elev_p95" | pos == "percentage_first_returns_above_2m" ~ 0.25, T ~ x)) %>%
+  left_join(keys$continuous, by = c("pos" = "variable")) %>%
+  mutate(var_long = case_when(is.na(var_long) ~ pos,
+                              T ~ var_long),
+         x = case_when(endsWith(pos, "DHI") ~ 3.85,
+                       T ~ x)) %>%
+  select(!c(divide_by, unit))
+
+plot_sem <- function(split_df) {
+  strata_name <- split_df %>%
+    pull(strata) %>%
+    unique()
+
+  fit <- fit_df %>%
+    filter(strata == strata_name)
+
+  plot_df <- split_df %>%
+    mutate(lhs = case_when(endsWith(lhs, "DHI") ~ "dhi", T ~ lhs)) %>%
+    left_join(arrows_both) %>%
+    mutate(lhs = case_when(lhs == "dhi" ~ var,
+                           T ~ lhs))
+
+  est_label_df <- plot_df %>%
+    group_by(var, mod_follows) %>%
+    mutate(xmid = ((xend + xstart) / 2),
+           ymid = ((yend + ystart) / 2) + 0.125) %>%
+    select(var, xmid, ymid, est)
+
+  plot_df %>%
+    ggplot() +
+    geom_segment(
+      aes(
+        x = xstart,
+        y = ystart,
+        xend = xend,
+        yend = yend,
+        lty = pvalue < 0.05,
+        col = est,
+      ),
+      arrow = arrow(),
+      linewidth = 2
+    ) +
+    # geom_segment(aes(x = xstart, y = ystart, xend = xend, yend = yend,
+    #                  col = pvalue < 0.05),
+    #              ) +
+    geom_label(data = label_df, aes(x = x, y = y, label = var_long)) +
+    geom_text(data = est_label_df, aes(
+      x = xmid,
+      y = ymid,
+      label = round(est, digits = 2)
+    )) +
+    facet_grid(rows = vars(var),
+               cols = vars(mod_follows)) +
+    theme_void() +
+    theme(strip.text = element_blank()) +
+    labs(lty = "Significant Path",
+         col = "Standardized Estimate",
+         title = strata_name) +
+    theme(legend.position = "bottom") +
+    scico::scale_colour_scico(palette = "hawaii",
+                              midpoint = 0,
+                              limits = c(-2, 2)) +
+    scale_linetype_manual(values = c("dashed", "solid")) +
+    geom_text(data = fit, aes(
+      x = 3.7,
+      y = 0.75,
+      label = paste0(
+        "CFI:",
+        round(cfi, digits = 3),
+        "\nRMSEA:",
+        round(rmsea, digits = 3),
+        "\nSRMR:",
+        round(srmr, digits = 3),
+        "\nP-Value:",
+        round(pvalue, digits = 3)
+      )
+    ))
+
+  ggsave(here::here("outputs", "strata_sem", paste0(strata_name, ".png")), dpi = 300,
+         width = unit(13, "inches"),
+         height = unit(11, "inches"),
+         bg = "white")
+
+}
+
+map(split_dfs, plot_sem)
+```
+
+```{r}
+# this cell is about aggregating to the thing. i also want to look at it by BEC
+# vs forest type
+
+vlce_classes <- vlce_colours %>%
+  pull(strata)
+
+sig_est_df <- par_df %>% 
+  ungroup() %>%
+  mutate(sig_est = ifelse(pvalue < 0.05, est, NA))
+
+all_est_df <- sig_est_df %>%
+  group_by(var, lhs, rhs, mod_follows) %>%
+  summarize(num_sig = sum(pvalue < 0.05),
+            per_sig = num_sig / n(),
+            mean_sig_est = mean(sig_est, na.rm = T)) %>%
+  arrange(mean_sig_est) %>%
+  mutate(strata = "All Strata")
+
+vlce_est_df <- sig_est_df %>%
+  filter(strata %in% vlce_classes) %>%
+  group_by(var, lhs, rhs, mod_follows) %>%
+  summarize(num_sig = sum(pvalue < 0.05),
+            per_sig = num_sig / n(),
+            mean_sig_est = mean(sig_est, na.rm = T)) %>%
+  arrange(mean_sig_est) %>%
+  mutate(strata = "Forest Type")
+
+bec_est_df <- sig_est_df %>%
+  filter(!(strata %in% vlce_classes)) %>%
+  group_by(var, lhs, rhs, mod_follows) %>%
+  summarize(num_sig = sum(pvalue < 0.05),
+            per_sig = num_sig / n(),
+            mean_sig_est = mean(sig_est, na.rm = T)) %>%
+  arrange(mean_sig_est) %>%
+  mutate(strata = "BEC Zones")
+```
+
+```{r}
+group_dfs <- list(all_est_df, vlce_est_df, bec_est_df)
+
+plot_sem_group <- function(group_df) {
+  strata <- group_df %>%
+    pull(strata) %>%
+    unique
+
+  plot_df <- group_df %>%
+    mutate(lhs = case_when(endsWith(lhs, "DHI") ~ "dhi", T ~ lhs)) %>%
+    left_join(arrows_both) %>%
+    mutate(lhs = case_when(lhs == "dhi" ~ var,
+                           T ~ lhs)) %>%
+    ungroup()
+
+  est_label_df <- plot_df %>%
+    group_by(var, mod_follows) %>%
+    mutate(xmid = (xend + xstart) / 2,
+           ymid = ((yend + ystart) / 2) + 0.125) %>%
+    select(var, xmid, ymid, mean_sig_est)
+
+  plot_df %>%
+    ggplot() +
+    geom_segment(aes(
+      x = xstart,
+      y = ystart,
+      xend = xend,
+      yend = yend,
+      linewidth = per_sig,
+      col = mean_sig_est,
+    ),
+    arrow = arrow()) +
+    # geom_segment(aes(x = xstart, y = ystart, xend = xend, yend = yend,
+    #                  col = pvalue < 0.05),
+    #              ) +
+    geom_label(data = label_df, aes(x = x, y = y, label = var_long)) +
+    geom_text(data = est_label_df, aes(x = xmid, y = ymid, label = round(mean_sig_est, digits = 2))) +
+    facet_grid(rows = vars(var),
+               cols = vars(mod_follows)) +
+    theme_void() +
+    theme(strip.text = element_blank()) +
+    labs(linewidth = "Proportion of Significant Paths",
+         col = "Standardized Estimate",
+         title = strata) +
+    theme(legend.position = "bottom") +
+    scico::scale_colour_scico(palette = "hawaii", 
+                              midpoint = 0,
+                              limits = c(-2, 2))
+
+  ggsave(here::here("outputs", "strata_sem", "aggregated", paste0(strata, ".png")), dpi = 300,
+         width = unit(13, "inches"),
+         height = unit(11, "inches"),
+         bg = "white")
+}
+
+map(group_dfs, plot_sem_group)
+```
+
+```{r map}
+sig_est_df %>% 
+  filter(!is.na(sig_est),
+         endsWith(lhs, "DHI")) %>%
+  group_by(strata, models) %>%
+  slice_max(abs(sig_est)) %>%
+  select(mod_follows, var, strata, rhs, est)
+
+```
+
+```{r}
+ver_model_latent <- '# latent variables
+     fs =~ elev_cv + total_biomass + elev_p95
+     elev_cv ~ elev_p95
+     total_biomass ~ elev_p95
+
+   # dhi follows fs
+    CumDHI ~ fs
+    MinDHI ~ fs
+    VarDHI ~ fs
+   # covariances
+    CumDHI ~~ MinDHI
+    MinDHI ~~ VarDHI
+    VarDHI ~~ CumDHI
+
+'
+
+hor_model_latent <- '# latent variables
+     fs =~ percentage_first_returns_above_2m + basal_area + percentage_first_returns_above_mean
+     basal_area ~ percentage_first_returns_above_2m
+     percentage_first_returns_above_mean ~ percentage_first_returns_above_2m
+
+   # dhi follows fs
+    CumDHI ~ fs
+    MinDHI ~ fs
+    VarDHI ~ fs
+   # covariances
+    CumDHI ~~ MinDHI
+    MinDHI ~~ VarDHI
+    VarDHI ~~ CumDHI
+
+'
+
+internal_model_latent <- '# latent variables
+     fs =~ elev_cv + elev_stddev
+
+   # dhi follows fs
+    CumDHI ~ fs
+    MinDHI ~ fs
+    VarDHI ~ fs
+   # covariances
+    CumDHI ~~ MinDHI
+    MinDHI ~~ VarDHI
+    VarDHI ~~ CumDHI
+
+'
+
+basic_model_latent <- '# latent variables
+     fs =~ percentage_first_returns_above_2m + elev_p95
+
+   # dhi follows fs
+    CumDHI ~ fs
+    MinDHI ~ fs
+    VarDHI ~ fs
+   # covariances
+    CumDHI ~~ MinDHI
+    MinDHI ~~ VarDHI
+    VarDHI ~~ CumDHI
+
+'
+
+derived_model_latent <- '# latent variables
+     fs =~ loreys_height + basal_area + elev_cv + elev_stddev + gross_stem_volume + total_biomass
+
+   # dhi follows fs
+    CumDHI ~ fs
+    MinDHI ~ fs
+    VarDHI ~ fs
+   # covariances
+    CumDHI ~~ MinDHI
+    MinDHI ~~ VarDHI
+    VarDHI ~~ CumDHI
+
+'
+
+all_model_latent <- '# latent variables
+     fs =~ basal_area + elev_cv + elev_mean + elev_stddev + loreys_height + percentage_first_returns_above_2m + percentage_first_returns_above_mean + total_biomass + gross_stem_volume + elev_p95
+
+   # dhi follows fs
+    CumDHI ~ fs
+    MinDHI ~ fs
+    VarDHI ~ fs
+   # covariances
+    CumDHI ~~ MinDHI
+    MinDHI ~~ VarDHI
+    VarDHI ~~ CumDHI
+
+'
+
+all_model_latent <- '# latent variables
+     fs =~ elev_cv + elev_mean + elev_stddev + loreys_height + percentage_first_returns_above_2m + percentage_first_returns_above_mean +  elev_p95
+
+   # dhi follows fs
+    CumDHI ~ fs
+    MinDHI ~ fs
+    VarDHI ~ fs
+   # covariances
+    CumDHI ~~ MinDHI
+    MinDHI ~~ VarDHI
+    VarDHI ~~ CumDHI
+
+'
+
+latent_models <- c(all_model_latent, derived_model_latent, basic_model_latent, internal_model_latent, hor_model_latent, ver_model_latent)
+
+```
+
+```{r}
+
+bec_normal <- bec_df %>%
+  mutate(across(basal_area:VarDHI,  .fns = ~ normal(.x)))
+
+splits <- bec_normal %>%
+  group_by(bec) %>%
+  group_split()
+
+latent_lavaans <- crossing(splits, latent_models) %>%
+  mutate(outs = map2(.x = latent_models, .y = splits, .f = sem))
+
+egg <- crossing(splits = list(bec_normal), latent_models) %>%
+  mutate(outs = map2(.x = latent_models, .y = splits, .f = sem))
+```
+
+```{r}
+egg$outs[[6]] %>%
+  graph_sem(layout = get_layout(., "layout_nicely"))
+```
+
+```{r}
+fitmeasures_list <- c("chisq", "df", "pvalue", "cfi", "rmsea", "aic", "srmr")
+
+myfits <- function(lavaan) {
+  try(fitmeasures(lavaan, fit.measures = fitmeasures_list), silent = T)
+}
+
+tib_out <- latent_lavaans %>%
+  mutate(strata = splits %>% map(pull, var = strata) %>% map(unique) %>% unlist(),
+         fits = outs %>% map(myfits),
+         pars = outs %>% map(parameterestimates) %>% map(tibble)#,
+         # interpreted = outs %>% map(effectsize::interpret)
+         ) %>%
+  relocate(splits, latent_models)
+
+non_converge <- tib_out %>%
+  filter(fits %>% map(class) == "try-error")
+
+converge <- tib_out %>%
+  # remove any models that don't converge
+  filter(!(fits %>% map(class) == "try-error"))
+
+converge_fm <- converge %>%
+  select(latent_models, strata, fits, pars) %>%
+  mutate(fits = fits %>% map(as.numeric) %>%
+           map(tibble) %>%
+           map(mutate, measures = fitmeasures_list) %>%
+           map(rename, value = `<dbl>`) %>%
+           map(pivot_wider, names_from = measures, values_from = value)) %>%
+  unnest(fits)
+```
+
+```{r}
+crossing(df = list(normal_df), latent_models) %>%
+  mutate(outs = map2(.x = latent_models, .y = df, .f = sem))
+```
+
+```{r}
+sem(latent_models[[1]], data = normal_df) %>%
+  summary()
+```
+
+```{r}
+for(s in clean_df %>%
+    pull(strata) %>%
+    unique) {
+  ggally_plot <- clean_df %>%
+    filter(strata == s) %>%
+    select(-bec,-category) %>%
+    select(strata:total_biomass) %>%
+    mutate(MinDHI = normal(MinDHI)) %>%
+    #select(CumDHI:total_biomass) %>%
+    GGally::ggpairs(aes(colour = strata, alpha = 0.2))
+
+  ggsave(
+    here::here("outputs", glue::glue("ggally_{s}.png")),
+    plot = ggally_plot,
+    dpi = 300,
+    height = 40,
+    width = 40
+  )
+}
+
+```
+
+```{r}
+# efa
+efa_data <- bec_normal %>%
+  select(basal_area:VarDHI)
+
+efa_fit <- efa(data = efa_data,
+               nfactors = 1:2)
+
+summary(efa_fit, nd = 3L, cutoff = 0.2, dot.cutoff = 0.05)
+```
+
+```{r}
+# cfa
+cfa_model <- "vert =~ elev_mean + elev_p95 + elev_stddev
+              hor =~ percentage_first_returns_above_2m + percentage_first_returns_above_mean
+
+              CumDHI ~ vert + hor + elev_cv
+              VarDHI ~ vert + hor + elev_cv
+              MinDHI ~ vert + hor + elev_cv"
+
+cfa_fit <- sem(cfa_model, bec_normal)
+
+summary(cfa_fit)
+
+lavaanPlot(cfa_fit, coefs = T)
+```
+
+```{r}
+bec_normal <- bec_df %>%
+  mutate(across(basal_area:VarDHI,  .fns = ~ normal(.x)))
+
+all_model_latent <- '# latent variables
+     fs =~ elev_cv + elev_mean + *elev_stddev + *loreys_height + percentage_first_returns_above_2m + *percentage_first_returns_above_mean +  *elev_p95
+
+   # dhi follows fs
+    CumDHI ~ fs
+    MinDHI ~ fs
+    VarDHI ~ fs
+   # covariances
+    CumDHI ~~ MinDHI
+    MinDHI ~~ VarDHI
+    VarDHI ~~ CumDHI
+
+'
+
+all_model_latent <- '# latent variables
+     fs =~ elev_cv + elev_p95 + percentage_first_returns_above_2m
+# add hierarchical model here
+
+   # dhi follows fs
+    CumDHI ~ fs
+    MinDHI ~ fs
+    VarDHI ~ fs
+   # covariances
+    CumDHI ~~ MinDHI
+    MinDHI ~~ VarDHI
+    VarDHI ~~ CumDHI
+
+'
+
+subs <- bec_normal %>%
+  slice_sample(n = 500)
+
+egg <- sem(all_model_latent, bec_normal) 
+egg %>% summary()
+
+egg %>% semPaths(what = "est")
+```
+
+```{r}
+ver_model_latent <- '# latent variables
+     fs =~ elev_cv + total_biomass + elev_p95
+     total_biomass ~ elev_p95
+
+   # dhi follows fs
+    CumDHI ~ fs
+    MinDHI ~ fs
+    VarDHI ~ fs
+   # covariances
+    CumDHI ~~ MinDHI
+    MinDHI ~~ VarDHI
+    VarDHI ~~ CumDHI
+
+'
+
+egg <- sem(ver_model_latent, bec_normal)
+
+egg %>% summary()
+
+egg %>% semPaths(what = "est")
+```
+
+Heatmaps
+
+```{r}
+library(dendextend)
+library(cluster)
+# clustering to select variables of interest
+
+bec_iter <- bec_normal %>%
+  filter(bec != "BG") %>%
+  pull(bec) %>%
+  unique()
+
+here::here("outputs", "heatmap") %>%
+  dir.create()
+for (s in bec_iter) {
+  spellman <- bec_normal %>%
+    filter(strata == s) %>%
+    select(basal_area:total_biomass) %>%
+    slice_sample(n = 236) %>%
+    as.matrix() %>%
+    Hmisc::rcorr()
+
+  spellman_p <- spellman %>%
+    .$P
+
+  spellman_cor <- spellman %>%
+    .$r
+
+  spellman_dist <- as.dist(1 - spellman_cor)
+
+  spellman_tree <- hclust(spellman_dist, method = "complete")
+
+  spellman_dend <- as.dendrogram(spellman_tree)
+
+  plot(color_branches(spellman_dend, k = 4))
+
+  spellman.kmedoids <- pam(spellman_dist, 4)
+
+  kclusters <- spellman.kmedoids$cluster
+
+  kclusters.reordered <- kclusters[order.dendrogram(spellman_dend)]
+  dend.colors <-
+    unique(get_leaves_branches_attr(color_branches(spellman_dend, k = 4), attr =
+                                      "col"))
+  plot(branches_attr_by_clusters(spellman_dend, kclusters.reordered , dend.colors))
+
+  color.scheme <-
+    rev(RColorBrewer::brewer.pal(10, "RdBu")) # generate the color scheme to use
+  color.scheme <- rev(scico::scico(n = 9, palette = "batlow"))
+
+  p_labs <- spellman_p %>%
+    cut(
+      breaks = c(0, 0.001, 0.01, 0.05, 1),
+      labels = c("***", "**", "*", "NS"),
+      right = F
+    ) %>%
+    matrix(nrow = nrow(spellman_p))
+
+  png(
+    filename = here::here("outputs", "heatmap", glue::glue("heatmap_{s}.png")),
+    height = 12,
+    width = 12,
+    units = "in",
+    res = 300
+  )
+  gplots::heatmap.2(
+    spellman_cor,
+    Rowv = ladderize(spellman_dend),
+    Colv = ladderize(spellman_dend),
+    dendrogram = "both",
+    revC = TRUE,
+    # rev column order of dendrogram so conforms to natural representation
+    trace = "none",
+    density.info = "none",
+    col = color.scheme,
+    key.title = NA,
+    key.xlab = "Correlation",
+    cellnote = p_labs
+    # cexRow = 0.5,
+    # cexCol = 0.5
+  )
+
+  dev.off()
+}
+```
+
+```{r}
+bec_normal <- bec_df %>%
+  mutate(across(basal_area:VarDHI,  .fns = ~ normal(.x)))
+
+all_model_latent <- '# latent variables
+     fs =~ elev_cv + elev_p95 + percentage_first_returns_above_2m
+
+   # dhi follows fs
+    CumDHI ~ fs
+
+   # covariances
+    CumDHI ~~ MinDHI
+    MinDHI ~~ VarDHI
+    VarDHI ~~ CumDHI
+
+'
+
+all_models <- c(all_model_latent, models)
+
+bec_normal %>%
+  filter(bec != "BG") %>%
+  group_by(bec) %>%
+  nest() %>%
+  crossing(all_models) %>%
+  mutate(out = map2(.x = all_models, .y = data, .f = sem))
+
+out <- sem(all_model_latent, data = bec_normal %>%
+             filter(bec == "BG"))
+
+summary(out)
+```
 
@@
 set.seed(69420)
+plot_bec_df <- bec_df %>%
+  filter(elev_cv <= 1000) %>%
+  group_by(bec) %>%
+  slice_sample(n = 172)
 
-zone_dfs <- extracted_df %>%
-  filter(!(zone %in% invalid_zones)) %>%
-  group_by(zone) %>%
-  group_split() %>%
-  map(slice_sample, n = downsample_no)
+plot_bec_df %>%
+  ggplot(aes(x = elev_p95, 
+             y = CumDHI, 
+             col = bec)) + 
+  geom_point(alpha = 0.5) +
+  geom_smooth(method = "lm") + 
+  facet_wrap(~bec) +
+  ggsci::scale_color_igv()
+
+plot_bec_df %>%
+  ggplot(aes(x = CumDHI, 
+             fill = bec)) + 
+  geom_histogram() +
+  facet_wrap(~bec) +
+  ggsci::scale_color_igv()
 ```
@@
 ```{r}
-library(Hmisc)
-library(corrplot)
-corr <- log_normal_df %>% 
-  ungroup() %>%
-  select(CumDHI:total_biomass) %>%
-  as.matrix() %>%
-  rcorr()
+dhi_vars <- c("CumDHI",
+              "VarDHI",
+              "MinDHI")
+models <- c(
+  paste0(
+    dhi_vars,
+    " ~ elev_p95 + total_biomass + elev_cv
+    elev_cv ~ elev_p95
+    total_biomass ~ elev_p95"
+  ),
+  paste0(
+    dhi_vars,
+    " ~ percentage_first_returns_above_2m + total_biomass + elev_cv
+    elev_cv ~ percentage_first_returns_above_2m
+    total_biomass ~ percentage_first_returns_above_2m"
+  )
+)
 
-corrplot(corr$r,
-         type = "upper",
-         tl.col = "black",
-         tl.srt = 45,
-         p.mat = corr$P, 
-         sig.level = 0.01)
+bec_normal <- bec_df %>%
+  mutate(across(basal_area:VarDHI,  .fns = ~ normal(.x)))
+
+lavaans <- bec_normal %>%
+  filter(bec != "BG") %>%
+  group_by(bec) %>%
+  nest() %>%
+  crossing(models) %>%
+  mutate(outs = map2(.x = models, .y = data, .f = sem))
+
 ```
@@
 ```{r}
-sample_names <- names(extracted_samples)
 
-dhi_vars <-
-  sample_names[(length(sample_names) - 2):length(sample_names)]
-
-v_models <-
-  paste0(
-    dhi_vars,
-    " ~ loreys_height + total_biomass + elev_stddev
-elev_stddev ~ loreys_height
-total_biomass ~ loreys_height"
-  )
-
-h_models <-
-  paste0(
-    dhi_vars,
-    " ~ basal_area + percentage_first_returns_above_2m + percentage_first_returns_above_mean
-percentage_first_returns_above_mean ~ basal_area
-percentage_first_returns_above_2m ~ basal_area"
-  )
-
-models <- c(v_models, h_models)
-```
-
-```{r some-weird-map-lol}
-bec_zones <- bec() %>%
-  mutate(area = st_area(.)) %>%
-  group_by(ZONE) %>%
-  dplyr::summarize(area = sum(area)) %>%
-  st_intersection(bc_bound())
-
-bec_counts_sf <- bec_zones %>%
-  left_join(bec_counts, by = c("ZONE" = "zone")) %>%
-  select(ZONE, n) %>%
-  mutate(ZONE = ifelse(!(ZONE %in% invalid_zones), ZONE, "Not Included"))
-
-ggplot(bec_counts_sf) +
-  geom_sf(aes(fill = ZONE), colour = "#00000000") +
-  ggsci::scale_fill_igv() +
-  scale_alpha_manual(values = c(0.5, 1)) +
-  theme_void()
-```
-
-```{r zonal-models}
-sem_extract <- function(split_df, model) {
-  zone <- split_df %>%
-    pull(zone) %>%
-    unique()
-
-  var <- str_split(model, pattern = " ")[[1]][1]
-  fit <- sem(model = model, data = split_df)
-
-  fits <-
-    fitmeasures(fit, c("chisq", "df", "pvalue", "cfi", "rmsea", "aic", "r2", "srmr"))
-  pars <- parameterestimates(fit) %>%
-    tibble() %>%
-    mutate(zone = zone, var = var, model = model)
-
-  fits <- tibble(fit = names(fits), value = as.numeric(fits)) %>%
-    pivot_wider(names_from = fit, values_from = value) %>%
-    mutate(zone = zone,
-           var = var, model = model)
-
-  interpreted <- fit %>%
-    effectsize::interpret() %>%
-    mutate(zone = zone, var = var, model = model)
-
-  return(list(
-    # fitmeasures = fits,
-    parameters = pars,
-    interpreted = interpreted,
-    fits = fits
-  ))
+fitmeasures_list <- c("chisq", "df", "pvalue", "cfi", "rmsea", "aic", "srmr")
+myfits <- function(lavaan) {
+  try(fitmeasures(lavaan, fit.measures = fitmeasures_list), silent = T)
 }
 
-crossed <- crossing(zone_dfs, models)
+tib_out <- lavaans %>%
+  mutate(var = word(models, start = 1),
+         #strata = data %>% map(pull, var = strata) %>% map(unique) %>% unlist(),
+         fits = outs %>% map(myfits),
+         pars = outs %>% map(parameterestimates) %>% map(tibble)#,
+         # interpreted = outs %>% map(effectsize::interpret)
+         ) %>%
+  relocate(data, models)
 
-dfs <- crossed %>% pull(zone_dfs)
-mods <- crossed %>% pull(models)
+non_converge <- tib_out %>%
+  filter(fits %>% map(class) == "try-error")
 
-outs <- map2(.x = dfs, .y = mods, .f = sem_extract)
+non_converge
 
-params <- map_df(outs, function(x) {
-    x$parameters
-  })
+converge <- tib_out %>%
+  # remove any models that don't converge
+  filter(!(fits %>% map(class) == "try-error"))
 
-# fitmeasures <- map_df(outs, function(x) {
-#   x$fitmeasures
-# })
+converge_fm <- converge %>%
+  select(bec, var, models, fits, pars) %>%
+  mutate(fits = fits %>% map(as.numeric) %>%
+           map(tibble) %>%
+           map(mutate, measures = fitmeasures_list) %>%
+           map(rename, value = `<dbl>`) %>%
+           map(pivot_wider, names_from = measures, values_from = value)) %>%
+  unnest(fits) %>%
+  mutate(mod_follows = str_remove(models, ".*DHI"))
 
-fitmeasures <- map_df(outs, function(x) {
-  x$interpreted
-})
+converge_fm %>% group_by(models) %>%
+  summarize(across(chisq:srmr, mean)) %>%
+  arrange(desc(cfi))
 
-fits <- map_df(outs, function(x) {
-  x$fits
-})
 ```
@@
 ```{r}
-vlce_samples <- here::here("data", "samples", "vlce") %>%
-  list.files(pattern = ".shp", full.names = T) %>%
-  map(read_sf) %>%
-  bind_rows()
-
-extracted_vlce <- extract(all_rasts, vlce_samples)
+converge_fm %>%
+  group_by(bec, var) %>%
+  mutate(aic_d = aic - min(aic)) %>%
+  group_by(mod_follows) %>%
+  filter(mod_follows == " ~ elev_p95 + total_biomass + elev_cv\n    elev_cv ~ elev_p95\n    total_biomass ~ elev_p95" & aic_d != 0)
 ```
@@
 ```{r}
-extracted_vlce %>%
-  st_drop_geometry() %>%
-  count(category)
+converge_unnest <- converge %>% 
+  select(bec, models, var, pars) %>%
+  unnest(pars)
+
+strongest_counts <- converge_unnest %>%
+  filter(pvalue < 0.05) %>%
+  filter(endsWith(lhs, "DHI")) %>%
+  filter(op == "~") %>% 
+  group_by(bec, models) %>%
+  slice_max(abs(est)) %>%
+  group_by(var, models, rhs) %>%
+  summarize(n = n(),
+            mean_est = mean(est)) %>%
+  mutate(model = ifelse(str_detect(models, "elev_p95"), "Height Model", "Cover Model"))
+
+strongest_counts %>%
+  group_by(var, models) %>%
+  mutate(per = n / sum(n)) %>%
+  left_join(keys$continuous, by = c("rhs" = "variable")) %>%
+  ggplot(aes(x = model, y = per, fill = var_long)) +
+  geom_col() +
+  facet_wrap(~var) +
+  labs(x = "Model",
+       y = "Proportion of Strongest Significant Predictor",
+       fill = "Predictor",
+       title = "Fig. (x): Proportion of strongest signficant forest structural predictors of yearly productivity metrics \nin models based on canopy cover vs canopy height."
+       ) +
+  scale_y_continuous(labels = scales::label_percent()) +
+  theme(legend.position = "bottom") +
+  scale_fill_discrete_c4a_cat("okabe")
+
+ggsave(here::here("outputs", "path_bars.png"))
+
 ```
@@
 ```{r}
-fitmeasures %>% 
+
+bec <- bec()
+
+bec_dissolve_loc <-
+  here::here("data", "shapefiles", "bec_dissolve.shp")
+
+if (!file.exists(bec_dissolve_loc)) {
+  dir.create(dirname(bec_dissolve_loc), showWarnings = F)
+
+  bec_dissolve <- bec %>%
+    st_intersection(bc_bound()) %>%
+    select(zone = ZONE,
+           geometry) %>%
     group_by(zone) %>%
-  dplyr::summarize(across(cfi:aic, mean))
+    summarize(geometry = st_union(geometry)) %>%
+    st_write(bec_dissolve_loc, append = F)
+}
+bec_dissolve <- read_sf(bec_dissolve_loc)
 
-fitmeasures %>%
-  mutate(v_h = ifelse(model %in% h_models, "h", "v")) %>%
-  group_by(model) %>%
-  dplyr::summarize(across(c(cfi:aic, chisq), mean))
+for_map <- converge_unnest %>%
+  filter(pvalue < 0.05) %>%
+  filter(endsWith(lhs, "DHI")) %>%
+  filter(op == "~") %>%
+  mutate(model = ifelse(str_detect(models, "elev_p95"), "Height Model", "Cover Model")) %>%
+  group_by(bec, var, model) %>%
+  slice_max(abs(est)) %>%
+  select(var, bec, model, rhs, est) %>%
+  left_join(keys$continuous %>%
+              select(variable, var_long),
+            by = c("rhs" = "variable")) %>%
+  mutate(var_long = as.factor(var_long)) %>%
+  ungroup() %>%
+  complete(var, bec, model) %>%
+  # filter(var == "CumDHI",
+  #        model == "Cover Model") %>%
+  left_join(bec_dissolve, by = c("bec" = "zone"))
 ```
@@
 ```{r}
-params %>%
-  filter(op == "~") %>%
-  group_by(model, lhs, rhs) %>%
-  dplyr::summarize(per_sig = (pvalue > 0.05) / n()) %>%
-  view()
+div_cols <- c(
+  c4a("troy", 5),
+  c4a("pi_yg", 5),
+  c4a("pu_or", 5),
+  c4a("div_tq_wh_pk", 5)
+)
+
+var_preds <- converge_unnest %>% pull(lhs) %>% unique() %>% str_subset(., "[^DHI]$")
+
+var_long <- keys$continuous %>%
+  filter(variable %in% var_preds) %>%
+  pull(var_long)
+
+col_join <- crossing(
+  var_long,
+  strength = c("a", "b", "c", "d", "e")) %>%
+  mutate(col = div_cols) %>%
+  mutate(col = ifelse(strength == "c", "grey80", col))
+
+col_join <- crossing(
+  var_long,
+  strength = c("a", "b", "c", "d", "e")) %>%
+  mutate(col = div_cols) %>%
+  mutate(col = ifelse(strength == "c", "grey80", col))
+
+legend_cols <- col_join %>%
+  pull(col) %>%
+  .[str_order(.)]
+map_legend <- col_join %>%
+  ggplot(aes(x = fct_rev(strength), y = var_long, fill = col)) +
+  geom_tile() +
+  scale_fill_manual(values = legend_cols) +
+  theme_void() +
+  theme(legend.position = "none",
+        axis.text = element_text())
+
+ggsave(here::here("outputs", "map_legend_biv.svg"), plot = map_legend, bg = "white", device = "svg")
+
+est_mm <- for_map %>% 
+  select(-geometry) %>% 
+  summarize(min = min(est, na.rm = T), 
+            max = max(est, na.rm = T))
+
+seq <- seq(0, 1,
+                 by = 0.01)
+
+myseq <- seq(est_mm$min * 100 %>% round(2), est_mm$max * 100 %>% round(2), by = 0.01)
+
+(alpha_legend <- crossing(var_preds, seq) %>%
+  left_join(keys$continuous, by = c("var_preds" = "variable")) %>%
+  ggplot(aes(x = seq, y ="1", fill = fct_rev(var_preds), alpha = seq)) +
+  geom_col() +
+  theme_void() +
+  theme(legend.position = "none",
+        axis.text.y = element_text()) +
+  facet_grid(cols = vars(var_long))
+
+)
+
+alpha_legend
+
+ggsave(here::here("outputs", "alpha_legend.png"), plot = alpha_legend, device = "png")
+
 ```
@@
 ```{r}
-test_mod <- "CumDHI ~ loreys_height + percentage_first_returns_above_2m + percentage_first_returns_above_mean
-percentage_first_returns_above_2m ~ total_biomass
-percentage_first_returns_above_mean ~ percentage_first_returns_above_2m + total_biomass
-total_biomass ~ loreys_height"
+added_cols <- for_map %>%
+  arrange(desc(est)) %>%
+  mutate(
+    strength = case_when(
+      est <= 1 & est > 0.6  ~ "a",
+      est <= 0.6 & est > 0.2 ~ "b",
+      est <= 0.2 & est > -0.2 ~ "c",
+      est <= -0.2 & est > -0.6 ~ "d",
+      est <= -0.6 & est > -1 ~ "e",
+      T ~ NA
+    )
+  ) %>%
+  left_join(col_join)
 
-data <- dfs[[19]] %>%
-  mutate(across(basal_area:total_biomass, scale),
-         across(basal_area:total_biomass, as.numeric))
+added_cols %>% 
+  drop_na() %>%
+  select(-geometry) %>% 
+  ggplot(aes(x = est, y = bec, fill = var_long)) + 
+  geom_col(position = "dodge") + 
+  facet_grid(cols = vars(var),
+             rows = vars(model)) +
+  labs(x = "Standardized Path Estimate for Strongest Predictor",
+       y = NULL,
+       fill = "Predictor") +
+  theme(legend.position = "bottom",
+        panel.grid = element_blank()) +
+  geom_vline(xintercept = 0, alpha = 0.5, col = "grey")
 
-(egg <- sem(test_mod, data = data))
+ggsave(here::here("outputs", "strongest_predictor_bar.png"))
 
-varTable(egg)
-fitmeasures(egg)
-parameterestimates(egg)
+plot_cols <- added_cols %>%
+  pull(col) %>%
+  unique() %>%
+  .[str_order(.)]
 
-graph_sem(egg)
+# bivariate plot
+
+plot <- ggplot() +
+  geom_sf(data = added_cols, aes(geometry = geometry, fill = col), col = "#00000000") +
+  scale_fill_manual(values = plot_cols) +
+  facet_grid(cols = vars(var),
+             rows = vars(model)) + 
+  theme_void() +
+  theme(legend.position = "none",
+        plot.title = element_text()) +
+  labs(title = "Fig (x): Map of strongest structural attribute predictor strength on yearly productivity metrics by BEC zone for the two proposed models.
+\nStrengths between -0.2 and 0.2 are shown in light grey.
+If there are no significant predictors, shown in dark grey")
+
+# alpha plots
+
+plot <- ggplot() +
+  geom_sf(data = added_cols, aes(geometry = geometry, fill = var_long, alpha = est), col = "#00000000") +
+  #scale_fill_manual(values = plot_cols) +
+  facet_grid(cols = vars(var),
+             rows = vars(model)) + 
+  theme_void() +
+  theme(legend.position = "none",
+        plot.title = element_text()) 
+# +
+#   labs(title = "Fig (x): Map of strongest structural attribute predictor strength on yearly productivity metrics by BEC zone for the two proposed models.
+\nStrengths between -0.2 and 0.2 are shown in light grey.
+If there are no significant predictors, shown in dark grey")
+
+ggsave(here::here("outputs", "path_map.png"), plot, bg = "white")
+```
+
+Exploratory Factor Analysis for eSEM
+```{r}
+efa_data <- bec_normal %>%
+  select(basal_area:total_biomass)
+efa_fit <- efa(
+  efa_data,
+  nfactors = 1:2, # global model maxes at 2 factors, 2 is min aic
+  rotation = "geomin",
+  rotation.args = list(geomin.epsilon = .1)
+)
+
+summ <- summary(efa_fit, cutoff = 0.5, dot.cutoff = 0.05)
+
+summ$fit.table
+
+loadings <- efa_fit$loadings %>%
+  as.data.frame() %>%
+  tibble() %>%
+  mutate(item = efa_fit$loadings %>% rownames()) %>%
+  pivot_longer(cols = c(f1, f2),
+               names_to = "latent",
+               values_to = "loading")
+
+anchors = c(f1 = "elev_p95", f2 = "percentage_first_returns_above_2m")
+# anchor using simple rs metrics
+
+loadings_filtered <- loadings %>%
+  filter(loading >= 0.5)
+esem_df <- loadings_filtered %>%
+  mutate(
+    anchor = case_when(
+      latent == "f1" & item == "elev_p95" ~ T,
+      latent == "f2" & item == "percentage_first_returns_above_2m" ~ T,
+      T ~ F
+    ),
+    syntax = ifelse(
+      anchor,
+      glue::glue("{loading}*{item}"),
+      glue::glue("start({loading})*{item}")
+    )
+  ) %>%
+  group_by(latent) %>%
+  summarize(esem = paste0(syntax, collapse = " + ")) %>%
+  mutate(latent = toupper(latent),
+         esem = glue::glue("{latent} =~ {esem}")) 
+esem_mod <- esem_df %>%
+  pull(esem) %>%
+  paste0(collapse = " \n ")
+
+dhi_esem <- paste0("CumDHI ~ elev_cv + F1 + F2
+        VarDHI ~ elev_cv + F1 + F2
+        MinDHI ~ elev_cv + F1 + F2
+       ",
+       esem_mod)
+
+dhi_esem_fit <- cfa(dhi_esem, data = bec_normal) 
 ```
@@
 ```{r}
-vlce_samples <- here::here("data", "samples", "vlce") %>%
-  list.files(full.names = T, pattern = ".shp$") %>%
-  map(read_sf) %>%
-  bind_rows()
+library(lavaanPlot)
 
-vlce_samples
+e_opts <-
+  lavaanPlot::formatting(list(color = "orange"),
+                         list(color = "red"),
+                         list(color = "blue"),
+                         type = "edge")
+
+dhi_esem_fit %>%
+  lavaanPlot::lavaanPlot(model = .,
+                         coefs = T,
+                         edge_options = list(color = "grey"),
+                         sig = 0.05,
+                         covs = T,
+                         graph_options = list(layout = "circo"))
+
+modificationIndices(dhi_esem_fit) %>%
+  tibble() %>% 
+  arrange(desc(epc))
+
+egg <- summary(dhi_esem_fit)
+
+egg$pe
 ```
+
+grouped esems
+
+```{r}
+splits <- bec_normal %>%
+  group_by(bec) %>%
+  group_split()
+
+get_efa <- function(split_df) {
+  # get bec zone to label output df
+  zone <- split_df %>%
+    pull(bec) %>%
+    unique()
+
+  # subset data
+  efa_bec_data <- split_df %>%
+    select(basal_area:total_biomass)
+
+  efas <- map(
+    1:4,
+    .f = function(x) {
+      efa(
+        efa_bec_data,
+        nfactors = x,
+        rotation = "geomin",
+        rotation.args = list(geomin.epsilon = .001)
+      )
+    }
+  )
+
+  summs <- map(
+    efas,
+    .f = function(x) {
+      # print(x)
+      tryCatch({
+        summary(x)
+      }, error = function(e) {
+        NULL
+      }, finally = NULL)
+    }
+  )
+
+  fitmeasures <- map_dfr(
+    summs,
+    .f = function(x) {
+      #nf <- x$nfac
+      x$fit.table %>% as.data.frame() %>% tibble() %>%
+        mutate(nf = x$nfactors)
+    }
+  )
+
+  nfactor_select <- fitmeasures %>%
+    slice_min(aic) %>%
+    pull(nf) %>%
+    as.numeric()
+
+  loadings_wide <- efas[[nfactor_select]]$loadings %>%
+    as.data.frame() %>%
+    tibble() %>%
+    mutate(item = efas[[nfactor_select]]$loadings %>%
+             rownames(),
+           bec = zone)
+
+  loadings_long <- loadings_wide %>%
+    pivot_longer(!c(item, bec),
+                 names_to = "latent",
+                 values_to = "loading")
+
+  item_max_load <- loadings_long %>%
+    group_by(item) %>%
+    slice_max(order_by = loading, n = 1)
+
+  anchors <- loadings_long %>%
+    group_by(item) %>%
+    slice_max(order_by = loading, n = 2) %>%
+    summarize(dif = max(loading) - min(loading)) %>%
+    left_join(item_max_load) %>%
+    group_by(latent) %>%
+    slice_max(order_by = dif, n = 1) %>%
+    select(item, latent, bec) %>%
+    mutate(anchor = T)
+
+  if(nfactor_select == 1) {
+    anchors <- loadings_long %>%
+      slice_max(order_by = loading, n = 1) %>%
+      mutate(anchor = T) %>%
+      select(!loading)
+  }
+
+  out_loading <- loadings_long %>%
+    left_join(anchors) %>%
+    mutate(anchor = ifelse(is.na(anchor), F, anchor))
+  # this contains anchors
+  # number of latent variables
+  # strong loadings for latent variables
+  # bec zone
+  # can also be used to get which 'item's (fs variables) are not included in the latents
+  return(list(efa = efas[[nfactor_select]], loadings = out_loading))
+}
+
+bec_efas <- map(splits, get_efa)
+
+efas <- bec_efas %>%
+  map(.f = function(x) {
+    x$efa
+  })
+
+all_loadings <- bec_efas %>%
+  map_dfr(.f = function(x) x$loading)
+```
+
+```{r loadings-grouped-plot}
+anchor_vars <- all_loadings %>% 
+  filter(anchor) %>% 
+  select(anchor_var = item, bec, latent)
+
+loadings_clean <- all_loadings %>% 
+  filter(loading >= 0.5) %>%
+  left_join(anchor_vars) %>% 
+  mutate(latent_labelled = case_when(anchor_var == "percentage_first_returns_above_2m" ~ "Canopy Cover",
+                                     anchor_var == "elev_cv" ~ "Structural Complexity",
+                                     T ~ "Height and Biomass")) %>%
+  left_join(keys$continuous, by = c("item" = "variable")) %>%
+  select(-divide_by, -unit)
+
+reorder_df <- loadings_clean %>%
+  distinct(var_long) %>%
+  arrange(var_long) %>%
+  bind_cols(order = c(4, 
+                      3, 
+                      1, 
+                      6, 
+                      9, 
+                      5, 
+                      7, 
+                      8, 
+                      2, 
+                      10)) %>%
+  arrange(order)
+
+loadings_clean_order <- left_join(loadings_clean, reorder_df)
+
+single_loads <- loadings_clean_order %>%
+  group_by(item, bec) %>%
+  filter(n() == 1)
+
+multi_loads <- loadings_clean_order %>%
+  group_by(item, bec) %>%
+  filter(n() != 1) %>%
+  group_by(bec) %>%
+  group_split()
+
+latent_plot <- ggplot(data = loadings_clean_order, aes(x = fct_reorder(var_long, order, .desc = T), y = bec)) +
+  geom_tile(data = single_loads, aes(fill = latent_labelled), width = 0.9, height = 0.9)
+
+for (loading in multi_loads) {
+  latent_plot = latent_plot + 
+    geom_tile(data = loading, aes(fill = fct_rev(latent_labelled)),
+            width = 0.90, height = 0.9,
+            position = "dodge")
+}
+
+latent_plot_all <- latent_plot +
+  scale_fill_discrete_c4a_cat(palette = "batlow", reverse = T) +
+  theme(panel.grid = element_blank(),
+        legend.position = "bottom") +
+  labs(x = NULL,
+       y = NULL,
+       fill = "Latent Groups") +
+  coord_flip()
+
+ggsave(here::here("outputs", "latent_groups.png"), plot = latent_plot_all)
+# title = "Fig (x): Groups of latent variables with loadings greater than or equal to 0.5 for each BEC zone.
+Loadings were determined using exploratory factor \nanalysis with up to four latent variables, and selecting the number of possible latent variables with the lowest AIC.
+All forest structure variables were \npotentially included as indicators in each latent variable."
+
+# Exploratory factor analysis shows three groups of latent variables (Figure (x).
+We assigned the groups to latent variables indicating canopy cover / biomass metrics, canopy height / biomass metrics, and structural complexity metrics.
+We assigned the indicators to latent variable groups by examining anchors for each latent variable.
+The anchors were calculated by choosing the indicator variable with the largest difference between the maximum value in a given loading compared to to said indicators loadings in all other latent variables.
+#Bunchgrass, notably, has one latent variable.
+It covers all indicators, except for structural complexity.
+This is possibly due to the small amount of forests in the Bunchgrass BEC zone.
+CDF and CWH are the only BEC zones to include elevation standard deviation in their structural complexity latent variable.
+Notably, these two are both coastal BEC zones, with large trees and high energy availability.
+A structural complexity latent variable is only found in five BEC zones.
+```
+
+```{r}
+splits <- bec_normal %>%
+  group_by(bec) %>%
+  group_split()
+
+get_efa <- function(split_df) {
+  # get bec zone to label output df
+  zone <- split_df %>%
+    pull(bec) %>%
+    unique()
+
+  # subset data
+  efa_bec_data <- split_df %>%
+    select(basal_area:VarDHI)
+
+  efas <- map(
+    1:5,
+    .f = function(x) {
+      efa(
+        efa_bec_data,
+        nfactors = x,
+        rotation = "geomin",
+        rotation.args = list(geomin.epsilon = .001)
+      )
+    }
+  )
+
+  summs <- map(
+    efas,
+    .f = function(x) {
+      # print(x)
+      tryCatch({
+        summary(x)
+      }, error = function(e) {
+        NULL
+      }, finally = NULL)
+    }
+  )
+
+  fitmeasures <- map_dfr(
+    summs,
+    .f = function(x) {
+      #nf <- x$nfac
+      x$fit.table %>% as.data.frame() %>% tibble() %>%
+        mutate(nf = x$nfactors)
+    }
+  )
+
+  nfactor_select <- fitmeasures %>%
+    slice_min(aic) %>%
+    pull(nf) %>%
+    as.numeric()
+
+  loadings_wide <- efas[[nfactor_select]]$loadings %>%
+    as.data.frame() %>%
+    tibble() %>%
+    mutate(item = efas[[nfactor_select]]$loadings %>%
+             rownames(),
+           bec = zone)
+
+  loadings_long <- loadings_wide %>%
+    pivot_longer(!c(item, bec),
+                 names_to = "latent",
+                 values_to = "loading")
+
+  item_max_load <- loadings_long %>%
+    group_by(item) %>%
+    slice_max(order_by = loading, n = 1)
+
+  anchors <- loadings_long %>%
+    group_by(item) %>%
+    slice_max(order_by = loading, n = 2) %>%
+    summarize(dif = max(loading) - min(loading)) %>%
+    left_join(item_max_load) %>%
+    group_by(latent) %>%
+    slice_max(order_by = dif, n = 1) %>%
+    select(item, latent, bec) %>%
+    mutate(anchor = T)
+
+  if(nfactor_select == 1) {
+    anchors <- loadings_long %>%
+      slice_max(order_by = loading, n = 1) %>%
+      mutate(anchor = T) %>%
+      select(!loading)
+  }
+
+  out_loading <- loadings_long %>%
+    left_join(anchors) %>%
+    mutate(anchor = ifelse(is.na(anchor), F, anchor))
+  # this contains anchors
+  # number of latent variables
+  # strong loadings for latent variables
+  # bec zone
+  # can also be used to get which 'item's (fs variables) are not included in the latents
+  return(list(efa = efas[[nfactor_select]], loadings = out_loading))
+}
+
+bec_efas <- map(splits, get_efa)
+
+efas <- bec_efas %>%
+  map(.f = function(x) {
+    x$efa
+  })
+
+all_loadings <- bec_efas %>%
+  map_dfr(.f = function(x) x$loading)
+```
+
+```{r loadings-grouped-plot}
+anchor_vars <- all_loadings %>% 
+  filter(anchor) %>% 
+  select(anchor_var = item, bec, latent)
+
+loadings_clean <- all_loadings %>% 
+  # filter(loading >= 0.5) %>%
+  left_join(anchor_vars) %>% 
+  mutate(latent_labelled = case_when(anchor_var == "percentage_first_returns_above_2m" ~ "Canopy Cover",
+                                     anchor_var == "elev_cv" ~ "Structural Complexity",
+                                     T ~ "Height and Biomass")) %>%
+  left_join(keys$continuous, by = c("item" = "variable")) %>%
+  select(-divide_by,-unit) %>%
+  mutate(var_long = ifelse(is.na(var_long), item, var_long))
+
+reorder_df <- all_loadings %>%
+  left_join(keys$continuous, by = c("item" = "variable")) %>%
+  select(-divide_by,-unit) %>%
+  mutate(var_long = ifelse(is.na(var_long), item, var_long)) %>%
+  distinct(var_long) %>%
+  arrange(var_long) %>%
+  bind_cols(order = c(4,
+                      3,
+                      1,
+                      6,
+                      11,
+                      9,
+                      5,
+                      7,
+                      8,
+                      12,
+                      2,
+                      10,
+                      13)) %>%
+  arrange(order)
+
+loadings_clean_order <- left_join(loadings_clean, reorder_df) %>%
+  filter(loading >= 0.5)
+
+single_loads <- loadings_clean_order %>%
+  group_by(item, bec) %>%
+  filter(n() == 1)
+
+multi_loads <- loadings_clean_order %>%
+  group_by(item, bec) %>%
+  filter(n() != 1) %>%
+  group_by(bec) %>%
+  group_split()
+
+grid_plot <- reorder_df %>%
+  crossing(bec = all_loadings %>% pull(bec))
+
+latent_plot <- ggplot(data = grid_plot, aes(x = fct_reorder(var_long, order, .desc = T), y = bec)) +
+  geom_tile(fill = "00000000") +
+  geom_tile(data = single_loads, aes(fill = latent_labelled), width = 0.9, height = 0.9)
+
+for (loading in multi_loads) {
+  latent_plot = latent_plot + 
+    geom_tile(data = loading, aes(fill = fct_rev(latent_labelled)),
+            width = 0.90, height = 0.9,
+            position = "dodge")
+}
+
+latent_dhis <- latent_plot +
+  scale_fill_discrete_c4a_cat(palette = "batlow", reverse = T) +
+  theme(panel.grid = element_blank(),
+        legend.position = "bottom") +
+  labs(x = NULL,
+       y = NULL,
+       fill = "Latent Groups") +
+  coord_flip()
+# Exploratory factor analysis shows three groups of latent variables (Figure (x).
+We assigned the groups to latent variables indicating canopy cover / biomass metrics, canopy height / biomass metrics, and structural complexity metrics.
+We assigned the indicators to latent variable groups by examining anchors for each latent variable.
+The anchors were calculated by choosing the indicator variable with the largest difference between the maximum value in a given loading compared to to said indicators loadings in all other latent variables.
+#Bunchgrass, notably, has one latent variable.
+It covers all indicators, except for structural complexity.
+This is possibly due to the small amount of forests in the Bunchgrass BEC zone.
+CDF and CWH are the only BEC zones to include elevation standard deviation in their structural complexity latent variable.
+Notably, these two are both coastal BEC zones, with large trees and high energy availability.
+A structural complexity latent variable is only found in five BEC zones.
+
+ggsave(here::here("outputs", "latent_dhis.png"), plot = latent_dhis, device = "png")
+```
+
+```{r}
+latent_boxplot <- loadings_clean %>%
+  left_join(reorder_df) %>%
+  ggplot(aes(x = fct_reorder(var_long, order, .desc = T), y = loading, col = var_long)) +
+  geom_boxplot(outlier.alpha = 0) +
+  geom_point(position = position_jitter(width = .25, height = 0), alpha = 0.5) +
+  geom_hline(yintercept = 0.5, col = "red", linetype = "dashed") +
+  labs(x = "Latent Variables",
+       y = "Loadings",
+       col = NULL) +
+
+  facet_wrap(~latent_labelled) +
+  coord_flip() +
+  theme(legend.position = "bottom",
+        panel.grid = element_blank()) +
+  scale_color_discrete_c4a_cat(palette = "poly.dark24")
+
+         # title = "Fig (x): Boxplots of latent variable loadings across BEC zones as determined by exploratory factor analysis.
+Red dashed lines show \nlatent variable indicator cutoff before being used as starting values for the SEM models.") +
+
+# Figure (x) shows boxplots of latent variable loadings from forest structure indicator variables.
+A maximum of three valid latent variables were identified across the sixteen BEC zones.
+We assigned indicators variables to latent variable groups by examining the anchoring variables for each latent variable.
+The anchoring variables were calculated by choosing the indicator variable with the largest difference between the maximum value in a given loading compared to said indicators loadings in all other latent variables.
+ggsave(here::here("outputs", "latent_boxplot.png"), plot = latent_boxplot, device = "png")
+```
+
+```{r}
+loadings_clean %>%
+  filter(loading >= 0.5) %>%
+  ggplot(aes(x = loading, y = latent_labelled, fill = var_long)) +
+  geom_col(position = "dodge") +
+  scale_color_discrete_c4a_cat(palette = "poly.dark24") +
+  facet_wrap(~bec)
+```
+
+```{r}
+num_latents <- all_loadings %>%
+  filter(loading >= 0.5) %>%
+  group_by(bec) %>%
+  count(latent) %>%
+  summarize(n = n())
+
+all_loadings %>%
+  filter(anchor)
+
+# check how many forest strcuture latent variables in each
+# are latent variables static or variable
+# see if anchors are static or variable
+
+loadings_filtered <- loadings %>%
+  filter(loading >= 0.5)
+
+bec_nest <- bec_normal %>% 
+  group_by(bec) %>% 
+  nest()
+
+for_esem <- all_loadings %>%
+  filter(loading >= 0.5)
+
+model_starts <- for_esem %>% 
+  distinct(bec, latent) %>% 
+  group_by(bec) %>% 
+  summarize(latents = paste0(latent, collapse = " + ")) %>%
+  mutate(model_start = glue::glue("CumDHI ~ {latents} \n VarDHI ~ {latents} \n MinDHI ~ {latents}")) %>%
+  select(!latents)
+
+bec_esem_syntax <- for_esem %>%
+  mutate(
+    loading = round(loading, 4),
+    syntax = ifelse(
+      anchor,
+      glue::glue("{loading}*{item}"),
+      glue::glue("start({loading})*{item}")
+    )
+  ) %>%
+  group_by(bec, latent) %>%
+  summarize(syntax = paste0(syntax, collapse = " + ")) %>%
+  mutate(syntax = glue::glue("{latent} =~ {syntax}")) %>%
+  summarize(syntax = paste0(syntax, collapse = " \n ")) %>%
+  left_join(model_starts) %>%
+  mutate(syntax = glue::glue("{model_start} \n {syntax}")) %>%
+  left_join(bec_nest) %>%
+  select(bec, data, syntax)
+
+bec_esems <- bec_esem_syntax %>% 
+  filter(bec != "IMA") %>%
+  mutate(out = map2(.x = syntax, .y = data, .f = sem))
+```
+
+```{r}
+
+esem_df <- loadings_filtered %>%
+  mutate(
+    anchor = case_when(
+      latent == "f1" & item == "elev_p95" ~ T,
+      latent == "f2" & item == "percentage_first_returns_above_2m" ~ T,
+      T ~ F
+    ),
+    syntax = ifelse(
+      anchor,
+      glue::glue("{loading}*{item}"),
+      glue::glue("start({loading})*{item}")
+    )
+  ) %>%
+  group_by(latent) %>%
+  summarize(esem = paste0(syntax, collapse = " + ")) %>%
+  mutate(latent = toupper(latent),
+         esem = glue::glue("{latent} =~ {esem}")) 
+esem_mod <- esem_df %>%
+  pull(esem) %>%
+  paste0(collapse = " \n ")
+
+dhi_esem <- paste0("CumDHI ~ elev_cv + F1 + F2
+        VarDHI ~ elev_cv + F1 + F2
+        MinDHI ~ elev_cv + F1 + F2
+       ",
+       esem_mod)
+
+dhi_esem_fit <- cfa(dhi_esem, data = bec_normal) 
+```
+
+```{r}
+cfa_syn <- "F1 =~ elev_p95 + elev_mean + elev_stddev + loreys_height
+            F2 =~ percentage_first_returns_above_2m + percentage_first_returns_above_mean
+            F3 =~ total_biomass + basal_area + gross_stem_volume"
+
+cfa_sem_syn <- "F1 =~ elev_p95 + elev_mean + elev_stddev + loreys_height
+            F2 =~ percentage_first_returns_above_2m + percentage_first_returns_above_mean
+            F3 =~ total_biomass + basal_area + gross_stem_volume
+            F4 =~ CumDHI + VarDHI + MinDHI
+            F4 ~ F1 + F2 + F3"
+
+cfa_fit <- cfa_syn %>%
+  cfa(data = bec_normal)
+
+cfa_sem_fit <- cfa_sem_syn %>%
+  sem(data = bec_normal)
+
+lavaanPlot(cfa_sem_fit,
+           coefs = T,
+           covs = T,
+           graph_options = list(layout = "circo"),
+           edge_options = list(color = "grey"))
+```
+
+```{r}
+one_latent_syn <- "F1 =~ elev_cv + total_biomass + elev_p95 + percentage_first_returns_above_2m
+CumDHI ~ F1
+VarDHI ~ F1
+MinDHI ~ F1"
+
+one_latent_fit <- sem(one_latent_syn, data = bec_normal)
+
+lavaanPlot(one_latent_fit,
+           coefs = T,
+           covs = T,
+           graph_options = list(layout = "circo"),
+           edge_options = list(color = "grey")
+           )
+```
+
+```{r}
+library(rdacca.hp)
+
+rdacca.hp(bec_normal$CumDHI, 
+          bec_normal %>% select(basal_area:total_biomass))
+
+rdacca.hp(bec_normal$VarDHI, 
+          bec_normal %>% select(basal_area:total_biomass))
+
+rdacca.hp(bec_normal$MinDHI, 
+          bec_normal %>% select(basal_area:total_biomass))
+```
+
+```{r}
+library(dplyr)
+
+data <- data.frame(data_id = c(letters[1:5], 
+                               letters[1:5]),
+                   data_group =  c(replicate(5, "Group1"),
+                                   replicate(5, "Group2")),
+                   data_value = c(0:4, replicate(5,2)))
+
+data_unique <- data %>% ## non-overlapping data
+  group_by(data_id, data_value) %>%
+  filter(n() == 1)
+
+data_shared <- data %>% ## overlapping data
+  group_by(data_id, data_value) %>%
+  filter(n() != 1)
+
+ggplot(data,
+       aes(x = data_value, y = data_id)) + 
+  geom_tile(data = data_unique, aes(fill = data_group, group = data_group), 
+            width = 0.4, height = 0.8) + ## non-overlapping data
+  geom_tile(data = data_shared, aes(fill = data_group, group = data_group), 
+            width = 0.4, height = 0.8, 
+            position = "dodge") ## non-overlapping data
+```

### 2023-03-20 scripts/manuscript/manuscript.qmd
first revisions to ncc 2023-03-17

@@
         corresponding: true
+  - name: Margaret E. Andrew
+    email: M.Andrew@murdoch.edu.au
+    affiliations: 
+        - id: murd
+          name: Murdoch University
+          department: Environmental and Conservation Sciences and Harry Butler Institute
+          city: Murdoch, WA, Australia
+          postal-code: 6150
   - name: Nicholas C. Coops
@@
           postal-code: V6T 1Z4
-  - name: Margaret E. Andrew
-    email: M.Andrew@murdoch.edu.au
-    affiliations: 
-        - id: murd
-          name: Murdoch University
-          department: Environmental and Conservation Sciences and Harry Butler Institute
-          city: Murdoch, WA, Australia
-          postal-code: 6150
   - name: Stephen S. Ban
@@
 abstract: |
-  insert abstract here
+  [insert abstract here - nicholas ignore the formatting on the authors for now, i can output to pdf and have it formatted nicely for submission]
 keywords: 
@@
   - forest productivity
+  - dynamic habitat indices
 date: last-modified
@@
     wrap: 72
+execute: 
+  echo: false
+fig-width: 300
 ---
@@
 information on the remaining two species EBV classes, this information
-is typically acquired with an *ad-hoc* approach. This collects high
-resolution, small spatial extent information, rather than the global or
-regional scales required for biodiversity assessment. Community
-composition falls into a similar dilemma, requiring species population
-information which necessitates high resolution spatial data.
+is typically acquired with an *ad-hoc* approach requiring high spatial
+and spectral resolution data. This data can often only be collected at
+local extents, rather than the global or regional scales required for
+biodiversity trend assessment [@valdez2023]. Community composition falls
+into a similar dilemma, requiring species population information which
+necessitates high resolution spatial data.
 
@@
 species diversity metrics [@macarthur1961; @goetz2007], however, other
-clades (and habitats), have been used [@davies2014; @nelson2005]. Many
-metrics derived from lidar remote sensing have been used as local
+clades (and habitats), have also been used [@davies2014; @nelson2005].
+Many metrics derived from lidar remote sensing have been used as local
 indicators of biodiversity, including canopy cover, canopy height,
@@
 
+Some vegetation structure metrics are simpler, and more accurate, to
+calculate than others [@coops2021]. These basic metrics, such as canopy
+height and canopy cover, can then be used to estimate additional
+structure metrics, such as basal area or total biomass. Canopy height
+and cover are commonly used as indicators of vertical and horizontal
+variation, respectively. Recently, more attention has been paid to
+internal structural complexity metrics, which can be more difficult and
+time consuming to generate [@coops2021; @ma2022], but have been shown to
+have stronger linkages with biodiversity [@guo2017], and productivity
+[@ali2019].
+
 Linkages between forest structure and productivity (namely, vegetation
@@
 metrics [@ali2019]. These relationships, their shapes, and their
-strengths have been attributed to multiple possible hypotheses and can
-vary based on environmental conditions [@ali2019]. The relationship
-between forest structural diversity metrics and annual productivity
-summaries has yet to be fully examined, including the DHIs.
+strengths have been attributed to multiple possible hypotheses and have
+been shown to vary based on environmental conditions [@ali2019]. The
+relationship between forest structural diversity metrics and annual
+productivity summaries has yet to be fully examined, including the DHIs.
 
-Some vegetation structure metrics are simpler, and more accurate, to
-calculate than others [@coops2021]. These basic metrics, such as canopy
-height and canopy cover,can then be used to estimate additional
-structure metrics, such as basal area or total biomass. Canopy height
-and cover are commonly used as an indicator of vertical and horizontal
-variation, respectively. Recently, more attention has been paid to
-internal structural complexity metrics, which can be more difficult and
-time consuming to generate [@coops2021; @ma2022] .
+Understanding the complementarity between potential EBVs is an integral
+component of their creation [@skidmore2021]. Biodiversity has been shown
+to be linked to both forest structure [@guo2017; @gao2014], and
+productivity [@radeloff2019] at a range of scales. Research has
+highlighted that the relationship between productivity and biodiversity
+is reciprocal [@worm2003], making productivity a suitable biodiversity
+indicator. In forested ecosystems, stand structural attributes have also
+been shown to influence productivity, with a range of responses
+dependent on environmental conditions [@ali2019].
 
-In this study, we seek to untangle the relationship between two EBVs:
-forest structure diversity metrics and yearly summaries of forest
-productivity. To accomplish this, we assess this relationship using path
-analysis to assess the direct and indirect (as mediated by more complex
-forest structural diversity metrics) effects of commonly collected
-forest structural metrics (canopy height and canopy cover) on yearly
-productivity summaries. Doing so will in turn assess their utility as
-complementary EBVs. We ran this analysis separately across four forested
-land covers and the forested ecosystems of British Columbia, Canada.
+Structural equation modelling and path analyses have been commonly used
+in ecology to assess the causal effects behind various hypotheses
+[@fan2016; @grace2010], including the relationship between forest
+structure, functioning, and biodiversity [@ali2019] . In this paper we
+seek to untangle the relationship between two EBVs: forest structure
+diversity metrics and yearly summaries of forest productivity. To
+accomplish this, we assess this relationship using path analysis to
+assess the direct and indirect (as mediated by more complex forest
+structural diversity metrics) effects of commonly collected forest
+structural metrics (canopy height and canopy cover) on yearly
+productivity summaries. Further, we use exploratory structural equation
+modelling (ESEM) to identify the latent variables driving the
+productivity metrics. We run this analysis both globally, and stratified
+by the forested ecosystems of British Columbia, Canada, and compare
+models using AIC scores and global fit measures across the ecosystems.
+The results from this study will assess the linkages and complementarity
+between two biodiversity indicators.
 
@@
 subzones, variants, and phases based on microclimate, precipitation, and
-topography.
+topography [@pojar1987].
+
+{insert figure showing ecosystems of interest} need to remake figure
+showing ecosystems, highlighting that bunchgrass isn't included
+
+{table showing elevations and climate ranges for each included bec zone}
 
@@
 
+### NTEMS
+
+### Dynamic Habitat Indices
+
+The Dynamic Habitat Indices are a set of satellite remote sensing
+derived productivity variables that summarize the cumulative amount of
+available energy, the minimum available energy, and the variation in
+available energy throughout a given year [@radeloff2019]. The DHIs have
+previously been produced at a global extent using MODIS imagery
+[@radeloff2019], however, recent studies have examined how these indices
+can be constructed at a finer spatial resolution using multi-annual
+Landsat imagery.
+
 ## Sampling
+
+We conducted model based sampling across the fifteen forest dominated
+ecosystems found within British Columbia \[BC figure reference, I have
+not made this figure yet\]. Samples were randomly selected within BEC
+zones of interest alongside multiple criteria. Each sampled pixel had to
+have a forested land cover class (coniferous, deciduous, mixed-wood, or
+wetland-treed), and be surrounded by the same land cover class.
+Additionally, each pixel had to have a coefficient of variation less
+than 0.5 in the Lorey's height and canopy cover forest structural
+metrics. These focal analyses were preprocessed in R and Python, and
+randomly sampled across a masked suitability raster for each BEC zone
+using the *sgsR* R package version 1.3.4 [@goodbody2023]. A maximum of
+3000 samples were taken from each BEC zone.
+
+\[note to ncc; I've stopped doing this sampling step as Meg said she
+didn't feel it was necessary\]
+
+The imputation method used to generate the forest structure layers leads
+to many duplicated values as there are a limited number of samples to
+predict from [@matasci2018]. A discussion of imputation approaches to
+large area lidar attribute modelling can be found in @coops2021. To
+account for duplicated input values in our modelling we averaged the
+endogenous variables within each ecosystem for each uniquely imputed set
+of forest attributes, forest type, and ecosystem. Due to variations in
+ecosystem size, differing numbers of samples were generated for each
+ecosystem. To account for this during model comparison, each ecosystem
+and land cover sample set was downsampled to the minimum number of
+samples for a single BEC zone (in this case, 170). For the forest type
+models, we attempted to select equal numbers of observations from each
+BEC zone. However, some forest types are rarely found in certain zones.
+To account for this, if an equal number of samples from each BEC zone
+were not available, the entire sample from that zone was taken. The
+amount of samples taken from other zones was then increased in order to
+collect a total number equal to the minimum number of samples in a
+single BEC zones (170). To meet the normality assumptions of path
+analysis, all variables were natural-log transformed and standardized,
+as per @grace2016. Variables containing zeros were natural-log plus one
+transformed.
 
@@
 
+### Path Analysis
+
+To determine the relationships between simple lidar derived forest
+structural attributes, complex/derived forest structural attributes, and
+forest productivity, we used path analysis to analyze two causal models
+(@fig-pathdag). To determine primary drivers of the three DHIs, we will
+summarize the predictors across ecosystems by counting the strongest
+predictor in each BEC zone. This will determine if the primary driver
+for each facet of the DHI is simple or complex, and allow us to assess
+ecosystem differences.
+
+```{r}
+#| label: fig-pathdag
+#| fig-cap: Proposed path diagrams
+#| warning: false
+#| echo: false
+
+knitr::include_graphics(here::here("outputs", "path_dags.png"))
+```
+
+### Exploratory Structural Equation Modelling
+
+Secondly, we will use exploratory structural equation modelling
+[@marsh2020; @asparouhov2009] to identify latent forest structural
+variables within the data, and use these latent variables to predict the
+dynamic habitat indices. Exploratory structural equation modelling is a
+combination of exploratory factor analysis (EFA) and structural equation
+modelling, which relaxes the strict requirement of zero cross-loadings
+in confirmatory factor analysis and allows for less strict measurement
+models to be used. ESEM is used when it is known that there is a latent
+structure to the data, but the specific indicator variables have not yet
+been determined.
+
+One advantage of ESEM is that it can create varying numbers of latent
+variables. For our analysis, we first ran exploratory factor analysis on
+all forest structural attributes both globally and by BEC zone with
+between 1-4 potential latent variables. We chose the most parsimonious
+EFA model with the lowest AIC scores for each ecosystem, leading to
+varying numbers of latent variables. We then determined the anchoring
+indicator for each latent variable. The anchors were calculated by
+choosing the indicator variable with the largest difference between the
+maximum value in a given loading compared to to said indicators loadings
+in all other latent variables. Each anchor variable was then assigned to
+a named group (Canopy Cover, Height and Biomass, and Structural
+Complexity).
+
+Following the EFA, we used structural equation modelling, with latent
+variables loadings determined by the EFA. We also filtered the loadings
+to be greater than or equal to 0.5, allowing us to examine the number
+and composition of latent variables found in each ecosystem. Following
+this, we predicted the DHIs in a single SEM with covariances between the
+DHIs. If a forest structural attribute did not end up in a latent
+variable, it was included in the structural equation model as a DHI
+predictor, without a latent variable as a mediator.
+
 # Results
+
+## Path Analysis
+
+```{r}
+#| label: fig-pathbar-prop
+#| fig-cap: Proportion of strongest signficant forest structural predictors of yearly productivity metrics in models based on canopy cover vs canopy height.
+#| warning: false
+#| echo: false
+
+knitr::include_graphics(here::here("outputs", "path_bars.png"))
+```
+
+```{r}
+#| label: fig-pathbar-pred
+#| fig-cap: Bar plot of strongest predictors in path analysis for each BEC zone and predicted DHI variable.
+#| warning: false
+#| echo: false
+
+knitr::include_graphics(here::here("outputs", "strongest_predictor_bar.png"))
+```
+
+```{r}
+#| label: fig-pathmap
+#| fig-cap: Map of strongest structural attribute predictor strength on yearly productivity metrics by BEC zone for the two proposed models. If there are no significant predictors, shown in dark grey
+#| warning: false
+#| echo: false
+
+knitr::include_graphics(here::here("outputs", "path_map_edit.png"))
+```
+
+## Exploratory Structural Equation Modelling
+
+I also want to run exploratory SEM with and without the DHI's to see
+what grouping it goes in. MinDHI has been breaking because some zones
+only have 0 as mindhi (makes sense, they are in the alpine).
+
+```{r}
+#| label: fig-esem-latent
+#| fig-cap: Groups of latent variables with loadings greater than or equal to 0.5 for each BEC zone.
+Loadings were determined using exploratory factor analysis with up to four latent variables, and selecting the number of possible latent variables with the lowest AIC.
+All forest structure variables were potentially included as indicators in each latent variable.
+#| warning: false
+#| echo: false
+
+knitr::include_graphics(here::here("outputs", "latent_groups.png"))
+```
+
+Exploratory factor analysis shows three groups of latent variables
+(@fig-esem-latent). We assigned the groups to latent variables
+indicating canopy cover / biomass metrics, canopy height / biomass
+metrics, and structural complexity metrics. Indicators to latent
+variable groups by examining anchors for each latent variable. The
+anchors were calculated by choosing the indicator variable with the
+largest difference between the maximum value in a given loading compared
+to to said indicators loadings in all other latent variables.
+
+Both the Ponderosa Pine (PP) and Sub-boreal Pine -- Spruce zones have
+one latent variable. Coastal Western Hemlock (CWH), Engelmann Spurce --
+Subalpine Fir (ESSF), and Montane Spruce (MS) are the only zones to have
+structural complexity based latent variables. CWH is the only zone to
+include elevation standard deviation in the structural complexity latent
+variable.
+
+```{r}
+#| label: fig-esem-latent-dhi
+#| fig-cap: Groups of latent variables with loadings greater than or equal to 0.5 for each BEC zone, including the DHIs.
+Loadings were determined using exploratory factor analysis with up to five latent variables, and selecting the number of possible latent variables with the lowest AIC.
+All forest structure variables were potentially included as indicators in each latent variable.
+#| warning: false
+#| echo: false
+
+knitr::include_graphics(here::here("outputs", "latent_dhis.png"))
+```
+
+Exploratory factor analysis shows three groups of latent variables. We
+assigned these groups to attributes based on the indicator variables in
+each latent variable, indicating canopy cover, height and biomass, and
+structural complexity. When including the DHIs in this EFA, it is rare
+that they are included with latent variables with large loadings. The
+cumulative DHI is an indicator for the canopy cover latent varible in
+the IDF (Interior Douglas-fir) and MS (Montane Spruce) BEC zones. The
+DHIs are not included in any other latent variable, indicating that they
+compose different information than the forest structural variables.
+
+```{r}
+#| label: fig-latent-boxplots
+#| fig-cap: Boxplots of latent variable loadings across BEC zones as determined by exploratory factor analysis.
+Red dashed lines show latent variable indicator cutoff before being used as starting values for the SEM models.
+#| warning: false
+#| echo: false
+
+knitr::include_graphics(here::here("outputs", "latent_boxplot.png"))
+```
+
+@fig-latent-boxplots shows boxplots of latent variable loadings from
+forest structure indicator variables. A maximum of three valid latent
+variables were identified across the sixteen BEC zones. We assigned
+indicators variables to latent variable groups by examining the
+anchoring variables for each latent variable. The anchoring variables
+were calculated by choosing the indicator variable with the largest
+difference between the maximum value in a given loading compared to said
+indicators loadings in all other latent variables. The only DHI variable
+that was included in the latent variables after subsetting the latent
+variables to only include strong loadings was the Cumulative DHI in the
+canopy cover latent variable.
+
+I need to actually make the DAGs and plot the parameters, but I haven't
+gotten there yet. I'll plot the global one to show what it looks like,
+then aggregate it in some way - probably similar to the path analysis. I
+also may include the variable partitioning that Meg talked about in the
+email if I can get it working.
 

### 2023-03-20 scripts/manuscript/manuscript.tex
first revisions to ncc 2023-03-17

@@
   pdfkeywords={structural equation modelling, remote
-sensing, landsat, forest structure, forest productivity},
+sensing, landsat, forest structure, forest productivity, dynamic habitat
+indices},
   colorlinks=true,
@@
 \begin{abstract}
-This is the abstract. Lorem ipsum dolor sit amet, consectetur adipiscing
-elit. Vestibulum augue turpis, dictum non malesuada a, volutpat eget
-velit. Nam placerat turpis purus, eu tristique ex tincidunt et. Mauris
-sed augue eget turpis ultrices tincidunt. Sed et mi in leo porta
-egestas. Aliquam non laoreet velit. Nunc quis ex vitae eros aliquet
-auctor nec ac libero. Duis laoreet sapien eu mi luctus, in bibendum leo
-molestie. Sed hendrerit diam diam, ac dapibus nisl volutpat vitae.
-Aliquam bibendum varius libero, eu efficitur justo rutrum at. Sed at
-tempus elit.
+{[}insert abstract here - nicholas ignore the formatting on the authors
+for now, i can output to pdf and have it formatted nicely for
+submission{]}
 \end{abstract}
@@
     structural equation modelling \sep remote
-sensing \sep landsat \sep forest structure \sep 
-    forest productivity
+sensing \sep landsat \sep forest structure \sep forest
+productivity \sep 
+    dynamic habitat indices
 \end{keyword}
 \end{frontmatter}
-    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, sharp corners, interior hidden, frame hidden, enhanced, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
+    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable, enhanced, sharp corners]}{\end{tcolorbox}}\fi
 
@@
 classes, this information is typically acquired with an \emph{ad-hoc}
-approach. This collects high resolution, small spatial extent
-information, rather than the global or regional scales required for
-biodiversity assessment. Community composition falls into a similar
-dilemma, requiring species population information which necessitates
-high resolution spatial data.
+approach requiring high spatial and spectral resolution data. This data
+can often only be collected at local extents, rather than the global or
+regional scales required for biodiversity trend assessment
+\citep{valdez2023}. Community composition falls into a similar dilemma,
+requiring species population information which necessitates high
+resolution spatial data.
 
@@
 species diversity metrics \citep{macarthur1961, goetz2007}, however,
-other clades (and habitats), have been used
+other clades (and habitats), have also been used
 \citep{davies2014, nelson2005}. Many metrics derived from lidar remote
@@
 
+Some vegetation structure metrics are simpler, and more accurate, to
+calculate than others \citep{coops2021}. These basic metrics, such as
+canopy height and canopy cover, can then be used to estimate additional
+structure metrics, such as basal area or total biomass. Canopy height
+and cover are commonly used as indicators of vertical and horizontal
+variation, respectively. Recently, more attention has been paid to
+internal structural complexity metrics, which can be more difficult and
+time consuming to generate \citep{coops2021, ma2022}, but have been
+shown to have stronger linkages with biodiversity \citep{guo2017}, and
+productivity \citep{ali2019}.
+
 Linkages between forest structure and productivity (namely, vegetation
@@
 relationships, their shapes, and their strengths have been attributed to
-multiple possible hypotheses and can vary based on environmental
-conditions \citep{ali2019}. The relationship between forest structural
-diversity metrics and annual productivity summaries has yet to be fully
-examined, including the DHIs.
+multiple possible hypotheses and have been shown to vary based on
+environmental conditions \citep{ali2019}. The relationship between
+forest structural diversity metrics and annual productivity summaries
+has yet to be fully examined, including the DHIs.
 
-Some vegetation structure metrics are simpler, and more accurate, to
-calculate than others \citep{coops2021}. These basic metrics, such as
-canopy height and canopy cover,can then be used to estimate additional
-structure metrics, such as basal area or total biomass. Canopy height
-and cover are commonly used as an indicator of vertical and horizontal
-variation, respectively. Recently, more attention has been paid to
-internal structural complexity metrics, which can be more difficult and
-time consuming to generate \citep{coops2021, ma2022} .
+Understanding the complementarity between potential EBVs is an integral
+component of their creation \citep{skidmore2021}. Biodiversity has been
+shown to be linked to both forest structure \citep{guo2017, gao2014},
+and productivity \citep{radeloff2019} at a range of scales. Research has
+highlighted that the relationship between productivity and biodiversity
+is reciprocal \citep{worm2003}, making productivity a suitable
+biodiversity indicator. In forested ecosystems, stand structural
+attributes have also been shown to influence productivity, with a range
+of responses dependent on environmental conditions \citep{ali2019}.
 
-In this study, we seek to untangle the relationship between two EBVs:
-forest structure diversity metrics and yearly summaries of forest
-productivity. To accomplish this, we assess this relationship using path
-analysis to assess the direct and indirect (as mediated by more complex
-forest structural diversity metrics) effects of commonly collected
-forest structural metrics (canopy height and canopy cover) on yearly
-productivity summaries. Doing so will in turn assess their utility as
-complementary EBVs. We ran this analysis separately across four forested
-land covers and the forested ecosystems of British Columbia, Canada.
+Structural equation modelling and path analyses have been commonly used
+in ecology to assess the causal effects behind various hypotheses
+\citep{fan2016, grace2010}, including the relationship between forest
+structure, functioning, and biodiversity \citep{ali2019} . In this paper
+we seek to untangle the relationship between two EBVs: forest structure
+diversity metrics and yearly summaries of forest productivity. To
+accomplish this, we assess this relationship using path analysis to
+assess the direct and indirect (as mediated by more complex forest
+structural diversity metrics) effects of commonly collected forest
+structural metrics (canopy height and canopy cover) on yearly
+productivity summaries. Further, we use exploratory structural equation
+modelling (ESEM) to identify the latent variables driving the
+productivity metrics. We run this analysis both globally, and stratified
+by the forested ecosystems of British Columbia, Canada, and compare
+models using AIC scores and global fit measures across the ecosystems.
+The results from this study will assess the linkages and complementarity
+between two biodiversity indicators.
 
@@
 subzones, variants, and phases based on microclimate, precipitation, and
-topography.
+topography \citep{pojar1987}.
+
+\{insert figure showing ecosystems of interest\} need to remake figure
+showing ecosystems, highlighting that bunchgrass isn't included
+
+\{table showing elevations and climate ranges for each included bec
+zone\}
 
@@
 
+\hypertarget{ntems}{%
+\subsubsection{NTEMS}\label{ntems}}
+
+\hypertarget{dynamic-habitat-indices}{%
+\subsubsection{Dynamic Habitat Indices}\label{dynamic-habitat-indices}}
+
+The Dynamic Habitat Indices are a set of satellite remote sensing
+derived productivity variables that summarize the cumulative amount of
+available energy, the minimum available energy, and the variation in
+available energy throughout a given year \citep{radeloff2019}. The DHIs
+have previously been produced at a global extent using MODIS imagery
+\citep{radeloff2019}, however, recent studies have examined how these
+indices can be constructed at a finer spatial resolution using
+multi-annual Landsat imagery.
+
 \hypertarget{sampling}{%
 \subsection{Sampling}\label{sampling}}
+
+We conducted model based sampling across the fifteen forest dominated
+ecosystems found within British Columbia {[}BC figure reference, I have
+not made this figure yet{]}. Samples were randomly selected within BEC
+zones of interest alongside multiple criteria. Each sampled pixel had to
+have a forested land cover class (coniferous, deciduous, mixed-wood, or
+wetland-treed), and be surrounded by the same land cover class.
+Additionally, each pixel had to have a coefficient of variation less
+than 0.5 in the Lorey's height and canopy cover forest structural
+metrics. These focal analyses were preprocessed in R and Python, and
+randomly sampled across a masked suitability raster for each BEC zone
+using the \emph{sgsR} R package version 1.3.4 \citep{goodbody2023}. A
+maximum of 3000 samples were taken from each BEC zone.
+
+{[}note to ncc; I've stopped doing this sampling step as Meg said she
+didn't feel it was necessary{]}
+
+The imputation method used to generate the forest structure layers leads
+to many duplicated values as there are a limited number of samples to
+predict from \citep{matasci2018}. A discussion of imputation approaches
+to large area lidar attribute modelling can be found in
+\citet{coops2021}. To account for duplicated input values in our
+modelling we averaged the endogenous variables within each ecosystem for
+each uniquely imputed set of forest attributes, forest type, and
+ecosystem. Due to variations in ecosystem size, differing numbers of
+samples were generated for each ecosystem. To account for this during
+model comparison, each ecosystem and land cover sample set was
+downsampled to the minimum number of samples for a single BEC zone (in
+this case, 170). For the forest type models, we attempted to select
+equal numbers of observations from each BEC zone. However, some forest
+types are rarely found in certain zones. To account for this, if an
+equal number of samples from each BEC zone were not available, the
+entire sample from that zone was taken. The amount of samples taken from
+other zones was then increased in order to collect a total number equal
+to the minimum number of samples in a single BEC zones (170). To meet
+the normality assumptions of path analysis, all variables were
+natural-log transformed and standardized, as per \citet{grace2016}.
+Variables containing zeros were natural-log plus one transformed.
 
@@
 
+\hypertarget{path-analysis}{%
+\subsubsection{Path Analysis}\label{path-analysis}}
+
+To determine the relationships between simple lidar derived forest
+structural attributes, complex/derived forest structural attributes, and
+forest productivity, we used path analysis to analyze two causal models
+(Figure~\ref{fig-pathdag}). To determine primary drivers of the three
+DHIs, we will summarize the predictors across ecosystems by counting the
+strongest predictor in each BEC zone. This will determine if the primary
+driver for each facet of the DHI is simple or complex, and allow us to
+assess ecosystem differences.
+
+\begin{figure}
+
+{\centering \includegraphics[width=1.6in,height=\textheight]{../../outputs/path_dags.png}
+
+}
+
+\caption{\label{fig-pathdag}Proposed path diagrams}
+
+\end{figure}
+
+\hypertarget{exploratory-structural-equation-modelling}{%
+\subsubsection{Exploratory Structural Equation
+Modelling}\label{exploratory-structural-equation-modelling}}
+
+Secondly, we will use exploratory structural equation modelling
+\citep{marsh2020, asparouhov2009} to identify latent forest structural
+variables within the data, and use these latent variables to predict the
+dynamic habitat indices. Exploratory structural equation modelling is a
+combination of exploratory factor analysis (EFA) and structural equation
+modelling, which relaxes the strict requirement of zero cross-loadings
+in confirmatory factor analysis and allows for less strict measurement
+models to be used. ESEM is used when it is known that there is a latent
+structure to the data, but the specific indicator variables have not yet
+been determined.
+
+One advantage of ESEM is that it can create varying numbers of latent
+variables. For our analysis, we first ran exploratory factor analysis on
+all forest structural attributes both globally and by BEC zone with
+between 1-4 potential latent variables. We chose the most parsimonious
+EFA model with the lowest AIC scores for each ecosystem, leading to
+varying numbers of latent variables. We then determined the anchoring
+indicator for each latent variable. The anchors were calculated by
+choosing the indicator variable with the largest difference between the
+maximum value in a given loading compared to to said indicators loadings
+in all other latent variables. Each anchor variable was then assigned to
+a named group (Canopy Cover, Height and Biomass, and Structural
+Complexity).
+
+Following the EFA, we used structural equation modelling, with latent
+variables loadings determined by the EFA. We also filtered the loadings
+to be greater than or equal to 0.5, allowing us to examine the number
+and composition of latent variables found in each ecosystem. Following
+this, we predicted the DHIs in a single SEM with covariances between the
+DHIs. If a forest structural attribute did not end up in a latent
+variable, it was included in the structural equation model as a DHI
+predictor, without a latent variable as a mediator.
+
 \hypertarget{results}{%
 \section{Results}\label{results}}
+
+\hypertarget{path-analysis-1}{%
+\subsection{Path Analysis}\label{path-analysis-1}}
+
+\begin{figure}
+
+{\centering \includegraphics[width=14.74in,height=\textheight]{../../outputs/path_bars.png}
+
+}
+
+\caption{\label{fig-pathbar-prop}Proportion of strongest signficant
+forest structural predictors of yearly productivity metrics in models
+based on canopy cover vs canopy height.}
+
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics[width=14.74in,height=\textheight]{../../outputs/strongest_predictor_bar.png}
+
+}
+
+\caption{\label{fig-pathbar-pred}Bar plot of strongest predictors in
+path analysis for each BEC zone and predicted DHI variable.}
+
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics[width=14.74in,height=\textheight]{../../outputs/path_map_edit.png}
+
+}
+
+\caption{\label{fig-pathmap}Map of strongest structural attribute
+predictor strength on yearly productivity metrics by BEC zone for the
+two proposed models. If there are no significant predictors, shown in
+dark grey}
+
+\end{figure}
+
+\hypertarget{exploratory-structural-equation-modelling-1}{%
+\subsection{Exploratory Structural Equation
+Modelling}\label{exploratory-structural-equation-modelling-1}}
+
+I also want to run exploratory SEM with and without the DHI's to see
+what grouping it goes in. MinDHI has been breaking because some zones
+only have 0 as mindhi (makes sense, they are in the alpine).
+
+\begin{figure}
+
+{\centering \includegraphics[width=14.74in,height=\textheight]{../../outputs/latent_groups.png}
+
+}
+
+\caption{\label{fig-esem-latent}Groups of latent variables with loadings
+greater than or equal to 0.5 for each BEC zone. Loadings were determined
+using exploratory factor analysis with up to four latent variables, and
+selecting the number of possible latent variables with the lowest AIC.
+All forest structure variables were potentially included as indicators
+in each latent variable.}
+
+\end{figure}
+
+Exploratory factor analysis shows three groups of latent variables
+(Figure~\ref{fig-esem-latent}). We assigned the groups to latent
+variables indicating canopy cover / biomass metrics, canopy height /
+biomass metrics, and structural complexity metrics. Indicators to latent
+variable groups by examining anchors for each latent variable. The
+anchors were calculated by choosing the indicator variable with the
+largest difference between the maximum value in a given loading compared
+to to said indicators loadings in all other latent variables.
+
+Both the Ponderosa Pine (PP) and Sub-boreal Pine -- Spruce zones have
+one latent variable. Coastal Western Hemlock (CWH), Engelmann Spurce --
+Subalpine Fir (ESSF), and Montane Spruce (MS) are the only zones to have
+structural complexity based latent variables. CWH is the only zone to
+include elevation standard deviation in the structural complexity latent
+variable.
+
+\begin{figure}
+
+{\centering \includegraphics[width=14.74in,height=\textheight]{../../outputs/latent_dhis.png}
+
+}
+
+\caption{\label{fig-esem-latent-dhi}Groups of latent variables with
+loadings greater than or equal to 0.5 for each BEC zone, including the
+DHIs. Loadings were determined using exploratory factor analysis with up
+to five latent variables, and selecting the number of possible latent
+variables with the lowest AIC. All forest structure variables were
+potentially included as indicators in each latent variable.}
+
+\end{figure}
+
+Exploratory factor analysis shows three groups of latent variables. We
+assigned these groups to attributes based on the indicator variables in
+each latent variable, indicating canopy cover, height and biomass, and
+structural complexity. When including the DHIs in this EFA, it is rare
+that they are included with latent variables with large loadings. The
+cumulative DHI is an indicator for the canopy cover latent varible in
+the IDF (Interior Douglas-fir) and MS (Montane Spruce) BEC zones. The
+DHIs are not included in any other latent variable, indicating that they
+compose different information than the forest structural variables.
+
+\begin{figure}
+
+{\centering \includegraphics[width=14.74in,height=\textheight]{../../outputs/latent_boxplot.png}
+
+}
+
+\caption{\label{fig-latent-boxplots}Boxplots of latent variable loadings
+across BEC zones as determined by exploratory factor analysis. Red
+dashed lines show latent variable indicator cutoff before being used as
+starting values for the SEM models.}
+
+\end{figure}
+
+Figure~\ref{fig-latent-boxplots} shows boxplots of latent variable
+loadings from forest structure indicator variables. A maximum of three
+valid latent variables were identified across the sixteen BEC zones. We
+assigned indicators variables to latent variable groups by examining the
+anchoring variables for each latent variable. The anchoring variables
+were calculated by choosing the indicator variable with the largest
+difference between the maximum value in a given loading compared to said
+indicators loadings in all other latent variables. The only DHI variable
+that was included in the latent variables after subsetting the latent
+variables to only include strong loadings was the Cumulative DHI in the
+canopy cover latent variable.
+
+I need to actually make the DAGs and plot the parameters, but I haven't
+gotten there yet. I'll plot the global one to show what it looks like,
+then aggregate it in some way - probably similar to the path analysis. I
+also may include the variable partitioning that Meg talked about in the
+email if I can get it working.
 

### 2023-04-12 scripts/3_extraction.qmd
return from australia

--- 2023-02-02 scripts/3_sem_visualization.qmd
@@
 ---
-title: "sem"
+title: "3_data_extraction"
 format: html
@@
 
-```{r}
-library(lavaan)
-
-theme_set(theme_light())
-```
-
 extract datasets
@@
 ```{r}
-extracted_samples <- extract(all_rasts, samples, bind = T)
+bec_sample_loc <- here::here("data", "samples", "bec_extracted.shp")
+vlce_sample_loc <- here::here("data", "samples", "vlce_extracted.shp")
+
+if (!file.exists(bec_sample_loc)) {
+  bec_samples_e <- extract(all_rasts, bec_samples, bind = T)
+  writeVector(bec_samples_e, bec_sample_loc, overwrite = T)
+}
+if (!file.exists(vlce_sample_loc)) {
+  vlce_samples_e <- extract(all_rasts, vlce_samples, bind = T)
+  writeVector(vlce_samples_e, vlce_sample_loc, overwrite = T)
+}
 ```
-
-)
-
-```{r}
-invalid_zones <- c("CMA", "IMA", "BAFA", "BG")
-# alpine (CMA, IMA, BAFA) & non-forest dominant (BG)
-
-extracted_df <- extracted_samples %>% 
-  st_as_sf() %>%
-  st_drop_geometry() %>%
-  filter(!(loreys_height < elev_mean | elev_p95 < elev_mean)) %>%
-  pivot_longer(basal_area:total_biomass, names_to = "variable") %>%
-  left_join(keys$continuous) %>%
-  mutate(value = value / divide_by) %>%
-  select(zone, category, ends_with("DHI"), variable, value) %>%
-  pivot_wider(names_from = "variable", values_from = "value") %>%
-  unchop(everything()) %>% # somehow does the fix i want
-  filter(elev_cv <= 1) %>% # remove erroneous values
-  drop_na() # remove anything w/ NA
-
-# groups by zone and all other structure variables 
-# takes the mean of each DHI dataset
-# effectively removing duplicates, but takes the mean instead of the first value
-extracted_df_b <- extracted_df %>%
-  group_by(across(c(zone, category, basal_area:total_biomass))) %>%
-  summarize(across(ends_with("DHI"), mean)) %>%
-  ungroup()
-```
-
-```{r}
-# downsampling procedure - allows for comparisons between models if sample size is the same (AIC and BIC are based on sample)
-
-downsample_df <- extracted_df_b %>%
-  count(zone) %>%
-  filter(!(zone %in% invalid_zones))
-
-downsample_no <- downsample_df %>%
-  pull(n) %>%
-  min()
-```
-
-```{r}
-set.seed(69420)
-
-clean_df <- extracted_df_b %>%
-  mutate(
-    overstory_per = percentage_first_returns_above_mean,
-    understory_per = percentage_first_returns_above_2m - percentage_first_returns_above_mean,
-    total_cover = percentage_first_returns_above_2m,
-    overstory_range = loreys_height - elev_mean,
-    understory_range = elev_mean - 2
-  ) %>%
-  relocate(zone,
-           category,
-           ends_with("DHI")) %>%
-  filter(!(zone %in% invalid_zones)) %>%
-  group_by(zone) %>%
-  slice_sample(n = downsample_no) %>%
-  select(!starts_with("perc"))
-```
-
-```{r}
-mylog <- function(data) {
-  if (any(data <= 0)) {
-    return(log1p(data))
-  } else {
-    return(log(data))
-  }
-}
-
-normal <- function(data) {
-  data %>%
-    mylog() %>%
-    scale() %>% 
-    as.numeric()
-}
-
-log_normal_df <- clean_df %>%
-  group_by(zone) %>%
-  mutate(across(CumDHI:understory_range, .fns = ~ normal(.x)))
-
-log_normal_df %>%
-  summarize(across(CumDHI:understory_range, .fns = ~ shapiro.test(.x)$p.value))
-```
-
-```{r}
-log_splits <- log_normal_df %>%
-  group_by(zone) %>%
-  group_split()
-```
-
-sample each bec zone down to the lowest number of samples
-
-```{r}
-set.seed(69420)
-
-zone_dfs <- extracted_df %>%
-  filter(!(zone %in% invalid_zones)) %>%
-  group_by(zone) %>%
-  group_split() %>%
-  map(slice_sample, n = downsample_no)
-```
-
-```{r}
-library(Hmisc)
-library(corrplot)
-corr <- log_normal_df %>% 
-  ungroup() %>%
-  select(CumDHI:total_biomass) %>%
-  as.matrix() %>%
-  rcorr()
-
-corrplot(corr$r,
-         type = "upper",
-         tl.col = "black",
-         tl.srt = 45,
-         p.mat = corr$P, 
-         sig.level = 0.01)
-```
-
-```{r}
-sample_names <- names(extracted_samples)
-
-dhi_vars <-
-  sample_names[(length(sample_names) - 2):length(sample_names)]
-
-v_models <-
-  paste0(
-    dhi_vars,
-    " ~ loreys_height + total_biomass + elev_stddev
-elev_stddev ~ loreys_height
-total_biomass ~ loreys_height"
-  )
-
-h_models <-
-  paste0(
-    dhi_vars,
-    " ~ basal_area + percentage_first_returns_above_2m + percentage_first_returns_above_mean
-percentage_first_returns_above_mean ~ basal_area
-percentage_first_returns_above_2m ~ basal_area"
-  )
-
-models <- c(v_models, h_models)
-```
-
-```{r some-weird-map-lol}
-bec_zones <- bec() %>%
-  mutate(area = st_area(.)) %>%
-  group_by(ZONE) %>%
-  dplyr::summarize(area = sum(area)) %>%
-  st_intersection(bc_bound())
-
-bec_counts_sf <- bec_zones %>%
-  left_join(bec_counts, by = c("ZONE" = "zone")) %>%
-  select(ZONE, n) %>%
-  mutate(ZONE = ifelse(!(ZONE %in% invalid_zones), ZONE, "Not Included"))
-
-ggplot(bec_counts_sf) +
-  geom_sf(aes(fill = ZONE), colour = "#00000000") +
-  ggsci::scale_fill_igv() +
-  scale_alpha_manual(values = c(0.5, 1)) +
-  theme_void()
-```
-
-```{r zonal-models}
-sem_extract <- function(split_df, model) {
-  zone <- split_df %>%
-    pull(zone) %>%
-    unique()
-
-  var <- str_split(model, pattern = " ")[[1]][1]
-  fit <- sem(model = model, data = split_df)
-
-  fits <-
-    fitmeasures(fit, c("chisq", "df", "pvalue", "cfi", "rmsea", "aic", "r2", "srmr"))
-  pars <- parameterestimates(fit) %>%
-    tibble() %>%
-    mutate(zone = zone, var = var, model = model)
-
-  fits <- tibble(fit = names(fits), value = as.numeric(fits)) %>%
-    pivot_wider(names_from = fit, values_from = value) %>%
-    mutate(zone = zone,
-           var = var, model = model)
-
-  interpreted <- fit %>%
-    effectsize::interpret() %>%
-    mutate(zone = zone, var = var, model = model)
-
-  return(list(
-    # fitmeasures = fits,
-    parameters = pars,
-    interpreted = interpreted,
-    fits = fits
-  ))
-}
-
-crossed <- crossing(zone_dfs, models)
-
-dfs <- crossed %>% pull(zone_dfs)
-mods <- crossed %>% pull(models)
-
-outs <- map2(.x = dfs, .y = mods, .f = sem_extract)
-
-params <- map_df(outs, function(x) {
-    x$parameters
-  })
-
-# fitmeasures <- map_df(outs, function(x) {
-#   x$fitmeasures
-# })
-
-fitmeasures <- map_df(outs, function(x) {
-  x$interpreted
-})
-
-fits <- map_df(outs, function(x) {
-  x$fits
-})
-```
-
-```{r}
-vlce_samples <- here::here("data", "samples", "vlce") %>%
-  list.files(pattern = ".shp", full.names = T) %>%
-  map(read_sf) %>%
-  bind_rows()
-
-extracted_vlce <- extract(all_rasts, vlce_samples)
-```
-
-```{r}
-extracted_vlce %>%
-  st_drop_geometry() %>%
-  count(category)
-```
-
-```{r}
-fitmeasures %>% 
-  group_by(zone) %>%
-  dplyr::summarize(across(cfi:aic, mean))
-
-fitmeasures %>%
-  mutate(v_h = ifelse(model %in% h_models, "h", "v")) %>%
-  group_by(model) %>%
-  dplyr::summarize(across(c(cfi:aic, chisq), mean))
-```
-
-```{r}
-params %>%
-  filter(op == "~") %>%
-  group_by(model, lhs, rhs) %>%
-  dplyr::summarize(per_sig = (pvalue > 0.05) / n()) %>%
-  view()
-```
-
-```{r}
-test_mod <- "CumDHI ~ loreys_height + percentage_first_returns_above_2m + percentage_first_returns_above_mean
-percentage_first_returns_above_2m ~ total_biomass
-percentage_first_returns_above_mean ~ percentage_first_returns_above_2m + total_biomass
-total_biomass ~ loreys_height"
-
-data <- dfs[[19]] %>%
-  mutate(across(basal_area:total_biomass, scale),
-         across(basal_area:total_biomass, as.numeric))
-
-(egg <- sem(test_mod, data = data))
-
-varTable(egg)
-fitmeasures(egg)
-parameterestimates(egg)
-
-graph_sem(egg)
-```
-
-```{r}
-vlce_samples <- here::here("data", "samples", "vlce") %>%
-  list.files(full.names = T, pattern = ".shp$") %>%
-  map(read_sf) %>%
-  bind_rows()
-
-vlce_samples
-```

### 2023-04-12 scripts/4_vegan.qmd
return from australia

---
title: "vegan_analysis"
format: html
editor: visual
---

```{r packages}
library("vegan")
```

```{r normal-fun}
mylog <- function(data) {
  if (any(data <= 0)) {
    return(log1p(data))
  } else {
    return(log(data))
  }
}

# normalization from grace et al 2016
# first does mylog (does natural logarithm. if 0s present, does log(one + x))
# then scales to z scores
# results it more normal looking data
normal <- function(data) {
  data %>%
    mylog() %>%
    scale() %>%
    as.numeric()
}
```

```{r data-prep}
bec_sample_loc <- here::here("data", "samples", "bec_extracted.shp")

if (!file.exists(bec_sample_loc)) {
  bec_samples_e <- extract(all_rasts, bec_samples, bind = T)
  writeVector(bec_samples_e, bec_sample_loc, overwrite = T)
}

bec_samples_e <- vect(bec_sample_loc)

bec_df <- bec_samples_e %>% 
  st_as_sf() %>%
  st_drop_geometry() %>%
  select(bec:VarDHI) %>%
  drop_na() %>%
  tibble() %>%
  filter(elev_cv <= 1000) %>%
  filter(elev_p95 >= 2000)

names_manual <- c(
  "bec",
  "category",
  "basal_area",
  "elev_cv",
  "elev_mean",
  "elev_p95",
  "elev_stddev",
  "gross_stem_volume",
  "loreys_height",
  "percentage_first_returns_above_2m",
  "percentage_first_returns_above_mean",
  "total_biomass",
  "CumDHI",
  "MinDHI",
  "VarDHI"
)

# names(bec_df) <- names(all_rasts)
names(bec_df) <- names_manual

bec_normal <- bec_df %>%
  mutate(across(basal_area:VarDHI,  .fns = ~ normal(.x)))

bec_normal_ndup <- bec_df %>%
  group_by(across(c(bec, category, basal_area:total_biomass))) %>%
  dplyr::summarize(across(ends_with("DHI"), mean)) %>%
  ungroup() %>%
  mutate(across(basal_area:VarDHI,  .fns = ~ normal(.x))) 
```

```{r global-model}
to_rda <- bec_normal %>%
  select(-elev_mean,
         -percentage_first_returns_above_mean,
         -elev_stddev)

dhi_tab <- to_rda %>%
  select(ends_with("DHI"))

extracted <- to_rda %>%
  select(elev_p95,
         elev_cv,
         # elev_stddev,
         percentage_first_returns_above_2m)

modelled <- to_rda %>%
  select(loreys_height,
         basal_area,
         total_biomass,
         gross_stem_volume)

rda_out <- 
  rda(dhi_tab ~ elev_p95 + percentage_first_returns_above_2m + elev_cv + # simple/raw/extracted
      basal_area + gross_stem_volume + loreys_height + total_biomass, # modelled
      data = to_rda)

rda_out

ordiplot(rda_out, choices = c(1, 2))

rda_eig <- rda_out$CCA$eig

# scores(rda_out, choices = c(1:3))$species
# scores(rda_out, choices = c(1:3))$biplot
# not actually the loadings, need to get those from correlations

summary(rda_out)$sites %>% 
  as.data.frame() %>% 
  select(RDA1) %>% 
  bind_cols(to_rda) %>%
  select(-bec, -category) %>% 
  cor() %>%
  as.data.frame() %>%
  select(RDA1) %>%
  arrange(desc(abs(RDA1)))

var_out <- varpart(dhi_tab, extracted, modelled)

varpart_out %>% 
  plot(Xnames = c("Derived", "Modelled"))

path_r2 <- var_out$part$indfract %>%
      mutate(type = c("extracted", "modelled", "overlap", "residual")) %>%
      select(type, Adj.R.squared) %>%
  tibble()
```

Metrics derived solely from the point cloud explain 10.6% of the variation in the DHIs, while metrics modelled based on the point cloud explain 3.4% of the variation in the DHIs. 85.9% of the variation is unexplained by structural attributes.

```{r r2-global}
RsquareAdj(rda_out)$adj.r.squared
```

This model shows very little explanatory power. (13.9%)

```{r anova-global}
anova.cca(rda_out, step = 1000)

anova_out <- rda_out  %>%
  anova(by = "axis")

anova_out$`Pr(>F)` %>%
  as.data.frame() %>%
  rename(p_value = 1) %>%
  drop_na() %>%
  mutate(sig = cut(p_value, 
                   breaks = c(0, 0.001, 0.01, 0.05, .10, 1),
                   labels = c("***", "**", "*", ".", "NS"))) %>%
  mutate(axis = paste0("RDA", row_number())) %>%
  relocate(axis)
```

The model is statistically significant (P \< 0.001). All of the axes are significant, although this may be related to the large sample size.

```{r bec-nest}
to_nest <- bec_normal

bec_nested <- to_nest %>%
  group_by(bec) %>%
  nest() %>%
  ungroup()
```

```{r rda-var-funs}
# function to take a tibble and get a list out outputs that can then be unnested and examined as needed. function(tibble) gets rda and varpart, as plottable objects ideally. then get the var_out$part$indfract. i'll also need to look at the significance using an anova at some point

my_rda_fun <- function(tibble) {
  dhi_tab <- tibble %>%
    select(ends_with("DHI"))

  rda(dhi_tab ~ elev_p95 + percentage_first_returns_above_2m + elev_cv + # simple/raw/extracted
      basal_area + gross_stem_volume + loreys_height + total_biomass, # modelled, put condition here for a conditional rda
      data = tibble)
}

my_var_fun <- function(tibble) {
  dhi_tab <- tibble %>%
    select(ends_with("DHI"))

  extracted <- tibble %>%
    select(elev_p95,
           elev_cv,
           # elev_stddev,
           percentage_first_returns_above_2m)

  modelled <- tibble %>%
    select(loreys_height,
           basal_area,
           total_biomass,
           gross_stem_volume)

  varpart(dhi_tab, extracted, modelled)
}
```

```{r rda-var-bec}
outs <- bec_nested %>%
  mutate(rda = map(data, my_rda_fun),
         var = map(data, my_var_fun)) 
```

```{r fracts-out}

fracts_out <- outs %>%
  mutate(indfracts = map(var, .f = function(x) {
    x$part$indfract %>%
      mutate(type = c("extracted", "modelled", "overlap", "residual")) %>%
      select(type, Adj.R.squared) %>%
      pivot_wider(names_from = type, values_from = Adj.R.squared)
    }
    )) %>%
  unnest(indfracts) %>%
  arrange(desc(residual)) %>%
  select(bec, extracted, modelled, overlap) %>%
  pivot_longer(c(extracted, modelled, overlap)) %>%
  mutate(value = ifelse(value > 0, value, 0)) %>%
  pivot_wider()
```

```{r raincloud-plot}
library(ggdist)
fracts_out %>%
  left_join(keys$bec, by = c("bec" = "zone")) %>%
  pivot_longer(extracted:overlap) %>%
  mutate(name = str_to_title(name)) %>%
  mutate(value = ifelse(value < 0, 0, value)) %>%
  ggplot(aes(x = value, y = name)) +
  stat_halfeye(
    # adjust bandwidth
    adjust = .5,
    # move to the right
    justification = -0.2,
    # remove the slub interval
    .width = 0,
    point_colour = NA
    ) +
  geom_boxplot(
    width = 0.12,
    # removing outliers
    outlier.color = NA,
    alpha = 0.5
  ) +
  stat_dots(
    # ploting on left side
    side = "left",
    # adjusting position
    justification = 1.1,
    # adjust grouping (binning) of observations
    binwidth = 0.0025,
    #position = "stack",
    aes(fill = group, col = group),
    alpha = 0.5
  ) +

  labs(y = NULL,
       x = "% Variance Explained",
       fill = "BEC Group",
       col = "BEC Group") +
  scale_x_continuous(labels = scales::label_percent()) +
  theme_bw() +
  theme(panel.grid = element_blank(),
        legend.position = c(0.8, .2))

```

```{r path-setup}
path_base <- tibble(x1 = c(0,
             0,
             0,
             0,
             1),
       y1 = c(1,
              -1,
              1,
              -1,
              0),
       x2 = c(1.85,
              1.85,
              .85,
              .85,
              1.85),
       y2 = c(.15,
              -.15,
              .15,
              -.15,
              0)) %>%
  ggplot(aes(x = x1, y = y1)) +
  geom_segment(aes(xend = x2, yend = y2), arrow = arrow()) +
  geom_text(aes(x = 0, y = -1.15), label = "Modelled") +
  geom_text(aes(x = 0, y = 1.15), label = "Extracted") +
  geom_text(aes(x = .65, y = 0), label = "Overlap") +
  geom_text(aes(x = 2, y = 0), label = "DHIs") +
  geom_point(data = tibble(x1 = c(.85, .85), 
                           y1 = c(-.5, .5)),
             size = 75,
             shape = 21,
             col = "grey") +
  theme_void() +
  scale_x_continuous(limits = c(-0.1, 2.1))

path_r2_plot <- tibble(type = c("extracted", "modelled", "overlap"),
       x1 = c(0.85, 0.85, 0.85),
       y1 = c(0.75, -0.75, 0))
```

```{r path-global}
var_oi <- c("VarDHI", "MinDHI", "CumDHI")
#var_oi <- c("VarDHI", "MinDHI")

var_out <- varpart(dhi_tab %>%
                     select(any_of(var_oi)),
                   extracted, modelled)

# plot(var_out, Xnames = c("Extracted", "Modelled"))

path_r2 <- var_out$part$indfract %>%
      mutate(type = c("extracted", "modelled", "overlap", "residual")) %>%
      select(type, Adj.R.squared) %>%
  tibble()

path_r2_clean <- path_r2_plot %>%
  left_join(path_r2) %>%
  mutate(r2_label = paste0(round(Adj.R.squared * 100, 2), "%"))

path_base +
  geom_text(data = path_r2_clean, aes(label = r2_label)) +
  labs(title = paste(var_oi, collapse = " "))
```

```{r group-climate-spatial}
climate_folder <- here::here("data", "climate")

clim_vars <- climate_folder %>%
  list.files(pattern = "csv$") %>%
  tools::file_path_sans_ext()

clim_vars <- clim_vars[!str_detect(clim_vars, "bec")]

group_clim_avg <-
  here::here(climate_folder, paste0(clim_vars, ".csv")) %>%
  read_csv() %>%
  mutate(var = clim_vars %>% rep(5) %>% sort) %>%
  select(group = BEC_CSV_GR, mean = MEAN, var)

```

```{r clim-plot-group}
fracts_clims <- fracts_out %>%
  left_join(keys$bec, by = c("bec" = "zone")) %>%
  group_by(group) %>%
  summarize(across(extracted:overlap, .fns = mean)) %>%
  left_join(group_clim_avg, multiple = "all")

clim_var = "tmax"

fracts_clims %>%
  pivot_longer(extracted:overlap) %>%
  filter(var == clim_var) %>%
  ggplot(aes(x = mean, y = value)) +
  geom_smooth(method = "lm", col = "red", lty = "dashed") +
  geom_point() +
  facet_wrap(~str_to_title(name)) +
  scale_y_continuous(labels = scales::label_percent()) +
  labs(x = clim_var,
       y = "% Variance Explained") +
  theme_bw() +
  theme(panel.grid = element_blank())
```

```{r clim-plot-bec}
clim_vars <- climate_folder %>%
  list.files(pattern = "csv$") %>%
  tools::file_path_sans_ext()

clim_vars <- clim_vars[!str_detect(clim_vars, "bec")]

bec_clim <- climate_folder %>%
  list.files(pattern = "bec.csv$",
             full.names = T) %>%
  read_csv() %>%
  mutate(var = clim_vars %>% rep(16) %>% sort) %>%
  select(bec = ZONE, mean = MEAN, var)

fracts_out %>%
  pivot_longer(extracted:overlap) %>%
  left_join(bec_clim, multiple = "all") %>%
  filter(var == "ppt") %>%
  ggplot(aes(x = mean, y = value)) +
  geom_point() +
  geom_smooth(method = "lm") + 
  facet_wrap(~name) 
```

```{r testing-idf-live}
to_rda <- bec_normal %>%
  filter(bec == "IDF")

dhi_tab <- to_rda %>%
  select(ends_with("DHI"))

extracted <- to_rda %>%
  select(elev_p95,
         elev_cv,
         percentage_first_returns_above_2m)

modelled <- to_rda %>%
  select(loreys_height,
         basal_area,
         total_biomass,
         gross_stem_volume)

rda_out <- 
  rda(dhi_tab ~ elev_p95 + percentage_first_returns_above_2m + elev_cv + # simple/raw/extracted
      basal_area + gross_stem_volume + loreys_height + total_biomass, # modelled
    data = to_rda)

rda_out

ordiplot(rda_out, choices = c(1, 2))

rda_eig <- rda_out$CCA$eig

var_out <- varpart(dhi_tab, extracted, modelled)

plot(var_out, Xnames = c("Extracted", "Modelled"))

path_r2 <- var_out$part$indfract %>%
      mutate(type = c("extracted", "modelled", "overlap", "residual")) %>%
      select(type, Adj.R.squared) %>%
  tibble()

path_r2_clean <- path_r2_plot %>%
  left_join(path_r2) %>%
  mutate(r2_label = paste0(round(Adj.R.squared * 100, 2), "%"))

path_base +
  geom_text(data = path_r2_clean, aes(label = r2_label)) +
  labs(title = "All")

summary(rda_out)$sites %>% 
  as.data.frame() %>% 
  select(RDA1) %>% 
  bind_cols(to_rda) %>%
  select(-bec, -category) %>% cor() %>%
  as.data.frame() %>%
  select(RDA1) %>%
  arrange(desc(abs(RDA1)))
```

```{r boxplots-dhi-bec}
bec_df %>%
  select(bec, ends_with("DHI")) %>%
  pivot_longer(cols = ends_with("DHI")) %>%
  group_by(name) %>% 
  left_join(keys$bec, by = c("bec" = "zone")) %>%
  mutate(bec = fct_reorder(bec, value)) %>%
  ggplot(aes(x = value, y = bec, fill = group)) +
  geom_boxplot() +
  facet_wrap(~name, scales = "free") +
  theme_bw() +
  theme(panel.grid = element_blank()) +
  labs(x = "DHI Value",
       y = NULL,
       fill = NULL) +
  scale_fill_discrete_c4a_cat(palette = "batlow")
```

```{r cdf-anova}
outs %>% 
  filter(bec == "CDF") %>%
  select(rda) %>% 
  pull(rda) %>%
  .[[1]] %>%
  anova()
```

```{r cum-min-hexplot}
bec_df %>% 
  select(bec, CumDHI, MinDHI) %>%
  left_join(keys$bec, by = c("bec" = "zone")) %>%
  ggplot(aes(x = CumDHI, y = MinDHI)) +
  geom_hex() +
  facet_wrap(~bec)
```

```{r eig-anova}
outs %>%
  mutate(eig = map(rda, .f = function(rda) {
    rda$CCA$eig
  })) %>%
  unnest_wider(eig) %>%
  mutate(prop21 = RDA1/RDA2) %>%
  arrange(desc(prop21)) %>%
  left_join(anova_a2) %>%
  select(bec, RDA1:sig)

anova_outs <- outs %>%
  mutate(
    anova = map(rda, .f = function(rda) {
      anova_out <- rda  %>%
        anova(by = "axis")

      anova_out$`Pr(>F)` %>%
        as.data.frame() %>%
        rename(anova_p = 1) %>%
        drop_na() %>%
        mutate(anova_sig = cut(
          anova_p,
          breaks = c(0, 0.001, 0.01, 0.05, .10, 1),
          labels = c("***", "**", "*", ".", "NS")))  %>%
        mutate(axis = paste0("RDA", row_number())) %>%
        relocate(axis)
  }))

anova_sig_axes <- anova_outs %>%
  select(bec, anova) %>%
  unnest(cols = anova) %>%
  filter(anova_sig != "NS")
```

```{r bec-loadings}
out_loadings <- outs %>%
  mutate(loadings = map(rda,.f = function(rda) {
    summary(rda)$sites %>%
      as.data.frame() %>%
      select(starts_with("RDA"))
  }),
  loadings = map2(.x = loadings, .y = data, .f = bind_cols),
  loadings = map(loadings, .f = function(x) {
    rcorr <- x %>% select(-category) %>%
      as.matrix() %>%
      Hmisc::rcorr()

    rvals <- rcorr$r %>%
      as.data.frame() %>%
      select(starts_with("RDA")) %>%
      mutate(var = row.names(.)) %>%
      pivot_longer(starts_with("RDA"),
                   values_to = "axis_loading",
                   names_to = "axis") %>%
      filter(!startsWith(var, "RDA"))

    pvals <- rcorr$P %>%
      as.data.frame() %>%
      select(starts_with("RDA")) %>%
      mutate(var = row.names(.)) %>%
      pivot_longer(starts_with("RDA"),
                   values_to = "axis_p",
                   names_to = "axis") %>%
      filter(!startsWith(var, "RDA"))

    left_join(rvals, pvals) %>%
      mutate(n = rcorr$n[[1, 1]])
  }

  )) %>%
  select(bec, loadings) %>%
  unnest(loadings)
```

```{r}
out_loadings %>% 
  mutate(axis_sig = cut(
          axis_p,
          breaks = c(0, 0.001, 0.01, 0.05, .10, 1),
          labels = c("***", "**", "*", ".", "NS"),
          right = F)) %>%
  inner_join(anova_sig_axes) %>%
  #filter(axis_sig == "NS") %>%
  arrange(axis_loading) %>%
  count(var)

# need to remove the correlations for variables that arent actually involved
# percentage_first_returns_above_mean, elev_mean, elev_stddev
out_loadings %>% count(var)
```

need to get significant axes from the anova and then filter this

get total rda eigenvalues. find proportion of total for each rda. multiple by 1 - residual. and thats gets path number b/w RD1/RD2 and DHIs. can just filter each by which exist since only a few. goes for path plot

grab the p values w/in zones as well as the correlation coefficients using Hmisc::rcorr(). only plot these plus loadings over threshold

check if deciduous and coniferous have signfiicant axes

### 2023-04-12 scripts/manuscript/manuscript.qmd
return from australia

@@
 date: last-modified
-bibliography: bibliography.bib
+bibliography: [bibliography.bib, packages.bib]
 csl: remote-sensing-of-environment.csl
@@
 
+```{r}
+#| warning: false
+#| echo: false
+#| output: false
+
+# include all libraries required by the R chunks here
+library(lavaan)
+library(sgsR)
+library(terra)
+library(sf)
+
+library(bcmaps)
+
+# generate package citations
+loaded_pkgs = (.packages())
+base_pkgs = getOption("defaultPackages")
+contrib_pkgs = loaded_pkgs[!(loaded_pkgs %in% base_pkgs)]
+pkgs_file = here::here("scripts", "manuscript", "packages.bib")
+knitr::write_bib(contrib_pkgs, pkgs_file)
+```
+
 # Introduction
@@
 biodiversity information, feasible to utilize, and complementary to one
-another [@skidmore2021].
+another [@skidmore2021]. There are six EBV classes, each of which
+correspond to a different facet of biodiversity including species
+populations, species traits, community composition, ecosystem structure,
+ecosystem function, and genetic composition [@pereira2013].
 
-There are six EBV classes, each corresponding to a different facet of
-biodiversity, including, genetic composition, species populations,
-species traits, community composition, ecosystem structure, and
-ecosystem function [@pereira2013]. @fernández2020 divides the six
-classes into two approaches, with one focusing on species biodiversity,
-and the other focusing on ecosystem diversity. Remote sensing has proven
-to be capable of measure five of the six classes, with the exception
-being genetic composition, which requires in-situ observation and
-samples [@skidmore2021]. Notably, while remote sensing can provide
-information on the remaining two species EBV classes, this information
-is typically acquired with an *ad-hoc* approach requiring high spatial
-and spectral resolution data. This data can often only be collected at
-local extents, rather than the global or regional scales required for
-biodiversity trend assessment [@valdez2023]. Community composition falls
-into a similar dilemma, requiring species population information which
-necessitates high resolution spatial data.
+Satellite remote sensing has proven to be capable of measuring five of
+the six EBV classes, the exception being genetic composition, which
+requires in-situ observation and sampling [@skidmore2021]. Species
+populations - and in turn community composition - can be assessed with
+very-high-resolution imagery to identify tree species at the tree-crown
+scale, however it is difficult and computationally expensive to extend
+these analyses to broader extents [@fassnacht2016; @graves2016] while
+species traits such as vegetation phenology have been observed at the
+single-tree scale using, for example, PlanetScope imagery and
+drone-based measurements [@wu2021]. Coarser spatial measurements, such
+as those taken from satellites by the MODIS [@zhang2003], Landsat
+imaging systems[@fisher2006], or Sentinel [@helfenstein2022;
+@darvishzadeh2019] can monitor these functional processes at broader
+extents, corresponding to ecosystem function, but the coarse spatial
+resolution removes the ability to relate these traits to individual
+trees. Remote sensing data has been shown to be arguably the most
+effective at monitoring ecosystem structure and function at regional to
+global extents through the use of optical imagery [@cohen2004] and
+active sensors such as lidar and radar datasets [@lefsky2002; @lang2021;
+@neuenschwander2019].
 
-The remaining two classes, ecosystem structure and function, are
-incredibly well suited to be examined at global or regional scales using
-mid-resolution satellite imagery, such as that provided by the Landsat
-series of satellites. Advances in satellite remote sensing processing
-have allowed 3d forest structure data to be imputed across wide spatial
-scales [@matasci2018; @coops2021] using data fusion approaches involving
-collected lidar data and optical/radar data. Other advances in image
-compositing have allowed yearly summaries of vegetation productivity to
-be calculated at regional to global scales, summarizing the yearly
-energy totals, minimums, and variations [@radeloff2019]. These datasets
-correspond quite well with the EBV classes ecosystem structure (forest
-structural diversity metrics), and ecosystem function (forest
-productivity metrics).
+Notably, while remote sensing can provide information on the
+species-level EBVs, this information is typically acquired with an
+*ad-hoc* approach requiring high spatial and spectral resolution data
+[@skidmore2021]. This data can often only be collected at small extents,
+rather than the global or regional scales required for biodiversity
+trend assessment [@valdez2023]. The two landscape-level EBVs (ecosystem
+structure and function) are well suited to be examined at global or
+regional scales using moderate-resolution satellite imagery, such as
+those provided by the Landsat or Sentinel series of satellites
+[@skidmore2021].
 
 Forest structural diversity has been linked to biodiversity at various
-scales [@guo2017; @bergen2009; @gao2014]. Increased structural
-complexity is hypothesized to create additional niches, leading to
-increased species diversity [@bergen2009]. The relationship between
-forest structure and biodiversity is commonly assessed using avian
-species diversity metrics [@macarthur1961; @goetz2007], however, other
-clades (and habitats), have also been used [@davies2014; @nelson2005].
-Many metrics derived from lidar remote sensing have been used as local
-indicators of biodiversity, including canopy cover, canopy height,
-vertical profiles, and aboveground biomass, while other 2nd order
-derived metrics such as canopy texture, height class distribution,
-edges, and patch metrics have been used to examine habitat and
-biodiversity at landscape scales [@bergen2009].
+scales [@guo2017; @bergen2009; @gao2014]. Many metrics derived from
+lidar remote sensing have been used as local indicators of biodiversity,
+including simple metrics such as canopy cover and canopy height as well
+as derived metrics including vertical profiles, aboveground biomass
+[@lefsky1999]. Other second order derived metrics such as canopy
+texture, height class distribution, edges, and patch metrics have also
+been used to examine habitat and biodiversity at landscape scales
+[@bergen2009]. Increased structural complexity has been hypothesized to
+create additional niches, leading to increased species diversity
+[@bergen2009], which has been frequently demonstrated using avian
+species diversity metrics [@macarthur1961]. For example: @herniman2020
+used spectral and lidar derived forest structure data to model avian
+habitat suitability; @clawges2008 found that lidar derived forest
+structural attributes are capable of identifying habitat types
+associated with avian species in pine/aspen forests; @goetz2007 used
+canopy structural diversity to predict bird species richness, finding
+that canopy vertical distribution was the strongest predictor of species
+richness. Forest structural metrics have also been used to study
+biodiversity in other clades as well [@davies2014; @nelson2005].
 
-The dynamic habitat indices (DHIs) are indicators of productivity
-calculated by summarizing vegetation indices over the course of one (or
-multiple) years [@radeloff2019]. These indices have been related to
-multiple facets of biodiversity at a range of scales, including species
-occurrence and abundance [@razenkova2020], alpha [@radeloff2019] and
-beta diversity [@andrew2012]. Hypotheses behind the biodiversity
-productivity relationships have been established, including the
-species-energy hypothesis, the environmental stress hypothesis, and the
-environmental stability hypothesis [@coops2019]. The cumulative DHI
+With respect to ecosystem function, energy availability in an ecosystem
+has shown to be a predictor of species richness and abundances at
+various scales [@chase2002; @radeloff2019; @coops2019; @razenkova], and
+is measurable using satellite remote sensing via the use of various
+vegetation indices [@huete2002; @radeloff2019]. Vegetation indices,
+which are indicative of photosynthetic activity, are commonly used as
+proxies of gross primary productivity [@huang2019]. These vegetation
+indices have also been used to assess patterns in biodiversity at single
+time points [@bonn2004], and more recently, through yearly summaries of
+productivity [@radeloff2019]. The relationship between energy
+availability and biodiversity occurs via various hypotheses, such as the
+available energy hypothesis, the environmental stress hypothesis, and
+the environmental stability hypothesis. These three hypotheses have in
+turn been linked to patterns of surface reflectance in remote sensing
+data [@radeloff2019]. The Dynamic Habitat Indices (DHIs) have been shown
+to be well suited to assess these hypotheses. The cumulative DHI
 calculates the total amount of energy available in a given pixel over
@@
 
-Some vegetation structure metrics are simpler, and more accurate, to
-calculate than others [@coops2021]. These basic metrics, such as canopy
-height and canopy cover, can then be used to estimate additional
-structure metrics, such as basal area or total biomass. Canopy height
-and cover are commonly used as indicators of vertical and horizontal
-variation, respectively. Recently, more attention has been paid to
-internal structural complexity metrics, which can be more difficult and
-time consuming to generate [@coops2021; @ma2022], but have been shown to
-have stronger linkages with biodiversity [@guo2017], and productivity
-[@ali2019].
+Within forested ecosystems, there are therefore many possible remotely
+sensed derived ecosystem structure and function metrics that can be
+assessed. Structural variables range in complexity from simple (canopy
+cover; canopy height), to more complex (elevation coefficient of
+variation; elevation standard deviation) to modelled (aboveground
+biomass; basal area), all of which can be assessed using lidar data
+[@coops2021]. Data on ecosystem functionality ranges from phenology
+metrics (date of leaf on/leaf off; length of growing season) to
+functional traits (chlorophyll content; leaf area index) to productivity
+estimates (GPP; total available energy) [@pettorelli2018]. Advances in
+satellite remote sensing processing have allowed 3D forest structure
+data to be imputed across wide spatial scales [@matasci2018; @coops2021]
+using data fusion approaches involving collected lidar data and
+optical/radar data. Other advances in image compositing have allowed
+yearly summaries of vegetation productivity to be calculated at regional
+to global scales, summarizing the yearly total, minimum, and variation
+in energy availability at similar spatial resolutions [@radeloff2019].
 
-Linkages between forest structure and productivity (namely, vegetation
-indices) have been examined for nearly 20 years [@huete2002,
-@knyazikhin1998; @myneni1994]. While there is significant theoretical
-and empirical evidence for their relationship at single time points
-(within a single image) [@myneni1994], various relationship directions
-and shapes have been found between forest structure and productivity
-metrics [@ali2019]. These relationships, their shapes, and their
-strengths have been attributed to multiple possible hypotheses and have
-been shown to vary based on environmental conditions [@ali2019]. The
-relationship between forest structural diversity metrics and annual
-productivity summaries has yet to be fully examined, including the DHIs.
+Linkages between forest structure and function have been examined within
+a remote sensine context for over 20 years [@huete2002, @knyazikhin1998;
+@myneni1994]. While there is significant theoretical and empirical
+evidence for their relationship at single time points (within a single
+image) [@myneni1994], various relationship directions and shapes have
+been found between forest structure and function metrics [@ali2019].
+These relationships, their shapes, and their strengths have been
+attributed to multiple possible hypotheses and have been shown to vary
+based on environmental conditions [@ali2019]. The relationship in
+particular between forest structural diversity metrics - which are now
+more accurately and comprehensively derived from lidar data - and
+temporal variation in functional metrics, specifically the metrics of
+ecosystem productivity via the DHI framework, have yet to be fully
+examined.
 
 Understanding the complementarity between potential EBVs is an integral
-component of their creation [@skidmore2021]. Biodiversity has been shown
-to be linked to both forest structure [@guo2017; @gao2014], and
-productivity [@radeloff2019] at a range of scales. Research has
-highlighted that the relationship between productivity and biodiversity
-is reciprocal [@worm2003], making productivity a suitable biodiversity
-indicator. In forested ecosystems, stand structural attributes have also
-been shown to influence productivity, with a range of responses
-dependent on environmental conditions [@ali2019].
+component of their creation [@skidmore2021]. Structural equation
+modelling (SEM) and path analyses have been commonly used in ecology to
+assess the causal effects behind various hypotheses [@fan2016;
+@grace2010], including the relationship between forest structure,
+functioning, and biodiversity [@ali2019]. For example, SEM has commonly
+been used to assess the role of structural diversity of carbon storage
+as compared to species diversity [@ali2016; @zhang2015].
 
-Structural equation modelling and path analyses have been commonly used
-in ecology to assess the causal effects behind various hypotheses
-[@fan2016; @grace2010], including the relationship between forest
-structure, functioning, and biodiversity [@ali2019] . In this paper we
-seek to untangle the relationship between two EBVs: forest structure
-diversity metrics and yearly summaries of forest productivity. To
-accomplish this, we assess this relationship using path analysis to
-assess the direct and indirect (as mediated by more complex forest
-structural diversity metrics) effects of commonly collected forest
-structural metrics (canopy height and canopy cover) on yearly
-productivity summaries. Further, we use exploratory structural equation
-modelling (ESEM) to identify the latent variables driving the
-productivity metrics. We run this analysis both globally, and stratified
-by the forested ecosystems of British Columbia, Canada, and compare
-models using AIC scores and global fit measures across the ecosystems.
-The results from this study will assess the linkages and complementarity
-between two biodiversity indicators.
+The overall goal of this paper is to characterize patterns of forest
+ecosystem structure and function and their interrelationships across a
+wide range of ecosystems encompassing significant environmental
+gradients. To do so, we synthesize data from moderate-scale
+remote-sensing derived metrics of structure, represented as both simple
+ALS metrics of canopy height and cover, as well as modeled forest
+structure variables including volume, and above ground biomass, with a
+remote sensing derived index on ecosystem function. Our first question
+is to examine how ecosystem structure and function vary across a large
+environmental gradient and then compare the simple and modeled
+representations of forest structure to different levels of ecosystem
+function. This question is important as it provides insights to the EBV
+community around redundancy in remote sensing metrics when describing
+structure and function on ecosystem and proposes a method to examine
+potential overlap when generating remote sensing EBV's.
+
+Our second question examines the independent and shared relationships of
+ecosystem structure height and cover, with modeled forest structure, on
+function. This provides insight into the choice of remote sensing
+attributes to use when developing EBV's. Remote sensing datasets can
+comprise relatively unprocessed, observations, in this case ALS measures
+of height and cover which are retrieval from the raw 3D point cloud, vs
+modeled attributes, such as biomass and volume, which involve the use
+the statistical relationships to transform the observations into more
+refined data products. Assessing which of these two (or combination of
+the two) approaches has more or less redundancy with estimates of
+function provides insights into the choice of data used to build EBV's.
+Lastly we examine how the primary and modeled structure attributes vary
+in their variance partitioning within key biomes across this large
+environmental range, examining to what extend ecosystem type impacts
+these interrelationships and thus the applicability of these results
+globally.
 
@@
 {insert figure showing ecosystems of interest} need to remake figure
-showing ecosystems, highlighting that bunchgrass isn't included
+showing ecosystems, highlighting that bunchgrass and alpine zones are
+not included
 
@@
 
-### NTEMS
+### Forest Structure
+
+We utilize a suite of forest structural diversity variables (canopy
+height, canopy cover, Lorey's height, overstory cover, basal area,
+aboveground biomass, gross stem volume, mean elevation, elevation
+standard deviation, and structural complexity \[coefficient of variation
+in elevation returns\]). This dataset is created at a 30 m spatial
+resolution according to @matasci2018. In brief, the method utilizes a
+set of lidar collections and field plots across Canada, and imputes the
+remaining pixels using a random forest k-Nearest Neighbour approach.
+Detailed information on the creation of this dataset can be found in
+@matasci2018.
 
@@
 
+The DHIs were calculated according to {RAZENKOVA LANDSAT DHI PAPER IN
+PRESS} for all of terrestrial British Columbia. In brief, they utilized
+Google Earth Engine [@gorelick2017] to obtain all valid Landsat pixels
+for a given study area by filtering out low quality pixels, then
+calculated the NDVI for each pixel. They use the median NDVI value for
+each month across the ten year time span (2011-2020) to calculate the
+sum, minimum, and coefficient of variation in NDVI values. More detailed
+information can be found in {Razenkova In Press}.
+
 ## Sampling
@@
 than 0.5 in the Lorey's height and canopy cover forest structural
-metrics. These focal analyses were preprocessed in R and Python, and
-randomly sampled across a masked suitability raster for each BEC zone
-using the *sgsR* R package version 1.3.4 [@goodbody2023]. A maximum of
-3000 samples were taken from each BEC zone.
-
-\[note to ncc; I've stopped doing this sampling step as Meg said she
-didn't feel it was necessary\]
-
-The imputation method used to generate the forest structure layers leads
-to many duplicated values as there are a limited number of samples to
-predict from [@matasci2018]. A discussion of imputation approaches to
-large area lidar attribute modelling can be found in @coops2021. To
-account for duplicated input values in our modelling we averaged the
-endogenous variables within each ecosystem for each uniquely imputed set
-of forest attributes, forest type, and ecosystem. Due to variations in
-ecosystem size, differing numbers of samples were generated for each
-ecosystem. To account for this during model comparison, each ecosystem
-and land cover sample set was downsampled to the minimum number of
-samples for a single BEC zone (in this case, 170). For the forest type
-models, we attempted to select equal numbers of observations from each
-BEC zone. However, some forest types are rarely found in certain zones.
-To account for this, if an equal number of samples from each BEC zone
-were not available, the entire sample from that zone was taken. The
-amount of samples taken from other zones was then increased in order to
-collect a total number equal to the minimum number of samples in a
-single BEC zones (170). To meet the normality assumptions of path
-analysis, all variables were natural-log transformed and standardized,
-as per @grace2016. Variables containing zeros were natural-log plus one
+metrics. A maximum of 3000 samples were sampled in each BEC zone. To
+meet the normality assumptions of structural equation modelling, all
+variables were natural-log transformed and standardized, as per
+@grace2016. Variables containing zeros were natural-log plus one
 transformed.
@@
 #| label: fig-pathdag
-#| fig-cap: Proposed path diagrams
+#| fig-cap: Proposed path diagrams.
 #| warning: false
@@
 predictor, without a latent variable as a mediator.
+
+## Software
+
+The processing for this paper was done in R version 4.2.2 [@R-base]. We
+used *sgsR* [@goodbody2023] for sampling, *terra* [@R-terra] for raster
+analyses, *sf* for vector analyses [@R-sf], and *lavaan* [@lavaan2012]
+for path analysis and structural equation modelling.
+
+Focal analyses were conducted in R and Python, and the sampling was
+conducted across a masked suitability raster for each BEC zone using the
+*sgsR* R package version 1.3.4 [@goodbody2023].
+
+All code associated with this analysis is available at
+https://github.com/emuise/code-structProdSem.
 

### 2023-04-12 scripts/manuscript/manuscript.tex
return from australia

@@
   pdftitle={Disentangling linkages between satellite derived forest structure and productivity across a range of forest types and ecosystems.},
-  pdfauthor={Evan R. Muise; Nicholas C. Coops; Txomin Hermosilla; A. Cole Burton; Margaret E. Andrew; Stephen S. Ban},
+  pdfauthor={Evan R. Muise; Margaret E. Andrew; Nicholas C. Coops; Txomin Hermosilla; A. Cole Burton; Stephen S. Ban},
   pdfkeywords={structural equation modelling, remote
@@
  \ead{evan.muise@student.ubc.ca} 
+\author[2]{Margaret E. Andrew%
+%
+}
+ \ead{M.Andrew@murdoch.edu.au} 
 \author[1]{Nicholas C. Coops%
@@
  \ead{nicholas.coops@ubc.ca} 
-\author[2]{Txomin Hermosilla%
+\author[3]{Txomin Hermosilla%
 %
@@
  \ead{cole.burton@ubc.ca} 
-\author[3]{Margaret E. Andrew%
-%
-}
- \ead{M.Andrew@murdoch.edu.au} 
 \author[4]{Stephen S. Ban%
@@
 Canada},postcode={V6T 1Z4},postcodesep={}}
-\affiliation[2]{organization={Natural Resources Canada, Canada Forest
+\affiliation[2]{organization={Murdoch University, Environmental and
+Conservation Sciences and Harry Butler Institute},city={Murdoch, WA,
+Australia},postcode={6150},postcodesep={}}
+\affiliation[3]{organization={Natural Resources Canada, Canada Forest
 Service (Pacific Forestry Centre)},addressline={506 Burnside Rd
 W},city={Victoria, BC, Canada},postcode={V8Z 1M5},postcodesep={}}
-\affiliation[3]{organization={Murdoch University, Environmental and
-Conservation Sciences and Harry Butler Institute},city={Murdoch, WA,
-Australia},postcode={6150},postcodesep={}}
 \affiliation[4]{organization={Ministry of Environment and Climate Change
@@
 \end{frontmatter}
-    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable, enhanced, sharp corners]}{\end{tcolorbox}}\fi
+    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, interior hidden, borderline west={3pt}{0pt}{shadecolor}, frame hidden, sharp corners, enhanced, boxrule=0pt]}{\end{tcolorbox}}\fi
+
+\begin{verbatim}
+This is lavaan 0.6-14
+lavaan is FREE software! Please report any bugs.
+\end{verbatim}
+
+\begin{verbatim}
+terra 1.7.3
+\end{verbatim}
+
+\begin{verbatim}
+Linking to GEOS 3.9.3, GDAL 3.5.2, PROJ 8.2.1; sf_use_s2() is TRUE
+\end{verbatim}
 
@@
 be global in scope, relevant to biodiversity information, feasible to
-utilize, and complementary to one another \citep{skidmore2021}.
+utilize, and complementary to one another \citep{skidmore2021}. There
+are six EBV classes, each of which correspond to a different facet of
+biodiversity including species populations, species traits, community
+composition, ecosystem structure, ecosystem function, and genetic
+composition \citep{pereira2013}.
 
-There are six EBV classes, each corresponding to a different facet of
-biodiversity, including, genetic composition, species populations,
-species traits, community composition, ecosystem structure, and
-ecosystem function \citep{pereira2013}. \citet{fernández2020} divides
-the six classes into two approaches, with one focusing on species
-biodiversity, and the other focusing on ecosystem diversity. Remote
-sensing has proven to be capable of measure five of the six classes,
-with the exception being genetic composition, which requires in-situ
-observation and samples \citep{skidmore2021}. Notably, while remote
-sensing can provide information on the remaining two species EBV
-classes, this information is typically acquired with an \emph{ad-hoc}
-approach requiring high spatial and spectral resolution data. This data
-can often only be collected at local extents, rather than the global or
-regional scales required for biodiversity trend assessment
-\citep{valdez2023}. Community composition falls into a similar dilemma,
-requiring species population information which necessitates high
-resolution spatial data.
+Satellite remote sensing has proven to be capable of measuring five of
+the six classes, with the exception being genetic composition, which
+requires in-situ observation and sampling \citep{skidmore2021}. Species
+populations - and in turn community composition - can be assessed with
+very-high-resolution imagery to identify tree species at the tree-crown
+scale, however it is difficult and computationally expensive to extend
+these analyses to broader extents \citep{fassnacht2016, graves2016}.
+Species traits such as phenology have been on a single-tree scale using
+PlanetScope imagery and drone-based measurements \citep{wu2021}. Coarser
+measurements, such as those taken by the MODIS \citep{zhang2003},
+Landsat \citep{fisher2006}, or Sentinel
+\citep{helfenstein2022, darvishzadeh2019} series of satellites can
+monitor these functional processes at larger extents, corresponding to
+ecosystem function, but the coarse spatial resolution removes the
+ability to relate these traits to individual trees. Remote sensing data
+has been shown to be effective at monitoring ecosystem structure at
+regional to global extents through the use of optical imagery
+\citep{cohen2004} and lidar datasets
+\citep{lefsky2002, lang2021, neuenschwander2019}.
 
-The remaining two classes, ecosystem structure and function, are
-incredibly well suited to be examined at global or regional scales using
-mid-resolution satellite imagery, such as that provided by the Landsat
-series of satellites. Advances in satellite remote sensing processing
-have allowed 3d forest structure data to be imputed across wide spatial
-scales \citep{matasci2018, coops2021} using data fusion approaches
-involving collected lidar data and optical/radar data. Other advances in
-image compositing have allowed yearly summaries of vegetation
-productivity to be calculated at regional to global scales, summarizing
-the yearly energy totals, minimums, and variations \citep{radeloff2019}.
-These datasets correspond quite well with the EBV classes ecosystem
-structure (forest structural diversity metrics), and ecosystem function
-(forest productivity metrics).
+Notably, while remote sensing can provide information on the
+species-level EBVs, this information is typically acquired with an
+\emph{ad-hoc} approach requiring high spatial and spectral resolution
+data \citep{skidmore2021}. This data can often only be collected at
+small extents, rather than the global or regional scales required for
+biodiversity trend assessment \citep{valdez2023}. The two
+landscape-level EBVs (ecosystem structure and function) are well suited
+to be examined at global or regional scales using moderate-resolution
+satellite imagery, such as those provided by the Landsat or Sentinel
+series of satellites \citep{skidmore2021}.
+
+Simple metrics to complex for structure.
 
@@
 complexity is hypothesized to create additional niches, leading to
-increased species diversity \citep{bergen2009}. The relationship between
-forest structure and biodiversity is commonly assessed using avian
-species diversity metrics \citep{macarthur1961, goetz2007}, however,
-other clades (and habitats), have also been used
-\citep{davies2014, nelson2005}. Many metrics derived from lidar remote
-sensing have been used as local indicators of biodiversity, including
-canopy cover, canopy height, vertical profiles, and aboveground biomass,
-while other 2nd order derived metrics such as canopy texture, height
-class distribution, edges, and patch metrics have been used to examine
-habitat and biodiversity at landscape scales \citep{bergen2009}.
+increased species diversity \citep{bergen2009}, which has been
+frequently demonstrated using avian species diversity metrics
+\citep{macarthur1961}. For example: \citet{herniman2020} used spectral
+and lidar derived forest structure data to model avian habitat
+suitability; \citet{clawges2008} found that lidar derived forest
+structural attributes are capable of identifying habitat types
+associated with avian species in pine/aspen forests; \citet{goetz2007}
+used canopy structural diversity to predict bird species richness,
+finding that canopy vertical distribution was the strongest predictor of
+species richness. Forest structural metrics have also been used to study
+other clades and habitats \citep{davies2014, nelson2005}. Many metrics
+derived from lidar remote sensing have been used as local indicators of
+biodiversity, including simple metrics such as canopy cover and canopy
+height as well as derived metrics including vertical profiles,
+aboveground biomass \citep{lefsky1999}. Other second order derived
+metrics such as canopy texture, height class distribution, edges, and
+patch metrics have also been used to examine habitat and biodiversity at
+landscape scales \citep{bergen2009}.
+
+Simple productivity (NDVI, indices) to complex aggregates (DHIs)
 
@@
 
-Some vegetation structure metrics are simpler, and more accurate, to
-calculate than others \citep{coops2021}. These basic metrics, such as
-canopy height and canopy cover, can then be used to estimate additional
-structure metrics, such as basal area or total biomass. Canopy height
-and cover are commonly used as indicators of vertical and horizontal
-variation, respectively. Recently, more attention has been paid to
-internal structural complexity metrics, which can be more difficult and
-time consuming to generate \citep{coops2021, ma2022}, but have been
-shown to have stronger linkages with biodiversity \citep{guo2017}, and
-productivity \citep{ali2019}.
+Biodiversity has been shown to be linked to both forest structure
+\citep{guo2017, gao2014}, and productivity \citep{radeloff2019} at a
+range of scales. Research has highlighted that the relationship between
+productivity and biodiversity is reciprocal \citep{worm2003}, making
+productivity a suitable biodiversity indicator. Within a forest
+environment, there are many possible ecosystem structure and function
+metrics that can be assessed. Structural variables range in complexity
+from simple (canopy cover; canopy height), to more complex (elevation
+coefficient of variation; elevation standard deviation) to modelled
+(aboveground biomass; basal area), all of which can be assessed using
+lidar data \citep{coops2021}. Data on ecosystem functionality ranges
+from phenology metrics (date of leaf on/leaf off; length of growing
+season) to functional traits (chlorophyll content; leaf area index) to
+productivity estimates (GPP; total available energy)
+\citep{pettorelli2018}. Advances in satellite remote sensing processing
+have allowed 3d forest structure data to be imputed across wide spatial
+scales \citep{matasci2018, coops2021} using data fusion approaches
+involving collected lidar data and optical/radar data. Other advances in
+image compositing have allowed yearly summaries of vegetation
+productivity to be calculated at regional to global scales, summarizing
+the yearly total, minimum, and variation in energy availability at
+similar spatial resolutions \citep{radeloff2019}.
 
-Linkages between forest structure and productivity (namely, vegetation
-indices) have been examined for nearly 20 years \citetext{\citealp[
-\citet{knyazikhin1998}]{huete2002}; \citealp{myneni1994}}. While there
-is significant theoretical and empirical evidence for their relationship
-at single time points (within a single image) \citep{myneni1994},
-various relationship directions and shapes have been found between
-forest structure and productivity metrics \citep{ali2019}. These
-relationships, their shapes, and their strengths have been attributed to
-multiple possible hypotheses and have been shown to vary based on
-environmental conditions \citep{ali2019}. The relationship between
-forest structural diversity metrics and annual productivity summaries
-has yet to be fully examined, including the DHIs.
+In forested ecosystems, stand structural attributes have also been shown
+to influence productivity, with a range of responses dependent on
+environmental conditions \citep{ali2019}. Linkages between forest
+structure and productivity (namely, vegetation indices) have been
+examined for nearly 20 years
+\citetext{\citealp[\citet{knyazikhin1998}]{huete2002}; \citealp{myneni1994}}.
+While there is significant theoretical and empirical evidence for their
+relationship at single time points (within a single image)
+\citep{myneni1994}, various relationship directions and shapes have been
+found between forest structure and productivity metrics \citep{ali2019}.
+These relationships, their shapes, and their strengths have been
+attributed to multiple possible hypotheses and have been shown to vary
+based on environmental conditions \citep{ali2019}. The relationship
+between forest structural diversity metrics and annual productivity
+summaries has yet to be fully examined, including the DHIs.
 
 Understanding the complementarity between potential EBVs is an integral
-component of their creation \citep{skidmore2021}. Biodiversity has been
-shown to be linked to both forest structure \citep{guo2017, gao2014},
-and productivity \citep{radeloff2019} at a range of scales. Research has
-highlighted that the relationship between productivity and biodiversity
-is reciprocal \citep{worm2003}, making productivity a suitable
-biodiversity indicator. In forested ecosystems, stand structural
-attributes have also been shown to influence productivity, with a range
-of responses dependent on environmental conditions \citep{ali2019}.
-
-Structural equation modelling and path analyses have been commonly used
-in ecology to assess the causal effects behind various hypotheses
-\citep{fan2016, grace2010}, including the relationship between forest
-structure, functioning, and biodiversity \citep{ali2019} . In this paper
-we seek to untangle the relationship between two EBVs: forest structure
-diversity metrics and yearly summaries of forest productivity. To
-accomplish this, we assess this relationship using path analysis to
-assess the direct and indirect (as mediated by more complex forest
-structural diversity metrics) effects of commonly collected forest
-structural metrics (canopy height and canopy cover) on yearly
-productivity summaries. Further, we use exploratory structural equation
-modelling (ESEM) to identify the latent variables driving the
-productivity metrics. We run this analysis both globally, and stratified
-by the forested ecosystems of British Columbia, Canada, and compare
-models using AIC scores and global fit measures across the ecosystems.
-The results from this study will assess the linkages and complementarity
-between two biodiversity indicators.
+component of their creation \citep{skidmore2021}. Structural equation
+modelling and path analyses have been commonly used in ecology to assess
+the causal effects behind various hypotheses \citep{fan2016, grace2010},
+including the relationship between forest structure, functioning, and
+biodiversity \citep{ali2019}. In this paper we seek to untangle the
+relationship between two EBVs: forest structure diversity metrics and
+yearly summaries of forest productivity. To accomplish this, we assess
+this relationship using path analysis to assess the direct and indirect
+(as mediated by more complex forest structural diversity metrics)
+effects of commonly collected forest structural metrics (canopy height
+and canopy cover) on yearly productivity summaries. Further, we use
+exploratory structural equation modelling (ESEM) to identify the latent
+variables driving the productivity metrics. We run this analysis both
+globally, and stratified by the forested ecosystems of British Columbia,
+Canada, and compare models using AIC scores and global fit measures
+across the ecosystems. The results from this study will assess the
+linkages and complementarity between two biodiversity indicators.
 
@@
 than 0.5 in the Lorey's height and canopy cover forest structural
-metrics. These focal analyses were preprocessed in R and Python, and
-randomly sampled across a masked suitability raster for each BEC zone
-using the \emph{sgsR} R package version 1.3.4 \citep{goodbody2023}. A
-maximum of 3000 samples were taken from each BEC zone.
+metrics. A maximum of 3000 samples were sampled in each BEC zone.
 
-{[}note to ncc; I've stopped doing this sampling step as Meg said she
-didn't feel it was necessary{]}
-
-The imputation method used to generate the forest structure layers leads
-to many duplicated values as there are a limited number of samples to
-predict from \citep{matasci2018}. A discussion of imputation approaches
-to large area lidar attribute modelling can be found in
-\citet{coops2021}. To account for duplicated input values in our
-modelling we averaged the endogenous variables within each ecosystem for
-each uniquely imputed set of forest attributes, forest type, and
-ecosystem. Due to variations in ecosystem size, differing numbers of
-samples were generated for each ecosystem. To account for this during
-model comparison, each ecosystem and land cover sample set was
-downsampled to the minimum number of samples for a single BEC zone (in
-this case, 170). For the forest type models, we attempted to select
-equal numbers of observations from each BEC zone. However, some forest
-types are rarely found in certain zones. To account for this, if an
-equal number of samples from each BEC zone were not available, the
-entire sample from that zone was taken. The amount of samples taken from
-other zones was then increased in order to collect a total number equal
-to the minimum number of samples in a single BEC zones (170). To meet
-the normality assumptions of path analysis, all variables were
+Due to variations in ecosystem size, differing numbers of samples were
+generated for each ecosystem. To account for this during model
+comparison, each ecosystem and land cover sample set was downsampled to
+the minimum number of samples for a single BEC zone (in this case, 170).
+To meet the normality assumptions of path analysis, all variables were
 natural-log transformed and standardized, as per \citet{grace2016}.
@@
 
+\hypertarget{software}{%
+\subsection{Software}\label{software}}
+
+The processing for this paper was done in R version 4.2.2
+\citep{R-base}. We used \emph{sgsR} \citep{goodbody2023} for sampling,
+\emph{terra} \citep{R-terra} for raster analyses, \emph{sf} for vector
+analyses \citep{R-sf}, and \emph{lavaan} \citep{lavaan2012} for path
+analysis and structural equation modelling.
+
+These focal analyses were preprocessed in R and Python, and randomly
+sampled across a masked suitability raster for each BEC zone using the
+\emph{sgsR} R package version 1.3.4 \citep{goodbody2023}.
+
 \hypertarget{results}{%
@@
 
-{\centering \includegraphics[width=14.74in,height=\textheight]{../../outputs/latent_groups.png}
+{\centering \includegraphics[width=7in,height=\textheight]{../../outputs/latent_groups.png}
 
@@
 \renewcommand\refname{References}
-  \bibliography{bibliography.bib}
+  \bibliography{bibliography.bib,packages.bib}
 

### 2023-04-18 scripts/4_vegan.qmd
added strata instead of just bec

@@
 library("vegan")
+library("tidyverse")
+library("cols4all")
+library(terra)
+library(sf)
+source(here::here("scripts", "get_keys.R"))
 ```
@@
 }
+
+```
+
+```{r theme}
+my_theme <- theme_bw() +
+  theme(panel.grid = element_blank())
+
+theme_set(my_theme)
 ```
@@
 
+to_rda <- bec_normal %>% 
+  group_by(category) %>% 
+  slice_sample(n = 3000)  %>% 
+  filter(category == "Mixed Wood") %>% 
+  ungroup()
+
 dhi_tab <- to_rda %>%
@@
 
-rda_out
-
 ordiplot(rda_out, choices = c(1, 2))
 
-rda_eig <- rda_out$CCA$eig
+rda_sum <- summary(rda_out)
+
+var_exp <- rda_sum$constr.chi / rda_sum$tot.chi
+
+rda_eig <- rda_out$CCA$eig %>%
+  as.data.frame() %>%
+  t() %>%
+  as.data.frame() %>%
+  tibble() %>%
+  pivot_longer(cols = everything()) %>%
+  mutate(prop = value / sum(value),
+         var_exp = prop * var_exp)
 
@@
   as.data.frame() %>% 
-  select(RDA1) %>% 
+  select(RDA1:RDA3) %>% 
   bind_cols(to_rda) %>%
@@
   as.data.frame() %>%
-  select(RDA1) %>%
-  arrange(desc(abs(RDA1)))
+  select(RDA1:RDA3) %>%
+  arrange(desc(abs(RDA1))) %>%
+  filter(!(startsWith(row.names(.), "RDA")))
 
@@
 
-varpart_out %>% 
+var_out %>% 
   plot(Xnames = c("Derived", "Modelled"))
@@
   nest() %>%
-  ungroup()
+  ungroup() %>%
+  rename(strata = bec)
+
+class_nested <- to_nest %>%
+  group_by(category) %>%
+  slice_sample(n = 3000) %>%
+  nest() %>% 
+  ungroup() %>%
+  rename(strata = category)
+
+strata_nested <- bind_rows(bec_nested, class_nested)
 ```
@@
 ```{r rda-var-bec}
-outs <- bec_nested %>%
-  mutate(rda = map(data, my_rda_fun),
-         var = map(data, my_var_fun)) 
+outs <- strata_nested %>%
+  mutate(rda = map(data, my_rda_fun, .progress = "RDA"),
+         var = map(data, my_var_fun, .progress = "VAR")) 
 ```
@@
   arrange(desc(residual)) %>%
-  select(bec, extracted, modelled, overlap) %>%
+  select(strata, extracted, modelled, overlap) %>%
   pivot_longer(c(extracted, modelled, overlap)) %>%
@@
   pivot_wider()
+
+fracts_out %>% 
+  mutate(total = extracted + modelled + overlap) %>%
+  arrange(strata %in% (bec_normal %>%
+           pull(category) %>%
+           unique()),
+          desc(total))
 ```
@@
 fracts_out %>%
-  left_join(keys$bec, by = c("bec" = "zone")) %>%
+  #left_join(keys$bec, by = c("bec" = "zone")) %>%
   pivot_longer(extracted:overlap) %>%
@@
     #position = "stack",
-    aes(fill = group, col = group),
+    # aes(fill = group, col = group),
     alpha = 0.5
@@
   geom_smooth(method = "lm") + 
-  facet_wrap(~name) 
+  facet_wrap(~name) +
+  theme_bw() +
+  theme(panel.grid = element_blank())
 ```
@@
 outs %>% 
-  filter(bec == "CDF") %>%
+  filter(bec == "IDF") %>%
   select(rda) %>% 
@@
   .[[1]] %>%
-  anova()
+  anova(by = "axis")
 ```
@@
 ```{r eig-anova}
-outs %>%
-  mutate(eig = map(rda, .f = function(rda) {
-    rda$CCA$eig
-  })) %>%
-  unnest_wider(eig) %>%
-  mutate(prop21 = RDA1/RDA2) %>%
-  arrange(desc(prop21)) %>%
-  left_join(anova_a2) %>%
-  select(bec, RDA1:sig)
+eig_outs <- outs %>%
+  mutate(eig = map(
+    rda,
+    .f = function(rda) {
+      rda_sum <- summary(rda)
+
+      var_exp <- rda_sum$constr.chi / rda_sum$tot.chi
+
+      rda_eig <- rda$CCA$eig %>%
+        as.data.frame() %>%
+        t() %>%
+        as.data.frame() %>%
+        tibble() %>%
+        pivot_longer(cols = everything(), names_to = "axis",
+                     values_to = "eigenvalue") %>%
+        mutate(prop = eigenvalue / sum(eigenvalue),
+               var_exp = prop * var_exp)
+    }
+  )) %>%
+  unnest(eig) %>%
+  select(strata, axis:var_exp)
 
@@
         relocate(axis)
-  }))
+  }, .progress = "anovas"))
 
@@
   select(bec, anova) %>%
-  unnest(cols = anova) %>%
-  filter(anova_sig != "NS")
+  unnest(cols = anova) # %>%
+  # filter(anova_sig != "NS")
+
+axis_varexp <- anova_sig_axes %>%
+  inner_join(eig_outs)
+
+axis_varexp %>%
+  left_join(keys$bec, by = c("bec" = "zone")) %>%
+  ggplot(aes(x = axis, y = var_exp, col = axis)) +
+  geom_jitter(height = 0, width = 0.05) +
+  theme(legend.position = "bottom") +
+  labs(x = NULL,
+       y = "% of Variance Explained",
+       col = NULL) +
+  scale_y_continuous(labels = scales::label_percent())
+# no secondary axis in alpine & coast.
+
+axis_varexp %>%
+  group_by(bec) %>%
+  # filter(n() == 2) %>%
+  select(bec, axis, var_exp) %>% 
+  pivot_wider(names_from = axis,
+              values_from = var_exp) %>%
+  mutate(ratio = RDA1 / RDA2)
+
+# smaller ratios present when the RDA1 doesn't explain much. when variance explained
+# is low, the ratio is low. this is because there isn't much variance explained
 ```
@@
 ```{r}
-out_loadings %>% 
-  mutate(axis_sig = cut(
+sig_loads <- out_loadings %>% 
+  mutate(loading_sig = cut(
           axis_p,
@@
   #filter(axis_sig == "NS") %>%
-  arrange(axis_loading) %>%
-  count(var)
+  arrange(desc(abs(axis_loading))) %>%
+  filter(!(var %in% c("percentage_first_returns_above_mean", 
+                    "elev_mean", 
+                    "elev_stddev"))) %>%
+  drop_na()# removes ima mindhi which has no variance
+
+sig_loads %>%
+  filter(endsWith(var, "DHI"),
+         axis == "RDA2") %>%
+  arrange(bec) %>%
+  select(bec, var, axis_loading) %>%
+  pivot_wider(names_from = var,
+              values_from = axis_loading)
+# when a second axis is present, it has a weak loading on the cumulative DHI
+# while the majoirty of the variability is contained within all three DHIs together, the second axis, when present, is more linked to the minmimum and variation DHI.
+
+sig_loads %>%
+  select(bec, var, axis, axis_loading, loading_sig) %>%
+  filter(endsWith(var, "DHI")) %>%
+  ggplot(aes(x = var, y = axis_loading, group = bec)) +
+  geom_point(aes(col = bec)) +
+  geom_line(aes(col = bec), alpha = 0.5, lty = "dashed") +
+  geom_hline(yintercept = 0, col = "red", alpha = 0.2) +
+  facet_wrap(~axis) +
+  theme_bw() +
+  theme(panel.grid = element_blank()) +
+  labs(x = NULL,
+       y = "Axis Loading")
+
+sig_loads %>%
+  select(bec, var, axis, axis_loading) %>%
+  filter(!endsWith(var, "DHI")) %>%
+  left_join(keys$continuous, by = c("var" = "variable")) %>%
+  ggplot(aes(x = var_long, y = abs(axis_loading), group = bec)) +
+  geom_point(aes(col = bec)) +
+  geom_line(aes(col = bec), alpha = 0.5, lty = "dashed") +
+  geom_hline(yintercept = 0, col = "red", alpha = 0.2) +
+  facet_wrap(~axis) +
+  theme(axis.text.x = element_text(angle = 45,
+                                   hjust = 1)) +
+  labs(x = NULL,
+       y = "Axis Loading (Absolute Value)")
+
+sig_loads %>%
+  filter(!endsWith(var, "DHI"),
+         loading_sig != "NS") %>%
+  group_by(bec, axis) %>%
+  mutate(loading_rank = data.table::frankv(abs(axis_loading), 
+                                           order = -1)) %>%
+  ggplot(aes(x = fct_reorder(var, loading_rank),
+             y = loading_rank, col = bec)) +
+  geom_jitter(height = 0) +
+  facet_wrap(~axis) +
+  theme(axis.text.x = element_text(angle = 45, 
+                                   hjust = 1)) +
+  labs(x = NULL,
+       y = "Axis Loading Rank")
 
@@
 
-need to get significant axes from the anova and then filter this
+```{r jenkies}
+library(rgeoda)
+sig_loads_f <- sig_loads %>%
+  filter(!endsWith(var, "DHI"),
+         loading_sig != "NS")
+
+jenkies <- sig_loads_f %>%
+  mutate(axis_loading = abs(axis_loading)) %>%
+  group_by(bec, axis) %>%
+  select(bec, axis, axis_loading) %>%
+  nest() %>%
+  mutate(breaks = map(data, natural_breaks, k = 2)) %>%
+  unnest(breaks) %>%
+  select(-data)
+
+sig_loads_f %>%
+  ggplot(aes(x = bec, y = abs(axis_loading))) +
+  facet_wrap(~axis) +
+  geom_point(data = jenkies, aes(y = breaks), col = "red", size = 3) +
+  geom_point(aes(shape = var))
+
+sig_loads_f %>%
+  left_join(jenkies) %>%
+  mutate(strong = abs(axis_loading) > breaks) %>%
+  ggplot(aes(x = bec, y = var)) +
+  geom_tile(aes(fill = strong, alpha = abs(axis_loading)))
+  geom_point(aes(col = strong))
+  left_join(keys$bec, by = c("bec" = "zone")) %>%
+  group_by(var) %>%
+  summarize(per_strong = sum(strong) / n())
+
+# We identified large contributors to each axis using a thresholding process.
+The threshold for each BEC zone and significant RDA axis was determined by using the Jenks natural breaks optimization method with k = 2 on significant loadings.
+We then summarized the most common large contributors across BEC zones and our BEC groupings.
+```
+
+```{r cutoffs}
+for_cutoff_plot <- sig_loads_f %>%
+  select(bec, axis, var, axis_loading) %>%
+  crossing(cutoff = seq(0, 1, 0.01)) %>%
+  mutate(over_cut = abs(axis_loading) > cutoff) %>%
+  group_by(var, cutoff) %>%
+  summarize(prop_over_cut = sum(over_cut) / n())
+
+for_cutoff_plot %>%
+  left_join(keys$continuous, by = c("var" = "variable")) %>%
+  ggplot(aes(x = cutoff, y = prop_over_cut, col = var_long)) +
+  geom_line() +
+  theme(legend.position = c(0.8, 0.6)) +
+  labs(x = "Cutoff Value",
+       y = "Proportion of loadings above cutoff",
+       col = "Variable") +
+  scale_y_continuous(labels = scales::label_percent()) +
+  scale_color_discrete_c4a_cat(palette = "met.tsimshian") +
+  geom_vline(xintercept = c(0.2, 0.25, 0.3), col = "red", lty = "dashed")
+```
+
+calculate the actual variance explained
 

### 2023-04-18 scripts/manuscript/manuscript.qmd
added strata instead of just bec

@@
 # include all libraries required by the R chunks here
-library(lavaan)
+library(vegan)
 library(sgsR)
@@
 community around redundancy in remote sensing metrics when describing
-structure and function on ecosystem and proposes a method to examine
-potential overlap when generating remote sensing EBV's.
+the structure and function of ecosystems and proposes a method to
+examine potential overlap when generating remote sensing EBV's.
 
@@
 subzones, variants, and phases based on microclimate, precipitation, and
-topography [@pojar1987].
+topography [@pojar1987]. To examine trends across the large
+environmental gradients, we group the BEC zones into five broad biomes,
+specifically, the south interior, northern, montane, alpine, and coastal
+groups similar to @hamann2006.
 
@@
 than 0.5 in the Lorey's height and canopy cover forest structural
-metrics. A maximum of 3000 samples were sampled in each BEC zone. To
-meet the normality assumptions of structural equation modelling, all
-variables were natural-log transformed and standardized, as per
-@grace2016. Variables containing zeros were natural-log plus one
-transformed.
+metrics. A maximum of 3000 samples were sampled in each BEC zone. All
+variables were natural-log transformed and standardized. Variables
+containing zeros were natural-log plus one transformed.
 
@@
 
-### Path Analysis
+### Partial Redundancy Analysis and Variation Partitioning
 
-To determine the relationships between simple lidar derived forest
-structural attributes, complex/derived forest structural attributes, and
-forest productivity, we used path analysis to analyze two causal models
-(@fig-pathdag). To determine primary drivers of the three DHIs, we will
-summarize the predictors across ecosystems by counting the strongest
-predictor in each BEC zone. This will determine if the primary driver
-for each facet of the DHI is simple or complex, and allow us to assess
-ecosystem differences.
-
-```{r}
-#| label: fig-pathdag
-#| fig-cap: Proposed path diagrams.
-#| warning: false
-#| echo: false
-
-knitr::include_graphics(here::here("outputs", "path_dags.png"))
-```
-
-### Exploratory Structural Equation Modelling
-
-Secondly, we will use exploratory structural equation modelling
-[@marsh2020; @asparouhov2009] to identify latent forest structural
-variables within the data, and use these latent variables to predict the
-dynamic habitat indices. Exploratory structural equation modelling is a
-combination of exploratory factor analysis (EFA) and structural equation
-modelling, which relaxes the strict requirement of zero cross-loadings
-in confirmatory factor analysis and allows for less strict measurement
-models to be used. ESEM is used when it is known that there is a latent
-structure to the data, but the specific indicator variables have not yet
-been determined.
-
-One advantage of ESEM is that it can create varying numbers of latent
-variables. For our analysis, we first ran exploratory factor analysis on
-all forest structural attributes both globally and by BEC zone with
-between 1-4 potential latent variables. We chose the most parsimonious
-EFA model with the lowest AIC scores for each ecosystem, leading to
-varying numbers of latent variables. We then determined the anchoring
-indicator for each latent variable. The anchors were calculated by
-choosing the indicator variable with the largest difference between the
-maximum value in a given loading compared to to said indicators loadings
-in all other latent variables. Each anchor variable was then assigned to
-a named group (Canopy Cover, Height and Biomass, and Structural
-Complexity).
-
-Following the EFA, we used structural equation modelling, with latent
-variables loadings determined by the EFA. We also filtered the loadings
-to be greater than or equal to 0.5, allowing us to examine the number
-and composition of latent variables found in each ecosystem. Following
-this, we predicted the DHIs in a single SEM with covariances between the
-DHIs. If a forest structural attribute did not end up in a latent
-variable, it was included in the structural equation model as a DHI
-predictor, without a latent variable as a mediator.
+Partial redundancy analysis and variation partitioning was used to
+relate the primary and modeled forest structure variables to ecosystem
+structure, to assess the inter-relationships on forest ecosystem
+function across the broad environment range. Redundancy analysis has
+widely been used in community ecology where environmental variables of
+interest are compared to species composition, and when the effect of
+space for example is assessed as a third covariate, in an approach known
+as partial redundancy analysis. While RDA is common in the ecological
+literature, this represents one of the first times this technique has
+been applied to assess redundancy in proposed essential biodiversity
+variables. Redundancy analysis utilizes multiple regression in that it
+models the effect of an explanatory matrix on a response matrix; with
+the key difference being the use of a response matrix rather than a
+single response variable. The first step is a multiple regression where
+each of the variables in the response and explanatory matrices are
+regressed resulting in a matrix of fitted values. PCA is then used in
+the second step to reduce dimensionality through eigenvectors and to
+allow the strength of relationships between assessed, taking into
+account the number of explanatory variables. Partial Redundancy Analysis
+(PRA) allows the use of additional explanatory variables, or
+co-variates. This is useful, as it allow the assessment of the effect of
+one set of variables on another while taking into account variations due
+to other third set that are not the focus of the study. The output is
+similar to standard RDA except it is also possible to examine how much
+variation is explained in the primary variables of interest, and the
+co-variates alone, and then when combined together. This variance
+partitioning is then shown typically through a Venn diagram, in which
+the percentage of explained variance by each explanatory data set is
+reported. We use ANOVAs to test which canonical axes are significant,
+and eigenvalues to determine the proprotion of variance explained by
+each axis.
 
@@
 used *sgsR* [@goodbody2023] for sampling, *terra* [@R-terra] for raster
-analyses, *sf* for vector analyses [@R-sf], and *lavaan* [@lavaan2012]
-for path analysis and structural equation modelling.
+analyses, *sf* for vector analyses [@R-sf], and *vegan* [@R-vegan] for
+redundancy analysis and variation partitioning.
 
@@
 
-## Path Analysis
-
-```{r}
-#| label: fig-pathbar-prop
-#| fig-cap: Proportion of strongest signficant forest structural predictors of yearly productivity metrics in models based on canopy cover vs canopy height.
-#| warning: false
-#| echo: false
-
-knitr::include_graphics(here::here("outputs", "path_bars.png"))
-```
-
-```{r}
-#| label: fig-pathbar-pred
-#| fig-cap: Bar plot of strongest predictors in path analysis for each BEC zone and predicted DHI variable.
-#| warning: false
-#| echo: false
-
-knitr::include_graphics(here::here("outputs", "strongest_predictor_bar.png"))
-```
-
-```{r}
-#| label: fig-pathmap
-#| fig-cap: Map of strongest structural attribute predictor strength on yearly productivity metrics by BEC zone for the two proposed models. If there are no significant predictors, shown in dark grey
-#| warning: false
-#| echo: false
-
-knitr::include_graphics(here::here("outputs", "path_map_edit.png"))
-```
-
-## Exploratory Structural Equation Modelling
-
-I also want to run exploratory SEM with and without the DHI's to see
-what grouping it goes in. MinDHI has been breaking because some zones
-only have 0 as mindhi (makes sense, they are in the alpine).
-
-```{r}
-#| label: fig-esem-latent
-#| fig-cap: Groups of latent variables with loadings greater than or equal to 0.5 for each BEC zone.
-Loadings were determined using exploratory factor analysis with up to four latent variables, and selecting the number of possible latent variables with the lowest AIC.
-All forest structure variables were potentially included as indicators in each latent variable.
-#| warning: false
-#| echo: false
-
-knitr::include_graphics(here::here("outputs", "latent_groups.png"))
-```
-
-Exploratory factor analysis shows three groups of latent variables
-(@fig-esem-latent). We assigned the groups to latent variables
-indicating canopy cover / biomass metrics, canopy height / biomass
-metrics, and structural complexity metrics. Indicators to latent
-variable groups by examining anchors for each latent variable. The
-anchors were calculated by choosing the indicator variable with the
-largest difference between the maximum value in a given loading compared
-to to said indicators loadings in all other latent variables.
-
-Both the Ponderosa Pine (PP) and Sub-boreal Pine -- Spruce zones have
-one latent variable. Coastal Western Hemlock (CWH), Engelmann Spurce --
-Subalpine Fir (ESSF), and Montane Spruce (MS) are the only zones to have
-structural complexity based latent variables. CWH is the only zone to
-include elevation standard deviation in the structural complexity latent
-variable.
-
-```{r}
-#| label: fig-esem-latent-dhi
-#| fig-cap: Groups of latent variables with loadings greater than or equal to 0.5 for each BEC zone, including the DHIs.
-Loadings were determined using exploratory factor analysis with up to five latent variables, and selecting the number of possible latent variables with the lowest AIC.
-All forest structure variables were potentially included as indicators in each latent variable.
-#| warning: false
-#| echo: false
-
-knitr::include_graphics(here::here("outputs", "latent_dhis.png"))
-```
-
-Exploratory factor analysis shows three groups of latent variables. We
-assigned these groups to attributes based on the indicator variables in
-each latent variable, indicating canopy cover, height and biomass, and
-structural complexity. When including the DHIs in this EFA, it is rare
-that they are included with latent variables with large loadings. The
-cumulative DHI is an indicator for the canopy cover latent varible in
-the IDF (Interior Douglas-fir) and MS (Montane Spruce) BEC zones. The
-DHIs are not included in any other latent variable, indicating that they
-compose different information than the forest structural variables.
-
-```{r}
-#| label: fig-latent-boxplots
-#| fig-cap: Boxplots of latent variable loadings across BEC zones as determined by exploratory factor analysis.
-Red dashed lines show latent variable indicator cutoff before being used as starting values for the SEM models.
-#| warning: false
-#| echo: false
-
-knitr::include_graphics(here::here("outputs", "latent_boxplot.png"))
-```
-
-@fig-latent-boxplots shows boxplots of latent variable loadings from
-forest structure indicator variables. A maximum of three valid latent
-variables were identified across the sixteen BEC zones. We assigned
-indicators variables to latent variable groups by examining the
-anchoring variables for each latent variable. The anchoring variables
-were calculated by choosing the indicator variable with the largest
-difference between the maximum value in a given loading compared to said
-indicators loadings in all other latent variables. The only DHI variable
-that was included in the latent variables after subsetting the latent
-variables to only include strong loadings was the Cumulative DHI in the
-canopy cover latent variable.
-
-I need to actually make the DAGs and plot the parameters, but I haven't
-gotten there yet. I'll plot the global one to show what it looks like,
-then aggregate it in some way - probably similar to the path analysis. I
-also may include the variable partitioning that Meg talked about in the
-email if I can get it working.
-
 # Discussion

### 2023-06-01 scripts/4_vegan.qmd
secondary sampling scheme, revised manuscript for ncc

--- 2023-04-12 scripts/4_vegan.qmd
@@
 ```{r packages}
-library("vegan")
+library(vegan)
+library(tidyverse)
+library(cols4all)
+library(terra)
+library(sf)
+library(patchwork)
+library(ggradar)
+source(here::here("scripts", "get_keys.R"))
 ```
@@
 bec_sample_loc <- here::here("data", "samples", "bec_extracted.shp")
-
-if (!file.exists(bec_sample_loc)) {
-  bec_samples_e <- extract(all_rasts, bec_samples, bind = T)
-  writeVector(bec_samples_e, bec_sample_loc, overwrite = T)
-}
 
@@
 
-bec_normal_ndup <- bec_df %>%
-  group_by(across(c(bec, category, basal_area:total_biomass))) %>%
-  dplyr::summarize(across(ends_with("DHI"), mean)) %>%
-  ungroup() %>%
-  mutate(across(basal_area:VarDHI,  .fns = ~ normal(.x))) 
 ```
@@
          -percentage_first_returns_above_mean,
-         -elev_stddev)
+         -elev_stddev,
+         -loreys_height)
+
+forest_types <- bec_normal %>%
+  pull(category) %>%
+  unique()
+
+# to_rda <- bec_normal %>% 
+#   group_by(category) %>% 
+#   slice_sample(n = 3000)  %>% 
+#   filter(category == "Mixed Wood") %>% 
+#   ungroup()
 
@@
 modelled <- to_rda %>%
-  select(loreys_height,
-         basal_area,
+  select(basal_area,
          total_biomass,
@@
   rda(dhi_tab ~ elev_p95 + percentage_first_returns_above_2m + elev_cv + # simple/raw/extracted
-      basal_area + gross_stem_volume + loreys_height + total_biomass, # modelled
+      basal_area + gross_stem_volume + total_biomass, # modelled
       data = to_rda)
-
-rda_out
 
@@
 
-rda_eig <- rda_out$CCA$eig
+rda_sum <- summary(rda_out)
+
+var_exp <- rda_sum$constr.chi / rda_sum$tot.chi
+
+rda_eig <- rda_out$CCA$eig %>%
+  as.data.frame() %>%
+  t() %>%
+  as.data.frame() %>%
+  tibble() %>%
+  pivot_longer(cols = everything()) %>%
+  mutate(prop = value / sum(value),
+         var_exp = prop * var_exp)
 
@@
 
-summary(rda_out)$sites %>% 
+rcorr <- summary(rda_out)$sites %>% 
   as.data.frame() %>% 
-  select(RDA1) %>% 
+  select(starts_with("RDA")) %>% 
   bind_cols(to_rda) %>%
   select(-bec, -category) %>% 
-  cor() %>%
+  as.matrix() %>%
+  Hmisc::rcorr()
+
+rvals <- rcorr$r %>%
   as.data.frame() %>%
-  select(RDA1) %>%
-  arrange(desc(abs(RDA1)))
+  select(starts_with("RDA")) %>%
+  mutate(var = row.names(.)) %>%
+  pivot_longer(starts_with("RDA"),
+               values_to = "axis_loading",
+               names_to = "axis") %>%
+  filter(!startsWith(var, "RDA"))
+
+pvals <- rcorr$P %>%
+  as.data.frame() %>%
+  select(starts_with("RDA")) %>%
+  mutate(var = row.names(.)) %>%
+  pivot_longer(starts_with("RDA"),
+               values_to = "axis_p",
+               names_to = "axis") %>%
+  filter(!startsWith(var, "RDA"))
+
+all_data_loadings <- left_join(rvals, pvals) %>%
+  filter(axis != "RDA3") %>%
+  mutate(sig = cut(axis_p, 
+                   breaks = c(0, 0.001, 0.01, 0.05, .10, 1),
+                   labels = c("***", "**", "*", ".", "NS"),
+                   right = F))
+
+for_radar_loads <- all_data_loadings %>%
+  filter(!endsWith(var, "DHI")) %>%
+  mutate(axis_loading = ifelse(sig == "NS", 0, abs(axis_loading))) %>%
+  select(var, axis, axis_loading) %>%
+  pivot_wider(names_from = var, values_from = axis_loading) %>%
+  mutate(cluster = paste(axis, "- All Data")) %>%
+  select(-axis) %>%
+  relocate(cluster)
 
@@
 ```{r bec-nest}
-to_nest <- bec_normal
+to_nest <- bec_normal %>%
+  select(-percentage_first_returns_above_mean,
+         -loreys_height,
+         -elev_stddev,
+         -elev_mean)
 
@@
   nest() %>%
+  ungroup() %>%
+  rename(strata = bec)
+
+# vlce sampling regime to get the samples up
+vlce_samples <- read_sf(here::here("data", "samples", "vlce_extracted.shp"))
+
+vlce_df <- vlce_samples %>%
+  st_drop_geometry() %>%
+  select(-category_1, -category_2) %>%
+  rename(category = strata) %>%
+  relocate(bec) %>%
+  drop_na() %>%
   ungroup()
+
+names(vlce_df) <- names_manual
+
+vlce_normal <- vlce_df %>%
+  mutate(across(basal_area:VarDHI,  .fns = ~ normal(.x)))
+
+to_nest <- vlce_normal %>%
+  select(-percentage_first_returns_above_mean,
+         -loreys_height,
+         -elev_stddev,
+         -elev_mean)
+
+class_nested <- to_nest %>%
+  group_by(category) %>%
+  slice_sample(n = 3000) %>%
+  nest() %>% 
+  ungroup() %>%
+  rename(strata = category)
+
+strata_nested <- bind_rows(bec_nested, class_nested)
 ```
@@
   rda(dhi_tab ~ elev_p95 + percentage_first_returns_above_2m + elev_cv + # simple/raw/extracted
-      basal_area + gross_stem_volume + loreys_height + total_biomass, # modelled, put condition here for a conditional rda
+      basal_area + gross_stem_volume + total_biomass, # modelled, put condition here for a conditional rda
       data = tibble)
@@
   modelled <- tibble %>%
-    select(loreys_height,
-           basal_area,
+    select(basal_area,
            total_biomass,
@@
   pivot_wider()
+
+fracts_out %>% 
+  mutate(total = extracted + modelled + overlap) %>%
+  arrange(strata %in% forest_types,
+          desc(total))
+```
+
+```{r}
+fract_all <- path_r2 %>% 
+  pivot_wider(names_from = type, values_from = Adj.R.squared) %>% 
+  select(-residual) %>%
+  mutate(strata = "All Data") %>% 
+  relocate(strata)
+
+fracts_all <- bind_rows(fracts_out,
+                        fract_all)
+
+fracts_tot <- fracts_all %>% 
+  mutate(total = extracted + modelled + overlap) %>%
+  select(strata, total)
+
+fracts_bar_tab <- fracts_all %>%
+  pivot_longer(-strata) %>%
+  left_join(keys$bec, by = c("strata" = "zone")) %>%
+  mutate(group = ifelse(is.na(group), "Forest Types", group),
+         group = ifelse(strata == "All Data", "All", group)) %>%
+  left_join(fracts_tot) %>%
+  mutate(zone_nm = ifelse(is.na(zone_nm), strata, zone_nm))
+
+fracts_bar_tab
+
+fracts_bar_plot <- fracts_bar_tab %>%
+  ggplot(aes(x = value, y = fct_reorder(zone_nm, total), fill = fct_rev(fct_relevel(str_to_title(name), "Overlap", after = 1)))) +
+  geom_col(position = "stack") +
+  facet_grid(rows = vars(fct_reorder(group, total, .desc = T)),
+             scales = "free_y",
+             space = "free") +
+  labs(x = "Variance Explained",
+       y = NULL,
+       fill = NULL) +
+  theme(legend.position = "bottom",
+        panel.spacing.y = unit(0, "lines")) +
+  guides(fill = guide_legend(reverse = T)) +
+  scale_x_continuous(labels = scales::label_percent())
+
+fracts_bar_plot
+ggsave(here::here("outputs", "fracts_bar_plot.png"), plot = fracts_bar_plot)
 ```
@@
 fracts_out %>%
-  left_join(keys$bec, by = c("bec" = "zone")) %>%
-  pivot_longer(extracted:overlap) %>%
+  mutate(total = extracted + modelled + overlap) %>%
+  #left_join(keys$bec, by = c("bec" = "zone")) %>%
+  pivot_longer(-strata) %>%
   mutate(name = str_to_title(name)) %>%
-  mutate(value = ifelse(value < 0, 0, value)) %>%
+  mutate(value = ifelse(value < 0, 0, value),
+         class = ifelse(strata %in% forest_types, 
+                        "Forest Types", 
+                        "BEC Zones")) %>%
   ggplot(aes(x = value, y = name)) +
@@
     #position = "stack",
-    aes(fill = group, col = group),
-    alpha = 0.5
+    # aes(fill = group, col = group),
+    alpha = 0.5,
+    aes(shape = class)
   ) +
-
   labs(y = NULL,
@@
   geom_segment(aes(xend = x2, yend = y2), arrow = arrow()) +
-  geom_text(aes(x = 0, y = -1.15), label = "Modelled") +
-  geom_text(aes(x = 0, y = 1.15), label = "Extracted") +
-  geom_text(aes(x = .65, y = 0), label = "Overlap") +
-  geom_text(aes(x = 2, y = 0), label = "DHIs") +
   geom_point(data = tibble(x1 = c(.85, .85), 
                            y1 = c(-.5, .5)),
-             size = 75,
+             size = 60,
              shape = 21,
              col = "grey") +
+  # geom_text(aes(x = 0, y = -1.15), label = "Modelled\nAttributes") +
+  # geom_text(aes(x = 0, y = 1.15), label = "Extracted\nAttributes") +
+  # geom_text(aes(x = .65, y = 0), label = "Overlap") +
+  # geom_text(aes(x = 2, y = 0), label = "Dynamic\nHabitat\nIndices") +
   theme_void() +
   scale_x_continuous(limits = c(-0.1, 2.1))
+
+ggsave(here::here("outputs", "path_base.png"), path_base)
 
@@
 path_base +
-  geom_text(data = path_r2_clean, aes(label = r2_label)) +
-  labs(title = paste(var_oi, collapse = " "))
-```
+  geom_text(data = path_r2_clean, aes(label = r2_label))
 
-```{r group-climate-spatial}
-climate_folder <- here::here("data", "climate")
-
-clim_vars <- climate_folder %>%
-  list.files(pattern = "csv$") %>%
-  tools::file_path_sans_ext()
-
-clim_vars <- clim_vars[!str_detect(clim_vars, "bec")]
-
-group_clim_avg <-
-  here::here(climate_folder, paste0(clim_vars, ".csv")) %>%
-  read_csv() %>%
-  mutate(var = clim_vars %>% rep(5) %>% sort) %>%
-  select(group = BEC_CSV_GR, mean = MEAN, var)
-
-```
-
-```{r clim-plot-group}
-fracts_clims <- fracts_out %>%
-  left_join(keys$bec, by = c("bec" = "zone")) %>%
-  group_by(group) %>%
-  summarize(across(extracted:overlap, .fns = mean)) %>%
-  left_join(group_clim_avg, multiple = "all")
-
-clim_var = "tmax"
-
-fracts_clims %>%
-  pivot_longer(extracted:overlap) %>%
-  filter(var == clim_var) %>%
-  ggplot(aes(x = mean, y = value)) +
-  geom_smooth(method = "lm", col = "red", lty = "dashed") +
-  geom_point() +
-  facet_wrap(~str_to_title(name)) +
-  scale_y_continuous(labels = scales::label_percent()) +
-  labs(x = clim_var,
-       y = "% Variance Explained") +
-  theme_bw() +
-  theme(panel.grid = element_blank())
-```
-
-```{r clim-plot-bec}
-clim_vars <- climate_folder %>%
-  list.files(pattern = "csv$") %>%
-  tools::file_path_sans_ext()
-
-clim_vars <- clim_vars[!str_detect(clim_vars, "bec")]
-
-bec_clim <- climate_folder %>%
-  list.files(pattern = "bec.csv$",
-             full.names = T) %>%
-  read_csv() %>%
-  mutate(var = clim_vars %>% rep(16) %>% sort) %>%
-  select(bec = ZONE, mean = MEAN, var)
-
-fracts_out %>%
-  pivot_longer(extracted:overlap) %>%
-  left_join(bec_clim, multiple = "all") %>%
-  filter(var == "ppt") %>%
-  ggplot(aes(x = mean, y = value)) +
-  geom_point() +
-  geom_smooth(method = "lm") + 
-  facet_wrap(~name) 
-```
-
-```{r testing-idf-live}
-to_rda <- bec_normal %>%
-  filter(bec == "IDF")
-
-dhi_tab <- to_rda %>%
-  select(ends_with("DHI"))
-
-extracted <- to_rda %>%
-  select(elev_p95,
-         elev_cv,
-         percentage_first_returns_above_2m)
-
-modelled <- to_rda %>%
-  select(loreys_height,
-         basal_area,
-         total_biomass,
-         gross_stem_volume)
-
-rda_out <- 
-  rda(dhi_tab ~ elev_p95 + percentage_first_returns_above_2m + elev_cv + # simple/raw/extracted
-      basal_area + gross_stem_volume + loreys_height + total_biomass, # modelled
-    data = to_rda)
-
-rda_out
-
-ordiplot(rda_out, choices = c(1, 2))
-
-rda_eig <- rda_out$CCA$eig
-
-var_out <- varpart(dhi_tab, extracted, modelled)
-
-plot(var_out, Xnames = c("Extracted", "Modelled"))
-
-path_r2 <- var_out$part$indfract %>%
-      mutate(type = c("extracted", "modelled", "overlap", "residual")) %>%
-      select(type, Adj.R.squared) %>%
-  tibble()
-
-path_r2_clean <- path_r2_plot %>%
-  left_join(path_r2) %>%
-  mutate(r2_label = paste0(round(Adj.R.squared * 100, 2), "%"))
-
-path_base +
-  geom_text(data = path_r2_clean, aes(label = r2_label)) +
-  labs(title = "All")
-
-summary(rda_out)$sites %>% 
-  as.data.frame() %>% 
-  select(RDA1) %>% 
-  bind_cols(to_rda) %>%
-  select(-bec, -category) %>% cor() %>%
-  as.data.frame() %>%
-  select(RDA1) %>%
-  arrange(desc(abs(RDA1)))
+ggsave(here::here("outputs", "path_venn.png"),
+        height = 3.5, width = 10)
 ```
@@
 
-```{r cdf-anova}
-outs %>% 
-  filter(bec == "CDF") %>%
-  select(rda) %>% 
-  pull(rda) %>%
-  .[[1]] %>%
-  anova()
-```
-
 ```{r cum-min-hexplot}
@@
         relocate(axis)
-  }))
+  }, .progress = "anovas"))
 
 anova_sig_axes <- anova_outs %>%
-  select(bec, anova) %>%
-  unnest(cols = anova) %>%
-  filter(anova_sig != "NS")
+  select(strata, anova) %>%
+  unnest(cols = anova) # %>%
+  # filter(anova_sig != "NS")
+
+axis_varexp <- anova_sig_axes %>%
+  inner_join(eig_outs)
+
+axis_varexp %>%
+  filter(anova_sig != "NS") %>%
+  filter(strata %in% forest_types)
+
+axis_varexp %>%
+  filter(anova_sig != "NS") %>%
+  mutate(category = ifelse(strata %in% forest_types, "Forest Types", "BEC Zones")) %>%
+  # left_join(keys$bec, by = c("bec" = "zone")) %>%
+  ggplot(aes(x = axis, y = var_exp, col = category, shape = category)) +
+  geom_jitter(height = 0, width = 0.05) +
+  theme(legend.position = "bottom") +
+  labs(x = NULL,
+       y = "% of Variance Explained",
+       col = NULL,
+       shape = NULL) +
+  scale_y_continuous(labels = scales::label_percent())
+# no secondary axis in alpine & coast.
+
+axis_varexp %>%
+  group_by(strata) %>%
+  # filter(n() == 2) %>%
+  select(strata, axis, var_exp) %>% 
+  pivot_wider(names_from = axis,
+              values_from = var_exp) %>%
+  mutate(ratio = RDA1 / RDA2)
+
+# smaller ratios present when the RDA1 doesn't explain much. when variance explained
+# is low, the ratio is low. this is because there isn't much variance explained
+```
+
+```{r varexp-coniferous}
+rda_varexp <- outs %>%
+  mutate(varexp = map(rda, .f = \(rda) {
+    rda_sum <- summary(rda)
+    var_exp <- rda_sum$constr.chi / rda_sum$tot.chi
+  })) %>%
+  unnest(varexp) %>%
+  arrange(desc(varexp)) %>%
+  select(strata, varexp)
+
+rda_varexp %>%
+  arrange(strata %in% forest_types,
+          desc(varexp))
+
+source(here::here("scripts", "get_bec_proportion_vlce.R"))
+
+prop_con <- bec_normal %>% 
+  select(bec, category) %>% 
+  group_by(bec) %>% 
+  summarize(per_sample_for = mean(category == "Coniferous")) %>% 
+  left_join(per_cons %>% 
+              select(-value)) %>% 
+  mutate(for_dif = per_sample_for - per_con) %>%
+  arrange(desc(abs(for_dif))) %>%
+  select(-for_dif) %>%
+  pivot_longer(starts_with("per"))
+
+con_var_plot <- prop_con %>%
+  filter(name == "per_con") %>%
+  left_join(rda_varexp, by = c("bec" = "strata")) %>% 
+  left_join(keys$bec, by = c("bec" = "zone")) %>%
+  mutate(name = ifelse(name == "per_con", 
+                       "Global Percent Coniferous", 
+                       "Sample Percent Coniferous")) %>%
+  ggplot(aes(x = value, y = varexp)) +
+  geom_point(aes(col = group)) +
+  geom_smooth(method = "lm") +
+  scale_x_continuous(labels = scales::label_percent()) +
+  scale_y_continuous(labels = scales::label_percent()) +
+  labs(x = "Percent Coniferous",
+       y = "Proportion of variance explained",
+       col = NULL) +
+  theme(legend.position = c(0.1, 0.85))
+
+ggsave(here::here("outputs", "con_var.png"), plot = con_var_plot)
 ```
@@
 out_loadings <- outs %>%
-  mutate(loadings = map(rda,.f = function(rda) {
+  mutate(loadings = map(rda, .f = function(rda) {
     summary(rda)$sites %>%
@@
   loadings = map(loadings, .f = function(x) {
-    rcorr <- x %>% select(-category) %>%
+    rcorr <- x %>% select(-any_of(c("bec", "category"))) %>%
       as.matrix() %>%
@@
   )) %>%
-  select(bec, loadings) %>%
+  select(strata, loadings) %>%
   unnest(loadings)
@@
   #filter(axis_sig == "NS") %>%
-  arrange(axis_loading) %>%
-  count(var)
+  arrange(desc(abs(axis_loading))) %>%
+  filter(!(var %in% c("percentage_first_returns_above_mean", 
+                    "elev_mean", 
+                    "elev_stddev"))) %>%
+  drop_na()# removes ima mindhi which has no variance
+
+sig_loads %>%
+  filter(anova_sig != "NS") %>%
+  filter(endsWith(var, "DHI"),
+         axis == "RDA2") %>%
+  arrange(strata) %>%
+  select(strata, var, axis_loading) %>%
+  pivot_wider(names_from = var,
+              values_from = axis_loading)
+# when a second axis is present, it has a weak loading on the cumulative DHI, as compared to the other DHIs
+# while the majoirty of the variability is contained within all three DHIs together, the second axis, when present, is more linked to the minmimum and variation DHI.
+
+sig_loads %>%
+  select(strata, var, axis, axis_loading, loading_sig) %>%
+  filter(endsWith(var, "DHI"),
+         loading_sig != "NS") %>%
+  ggplot(aes(x = var, y = abs(axis_loading), group = strata)) +
+  geom_point(aes(col = strata)) +
+  geom_line(aes(col = strata), alpha = 0.5, lty = "dashed") +
+  geom_hline(yintercept = 0, col = "red", alpha = 0.2) +
+  facet_wrap(~axis) +
+  theme_bw() +
+  theme(panel.grid = element_blank()) +
+  labs(x = NULL,
+       y = "Axis Loading")
+
+sig_loads %>%
+  select(strata, var, axis, axis_loading) %>%
+  filter(!endsWith(var, "DHI")) %>%
+  left_join(keys$continuous, by = c("var" = "variable")) %>%
+  ggplot(aes(x = var_long, y = abs(axis_loading), group = strata)) +
+  geom_point(aes(col = strata)) +
+  geom_line(aes(col = strata), alpha = 0.5, lty = "dashed") +
+  geom_hline(yintercept = 0, col = "red", alpha = 0.2) +
+  facet_wrap(~axis) +
+  theme(axis.text.x = element_text(angle = 45,
+                                   hjust = 1)) +
+  labs(x = NULL,
+       y = "Axis Loading (Absolute Value)")
+
+sig_loads %>%
+  filter(!endsWith(var, "DHI"),
+         loading_sig != "NS") %>%
+  group_by(strata, axis) %>%
+  mutate(loading_rank = data.table::frankv(abs(axis_loading), 
+                                           order = -1)) %>%
+  ggplot(aes(x = fct_reorder(var, loading_rank),
+             y = loading_rank, col = strata)) +
+  geom_jitter(height = 0) +
+  facet_wrap(~axis) +
+  theme(axis.text.x = element_text(angle = 45, 
+                                   hjust = 1)) +
+  labs(x = NULL,
+       y = "Axis Loading Rank")
 
@@
 out_loadings %>% count(var)
+
+clean_loads <- sig_loads %>%
+  filter(anova_sig != "NS",
+         loading_sig != "NS") %>%
+  bind_rows(all_data_loadings %>%
+              mutate(strata = "All Data") %>%
+              rename(loading_sig = sig)) %>%
+  select(strata:axis_loading) %>%
+  mutate(axis_loading = abs(axis_loading))
 ```
 
-need to get significant axes from the anova and then filter this
+##### run to here for analysis to be done
+radar chart based on groups, not hierarchical clustering
 
-get total rda eigenvalues.
-find proportion of total for each rda.
-multiple by 1 - residual.
-and thats gets path number b/w RD1/RD2 and DHIs.
-can just filter each by which exist since only a few.
-goes for path plot
+```{r radar}
+gg_color_hue <- function(n) {
+  hues = seq(15, 375, length = n + 1)
+  hcl(h = hues, l = 65, c = 100)[1:n]
+}
 
-grab the p values w/in zones as well as the correlation coefficients using Hmisc::rcorr(). only plot these plus loadings over threshold
+n = 8
+cols = c4a(palette = "rainbow", n = 8)
 
-check if deciduous and coniferous have signfiicant axes
+radar_loads <- clean_loads %>%
+  pivot_wider(names_from = var, 
+              values_from = axis_loading) %>%
+  left_join(keys$bec, by = c("strata" = "zone")) %>%
+  mutate(group = ifelse(is.na(group), "Forest Types", group),
+         group = ifelse(strata == "All Data", "All Data", group),
+         group = ifelse(axis == "RDA2", "RDA2", group)) %>%
+  group_by(group) %>%
+  summarize(across(CumDHI:elev_p95, \(x) mean(x, na.rm = T))) %>%
+  mutate(group = fct_relevel(group, "RDA2", after = Inf),
+         group = fct_relevel(group, "Forest Types", after = 1))
+
+radar_labels <- radar_loads %>%
+  select(percentage_first_returns_above_2m:elev_p95) %>%
+  colnames() %>%
+  tibble(variable = .) %>%
+  left_join(keys$continuous) %>%
+  pull(var_long)
+
+in_radar <- radar_loads %>%
+  select(!ends_with("DHI")) %>%
+  replace(is.na(.), 0) %>%
+  ggradar(axis.labels = radar_labels,
+          legend.title = "Cluster",
+          values.radar = c("", "", 1),
+          base.size = 1,
+          background.circle.transparency = 0,
+          group.point.size = 2,
+          group.colours = cols,
+          axis.label.size = 3,
+          plot.extent.x.sf = 1.5,
+          group.line.width = 1) +
+  theme(legend.position = "none")
+
+out_radar <- radar_loads %>%
+  select(group, ends_with("DHI")) %>%
+  replace(is.na(.), 0) %>%
+
+  ggradar(legend.title = "Cluster",
+          values.radar = c("", "", 1),
+          base.size = 1,
+          background.circle.transparency = 0,
+          group.point.size = 2,
+          group.colours = cols,
+          axis.label.size = 3,
+          plot.extent.x.sf = 1.5,
+          group.line.width = 1) +
+  theme(legend.position = "none")
+
+boxplots <- clean_loads %>% 
+  left_join(keys$continuous, by = c("var" = "variable")) %>%
+  mutate(var_long = ifelse(is.na(var_long), var, var_long)) %>%
+  select(!(divide_by:unit)) %>%
+  left_join(keys$bec, by = c("strata" = "zone")) %>%
+  mutate(group = ifelse(is.na(group), "Forest Types", group),
+         group = ifelse(axis == "RDA2", "RDA2", group),
+         group = ifelse(strata == "All Data", "All Data", group),
+         inout = ifelse(endsWith(var, "DHI"), "Response", "Predictor")) %>%
+  mutate(group = fct_relevel(group, "RDA2", after = Inf),
+         group = fct_relevel(group, "Forest Types", after = 1)) %>%
+  ggplot(aes(x = abs(axis_loading), y = fct_rev(fct_reorder(var_long, abs(axis_loading))))) +
+  geom_boxplot(outlier.colour = "#00000000", width = .5) +
+  geom_jitter(height = 0.1, aes(col = group, shape = group)) +
+  facet_grid(cols = vars(axis),
+             rows = vars(inout),
+             scales = "free_y",
+             space = "free_y") +
+  labs(x = "Axis Loading (absolute)",
+       y = NULL,
+       col = NULL,
+       shape = NULL) +
+  theme(panel.spacing.y = unit(0, "lines"),
+        legend.position = "right") +
+  scale_shape_manual(values = c(17, 15, 16, 16, 16, 16, 16, 16)) +
+  scale_colour_manual(values = cols)
+
+radar_boxplot <- ((in_radar / out_radar) | boxplots)  +
+  plot_annotation(tag_levels = "A") +
+  plot_layout(guides = "collect") &
+  theme(plot.tag = element_text(size = 12))
+
+ggsave(here::here("outputs", "radar_boxplot.png"),
+       plot = radar_boxplot,
+       device = "png")
+```
+
+```{r drawio}
+
+# make dir
+drawio_dir <- here::here("outputs", "drawio")
+
+dir.create(drawio_dir, showWarnings = F)
+
+# templates
+one_rda <-
+  read_lines(here::here(drawio_dir, "_rda_path-text_replace-RDA1.drawio"))
+
+two_rda <-
+  read_lines(here::here(drawio_dir, "_rda_path-text_replace-RDA2.drawio"))
+
+# list to iterate
+stratum <- clean_loads %>%
+  pull(strata) %>%
+  unique()
+
+# make sure everything gets change in rda fig
+possibles <- clean_loads %>%
+  select(var, axis) %>%
+  complete() %>%
+  mutate(possible = paste(var, axis, sep = "-")) %>%
+  pull(possible) %>%
+  unique()
+
+# fractions dfs
+fract_all <- path_r2 %>% 
+  pivot_wider(names_from = type, values_from = Adj.R.squared) %>% 
+  select(-residual) %>%
+  mutate(strata = "All Data") %>% 
+  relocate(strata)
+
+fracts_all <- fracts_out %>%
+  bind_rows(fract_all)
+
+# variance explained by axis
+
+all_varexp <- rda_eig %>%
+  filter(name != "RDA3") %>%
+  mutate(strata = "All Data") %>%
+  rename(axis = name) %>%
+  select(-value)
+
+varexp_all <- axis_varexp %>%
+  filter(anova_sig != "NS") %>%
+  select(strata, axis, prop, var_exp) %>%
+  bind_rows(all_varexp)
+
+for (s in stratum) {
+  strata_tp <- clean_loads %>%
+    filter(strata == s) %>%
+    mutate(textreplace = paste(var, axis, sep = "-"),
+           axis_loading = as.character(round(axis_loading, digits = 2)))
+
+  if (strata_tp %>% pull(axis) %>% unique() %>% length() == 1) {
+    strata_drawio <- one_rda
+  } else {
+    strata_drawio <- two_rda
+  }
+
+  textreplace <- strata_tp %>%
+    pull(textreplace)
+
+  replaceval <- strata_tp %>%
+    pull(axis_loading)
+
+  for (i in 1:length(textreplace)) {
+    text <- textreplace[[i]]
+    val <- replaceval[[i]]
+
+    strata_drawio <- str_replace(strata_drawio, text, val)
+  }
+
+  strata_drawio <- str_replace(strata_drawio, "title_strata", s)
+
+  #strata_drawio <- strata_drawio[!str_detect(strata_drawio, "RDA")]
+  # filter through all possibilities remaining, replace them with NS
+
+  for(p in possibles) {
+    strata_drawio <- str_replace(strata_drawio, p, "NS")
+  }
+
+  strata_fract <- fracts_all %>%
+    filter(strata == s) %>%
+    pivot_longer(extracted:overlap) %>%
+    mutate(value = (value * 100) %>%
+             round(1) %>%
+             paste0(., "%"))
+
+  fract_names <- strata_fract %>%
+    pull(name)
+
+  fract_value <- strata_fract %>%
+    pull(value)
+
+  for(i in 1:length(fract_names)) {
+    strata_drawio <- str_replace(strata_drawio, fract_names[[i]], fract_value[[i]])
+  }
+
+  strata_varexp <- varexp_all %>%
+    filter(strata == s) %>%
+    mutate(var_exp = (var_exp * 100) %>%
+             round(1) %>%
+             paste0(., "%"))
+
+  axis_names <- strata_varexp %>%
+    pull(axis) %>%
+    paste("var_exp", ., sep = "-")
+
+  axis_vals <- strata_varexp %>%
+    pull(var_exp)
+
+  for(i in 1:length(axis_names)) {
+    strata_drawio <- str_replace(strata_drawio, axis_names[[i]], axis_vals[[i]])
+  }
+
+  write_lines(strata_drawio, here::here(drawio_dir, paste0(s, ".drawio")))
+}
+```
+
+```{r complexity-iqr}
+strata_nested %>% 
+  unnest(data) %>% 
+  ggplot(aes(x = fct_reorder(strata, elev_cv, IQR), y = elev_cv)) + 
+  geom_boxplot() + 
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+```
+
+# rgb map
+
+```{r}
+bec_dissolve <- read_sf(here::here("data", "shapefiles", "bec_dissolve.shp")) %>%
+  janitor::clean_names() %>%
+  select(zone)
+
+# non normalized
+loadings_rgb <- clean_loads %>% 
+  group_by(axis) %>%
+  pivot_wider(names_from = var,
+              values_from = axis_loading) %>%
+  ungroup() %>%
+  replace(is.na(.), 0) %>%
+  mutate(across(elev_cv:elev_p95, \(x) ifelse(x > 0.5, 0.5, x))) %>%
+  mutate(dhi_rgb = rgb(CumDHI, VarDHI, MinDHI),
+         pri_rgb = rgb(elev_p95, percentage_first_returns_above_2m, elev_cv, maxColorValue = 1),
+         mod_rgb = rgb(basal_area, total_biomass, gross_stem_volume, maxColorValue = 1)) %>%
+  select(strata, axis, percentage_first_returns_above_2m, elev_p95, elev_cv, ends_with("rgb")) %>%
+  pivot_longer(ends_with("rgb"))
+
+# normalized to max
+loadings_rgb <- clean_loads %>% 
+  group_by(var) %>%
+  mutate(axis_loading = axis_loading / max(axis_loading)) %>%
+  group_by(axis) %>%
+  pivot_wider(names_from = var,
+              values_from = axis_loading) %>%
+  ungroup() %>%
+  replace(is.na(.), 0) %>%
+  #mutate(across(elev_cv:elev_p95, \(x) ifelse(x > 0.5, 0.5, x))) %>%
+  mutate(dhi_rgb = rgb(CumDHI, VarDHI, MinDHI),
+         pri_rgb = rgb(elev_p95, percentage_first_returns_above_2m, elev_cv, maxColorValue = 1),
+         mod_rgb = rgb(basal_area, total_biomass, gross_stem_volume, maxColorValue = 1)) %>%
+  #select(strata, axis, percentage_first_returns_above_2m, elev_p95, elev_cv, ends_with("rgb")) %>%
+  pivot_longer(ends_with("rgb"))
+
+becs <- bec_normal %>%
+  pull(bec) %>%
+  unique()
+
+rgb_loads_split <- loadings_rgb %>%
+  filter(strata %in% becs) %>%
+  group_by(axis, name) %>%
+  group_split()
+```
+
+```{r}
+working <- rgb_loads_split[[1]]
+
+make_rgb_map <- function(tibble) {
+
+  rgb_cols = tibble %>%
+    pull(value) %>%
+    sort()
+
+  bec_dissolve %>%
+    left_join(tibble, by = c("zone" = "strata")) %>%
+    mutate(ifelse(is.na(value), "grey", value)) %>%
+    ggplot() +
+    geom_sf(aes(geometry = geometry, fill = value), colour = NA) +
+    scale_fill_manual(values = rgb_cols) +
+    theme_void() +
+    theme(legend.position = "none")
+}
+
+rgb_maps <- map(rgb_loads_split, make_rgb_map)
+
+patchwork_rgbplots <- (rgb_maps[[3]] + rgb_maps[[2]]) / 
+                      (rgb_maps[[6]] + rgb_maps[[5]]) &
+  plot_annotation(tag_levels = "A")
+
+ggsave(here::here("outputs", "rgbplots_norm.png"), patchwork_rgbplots,
+       height = 6,
+       width = 8)
+
+print("png saved")
+
+ggsave(here::here("outputs", "rgbplots_norm.pdf"), patchwork_rgbplots,
+       height = 6,
+       width = 8)
+
+print("pdf saved")
+```
+
+```{r study-map}
+library(geodata)
+
+provinces <- geodata::gadm(country = "Canada", level = 1, path = here::here("data", "province"))
+
+prov_sf <- provinces %>%
+  st_as_sf() %>%
+  st_transform(3005)
+
+bbox <- prov_sf %>%
+  filter(NAME_1 == "British Columbia") %>%
+  st_bbox()
+
+usa <- geodata::gadm(country = "USA", level = 0, path = here::here("data", "usa")) %>%
+  st_as_sf()
+
+countries <- geodata::world(level = 0, path = here::here("data", "world")) %>%
+  st_as_sf()
+
+na_bbox <- countries %>%
+  filter(GID_0 == "USA" |
+         NAME_0 == "Canada") %>%
+  st_bbox()
+
+cad_bbox <- prov_sf %>%
+  st_transform(3348) %>%
+  st_bbox()
+
+countries_map <- countries %>%
+  filter(GID_0 == "USA" |
+         NAME_0 == "Greenland") %>%
+  st_transform(3348)
+
+lakes <- read_sf(here::here("data", "great_lakes", "great_lakes.shp")) %>%
+  st_transform(3348)
+
+bec_map <- ggplot() +
+  geom_sf(data = bcmaps::bc_bound(),
+          fill = "grey70", col = NA) +
+  geom_sf(
+    data = prov_sf %>% filter(NAME_1 != "British Columbia"),
+    fill = "grey70"
+  ) +
+  geom_sf(data = usa,
+          fill = "grey50") +
+  geom_sf(data = bec_dissolve %>%
+            left_join(keys$bec),
+          aes(fill = fct_reorder(zone_nm, zone)),
+          colour = NA) +
+  ggsci::scale_fill_igv() +
+  labs(fill = NULL,
+       linetype = NULL) +
+  coord_sf(
+    xlim = c(bbox$xmin - 300000, bbox$xmax),
+    ylim = c(bbox$ymin, bbox$ymax),
+    crs = 3005,
+    datum = 4326
+  ) +
+  theme_grey() +
+  theme(
+    panel.background = element_rect(fill = "#525252"),
+    panel.border = element_rect(colour = "black", fill = NA),
+    legend.position = c(.15, .4),
+    legend.background = element_rect(fill = NA),
+    legend.box.background = element_rect(fill = "white")
+  )
+
+#ggsave(here::here("outputs", "bec_map_no_inset.png"), bec_map, device = "png")
+
+inset = ggplot() +
+  geom_sf(data = countries_map, fill = "grey50") +
+  geom_sf(data = prov_sf %>% st_transform(3348) %>% filter(NAME_1 != "British Columbia"), fill = "grey70", size = 0.5) +
+  geom_sf(data = prov_sf %>% st_transform(3348) %>% filter(NAME_1 == "British Columbia"), fill = "#e5e5e5", size = 0.5) +
+  geom_sf(data = lakes, fill = "#525252") +
+  geom_sf(data = bbox %>% st_as_sfc %>% st_transform(3348), fill = NA, colour = "red", size = 2) +
+  coord_sf(xlim = c(cad_bbox$xmin - 300000, cad_bbox$xmax),
+           ylim = c(cad_bbox$ymin, cad_bbox$ymax)) +
+  theme_void() +
+  theme(panel.border = element_rect(colour = "black"),
+        panel.background = element_rect(fill = "#525252"))
+
+# map_inset <- ggdraw(bec_map) +
+#   draw_plot(inset, x = 0.7, y = 0.7, width = .28, height = .28)
+
+map_inset <- bec_map + inset_element(inset, left = 0.71,
+                                            bottom = 0.7,
+                                            right = 1,
+                                            top = 0.98, 
+                                     align_to = "plot")
+
+ggsave(here::here("outputs", "bec_map.png"), plot = map_inset, device = "png", height = 7, width = 10)
+
+```
+
+```{r bec-clim-tab}
+climate_folder <- here::here("data", "climate")
+clim_vars <- climate_folder %>%
+  list.files(pattern = "csv$") %>%
+  tools::file_path_sans_ext()
+
+clim_vars <- clim_vars[!str_detect(clim_vars, "bec")]
+
+bec_clim <- climate_folder %>%
+  list.files(pattern = "bec.csv$",
+             full.names = T) %>%
+  read_csv() %>%
+  mutate(var = clim_vars %>% rep(16) %>% sort) %>%
+  select(bec = ZONE, mean = MEAN, var) %>%
+  pivot_wider(values_from = mean,
+              names_from = var)
+
+bec_tab <- keys$bec %>%
+  left_join(bec_clim, by = c("zone" = "bec")) %>%
+  select(-order) %>%
+  arrange(group, zone) %>%
+  relocate(group)
+
+write_csv(bec_tab, here::here("data", "climate.csv"))
+
+knitr::kable(bec_tab)
+```
+
+```{r}
+bec_normal %>%
+  select(-elev_stddev,-loreys_height,-percentage_first_returns_above_mean,-elev_mean) %>%
+  left_join(keys$bec, by = c("bec" = "zone")) %>%
+  pivot_longer(basal_area:VarDHI) %>%
+  ggplot(aes(x = value, y = bec, fill = group)) +
+  geom_boxplot() +
+  facet_wrap( ~ name)
+```

### 2023-06-01 scripts/manuscript/manuscript.qmd
secondary sampling scheme, revised manuscript for ncc

@@
 ---
-title: Disentangling linkages between satellite derived forest structure and productivity across a range of forest types and ecosystems.
+title: Disentangling linkages between satellite derived forest structure and productivity essential biodiversity variables.
 author:
@@
           postal-code: V8V 1T7
-abstract: |
-  [insert abstract here - nicholas ignore the formatting on the authors for now, i can output to pdf and have it formatted nicely for submission]
+abstract: In today's changing climate and in light of the biodiversity crises, it is integral to be able to monitor the change in biodiversity at large scales.
+Key to this is the development of variables capable of monitoring biodiversity trends.
+One framework developed to tackle this problem is the Essential Biodviersity Variables (EBV), which is designed to be analogous to the Essential Climate variables, and capable of monitoring biodiversity globally while also being complementary to one another, within EBV classes.
+The EBV framework contains six classes, some of which are extremely well suited to be monitoring from satellite remote sensing.
+Among these with the potential to be satellite derived, the ecosystem structure and function classes stand out.
+In this paper, we seek to assess the complementarity of forest structure, as imputed across the entirety of British Columbia, Canada, with the Dynamic Habitat Indices, a yearly summary of productivity indices.
+Both datasets have been previously linked with biodiversity metrics across a range of scales.
+Using redundancy analysis, we find that forest structural attributes and the DHIs are decoupled from one another, with the forest structure datasets explaining 14% of the variation in the DHIs.
+Further, we explore how the proportion of variance explained in the DHIs from primary structural attributres such as canopy height and canopy cover and modelled structural attributes such as aboveground biomass varies.
+We find there is generally large overlap between the two forest structure datasets.
+Overall, we find that forest structure as an ecosystem structure EBV and the DHIs as an ecosystem function EBV are well suited to be complementary EBVs across large environmental gradients.
 keywords: 
-  - structural equation modelling
   - remote sensing
@@
   - dynamic habitat indices
+  - essential biodiversity variables
+  - redundancy analysis
 date: last-modified
@@
 library(sf)
+library(gt)
+library(tidyverse)
 
@@
 knitr::write_bib(contrib_pkgs, pkgs_file)
+
+bec_tab <- read_csv(here::here("data", "climate.csv")) %>%
+  mutate(across(.cols = c(tmax, tmin), \(x) round(x, digits = 1)))
 ```
@@
 
-Biodiversity is the summation of variation in biological life, across
-genes, species, communities, and ecosystems. Currently, biodiversity is
-in decline, facing extinction rates above the background extinction rate
-[@thomas2004; @urban2015], and homogenization of communities at various
-scales [@mcgill2015]. In response, the global biodiversity community is
-making efforts to assess and halt the degradation of biodiversity on
-earth. The Group for Earth Observation Biodiversity Observation Network
-has developed the Essential Biodiversity Variables [EBVs, @pereira2013],
-designed as an analog to the Essential Climate Variables framework
-[@bojinski2014]. EBVs are designed to be global in scope, relevant to
-biodiversity information, feasible to utilize, and complementary to one
-another [@skidmore2021]. There are six EBV classes, each of which
-correspond to a different facet of biodiversity including species
-populations, species traits, community composition, ecosystem structure,
-ecosystem function, and genetic composition [@pereira2013].
+With biodiversity being in decline, and facing extinction rates above
+the background extinction rate [@thomas2004; @urban2015], as well as the
+homogenization of communities at various scales [@mcgill2015] it is
+integral to be able to monitor how biodiversity is changing across the
+globe. In response, the global biodiversity community is making efforts
+to assess and halt the degradation of biodiversity. The Group for Earth
+Observation Biodiversity Observation Network has developed the Essential
+Biodiversity Variables [EBVs, @pereira2013], designed as an analog to
+the Essential Climate Variables framework [@bojinski2014]. EBVs are
+designed to be global in scope, relevant to biodiversity information,
+feasible to use, and complementary to one another [@skidmore2021]. While
+it can be incredibly difficult, time consuming, and expensive to collect
+data on biodiversity across wide swaths of land and varying ecosystems,
+EBVs, which can be correlated to sampled biodiversity information, allow
+for the monitoring and assessment of protected area effectiveness and
+ecosystem health at large spatial scales [@hansen2021]. There are six
+EBV classes, each of which correspond to a different facet of
+biodiversity including species populations, species traits, community
+composition, ecosystem structure, ecosystem function, and genetic
+composition [@pereira2013].
 
@@
 single-tree scale using, for example, PlanetScope imagery and
-drone-based measurements [@wu2021]. Coarser spatial measurements, such
-as those taken from satellites by the MODIS [@zhang2003], Landsat
-imaging systems[@fisher2006], or Sentinel [@helfenstein2022;
-@darvishzadeh2019] can monitor these functional processes at broader
-extents, corresponding to ecosystem function, but the coarse spatial
-resolution removes the ability to relate these traits to individual
-trees. Remote sensing data has been shown to be arguably the most
-effective at monitoring ecosystem structure and function at regional to
-global extents through the use of optical imagery [@cohen2004] and
-active sensors such as lidar and radar datasets [@lefsky2002; @lang2021;
-@neuenschwander2019].
+drone-based measurements [@wu2021]. However, the spatially limited and
+often *ad-hoc* data collection approaches associated with monitoring
+individuals is not conducive to the global or regional scales required
+for biodiversity trend assessment [@valdez2023].
 
-Notably, while remote sensing can provide information on the
-species-level EBVs, this information is typically acquired with an
-*ad-hoc* approach requiring high spatial and spectral resolution data
-[@skidmore2021]. This data can often only be collected at small extents,
-rather than the global or regional scales required for biodiversity
-trend assessment [@valdez2023]. The two landscape-level EBVs (ecosystem
-structure and function) are well suited to be examined at global or
-regional scales using moderate-resolution satellite imagery, such as
-those provided by the Landsat or Sentinel series of satellites
-[@skidmore2021].
+The two landscape-level EBVs (ecosystem structure and function) are well
+suited to be examined at large spatial scales using coarser spatial
+measurements, such as those taken from satellites by the Moderate
+Resolution Imaging Spectroradiometer [MODIS; @zhang2003], the Landsat
+imaging systems [@fisher2006], or Sentinel-2 [@helfenstein2022;
+@darvishzadeh2019] programs. These mid-resolution satellites can monitor
+processes at broader extents but the coarse spatial resolution removes
+the ability to relate these traits to individual organisms. As a result,
+satellite remote sensing data has been shown to be arguably the most
+effective at monitoring ecosystem based EBVs focused on structure and
+function. These EBVs classes can be monitored at regional to global
+extents through the use of optical imagery [@cohen2004], as well as
+active sensors such as lidar (light detection and ranging) and radar
+[@guo2017; @lefsky2002; @lang2021; @neuenschwander2019; @coops2016].
+
+## EBV - Ecosystem Structure
 
 Forest structural diversity has been linked to biodiversity at various
-scales [@guo2017; @bergen2009; @gao2014]. Many metrics derived from
-lidar remote sensing have been used as local indicators of biodiversity,
-including simple metrics such as canopy cover and canopy height as well
-as derived metrics including vertical profiles, aboveground biomass
-[@lefsky1999]. Other second order derived metrics such as canopy
-texture, height class distribution, edges, and patch metrics have also
-been used to examine habitat and biodiversity at landscape scales
-[@bergen2009]. Increased structural complexity has been hypothesized to
-create additional niches, leading to increased species diversity
-[@bergen2009], which has been frequently demonstrated using avian
-species diversity metrics [@macarthur1961]. For example: @herniman2020
-used spectral and lidar derived forest structure data to model avian
-habitat suitability; @clawges2008 found that lidar derived forest
-structural attributes are capable of identifying habitat types
-associated with avian species in pine/aspen forests; @goetz2007 used
-canopy structural diversity to predict bird species richness, finding
-that canopy vertical distribution was the strongest predictor of species
-richness. Forest structural metrics have also been used to study
-biodiversity in other clades as well [@davies2014; @nelson2005].
+scales [@guo2017; @bergen2009; @gao2014]. Structural attributes range in
+complexity from simple (canopy cover; canopy height), to more complex
+(vertical and horizontal structural complexity) to modelled (aboveground
+biomass; basal area), all of which can be assessed using lidar data
+[@coops2021]. A suite of these lidar-derived attributes have been used
+as local indicators of biodiversity, including simple metrics such as
+canopy cover and canopy height as well as derived metrics including
+vertical profiles, aboveground biomass [@lefsky1999; @guo2017;
+@coops2016]. Other second order derived metrics such as canopy texture,
+height class distribution, edges, and patch metrics have also been used
+to examine habitat and biodiversity at landscape scales [@bergen2009].
+Advances in satellite remote sensing processing have allowed 3D forest
+structure data to be imputed across wide spatial scales [@matasci2018;
+@coops2021] using data fusion approaches involving collected lidar data
+and optical/radar data.
+
+Increased forest structural complexity has been hypothesized to create
+additional niches, leading to increased species diversity [@bergen2009],
+which has been frequently demonstrated using avian species diversity
+metrics [@macarthur1961]. For example: @herniman2020 used spectral and
+lidar derived forest structure data to model avian habitat suitability;
+@clawges2008 found that lidar derived forest structural attributes are
+capable of identifying habitat types associated with avian species in
+pine/aspen forests; @goetz2007 used canopy structural diversity to
+predict bird species richness, finding that canopy vertical distribution
+was the strongest predictor of species richness. Forest structural
+metrics have also been used to study biodiversity in other clades as
+well [@davies2014; @nelson2005].
+
+## EBV - Ecosystem Function
 
@@
 has shown to be a predictor of species richness and abundances at
-various scales [@chase2002; @radeloff2019; @coops2019; @razenkova], and
-is measurable using satellite remote sensing via the use of various
+various scales [@chase2002; @radeloff2019; @coops2019; @razenkova2023],
+and is measurable using satellite remote sensing via the use of various
 vegetation indices [@huete2002; @radeloff2019]. Vegetation indices,
@@
 time points [@bonn2004], and more recently, through yearly summaries of
-productivity [@radeloff2019]. The relationship between energy
-availability and biodiversity occurs via various hypotheses, such as the
-available energy hypothesis, the environmental stress hypothesis, and
-the environmental stability hypothesis. These three hypotheses have in
-turn been linked to patterns of surface reflectance in remote sensing
-data [@radeloff2019]. The Dynamic Habitat Indices (DHIs) have been shown
-to be well suited to assess these hypotheses. The cumulative DHI
-calculates the total amount of energy available in a given pixel over
-the course of a year. Cumulative DHI is strongly linked to the available
-energy hypothesis, which suggests that with greater available energy
-species richness will increase [@wright1983]. The minimum DHI, which
-calculates the lowest productivity over the course of a year can be
-matched to the environmental stress hypothesis, which proposes that
-higher levels of minimum available energy will lead to higher species
-richness [@currie2004]. Finally, the variation DHI, which calculates the
-coefficient of variance in a vegetation index through the course of a
-year, corresponds to the environmental stability hypothesis which states
-that lower energy variation throughout a year will lead to increased
-species richness [@williams2008].
+productivity [@berry2007; @radeloff2019]. The relationship between
+energy availability and biodiversity occurs via various hypothesized
+mechanisms, such as the available energy hypothesis [@currie2004;
+@wright1983], the environmental stress hypothesis [@currie2004], and the
+environmental stability hypothesis [@williams2008]. These three
+hypotheses have in turn been linked to patterns of annual surface
+reflectance in remote sensing data [@berry2007; @radeloff2019].
 
-Within forested ecosystems, there are therefore many possible remotely
-sensed derived ecosystem structure and function metrics that can be
-assessed. Structural variables range in complexity from simple (canopy
-cover; canopy height), to more complex (elevation coefficient of
-variation; elevation standard deviation) to modelled (aboveground
-biomass; basal area), all of which can be assessed using lidar data
-[@coops2021]. Data on ecosystem functionality ranges from phenology
-metrics (date of leaf on/leaf off; length of growing season) to
-functional traits (chlorophyll content; leaf area index) to productivity
-estimates (GPP; total available energy) [@pettorelli2018]. Advances in
-satellite remote sensing processing have allowed 3D forest structure
-data to be imputed across wide spatial scales [@matasci2018; @coops2021]
-using data fusion approaches involving collected lidar data and
-optical/radar data. Other advances in image compositing have allowed
-yearly summaries of vegetation productivity to be calculated at regional
-to global scales, summarizing the yearly total, minimum, and variation
-in energy availability at similar spatial resolutions [@radeloff2019].
+@berry2007 first explored this idea by proposing the linkage of
+intra-annual summaries of MODIS derived GPP to dispersive bird species.
+This idea was further refined into the Dynamic Habitat Indices [DHIs;
+@coops2008], which have now been shown to be well suited to assess the
+three aforementioned hypotheses at global scales [@radeloff2019]. The
+cumulative DHI calculates the total amount of energy available in a
+given pixel over the course of a year. Cumulative DHI is strongly linked
+to the available energy hypothesis, which suggests that with greater
+available energy species richness will increase [@wright1983]. The
+minimum DHI, which calculates the lowest productivity over the course of
+a year can be matched to the environmental stress hypothesis, which
+proposes that higher levels of minimum available energy will lead to
+higher species richness [@currie2004]. Finally, the variation DHI, which
+calculates the coefficient of variance in a vegetation index through the
+course of a year, corresponds to the environmental stability hypothesis
+which states that lower energy variation throughout a year will lead to
+increased species richness [@williams2008].
 
-Linkages between forest structure and function have been examined within
-a remote sensine context for over 20 years [@huete2002, @knyazikhin1998;
-@myneni1994]. While there is significant theoretical and empirical
-evidence for their relationship at single time points (within a single
-image) [@myneni1994], various relationship directions and shapes have
-been found between forest structure and function metrics [@ali2019].
-These relationships, their shapes, and their strengths have been
-attributed to multiple possible hypotheses and have been shown to vary
-based on environmental conditions [@ali2019]. The relationship in
-particular between forest structural diversity metrics - which are now
-more accurately and comprehensively derived from lidar data - and
-temporal variation in functional metrics, specifically the metrics of
-ecosystem productivity via the DHI framework, have yet to be fully
-examined.
+## Biodiversity Monitoring with EBVs
 
-Understanding the complementarity between potential EBVs is an integral
-component of their creation [@skidmore2021]. Structural equation
-modelling (SEM) and path analyses have been commonly used in ecology to
-assess the causal effects behind various hypotheses [@fan2016;
-@grace2010], including the relationship between forest structure,
-functioning, and biodiversity [@ali2019]. For example, SEM has commonly
-been used to assess the role of structural diversity of carbon storage
-as compared to species diversity [@ali2016; @zhang2015].
+Biodiversity monitoring programs often require a range of information in
+order to accurately assess changes in ecosystem integrity
+[@lindenmayer2010]. Choosing datasets that are most closely related to
+the phenomenon of interest in a given application allows for direct
+connections to monitoring results and management actions [@pressey2021].
+With the advent of large-extent monitoring methods like satellite remote
+sensing, and a proliferation of potential EBVs datasets, it becomes
+important to assess the interrelationships between these datasets, and
+assess their complementary of the information to reduce duplication of
+efforts [@pereira2013; @skidmore2021]. When strong relationships are
+present between EBV classes, it becomes possible to assess the
+ecological relationships between potential EBVs. On the other hand, when
+datasets do not appear related, they may be well suited to be used in
+monitoring programs together, as complementary EBVs.
 
-The overall goal of this paper is to characterize patterns of forest
-ecosystem structure and function and their interrelationships across a
-wide range of ecosystems encompassing significant environmental
-gradients. To do so, we synthesize data from moderate-scale
-remote-sensing derived metrics of structure, represented as both simple
-ALS metrics of canopy height and cover, as well as modeled forest
-structure variables including volume, and above ground biomass, with a
-remote sensing derived index on ecosystem function. Our first question
-is to examine how ecosystem structure and function vary across a large
-environmental gradient and then compare the simple and modeled
-representations of forest structure to different levels of ecosystem
-function. This question is important as it provides insights to the EBV
-community around redundancy in remote sensing metrics when describing
-the structure and function of ecosystems and proposes a method to
-examine potential overlap when generating remote sensing EBV's.
+Linkages between forest ecosystem structure and function have been
+examined within a remote sensing context for over 20 years [@huete2002,
+@knyazikhin1998; @myneni1994]. While there is significant theoretical
+and empirical evidence for their relationship at single time points
+(within a single image) [@myneni1994], various relationship directions
+and shapes have been found between forest structure and function metrics
+[@ali2019]. Hypothesized mechanisms such as niche complementary have
+shown that aboveground biomass increases with stand structure
+[@zhang2012], while asymmetric competition for light can reduce forest
+productivity with increased structural complexity [@bourdier2016]. The
+relationship in particular between forest structural diversity metrics -
+which are now more accurately and comprehensively derived from lidar
+data - and temporal variation in functional metrics, specifically the
+metrics of ecosystem productivity via the DHI framework, have yet to be
+fully examined.
+
+The overall goal of this paper is to assess patterns of forest ecosystem
+structure and function and their complementarity across a wide range of
+ecosystems encompassing significant environmental gradients. To do so,
+we synthesize data from moderate-scale remote-sensing derived metrics of
+forest structure, represented as both simple ALS-extracted metrics of
+canopy height, cover and vertical complexity, as well as modelled forest
+structure attributes including volume, and aboveground biomass, with a
+well-established remote sensing derived index on ecosystem function. Our
+first question is to examine how ecosystem structure and function
+complement one another across a large environmental gradient and then
+compare the simple and modelled representations of forest structure to
+different levels of ecosystem function. This question is important as it
+provides insights to the EBV community around complementarity of remote
+sensing metrics when describing the structure and function of ecosystems
+and proposes a method to examine potential overlap when generating
+remote sensing EBVs.
 
 Our second question examines the independent and shared relationships of
-ecosystem structure height and cover, with modeled forest structure, on
-function. This provides insight into the choice of remote sensing
-attributes to use when developing EBV's. Remote sensing datasets can
-comprise relatively unprocessed, observations, in this case ALS measures
-of height and cover which are retrieval from the raw 3D point cloud, vs
-modeled attributes, such as biomass and volume, which involve the use
-the statistical relationships to transform the observations into more
-refined data products. Assessing which of these two (or combination of
-the two) approaches has more or less redundancy with estimates of
-function provides insights into the choice of data used to build EBV's.
-Lastly we examine how the primary and modeled structure attributes vary
-in their variance partitioning within key biomes across this large
-environmental range, examining to what extend ecosystem type impacts
-these interrelationships and thus the applicability of these results
-globally.
+ecosystem structure height and cover, with modelled forest structure, on
+ecosystem function. This provides insight into the choice of remote
+sensing attributes to use when developing EBVs within a single EBV
+class. Remote sensing datasets can comprise relatively unprocessed
+observations, in this case ALS measures of height and cover which are
+derived from the raw 3D point cloud vs modelled attributes, such as
+biomass and volume, which involve the use the statistical relationships
+with field data to transform the observations into more refined data
+products. Assessing which of these two (or combination of the two)
+approaches has stronger or weaker correlations with estimates of
+function provides insights into the choice of data used to build EBVs.
+Lastly, we examine how the primary and modelled structure attributes
+partition the variance of the DHIs within key biomes and forest types
+across a large environmental range, examining to what extent ecosystem
+and forest types impacts these relationships and thus providing insight
+into the applicability of these results globally.
 
@@
 
-## Study Area
+## Study Area {#sec-study-area}
 
 British Columbia is the westernmost province of Canada, and is home to a
-variety of terrestrial ecosystems [@pojar1987]. Approximately 64% of the
-province is forested [@bcministryofforests2003]. There is a large amount
-of ecosystem variation in the province, with large climate and
-topographic gradients. The Biogeoclimatic Ecosystem Classification (BEC)
-system identifies 16 zones based on the dominant tree species and the
-ecosystems general climate. These zones can be further split into
-subzones, variants, and phases based on microclimate, precipitation, and
-topography [@pojar1987]. To examine trends across the large
-environmental gradients, we group the BEC zones into five broad biomes,
-specifically, the south interior, northern, montane, alpine, and coastal
-groups similar to @hamann2006.
+variety of terrestrial ecosystems. Approximately 64% of the province is
+forested, with large environmental and topographic gradients
+[@pojar1987; @bcministryofforests2003]. The Biogeoclimatic Ecosystem
+Classification (BEC) system identifies 16 zones based on the dominant
+tree species and the ecosystems general climate (@fig-study-map).
 
-{insert figure showing ecosystems of interest} need to remake figure
-showing ecosystems, highlighting that bunchgrass and alpine zones are
-not included
+```{r}
+#| label: fig-study-map
+#| fig-cap: Location of biogeoclimatic ecosystem classification (BEC) zones in British Columbia.
 
-{table showing elevations and climate ranges for each included bec zone}
+knitr::include_graphics(here::here("outputs", "bec_map.png"))
+```
+
+These zones can be further split into subzones, variants, and phases
+based on microclimate, precipitation, and topography [@pojar1987]. To
+examine trends across the large environmental gradients, we group the
+BEC zones into five broad biomes, specifically, the southern interior,
+northern interior, montane, alpine, and coastal groups similar to
+@hamann2006. We also report each BEC zone's average climate data,
+according to @wang2020 (@tbl-bec-group).
+
+```{r}
+#| label: tbl-bec-group
+#| tbl-cap: BEC Zones, their aggregate groups, and their average climate values for precipitation, maximum temperature, and minimum temperature. Climate data from Wang et al. 2016.
+
+# bec_tab %>%
+#   gt() %>%
+#   cols_label(
+#     group = "BEC Group",
+#     zone = "BEC Zone",
+#     zone_nm = "Full Name",
+#     ppt = "Precipitation (mm/yr)",
+#     tmax = "Max Temperature (°C)",
+#     tmin = "Min Temperature (°C)"
+#   ) %>%
+#   as_word()
+
+knitr::kable(
+  bec_tab,
+  col.names = c(
+    "BEC Group",
+    "BEC Zone",
+    "Full Name",
+    "Precipitation (mm/yr)",
+    "Max Temperature (°C)",
+    "Min Temperature (°C)"
+  )
+)
+```
 
@@
 
-We utilize a suite of forest structural diversity variables (canopy
-height, canopy cover, Lorey's height, overstory cover, basal area,
-aboveground biomass, gross stem volume, mean elevation, elevation
-standard deviation, and structural complexity \[coefficient of variation
-in elevation returns\]). This dataset is created at a 30 m spatial
-resolution according to @matasci2018. In brief, the method utilizes a
-set of lidar collections and field plots across Canada, and imputes the
-remaining pixels using a random forest k-Nearest Neighbour approach.
-Detailed information on the creation of this dataset can be found in
-@matasci2018.
+We used a suite of forest structural attributes (canopy height, canopy
+cover, overstory cover, basal area, aboveground biomass, gross stem
+volume, mean elevation, elevation standard deviation, and structural
+complexity \[coefficient of variation in elevation returns\]). This
+dataset was created for the year 2015 at a 30 m spatial resolution
+according to @matasci2018. In brief, the method used a set of lidar
+collections and field plots across Canada, and imputed the remaining
+pixels using a random forest k-Nearest Neighbour approach on
+Landsat-derived surface reflectance and auxiliary data such as
+topography. Detailed information on the creation of this dataset can be
+found in @matasci2018.
 
@@
 
-The Dynamic Habitat Indices are a set of satellite remote sensing
-derived productivity variables that summarize the cumulative amount of
-available energy, the minimum available energy, and the variation in
-available energy throughout a given year [@radeloff2019]. The DHIs have
-previously been produced at a global extent using MODIS imagery
-[@radeloff2019], however, recent studies have examined how these indices
-can be constructed at a finer spatial resolution using multi-annual
-Landsat imagery.
+We use an established set of indices of annual productivity shown to be
+related to global biodvierstiy trends: the Dynamic Habitat Indices
+[@radeloff2019]. The DHIs are a set of satellite remote sensing derived
+productivity variables that summarize the cumulative amount of available
+energy, the minimum available energy, and the variation in available
+energy throughout a given year [@berry2007; @radeloff2019]. The DHIs
+have previously been produced at a global extent using MODIS imagery ,
+and have been used to assess alpha [@radeloff2019] and beta[@andrew2012]
+diversity, species abundances [@razenkova2023], and construct novel
+ecoregionalizations [@coops2009]; @andrew2013\] . Recent studies have
+began to examine how these indices can be constructed at a finer spatial
+resolution by using multi-annual Landsat imagery to generate a single
+synthetic year of monthly observations [@razenkova2022].
 
-The DHIs were calculated according to {RAZENKOVA LANDSAT DHI PAPER IN
-PRESS} for all of terrestrial British Columbia. In brief, they utilized
-Google Earth Engine [@gorelick2017] to obtain all valid Landsat pixels
-for a given study area by filtering out low quality pixels, then
-calculated the NDVI for each pixel. They use the median NDVI value for
-each month across the ten year time span (2011-2020) to calculate the
-sum, minimum, and coefficient of variation in NDVI values. More detailed
-information can be found in {Razenkova In Press}.
+The DHIs were calculated according to [@razenkova2022; Razenkova et al.,
+In Press] for all of terrestrial British Columbia. In brief, Google
+Earth Engine [@gorelick2017] was used to obtain all valid Landsat pixels
+for a given study area, filtering out pixels containing shadows, slouds,
+and cloud shadows within each image [@zhu2012], then calculated the NDVI
+for each pixel in each image. They then calculated the median NDVI value
+for each month across the ten year time span (2011-2020) to generate a
+synthetic year of monthly data. The sum, minimum, and coefficient of
+variation across this synthetic year of NDVI values is then calculated.
+More detailed information can be found in Razenkova et al. (In Press).
 
@@
 
-We conducted model based sampling across the fifteen forest dominated
-ecosystems found within British Columbia \[BC figure reference, I have
-not made this figure yet\]. Samples were randomly selected within BEC
-zones of interest alongside multiple criteria. Each sampled pixel had to
+We conducted model-based sampling across the sixteen ecosystems found
+within British Columbia (@fig-study-map). Samples were randomly selected
+within each BEC zone, in undisturbed pixels. Each sampled pixel had to
 have a forested land cover class (coniferous, deciduous, mixed-wood, or
-wetland-treed), and be surrounded by the same land cover class.
-Additionally, each pixel had to have a coefficient of variation less
-than 0.5 in the Lorey's height and canopy cover forest structural
-metrics. A maximum of 3000 samples were sampled in each BEC zone. All
-variables were natural-log transformed and standardized. Variables
-containing zeros were natural-log plus one transformed.
+wetland-treed), and be surrounded by the same land cover class. The land
+cover mask was generated following [@hermosilla2022] using a
+best-available-pixel composite, and an inverse-distance weighted random
+forest approach across Canada. Additionally, each pixel had to have a
+coefficient of variation less than 0.5 in surrounding pixels in the two
+simplest forest structural attributes, canopy height and canopy cover. A
+maximum of 3000 samples were sampled in each BEC zone with a 1 km
+minimum sampling distance to reduce the effects of spatial
+autocorrelation. All variables were natural-log transformed and
+standardized. Variables containing zeros were natural-log plus one
+transformed. Sampling was conducted in R [@R-base] version 4.2.2 using
+the **sgsR** package [@R-sgsR]. Focal analyses for the land cover
+classes and coefficient of variation were calculated in Python version
+3.9.
+
+# Sampling 2
+
+We implement two sampling schemes -- one for ecosystems, and one for
+forest types -- across the terrestrial ecosystems of British Columbia
+(@fig-study-map). Both have samples randomly selected. For both sampling
+schemes, each individual sample was taken from a forested pixel which
+was surrounded by the same land cover class. The land cover class map
+was generated for the year 2015 following [@hermosilla2022] using a
+best-available-pixel composite, and an inverse-distance weighted random
+forest approach across Canada. Additionally, each pixel had to have a
+coefficient of variation less than 0.5 in surrounding pixels in the two
+simplest forest structural attributes, canopy height and canopy cover. A
+minimum sampling distance of 1 km was implemented to reduce the effects
+of spatial autocorrelation, and samples that had been disturbed in the
+last 30 years were discarded by using a disturbance mask generated for
+the forested ecosystems of Canada by @hermosilla2016.
+
+The ecosystem sampling scheme was implemented by randomly sampling
+pixels matching the above criteria within each BEC zone of British
+Columbia, up to a maximum of 3000 pixels, regardless of land cover
+class. The forest type sampling scheme was similar in that each BEC zone
+was sampled, however, it implemented stratified sampling on each forest
+type, up to a maximum of 500 pixels per BEC zone and forest type.
+
+Each sampling scheme was natural-log transformed and standardized to
+Z-scores. Variables containing zeros were natural-log plus one
+transformed. Sampling was conducted in R [@R-base] version 4.2.2 using
+the **sgsR** package [@R-sgsR]. Focal analyses for the land cover
+classes and coefficient of variations of canopy height and cover were
+calculated in Python version 3.9.
 
@@
 
-### Partial Redundancy Analysis and Variation Partitioning
+### Redundancy Analysis and Variation Partitioning
 
-Partial redundancy analysis and variation partitioning was used to
-relate the primary and modeled forest structure variables to ecosystem
-structure, to assess the inter-relationships on forest ecosystem
-function across the broad environment range. Redundancy analysis has
-widely been used in community ecology where environmental variables of
-interest are compared to species composition, and when the effect of
-space for example is assessed as a third covariate, in an approach known
-as partial redundancy analysis. While RDA is common in the ecological
-literature, this represents one of the first times this technique has
-been applied to assess redundancy in proposed essential biodiversity
-variables. Redundancy analysis utilizes multiple regression in that it
-models the effect of an explanatory matrix on a response matrix; with
-the key difference being the use of a response matrix rather than a
-single response variable. The first step is a multiple regression where
-each of the variables in the response and explanatory matrices are
-regressed resulting in a matrix of fitted values. PCA is then used in
-the second step to reduce dimensionality through eigenvectors and to
-allow the strength of relationships between assessed, taking into
-account the number of explanatory variables. Partial Redundancy Analysis
-(PRA) allows the use of additional explanatory variables, or
-co-variates. This is useful, as it allow the assessment of the effect of
-one set of variables on another while taking into account variations due
-to other third set that are not the focus of the study. The output is
-similar to standard RDA except it is also possible to examine how much
-variation is explained in the primary variables of interest, and the
-co-variates alone, and then when combined together. This variance
-partitioning is then shown typically through a Venn diagram, in which
-the percentage of explained variance by each explanatory data set is
-reported. We use ANOVAs to test which canonical axes are significant,
-and eigenvalues to determine the proprotion of variance explained by
-each axis.
+Redundancy analysis (RDA) and variation partitioning were used to relate
+the primary and modelled forest structure attributes to ecosystem
+function across a broad environment range. Redundancy analysis functions
+similarly to a multiple linear regression, except it is capable of
+predicting multiple response variables. It accomplishes this by first
+running a multiple linear regression of each predictor variable on each
+response variable, then running a principle component analysis on the
+residuals from each multiple linear regression. This reduces the
+dimensionality of the output, and allows the relationship strength to be
+assessed by calculating the loadings of both predictor and response
+variables on the RDA axes. Partial redundancy analysis function
+similarly, except also considers co-variates [@legendre2012]. Redundancy
+analysis has widely been used in community ecology where environmental
+variables of interest are compared to species composition
+[@blanchet2014; @kleyer2012] and has similarities to partial least
+squares regression -- namely, their multivariate approach, usage of
+dimensionality reduction, and linearity assumptions -- which is commonly
+used in remote sensing literature [@roelofsen2014; @burnett2021].
 
-## Software
+Following the RDA, we employ ANOVAs to determine which axes are
+significant, and calculate the proportion of variance attributable to
+each axis using the eigenvalues generated from the RDA. We calculate
+axis loadings for both predictor and response variables by calculating
+the correlation between the variables and the RDA axes. Axis loadings
+represent the relationship between a given variable and the RDA axis. We
+only consider and display significant axes. To visualize the RDA for
+both predictor and response variables, we display the results as path
+diagrams, with loadings from each predictor to the RDA axis to the
+response variables. The variance explained by each axis is also
+displayed in the RDA box (@fig-rda-var). All RDA calculations were done
+in R [@R-base] version 4.2.2 using the **vegan** package [@R-vegan].
 
-The processing for this paper was done in R version 4.2.2 [@R-base]. We
-used *sgsR* [@goodbody2023] for sampling, *terra* [@R-terra] for raster
-analyses, *sf* for vector analyses [@R-sf], and *vegan* [@R-vegan] for
-redundancy analysis and variation partitioning.
+Variation partitioning is an extension of partial RDA which can assess
+the overlap between the explanatory power of two datasets by utilizing
+multiple partial RDAs and exchanging which datasets are considered the
+predictor, and which is considered the co-variate [@legendre2012].
+Variation partitioning is traditionally displayed using a Venn diagram,
+in which the percentage of variance explained by each dataset is in a
+circle, and the overlap between circles represents the overlap in
+variance explained. All variation partition analyses were done in R
+[@R-base] version 4.2.2 using the **vegan** package [@R-vegan].
 
-Focal analyses were conducted in R and Python, and the sampling was
-conducted across a masked suitability raster for each BEC zone using the
-*sgsR* R package version 1.3.4 [@goodbody2023].
+RDA and variation partitioning analyses were conducted for all samples
+in the ecosystem sampling scheme, as well as individually across each
+ecosystem and forest type. The results were aggregated to BEC zone
+groups (see @sec-study-area).
 
-All code associated with this analysis is available at
+All code associated with the processing and analysis is available at
 https://github.com/emuise/code-structProdSem.
@@
 
+```{r}
+#| label: fig-rda-var
+#| fig-cap: A) Axis loadings from redundancy analysis of primary and modelled forest structural attributes on the dynamic habitat indices.
+B) Results from variation partitioning of primary and modelled forest structural attributes on the DHIs.
+Both visualized analyses are across all collected samples.
+See supplementary information for results from each BEC zone and forest type.
+
+knitr::include_graphics(here::here("outputs", "all_data_figure.drawio.png"))
+```
+
+To examine the relationship between ecosystem structure and function
+across a large environmental gradient, we present the results of a
+redundancy analysis of forest structural attributes on the dynamic
+habitat indices across the entire sampled dataset in @fig-rda-var A.
+While there are three RDA axes associated with the full dataset, the
+third axis explains 0.05% of the variance in the DHIs, and as a result
+it is not shown. The first axis is strongly represents all the DHIs
+(loadings \> 0.85 for all DHIs), and has the strongest loadings from
+canopy cover, basal area, aboveground biomass, and gross stem volume.
+The other input attributes (canopy height, structural complexity) have
+smaller loadings. The second axis primarily represents the seasonality
+(Minimum and Variation DHI) of the DHIs, with a very small loading on
+the Cumulative DHI, and has the largest loadings in canopy cover and
+complexity, with the rest of the forest structural attributes having
+negligible loadings (@fig-rda-var A).
+
+The results from the variation partitioning analysis (@fig-rda-var B)
+show that the majority of the variance explained by the input datasets
+is shared across both primary and modelled attributes. In total, only
+13.4% of the variation in the DHIs is explained by the structural
+information, with 9.4% of this being from overlap between the datasets.
+The primary and modelled attributes explain 2.7 and 1.3% of the
+variation on their own, respectively (@fig-rda-var B).
+
+```{r}
+#| label: fig-radar
+#| fig-cap: Radar plots of average loadings strength by group.
+A and B show input and response loadings, respectively.
+C) Boxplots of BEC zone, forest types, and all data loadings for predictor and response variables.
+
+knitr::include_graphics(here::here("outputs", "radar_boxplot.png"))
+```
+
+Further, we wanted to examine the strength of the relationship between
+the DHIs and primary structural attributes vs modelled structural
+attributes across the forest types and ecosystems of the province.
+@fig-radar A and B show the axis loadings for structural attributes and
+the DHIs, respectively. Across the BEC zone groups, the loadings are
+generally similar in the structural attributes. Southern ecosystems have
+low or non-significant structural complexity loadings, while the forest
+types have structural complexity as their largest loading. The loadings
+for the DHIs in the first axis are generally larger than the loadings
+for the structural attributes, with the cumulative DHI often being the
+strongest loading overall. In alpine BEC zones, the minimum DHI loading
+is smaller than the variation and cumulative DHIs. In contrast to the
+first RDA axis, the secondary RDA axis is primarily driven by variation
+in the minimum DHI, with medium loadings in the variation DHI, and small
+loadings in the cumulative DHI (@fig-radar A & B). @fig-radar C shows
+boxplots of individual BEC zone loadings for structural attributes and
+the DHIs. Across the BEC zones, canopy cover is generally the strongest
+predictor loading of the first RDA axis, and structural complexity is
+generally the weakest predictor loading. The exception is in the forest
+types, where structural complexity is often the largest driver of the
+first axis. When significant, the second RDA axis always has small
+structural attribute loadings.
+
+```{r}
+#| label: fig-fracts-bar
+#| fig-cap: Stacked bar plot proportion of DHI variation explained by extracted, modelled, and the overlap between primary and modelled structural attributes.
+
+knitr::include_graphics(here::here("outputs", "fracts_bar_plot.png"))
+```
+
+@fig-fracts-bar shows stacked bar plots of the proportion of variation
+explained in the DHIs by primary, modelled, and the overlap between
+primary and modelled attributes. The forest types sampling scheme
+generally results in higher amounts of variance explained, which is
+especially prevalent in broadleaf and mixed wood forests. The overlap
+between modelled and primary attributes is generally higher than each
+dataset on their own, with the exception being mixed wood and broadleaf
+forests and the Coastal Western Hemlock and Engellmann Spruce - -
+Subalpine Fir BEC zone zones. The Interior Mountain-heather Alpine and
+Montane Spruce ecosystems have larger amounts of DHI variance explained
+than the rest of their ecosystem groups. Notably, the variation
+explained by the primary and modelled attributes is commonly presented
+through a single RDA axis, which generally corresponds to overall
+productivity through the year (@fig-radar). Generally, the DHIs are
+decoupled from the structural attributes, as shown by the overall
+variance explained being under 30%, regardless of whether the attribute
+is directly extracted from the point cloud or is a modelled variable.
+
+```{r}
+#| label: fig-fcc
+#| fig-cap: False colour maps of axis loadings for the first RDA axis (top) and second RDA axis (bottom).
+All colour values are normalized to the maximum loading of each variable.
+A and D show axis loadings for canopy height (r), canopy cover (g) and structural complexity (b).
+B and E show axis loadings for basal area(r), total biomass (g) and gross stem volume (b).
+C and F show axis loadings for the cumulative DHI (r), variation DHI (g) and minimum DHI (b).
+
+knitr::include_graphics(here::here("outputs", "rgb_norm_edited.png"))
+```
+
+@fig-fcc shows false colour composites of the primary (A & C) and
+modelled (B & D) loadings across the ecosystems of British Columbia. In
+the first RDA axis there is spatial variation in the primary attributes,
+with interior zones being primarily driven by canopy cover (green), the
+coastal regions having strong structural complexity and canopy height
+loadings, and the boreal in the northwest having the strongest loading
+in canopy height (@fig-fcc A). The modelled attributes generally show
+gray-scale colour, indicating that basal area, total biomass, and gross
+stem volume explain similar variation in the DHIs (@fig-fcc B). Interior
+Cedar -- Hemlock, Boreal White and Black Spruce, Interior Douglas-fir,
+Mountain Hemlock, Montane Spruce and Sub-Boreal Spruce have a secondary
+axis, shown in @fig-radar C & D. All zones with a secondary axis are
+located away from the Pacific coast. Canopy cover has the strongest
+secondary axis loading in the boreal white and black spruce, while the
+secondary axis in the remaining four zones have the highest loadings in
+canopy height (@fig-fcc C). Again, the modeled attributes are in
+grayscale, indicating similar loadings across the three modelled
+attributes (@fig-fcc E).
+
 # Discussion
 
-# Conclusion
+Complementarity plays a crucial role in the context of Essential
+Biodiversity Variables (EBVs) and holds significant importance for
+monitoring and understanding biodiversity [@pereira2013; @skidmore2021].
+At the ecosystem scale, the interconnectedness and interdependance of
+ecosystem structure and function is important to recognize and
+disentangle. In forested environments, forest structure and productivity
+have been shown to be linked in multiple studies spanning three decades
+[@ali2019; @myneni1994], however, a linkage between intra-annual
+production and forest structure has yet to be shown. In this study, we
+use statistical analyses from community ecology - namely redundancy
+analysis and variation partitioning - to assess the complementarity of
+forest structure and yearly productivity summaries. While RDA is common
+in the ecological literature, this analysis represents one of the first
+times this technique has been applied to assess the complimentary of
+proposed satellite-derived EBVs. We find that the datasets do not
+strongly overlap, with forest structure explaining 13.4% of the
+variation in the DHIs in samples taken across the entirety of British
+Columbia in the ecosystem sampling scheme (@fig-rda-var B). This
+indicates that they are suitable to be used in tandem with one another
+when used as ecosystem EBVs.
+
+Across most of British Columbia's ecosystem, we identified a single RDA
+axis associated with the DHIs, encompassing the variation in yearly
+productivity (@fig-rda-var A). Within this first axis, the strongest
+loadings were canopy cover and the modelled attributes. In Alpine
+ecosystems the relationship between the first axis and the Minimum DHI
+was lowest, likely due to the low amount of variation in the Minimum DHI
+in these ecosystems. When a second axis was significant, it consistently
+had strong loadings on the minimum and variation DHIs, which indicates a
+complex internal productivity relationship in certain ecosystems. This
+secondary axis has smaller axis loadings associated with the primary and
+modelled attributes, with the strongest loadings being canopy cover and
+structural complexity across the entire dataset (@fig-rda-var A).
+
+Our second question sought to explore whether modelled attributes (basal
+area, gross stem volume, aboveground biomass) add additional explanatory
+information when predicting the DHIs, as compared to primary forest
+structural attributes. We generally found that canopy cover had the
+largest axis loadings in the first RDA axis (@fig-radar). Modelled
+attributes such as basal area, aboveground biomass, and gross stem
+volume shared similar loading magnitudes across the range of studied
+ecosystems, indicating they do not add additional value when utilized
+with one another (@fig-radar C; @fig-fcc B). The loadings between the
+modelled attributes and canopy cover are often similar, and as such, we
+recommend utilizing the attributes derived directly from the point cloud
+in the case of ALS data, or selecting a single modelled attribute.
+
+During the analysis we found a high amount of total variance explained
+in deciduous forests (mixed-wood and broadleaf) when compared to the
+other two forest types and most BEC zones (@fig-fracts-bar). This could
+potentially be due to the temporal linkages between canopy cover in
+deciduous forests and productivity. In evergreen coniferous forests
+where canopy cover is present year-round, there is likely a temporal
+mismatch between forest structural attributes and the DHIs. Further, the
+strongest loadings in these two forest types was vertical structural
+complexity, rather than canopy cover. This indicates that the DHIs are
+more closely related to forest structure in deciduous forests, however
+even there they are still decoupled from one another.
+
+Finally, we explored the amount of variation explained by primary vs
+modelled attributes, as well as their overlap (@fig-fracts-bar). We
+generally found that the overlap between the primary and modelled
+attributes explained most of the variation, with some exceptions,
+indicating that using either set of forest structural attributes may be
+suitable when monitoring biodiversity. The exceptions are especially
+prominent in the broadleaf and mixed wood forest classes, and the
+Coastal Western Hemlock ecosystem, where there is no variation explained
+by the overlap between the two datasets. The majority of the variance
+explained in the broadleaf and mixed wood forests was in the forest
+structural attributes derived directly from the point cloud. This was
+expected, as the modelled attributes are calculated based on the derived
+attributes as well as site index and ancillary information [@coops2021].
+
+Recent advances in creating synthetic yearly observations have allowed
+the DHIs to be generated at a finer scale (30 m; Razenkova et al., In
+Press), rather than the previously used 1 km DHIs derived from MODIS
+\[@radeloff2019; @razenkova2022\]. This represents a significant
+advancement when assessing the utility of EBVs, as the 30 m scale is
+well suited to examine a range of ecological applications, including
+forest structure and productivity, and more closely approximates the
+scale at which species perceive habitat [@cohen2004; @kennedy2014]. In
+addition to more closely matching the scale of ecological phenomenon,
+this allows these datasets to be matched and analyzed with other
+datasets generated from the Landsat archive.
+
+In conclusion, we used redundancy analysis and variation partitioning to
+assess the complementarity of two potential EBV datasets - forest
+structure and the DHIs. We also separated the forest structure datasets
+into primary and modelled attributes in order to assess the need to
+develop more complex structural attributes, or if data derived directly
+from lidar datasets was suitable. We found that the structural
+attributes are not strongly related to the DHIs, indicating that they
+are suitable to be used together as ecosystem scale EBVs when monitoring
+forest environments. We also found that variation explained by the
+overlap between primary and modelled attributes was often higher than
+the variation explained by either individually.
 

### 2023-06-01 scripts/manuscript/manuscript.tex
secondary sampling scheme, revised manuscript for ncc

--- 2023-02-02 scripts/manuscript/manuscript.tex
@@
 \hypersetup{
-  pdftitle={Disentangling linkages between satellite derived forest structure and productivity across a range of forest types and ecosystems.},
-  pdfauthor={Evan R. Muise; Nicholas C. Coops; Txomin Hermosilla; A. Cole Burton; Margaret E. Andrew; Stephen S. Ban},
-  pdfkeywords={structural equation modelling, remote
-sensing, landsat, forest structure, forest productivity},
+  pdftitle={Disentangling linkages between satellite derived forest structure and productivity essential biodiversity variables.},
+  pdfauthor={Evan R. Muise; Margaret E. Andrew; Nicholas C. Coops; Txomin Hermosilla; A. Cole Burton; Stephen S. Ban},
+  pdfkeywords={remote sensing, landsat, forest structure, forest
+productivity, dynamic habitat indices, essential biodiversity
+variables},
   colorlinks=true,
@@
 \title{Disentangling linkages between satellite derived forest structure
-and productivity across a range of forest types and ecosystems.}
+and productivity essential biodiversity variables.}
 \author[1]{Evan R. Muise%
@@
 \begin{keyword}
-    structural equation modelling \sep remote
-sensing \sep landsat \sep forest structure \sep 
-    forest productivity
+    remote sensing \sep landsat \sep forest structure \sep forest
+productivity \sep dynamic habitat indices \sep 
+    essential biodiversity variables
 \end{keyword}
 \end{frontmatter}
-    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, sharp corners, interior hidden, frame hidden, enhanced, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
+    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, breakable, interior hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}, frame hidden, sharp corners]}{\end{tcolorbox}}\fi
 
@@
 
-Biodiversity is the summation of variation in biological life, across
-genes, species, communities, and ecosystems. Currently, biodiversity is
-in decline, facing extinction rates above the background extinction rate
-\citep{thomas2004, urban2015}, and homogenization of communities at
-various scales \citep{mcgill2015}. In response, the global biodiversity
-community is making efforts to assess and halt the degradation of
-biodiversity on earth. The Group for Earth Observation Biodiversity
-Observation Network has developed the Essential Biodiversity Variables
-\citep[EBVs,][]{pereira2013}, designed as an analog to the Essential
-Climate Variables framework \citep{bojinski2014}. EBVs are designed to
-be global in scope, relevant to biodiversity information, feasible to
-utilize, and complementary to one another \citep{skidmore2021}.
+With biodiversity being in decline, and facing extinction rates above
+the background extinction rate \citep{thomas2004, urban2015}, as well as
+the homogenization of communities at various scales \citep{mcgill2015}
+it is integral to be able to monitor how biodiversity is changing across
+the globe. In response, the global biodiversity community is making
+efforts to assess and halt the degradation of biodiversity. The Group
+for Earth Observation Biodiversity Observation Network has developed the
+Essential Biodiversity Variables \citep[EBVs,][]{pereira2013}, designed
+as an analog to the Essential Climate Variables framework
+\citep{bojinski2014}. EBVs are designed to be global in scope, relevant
+to biodiversity information, feasible to use, and complementary to one
+another \citep{skidmore2021}. While it can be incredibly difficult, time
+consuming, and expensive to collect data on biodiversity across wide
+swaths of land and varying ecosystems, EBVs, which can be correlated to
+sampled biodiversity information, allow for the monitoring and
+asssessment of protected area effectiveness and ecosystem health at
+large spatial scales \citep{hansen2021}. There are six EBV classes, each
+of which correspond to a different facet of biodiversity including
+species populations, species traits, community composition, ecosystem
+structure, ecosystem function, and genetic composition
+\citep{pereira2013}.
 
-There are six EBV classes, each corresponding to a different facet of
-biodiversity, including, genetic composition, species populations,
-species traits, community composition, ecosystem structure, and
-ecosystem function \citep{pereira2013}. \citet{fernández2020} divides
-the six classes into two approaches, with one focusing on species
-biodiversity, and the other focusing on ecosystem diversity. Remote
-sensing has proven to be capable of measure five of the six classes,
-with the exception being genetic composition, which requires in-situ
-observation and samples \citep{skidmore2021}. Notably, while remote
-sensing can provide information on the remaining two species EBV
-classes, this information is typically acquired with an \emph{ad-hoc}
-approach. This collects high resolution, small spatial extent
-information, rather than the global or regional scales required for
-biodiversity assessment. Community composition falls into a similar
-dilemma, requiring species population information which necessitates
-high resolution spatial data.
+Satellite remote sensing has proven to be capable of measuring five of
+the six EBV classes, the exception being genetic composition, which
+requires in-situ observation and sampling \citep{skidmore2021}. Species
+populations - and in turn community composition - can be assessed with
+very-high-resolution imagery to identify tree species at the tree-crown
+scale, however it is difficult and computationally expensive to extend
+these analyses to broader extents \citep{fassnacht2016, graves2016}
+while species traits such as vegetation phenology have been observed at
+the single-tree scale using, for example, PlanetScope imagery and
+drone-based measurements \citep{wu2021}. However, the spatially limited
+and often \emph{ad-hoc} data collection approaches associated with
+monitoring individuals is not conducive to the global or regional scales
+required for biodiversity trend assessment \citep{valdez2023}.
 
-The remaining two classes, ecosystem structure and function, are
-incredibly well suited to be examined at global or regional scales using
-mid-resolution satellite imagery, such as that provided by the Landsat
-series of satellites. Advances in satellite remote sensing processing
-have allowed 3d forest structure data to be imputed across wide spatial
-scales \citep{matasci2018, coops2021} using data fusion approaches
-involving collected lidar data and optical/radar data. Other advances in
-image compositing have allowed yearly summaries of vegetation
-productivity to be calculated at regional to global scales, summarizing
-the yearly energy totals, minimums, and variations \citep{radeloff2019}.
-These datasets correspond quite well with the EBV classes ecosystem
-structure (forest structural diversity metrics), and ecosystem function
-(forest productivity metrics).
+The two landscape-level EBVs (ecosystem structure and function) are well
+suited to be examined at large spatial scales using coarser spatial
+measurements, such as those taken from satellites by the Moderate
+Resolution Imaging Spectroradiometer {[}MODIS; \citet{zhang2003}{]}, the
+Landsat imaging systems \citep{fisher2006}, or Sentinel-2
+\citep{helfenstein2022, darvishzadeh2019} programs. These mid-resolution
+satellites can monitor processes at broader extents but the coarse
+spatial resolution removes the ability to relate these traits to
+individual organisms. As a result, satellite remote sensing data has
+been shown to be arguably the most effective at monitoring ecosystem
+based EBVs focused on structure and function. These EBVs classes can be
+monitored at regional to global extents through the use of optical
+imagery \citep{cohen2004}, as well as active sensors such as lidar
+(light detection and ranging) and radar
+\citep{guo2017, lefsky2002, lang2021, neuenschwander2019, coops2016}.
+
+\hypertarget{ebv---ecosystem-structure}{%
+\subsection{EBV - Ecosystem Structure}\label{ebv---ecosystem-structure}}
 
 Forest structural diversity has been linked to biodiversity at various
-scales \citep{guo2017, bergen2009, gao2014}. Increased structural
-complexity is hypothesized to create additional niches, leading to
-increased species diversity \citep{bergen2009}. The relationship between
-forest structure and biodiversity is commonly assessed using avian
-species diversity metrics \citep{macarthur1961, goetz2007}, however,
-other clades (and habitats), have been used
-\citep{davies2014, nelson2005}. Many metrics derived from lidar remote
-sensing have been used as local indicators of biodiversity, including
-canopy cover, canopy height, vertical profiles, and aboveground biomass,
-while other 2nd order derived metrics such as canopy texture, height
-class distribution, edges, and patch metrics have been used to examine
-habitat and biodiversity at landscape scales \citep{bergen2009}.
+scales \citep{guo2017, bergen2009, gao2014}. Structural attributes range
+in complexity from simple (canopy cover; canopy height), to more complex
+(vertical and horizontal structural complexity) to modelled (aboveground
+biomass; basal area), all of which can be assessed using lidar data
+\citep{coops2021}. A suite of these lidar-derived attributes have been
+used as local indicators of biodiversity, including simple metrics such
+as canopy cover and canopy height as well as derived metrics including
+vertical profiles, aboveground biomass
+\citep{lefsky1999, guo2017, coops2016}. Other second order derived
+metrics such as canopy texture, height class distribution, edges, and
+patch metrics have also been used to examine habitat and biodiversity at
+landscape scales \citep{bergen2009}. Advances in satellite remote
+sensing processing have allowed 3D forest structure data to be imputed
+across wide spatial scales \citep{matasci2018, coops2021} using data
+fusion approaches involving collected lidar data and optical/radar data.
 
-The dynamic habitat indices (DHIs) are indicators of productivity
-calculated by summarizing vegetation indices over the course of one (or
-multiple) years \citep{radeloff2019}. These indices have been related to
-multiple facets of biodiversity at a range of scales, including species
-occurrence and abundance \citep{razenkova2020}, alpha
-\citep{radeloff2019} and beta diversity \citep{andrew2012}. Hypotheses
-behind the biodiversity productivity relationships have been
-established, including the species-energy hypothesis, the environmental
-stress hypothesis, and the environmental stability hypothesis
-\citep{coops2019}. The cumulative DHI calculates the total amount of
+Increased forest structural complexity has been hypothesized to create
+additional niches, leading to increased species diversity
+\citep{bergen2009}, which has been frequently demonstrated using avian
+species diversity metrics \citep{macarthur1961}. For example:
+\citet{herniman2020} used spectral and lidar derived forest structure
+data to model avian habitat suitability; \citet{clawges2008} found that
+lidar derived forest structural attributes are capable of identifying
+habitat types associated with avian species in pine/aspen forests;
+\citet{goetz2007} used canopy structural diversity to predict bird
+species richness, finding that canopy vertical distribution was the
+strongest predictor of species richness. Forest structural metrics have
+also been used to study biodiversity in other clades as well
+\citep{davies2014, nelson2005}.
+
+\hypertarget{ebv---ecosystem-function}{%
+\subsection{EBV - Ecosystem Function}\label{ebv---ecosystem-function}}
+
+With respect to ecosystem function, energy availability in an ecosystem
+has shown to be a predictor of species richness and abundances at
+various scales \citep{chase2002, radeloff2019, coops2019, razenkova},
+and is measurable using satellite remote sensing via the use of various
+vegetation indices \citep{huete2002, radeloff2019}. Vegetation indices,
+which are indicative of photosynthetic activity, are commonly used as
+proxies of gross primary productivity \citep{huang2019}. These
+vegetation indices have also been used to assess patterns in
+biodiversity at single time points \citep{bonn2004}, and more recently,
+through yearly summaries of productivity
+\citep{berry2007, radeloff2019}. The relationship between energy
+availability and biodiversity occurs via various hypothesized
+mechanisms, such as the available energy hypothesis
+\citep{currie2004, wright1983}, the environmental stress hypothesis
+\citep{currie2004}, and the environmental stability hypothesis
+\citep{williams2008}. These three hypotheses have in turn been linked to
+patterns of annual surface reflectance in remote sensing data
+\citep{berry2007, radeloff2019}.
+
+\citet{berry2007} first explored this idea by proposing the linkage of
+intra-annual summaries of MODIS derived GPP to dispersive bird species.
+This idea was further refined into the Dynamic Habitat Indices {[}DHIs;
+\citet{coops2008}{]}, which have now been shown to be well suited to
+assess the three aforementioned hypotheses at global scales
+\citep{radeloff2019}. The cumulative DHI calculates the total amount of
 energy available in a given pixel over the course of a year. Cumulative
@@
 
-Linkages between forest structure and productivity (namely, vegetation
-indices) have been examined for nearly 20 years \citetext{\citealp[
+\hypertarget{biodiversity-monitoring-with-ebvs}{%
+\subsection{Biodiversity Monitoring with
+EBVs}\label{biodiversity-monitoring-with-ebvs}}
+
+Biodiversity monitoring programs often require a range of information in
+order to accurately assess changes in ecosystem integrity
+\citep{lindenmayer2010}. Choosing datasets that are most closely related
+to the phenomenon of interest in a given application allows for direct
+connections to monitoring results and management actions
+\citep{pressey2021}. With the advent of large-extent monitoring methods
+like satellite remote sensing, and a proliferation of potential EBVs
+datasets, it becomes important to assess the interrelationships between
+these datasets, and assess their complementary of the information to
+reduce duplication of efforts \citep{pereira2013, skidmore2021}. When
+strong relationships are present between EBV classes, it becomes
+possible to assess the ecological relationships between potential EBVs.
+On the other hand, when datasets do not appear related, they may be well
+suited to be used in monitoring programs together, as complementary
+EBVs.
+
+Linkages between forest ecosystem structure and function have been
+examined within a remote sensing context for over 20 years
+\citetext{\citealp[
 \citet{knyazikhin1998}]{huete2002}; \citealp{myneni1994}}. While there
@@
 various relationship directions and shapes have been found between
-forest structure and productivity metrics \citep{ali2019}. These
-relationships, their shapes, and their strengths have been attributed to
-multiple possible hypotheses and can vary based on environmental
-conditions \citep{ali2019}. The relationship between forest structural
-diversity metrics and annual productivity summaries has yet to be fully
-examined, including the DHIs.
+forest structure and function metrics \citep{ali2019}. Hypothesized
+mechanisms such as niche complementary have shown that aboveground
+biomass increases with stand structure \citep{zhang2012}, while
+asymmetric competition for light can reduce forest productivity with
+increased structural complexity \citep{bourdier2016}. The relationship
+in particular between forest structural diversity metrics - which are
+now more accurately and comprehensively derived from lidar data - and
+temporal variation in functional metrics, specifically the metrics of
+ecosystem productivity via the DHI framework, have yet to be fully
+examined.
 
-Some vegetation structure metrics are simpler, and more accurate, to
-calculate than others \citep{coops2021}. These basic metrics, such as
-canopy height and canopy cover,can then be used to estimate additional
-structure metrics, such as basal area or total biomass. Canopy height
-and cover are commonly used as an indicator of vertical and horizontal
-variation, respectively. Recently, more attention has been paid to
-internal structural complexity metrics, which can be more difficult and
-time consuming to generate \citep{coops2021, ma2022} .
+The overall goal of this paper is to assess patterns of forest ecosystem
+structure and function and their complementarity across a wide range of
+ecosystems encompassing significant environmental gradients. To do so,
+we synthesize data from moderate-scale remote-sensing derived metrics of
+forest structure, represented as both simple ALS-extracted metrics of
+canopy height, cover and vertical complexity, as well as modelled forest
+structure attributes including volume, and aboveground biomass, with a
+well-established remote sensing derived index on ecosystem function. Our
+first question is to examine how ecosystem structure and function
+complement one another across a large environmental gradient and then
+compare the simple and modelled representations of forest structure to
+different levels of ecosystem function. This question is important as it
+provides insights to the EBV community around complementarity of remote
+sensing metrics when describing the structure and function of ecosystems
+and proposes a method to examine potential overlap when generating
+remote sensing EBVs.
 
-In this study, we seek to untangle the relationship between two EBVs:
-forest structure diversity metrics and yearly summaries of forest
-productivity. To accomplish this, we assess this relationship using path
-analysis to assess the direct and indirect (as mediated by more complex
-forest structural diversity metrics) effects of commonly collected
-forest structural metrics (canopy height and canopy cover) on yearly
-productivity summaries. Doing so will in turn assess their utility as
-complementary EBVs. We ran this analysis separately across four forested
-land covers and the forested ecosystems of British Columbia, Canada.
+Our second question examines the independent and shared relationships of
+ecosystem structure height and cover, with modelled forest structure, on
+ecosystem function. This provides insight into the choice of remote
+sensing attributes to use when developing EBVs within a single EBV
+class. Remote sensing datasets can comprise relatively unprocessed
+observations, in this case ALS measures of height and cover which are
+derived from the raw 3D point cloud vs modelled attributes, such as
+biomass and volume, which involve the use the statistical relationships
+with field data to transform the observations into more refined data
+products. Assessing which of these two (or combination of the two)
+approaches has stronger or weaker correlations with estimates of
+function provides insights into the choice of data used to build EBVs.
+Lastly, we examine how the primary and modelled structure attributes
+partition the variance of the DHIs within key biomes and forest types
+across a large environmental range, examining to what extent ecosystem
+and forest types impacts these relationships and thus providing insight
+into the applicability of these results globally.
 
@@
 
-\hypertarget{study-area}{%
-\subsection{Study Area}\label{study-area}}
+\hypertarget{sec-study-area}{%
+\subsection{Study Area}\label{sec-study-area}}
 
 British Columbia is the westernmost province of Canada, and is home to a
-variety of terrestrial ecosystems \citep{pojar1987}. Approximately 64\%
-of the province is forested \citep{bcministryofforests2003}. There is a
-large amount of ecosystem variation in the province, with large climate
-and topographic gradients. The Biogeoclimatic Ecosystem Classification
-(BEC) system identifies 16 zones based on the dominant tree species and
-the ecosystems general climate. These zones can be further split into
-subzones, variants, and phases based on microclimate, precipitation, and
-topography.
+variety of terrestrial ecosystems. Approximately 64\% of the province is
+forested, with large environmental and topographic gradients
+\citep{pojar1987, bcministryofforests2003}. The Biogeoclimatic Ecosystem
+Classification (BEC) system identifies 16 zones based on the dominant
+tree species and the ecosystems general climate. These zones can be
+further split into subzones, variants, and phases based on microclimate,
+precipitation, and topography \citep{pojar1987}. To examine trends
+across the large environmental gradients, we group the BEC zones into
+five broad biomes, specifically, the south interior, northern, montane,
+alpine, and coastal groups similar to \citet{hamann2006}.
+
+\{insert figure showing ecosystems of interest and grouped bec zones\}
+need to remake figure showing ecosystems
+
+\{table showing elevations and climate ranges for each included bec
+zone\}
 
@@
 
+\hypertarget{forest-structure}{%
+\subsubsection{Forest Structure}\label{forest-structure}}
+
+We used a suite of forest structural attributes (canopy height, canopy
+cover, Lorey's height, overstory cover, basal area, aboveground biomass,
+gross stem volume, mean elevation, elevation standard deviation, and
+structural complexity {[}coefficient of variation in elevation
+returns{]}). This dataset was created at a 30 m spatial resolution
+according to \citet{matasci2018}. In brief, the method used a set of
+lidar collections and field plots across Canada, and imputed the
+remaining pixels using a random forest k-Nearest Neighbour approach on
+Landsat-derived surface reflectance and auxiliary data such as
+topography. Detailed information on the creation of this dataset can be
+found in \citet{matasci2018}.
+
+\hypertarget{dynamic-habitat-indices}{%
+\subsubsection{Dynamic Habitat Indices}\label{dynamic-habitat-indices}}
+
+We use an established set of indices of annual productivity shown to be
+related to global biodvierstiy trends: the Dynamic Habitat Indices
+{[}\citet{radeloff2019}{]}. The DHIs are a set of satellite remote
+sensing derived productivity variables that summarize the cumulative
+amount of available energy, the minimum available energy, and the
+variation in available energy throughout a given year
+{[}\citet{berry2007}; \citet{radeloff2019}{]}. The DHIs have previously
+been produced at a global extent using MODIS imagery , and have been
+used to assess alpha \citep{radeloff2019} and
+beta{[}\citet{andrew2012}{]} diversity, species abundances
+{[}\citet{razenkova}{]}, and construct novel ecoregionalizations
+\citep{coops2009, andrew2013} . Recent studies have began to examine how
+these indices can be constructed at a finer spatial resolution by using
+multi-annual Landsat imagery to generate a single synthetic year of
+monthly observations {[}\citet{razenkova2022}{]}.
+
+The DHIs were calculated according to {[}\citet{razenkova2022};
+RAZENKOVA LANDSAT DHI PAPER IN PRESS{]} for all of terrestrial British
+Columbia. In brief, Google Earth Engine \citep{gorelick2017} was used to
+obtain all valid Landsat pixels for a given study area, filtering out
+pixels containing shadows, slouds, and cloud shadows within each image
+\citep{zhu2012}, then calculated the NDVI for each pixel in each image.
+They then calculated the median NDVI value for each month across the ten
+year time span (2011-2020) to generate a synthetic year of monthly data.
+The sum, minimum, and coefficient of variation across this synthetic
+year of NDVI values is then calculated. More detailed information can be
+found in \{Razenkova In Press\}.
+
 \hypertarget{sampling}{%
 \subsection{Sampling}\label{sampling}}
+
+We conducted model-based sampling across the sixteen forest-dominated
+ecosystems found within British Columbia {[}BC figure reference, I have
+not made this figure yet{]}. Samples were randomly selected within each
+BEC zone, in undisturbed pixels. Each sampled pixel had to have a
+forested land cover class (coniferous, deciduous, mixed-wood, or
+wetland-treed), and be surrounded by the same land cover class. The land
+cover mask was generated by \citet[need to add to
+zotero]{hermosilla2022} using a best-available-pixel composite, and an
+inverse-distance weighted random forest approach across Canada.
+Additionally, each pixel had to have a coefficient of variation less
+than 0.5 in surrounding pixels in the two simplest forest structural
+attributes, canopy height and canopy cover. A maximum of 3000 samples
+were sampled in each BEC zone with a 1 km minimum sampling distance to
+reduce the effects of spatial autocorrelation. All variables were
+natural-log transformed and standardized. Variables containing zeros
+were natural-log plus one transformed. Sampling was conducted in R
+\citep{R-base} version 4.2.2 using the \textbf{sgsR} package
+\citep{R-sgsR}. Focal analyses for the land cover classes and
+coefficient of variation were calculated in Python version 3.9.
 
@@
 
+\hypertarget{redundancy-analysis-and-variation-partitioning}{%
+\subsubsection{Redundancy Analysis and Variation
+Partitioning}\label{redundancy-analysis-and-variation-partitioning}}
+
+Redundancy analysis (RDA) and variation partitioning were used to relate
+the primary and modelled forest structure attributes to ecosystem
+function across a broad environment range. Redundancy analysis functions
+similarly to a multiple linear regression, except it is capable of
+predicting multiple response variables. It accomplishes this by first
+running a multiple linear regression of each predictor variable on each
+response variable, then running a principle component analysis on the
+residuals from each multiple linear regression. This reduces the
+dimensionality of the output, and allows the relationship strength to be
+assessed by calculating the loadings of both predictor and response
+variables on the RDA axes. Partial redundancy analysis function
+similarly, except also considers co-variates \citep{legendre2012}.
+Redundancy analysis has widely been used in community ecology where
+environmental variables of interest are compared to species composition
+\citep{blanchet2014, kleyer2012}. While RDA is common in the ecological
+literature, this analysis represents one of the first times this
+technique has been applied to assess the complimentary of proposed EBVs.
+Following the RDA, we employ ANOVAs to determine which axes are
+significant, and calculate the proportion of variance attributable to
+each axis using the eigenvalues generated from the RDA. We calculate
+axis loadings for both predictor and response variables by calculating
+the correlation between the variables and the RDA axes. Axis loadings
+represent the relationship between a given variable and the RDA axis. We
+only consider and display significant axes. To visualize the RDA for
+both predictor and response variables, we display the results as path
+diagrams, with loadings from each predictor to the RDA axis to the
+response variables. The variance explained by each axis is also
+displayed in the RDA box Figure~\ref{fig-rda-var}. All RDA calculations
+were done in R \citep{R-base} version 4.2.2 using the \textbf{vegan}
+package \citep{R-vegan}.
+
+Variation partitioning is an extension of partial RDA which can assess
+the overlap between the explanatory power of two datasets by utilizing
+multiple partial RDAs and exchanging which datasets are considered the
+predictor, and which is considered the co-variate \citep{legendre2012}.
+Variation partitioning is traditionally displayed using a Venn diagram,
+in which the percentage of variance explained by each dataset is in a
+circle, and the overlap between circles represents the overlap in
+variance explained. All variation partition analyses were done in R
+\citep{R-base} version 4.2.2 using the \textbf{vegan} package
+\citep{R-vegan}.
+
+RDA and variation partitioning analyses were conducted for all samples,
+as well as individually across each BEC zone and forest type. The
+results were aggregated to BEC zone groups see~\ref{sec-study-area} for
+visualization.
+
+All code associated with the processing and analysis is available at
+https://github.com/emuise/code-structProdSem.
+
 \hypertarget{results}{%
 \section{Results}\label{results}}
+
+\begin{figure}
+
+{\centering \includegraphics[width=2.66in,height=\textheight]{../../outputs/all_data_figure.drawio.png}
+
+}
+
+\caption{\label{fig-rda-var}A) Axis loadings from redundancy analysis of
+primary and modelled forest structural attributes on the dynamic habitat
+indices. B) Results from variation partitioning of primary and modelled
+forest structural attributes on the DHIs. Both visualized analyses are
+across all collected samples. See supplementary information for results
+from each BEC zone and forest type.}
+
+\end{figure}
+
+Figure~\ref{fig-rda-var} A shows the results from redundancy analysis of
+forest structural attributes on the dynamic habitat indices across the
+entire sampled dataset. While there are three RDA axes associated with
+the full dataset, the third axis explains 0.05\% of the variance in the
+DHIs, and as such we do not show it. The first axis is strongly
+represents all the DHIs (loadings \textgreater{} 0.85 for all DHIs), and
+has the strongest loadings from canopy cover, basal area, aboveground
+biomass, and gross stem volume. The other input attributes (canopy
+height, structural complexity, and Lorey's height) have smaller
+loadings. The second axis primarily represents the seasonality (Minimum
+and Variation DHI) of the DHIs, with a very small loading on the
+Cumulative DHI, and is primarily driven by canopy cover and complexity
+(Figure~\ref{fig-rda-var} A).
+
+The results from the variation partitioning (Figure~\ref{fig-rda-var} B)
+show that the majority of the variance explained by the input datasets
+is shared across both primary and modelled attributes. 14\% of the
+variation in the DHIs is explained by primary and modelled attributes,
+with 8.2\% of this being from overlap between the datasets. The primary
+and modelled attributes explain 3 and 2.8\% of the variation on their
+own, respectively (Figure~\ref{fig-rda-var} B).
+
+\begin{figure}
+
+{\centering \includegraphics[width=13.06in,height=\textheight]{../../outputs/radar_boxplot.png}
+
+}
+
+\caption{\label{fig-radar}Radar plots of average loadings strength by
+group. A and B show input and response loadings, respectively. C)
+Boxplots of BEC zone, forest types, and all data loadings for predictor
+and response variables.}
+
+\end{figure}
+
+Figure~\ref{fig-radar} A and B show the axis loadings for predictor and
+response variables, respectively. Across the BEC zone groups, the
+loadings are generally similar in the predictor variables.
+Figure~\ref{fig-radar} C shows the individual BEC zone loadings. Canopy
+cover is generally the strongest predictor loading, and structural
+complexity is generally the weakest predictor loading. The exception is
+in deciduous forest types (mixed-wood and broadleaf), where structural
+complexity is the largest axis loading, alongside canopy cover
+(Figure~\ref{fig-radar} C). The second axis has smaller significant
+loadings.
+
+For predictor variables, there are differences between the two axes. The
+first axis generally has large loadings across all DHIs, with the
+exception being alpine ecosystems, where the minimum DHI has lower
+loadings. The secondary RDA axis is primarily driven by variation in the
+minimum DHI, with medium loadings in the variation DHI, and small
+loadings in the cumulative DHI (Figure~\ref{fig-radar} B \& C).
+
+\begin{figure}
+
+{\centering \includegraphics[width=13.06in,height=\textheight]{../../outputs/raincloud.png}
+
+}
+
+\caption{\label{fig-raincloud}Raincloud plot of proportion of DHI
+variation explained by total, the overlap between primary and modelled
+structural attributes, and the overlap between primary and modelled
+attributes.}
+
+\end{figure}
+
+Figure~\ref{fig-raincloud} shows raincloud plots of the percentage of
+variation explained by predictor attributes in total, unique to modelled
+attributes, unique to primary attributes, and the overlap between
+primary and modelled attributes. Running the analysis by forest types
+generally results in higher amounts of variance explained, this is
+especially prevalent in deciduous forest types (mixed-wood and
+broadleaf). The overlap between modelled and primary attributes is
+generally higher than each dataset on their own
+(Figure~\ref{fig-raincloud}). When the overlap explains a low amount
+variance, the information represented by the primary and modelled
+attributes are generally similar (i.e.~in the all data analysis, 14\% of
+the DHI's variance is explained by the strucutral data, with 8\% being
+from the overlap, and 3\% being from the primary and modelled
+attributes, respectively; Figure~\ref{fig-rda-var} A). Notably, the
+variation explained by the primary and modelled attributes is commonly
+presented through a single RDA axis, which generally corresponds to
+overall productivity through the year (Figure~\ref{fig-radar}).
+
+\begin{figure}
+
+{\centering \includegraphics[width=11.81in,height=\textheight]{../../outputs/con_var.png}
+
+}
+
+\caption{\label{fig-con-var}Proportion of variance explained by
+proportion of BEC zone's forest that are coniferous.}
+
+\end{figure}
+
+Figure~\ref{fig-con-var} shows the proportion of variance explained as a
+proportion of a given BEC zone's forest that is coniferous. As the
+proportion of coniferous forest increases, so too does the amount of
+variance explained, while analyses ran on each forest type do not show a
+similar pattern (Figure~\ref{fig-raincloud}).
+
+\begin{figure}
+
+{\centering \includegraphics[width=8in,height=\textheight]{../../outputs/rgbplots_manual_edit.png}
+
+}
+
+\caption{\label{fig-fcc}False colour maps of axis loadings for the first
+RDA axis (top) and second RDA axis (bottom). A and D show axis loadings
+for canopy height (r), canopy cover (g) and structural complexity (b). B
+and E show axis loadings for basal area(r), total biomass (g) and gross
+stem volume (b). C and F show axis loadings for the cumulative DHI (r),
+variation DHI (g) and minimum DHI (b).}
+
+\end{figure}
+
+Figure~\ref{fig-fcc} shows false colour composites of primary (A \& F),
+modelled (B \& E), and DHI (C \& F) loadings across the BEC zones of
+British Columbia. In the first RDA axis there is spatial variation in
+the primary attributes, with the centre of the province's axis being
+primarily driven by canopy cover (green), the coastal western hemlock
+zone (southwest coast) having a strong structural complexity loading,
+and the boreal in the northwest having the strongest loadings in canopy
+height (Figure~\ref{fig-fcc} A). The modelled attributes generally show
+gray-scale colour, indicating that basal area, total biomass, and gross
+stem volume do not explain additional variation in the DHIs
+(Figure~\ref{fig-fcc} B). Figure~\ref{fig-fcc} C shows the loadings of
+the DHI, which are generally light tones, indicating high loadings
+across the province. Yellow shows a smaller minimum DHI loading, with
+higher loadings in the cumulative and variation DHIs.
+
+Only four zones have a secondary RDA axis (Figure~\ref{fig-radar} C).
+These zones have been spatially represented in (Figure~\ref{fig-fcc}
+D-F). Boreal white and black spruce, interior Douglas-fir, mountain
+hemlock, and montane spruce have a secondary axis, with no spatial
+pattern associated with them. Canopy cover has the strongest loading in
+the boreal white and black spruce, while the secondary axis in the
+remaining three zones have the highest loadings in canopy height
+(Figure~\ref{fig-fcc} D). Again, the modeleed attributes are in
+grayscale, indicating similar loadings across the three modelled
+attributes (Figure~\ref{fig-fcc} E). The DHIs in the second axis are
+primarily driven by the minimum DHI, with low loadings on the cumulative
+DHI, and moderate loadings on the variation DHI ( Figure~\ref{fig-fcc}
+F; Figure~\ref{fig-radar} C).
 
@@
 
-\hypertarget{conclusion}{%
-\section{Conclusion}\label{conclusion}}
+It is integral to consider the complementarity of potential EBVs
+\citep{pereira2013, skidmore2021} . Forest structure and productivity
+have been shown to be linked in multiple studies spanning three decades
+\citep{ali2019, myneni1994}, however, a linkage between intra-annual
+production and forest structure has not been shown. In this study, we
+use statistical analyses from community ecology - namely redundancy
+analysis and variation partitioning - to assess the complementarity of
+forest structure and yearly productivity summaries. We find that the
+datasets do not strongly overlap, with forest structure explaining 14\%
+of the variation in the dynamic habitat indices in samples taken across
+the entirety of British Columbia (Figure~\ref{fig-rda-var} B). This
+indicates that they are suitable to be used in tandem with one another
+when used as ecosystem EBVs.
+
+Across most of British Columbia's ecosystem, we identified a single RDA
+axis associated with the DHIs, encompassing the overall productivity
+(Figure~\ref{fig-rda-var} A). Within this first axis, the strongest
+loadings were canopy cover and the modelled attributes. When a second
+axis was significant, it consistently had strong loadings on the minimum
+and variation DHIs indicating a complex relationship within the DHIs in
+certain ecosystems. This secondary axis has smaller axis loadings
+associated with the primary and modelled attributes, with the strongest
+loadings being canopy cover and structural complexity across the entire
+dataset (Figure~\ref{fig-rda-var} A).
+
+We also sought to explore whether modelled attributes (basal area, gross
+stem volume, aboveground biomass, and Lorey's height) add additional
+explanatory information when predicting the DHIs, as compared to primary
+forest structural attributes. We generally found that canopy cover had
+the largest axis loadings in the first RDA axis
+(Figure~\ref{fig-radar}), which corresponds with canopy cover and LAI
+intercepting radiation from the sun influencing the amount of energy
+available in the environment \citep{knyazikhin1998}. Modelled attributes
+such as basal area, aboveground biomass, and gross stem volume shared
+similar loading magnitudes across the range of studied ecosystems,
+indicating they do not add additional value when compared to one another
+(Figure~\ref{fig-radar} C; Figure~\ref{fig-fcc} B). The loadings between
+the modelled attributes and canopy cover are often similar, and as such,
+we recommend utilizing the attributes derived directly from the point
+cloud in the case of ALS data, or selecting a single modelled attribute.
+
+During the analysis we noticed a high amount of total variance explained
+in deciduous forests (mixed-wood and broadleaf) when compared to the
+other two forest types and most BEC zones. This could possibly be due to
+the loss of canopy cover during the winter being linked more closely to
+the DHIs, which are an annual summary of productivity, and not taken at
+a single time point, reducing the potential temporal mismatch between
+the two datasets. Further, the strongest loadings in these two forest
+types was vertical structural complexity, rather than canopy cover. This
+could indicate that in deciduous forests, additional leaf density at
+multiple layers is more important for productivity than canopy cover.
+
+BEC zones dominated by coniferous forests tended to have higher amounts
+of variance explained by the structural attributes
+(Figure~\ref{fig-con-var}). This appears to be contrary to the
+observation that mixed-wood and broadleaf forests have much higher
+amounts of variation explained, and an analysis consisting solely of
+coniferous pixels has relatively low amounts of variance explained
+(Figure~\ref{fig-raincloud}). This could be due to the large range in
+ecosystem variation across the province in coniferous forests. Running a
+similar analysis across each forest type and each BEC zone could
+increase the amount of explained variation, as the structural datasets
+would likely have less internal variation.
+
+Finally, we explored the amount of variation explained by primary vs
+modelled attributes, as well as their overlap
+(Figure~\ref{fig-raincloud}). We generally found that the overlap
+between the primary and modelled attributes explained most of the
+variation, with some exceptions, indicating that using either set of
+forest structural attributes is suitable when monitoring biodiversity.
+
+Recent advances in creating synthetic yearly observations have allowed
+the DHIs to be generated at a Landsat scale (30 m), rather than the
+previously used 250 m DHIs derived from MODIS {[}is the razenkova paper
+out yet???; \citet{radeloff2019}; \citet{razenkova2022}{]}. This allows
+these datasets to be matched and analyzed with other datasets generated
+from the Landsat archive. This represents a significant advancement when
+assessing the utility of EBVs, as the 30 m scale is well suited to
+examine a range of ecological applications, including forest structure
+and productivity \citep{cohen2004}.
+
+In conclusion, we used redundancy analysis and variation partitioning to
+assess the complimentarity of two potential EBV datasets - forest
+structure and the DHIs. We also separated the forest structure datasets
+into primary and modelled attributes in order to assess the need to
+develop more complex structural attributes, or if base data derived
+directly from lidar datasets was suitable. We found that the structural
+attributes are not strongly related to the DHIs, indicating that they
+are suitable to be used together as ecosystem scale EBVs when monitoring
+forest environments. We also found that variation explained by the
+overlap between primary and modelled attributes was often higher than
+the variation explained by either individually.
+
+Further research could assess the importance on intra-annual
+productivity vs the single time point of forest strucutre, which does
+not strongly change throughout the yaer (barring large disturbances).
 

### 2023-06-05 scripts/4_vegan.qmd
full draft to tx & mea

@@
   arrange(group, zone) %>%
-  relocate(group)
+  relocate(group) %>%
+  mutate(across(ppt:tmin, \(x) round(x, digits = 1)))
 

### 2023-06-05 scripts/manuscript/manuscript.qmd
full draft to tx & mea

@@
           postal-code: V8V 1T7
-abstract: In today's changing climate and in light of the biodiversity crises, it is integral to be able to monitor the change in biodiversity at large scales.
-Key to this is the development of variables capable of monitoring biodiversity trends.
-One framework developed to tackle this problem is the Essential Biodviersity Variables (EBV), which is designed to be analogous to the Essential Climate variables, and capable of monitoring biodiversity globally while also being complementary to one another, within EBV classes.
-The EBV framework contains six classes, some of which are extremely well suited to be monitoring from satellite remote sensing.
-Among these with the potential to be satellite derived, the ecosystem structure and function classes stand out.
-In this paper, we seek to assess the complementarity of forest structure, as imputed across the entirety of British Columbia, Canada, with the Dynamic Habitat Indices, a yearly summary of productivity indices.
-Both datasets have been previously linked with biodiversity metrics across a range of scales.
-Using redundancy analysis, we find that forest structural attributes and the DHIs are decoupled from one another, with the forest structure datasets explaining 14% of the variation in the DHIs.
-Further, we explore how the proportion of variance explained in the DHIs from primary structural attributres such as canopy height and canopy cover and modelled structural attributes such as aboveground biomass varies.
-We find there is generally large overlap between the two forest structure datasets.
-Overall, we find that forest structure as an ecosystem structure EBV and the DHIs as an ecosystem function EBV are well suited to be complementary EBVs across large environmental gradients.
+abstract: Under a changing climate and in light of the biodiversity crises, monitoring change in biodiversity at broad scales has become ever more imperative.
+Key to regular biodiversity monitoring over time and space is the definition and selection of variables capable of accurately monitoring biodiversity trends.
+One framework is the Essential Biodiversity Variables (EBV), which are designed to be analogous to the Essential Climate variables, and capable of monitoring biodiversity globally while also being complementary to one another, within EBV classes.
+Five of six EBV classes have been shown to be well suited to be monitored using satellite remote sensing, with particular focus warranted on the ecosystem structure and function classes.
+In this paper, we seek to address the complementarity of variables describing forest structure, as imputed across the entirety of British Columbia, Canada, with variables representing ecosystem functioning, namely the Dynamic Habitat Indices, a yearly summary of productivity indices.
+Both sets of variables have previously been linked with biodiversity metrics across a range of scales.
+Using redundancy analysis, we find that forest structural attributes and the DHIs are essentially decoupled from each other, with the forest structure variables explaining 14% of the variation in the DHIs.
+Further, we explore how the proportion of variance explained in the DHIs varies with either primary structure variables, such as canopy height and cover, vs more complex modelled forest structural attributes, such as aboveground biomass and volume.
+Results suggest there is large overlap between the two types of forest structure variables suggesting little added benefit of derived attributes beyond height and cover, in this context.
+Overall, we find that forest structure as an ecosystem structure EBV and the DHIs as an ecosystem function EBV represent markedly different processes as observed by remote sensing and they are highly complementary across large environmental gradients, suggesting that in an earth-observation based monitoring program both are needed.
 keywords: 
@@
 feasible to use, and complementary to one another [@skidmore2021]. While
-it can be incredibly difficult, time consuming, and expensive to collect
-data on biodiversity across wide swaths of land and varying ecosystems,
-EBVs, which can be correlated to sampled biodiversity information, allow
-for the monitoring and assessment of protected area effectiveness and
+it can be difficult, time consuming, and expensive to collect data on
+biodiversity across wide swaths of land and varying ecosystems, EBVs,
+which can be correlated to sampled biodiversity information, allow for
+the monitoring and assessment of protected area effectiveness and
 ecosystem health at large spatial scales [@hansen2021]. There are six
 EBV classes, each of which correspond to a different facet of
-biodiversity including species populations, species traits, community
+biodiversity, including species populations, species traits, community
 composition, ecosystem structure, ecosystem function, and genetic
@@
 measurements, such as those taken from satellites by the Moderate
-Resolution Imaging Spectroradiometer [MODIS; @zhang2003], the Landsat
+Resolution Imaging Spectroradiometer [MODIS, @zhang2003], the Landsat
 imaging systems [@fisher2006], or Sentinel-2 [@helfenstein2022;
@@
 canopy cover and canopy height as well as derived metrics including
-vertical profiles, aboveground biomass [@lefsky1999; @guo2017;
-@coops2016]. Other second order derived metrics such as canopy texture,
-height class distribution, edges, and patch metrics have also been used
-to examine habitat and biodiversity at landscape scales [@bergen2009].
-Advances in satellite remote sensing processing have allowed 3D forest
-structure data to be imputed across wide spatial scales [@matasci2018;
-@coops2021] using data fusion approaches involving collected lidar data
-and optical/radar data.
+vertical profiles, aboveground biomass, or lidar derived indices
+[@lefsky1999; @guo2017; @coops2016]. Other second order derived metrics
+such as canopy texture, height class distribution, edges, and patch
+metrics have also been used to examine habitat and biodiversity at
+landscape scales [@bergen2009]. Advances in satellite remote sensing
+processing have now allowed 3D forest structure data to be imputed
+across wide spatial scales [@matasci2018; @coops2021] using data fusion
+approaches involving collected lidar data and optical/radar data.
 
@@
 intra-annual summaries of MODIS derived GPP to dispersive bird species.
-This idea was further refined into the Dynamic Habitat Indices [DHIs;
+This idea was further refined into the Dynamic Habitat Indices [DHIs,
 @coops2008], which have now been shown to be well suited to assess the
@@
 Biodiversity monitoring programs often require a range of information in
-order to accurately assess changes in ecosystem integrity
+order to accurately assess changes in ecological integrity
 [@lindenmayer2010]. Choosing datasets that are most closely related to
@@
 important to assess the interrelationships between these datasets, and
-assess their complementary of the information to reduce duplication of
-efforts [@pereira2013; @skidmore2021]. When strong relationships are
-present between EBV classes, it becomes possible to assess the
-ecological relationships between potential EBVs. On the other hand, when
-datasets do not appear related, they may be well suited to be used in
-monitoring programs together, as complementary EBVs.
+assess their complementary of the information to reduce the duplication
+of efforts [@pereira2013; @skidmore2021]. When strong relationships are
+present between EBVs, it is possible to assess the ecological
+relationships between potential EBVs. On the other hand, when datasets
+do not appear related, they may be well suited to be used in monitoring
+programs together, as complementary EBVs.
 
@@
 The overall goal of this paper is to assess patterns of forest ecosystem
-structure and function and their complementarity across a wide range of
-ecosystems encompassing significant environmental gradients. To do so,
-we synthesize data from moderate-scale remote-sensing derived metrics of
-forest structure, represented as both simple ALS-extracted metrics of
-canopy height, cover and vertical complexity, as well as modelled forest
-structure attributes including volume, and aboveground biomass, with a
-well-established remote sensing derived index on ecosystem function. Our
-first question is to examine how ecosystem structure and function
-complement one another across a large environmental gradient and then
-compare the simple and modelled representations of forest structure to
-different levels of ecosystem function. This question is important as it
-provides insights to the EBV community around complementarity of remote
-sensing metrics when describing the structure and function of ecosystems
-and proposes a method to examine potential overlap when generating
-remote sensing EBVs.
+structure and function, as observed using remote sensing technologies,
+and their complementarity across a wide range of ecosystems encompassing
+significant environmental gradients. To do so, we synthesize data from
+moderate-scale remote-sensing derived metrics of forest structure,
+represented as both simple ALS-extracted metrics of canopy height, cover
+and vertical complexity, as well as modelled forest structure attributes
+including volume, and aboveground biomass, with a well-established
+remote sensing derived index on ecosystem function. First, we examine
+how ecosystem structure and function complement one another across a
+large environmental gradient and then compare the simple and modelled
+representations of forest structure to different levels of ecosystem
+function. This question is important as it provides insights to the EBV
+community around complementarity of remote sensing metrics when
+describing the structure and function of ecosystems and proposes a
+method to examine potential overlap when generating remote sensing EBVs.
 
@@
 products. Assessing which of these two (or combination of the two)
-approaches has stronger or weaker correlations with estimates of
-function provides insights into the choice of data used to build EBVs.
-Lastly, we examine how the primary and modelled structure attributes
-partition the variance of the DHIs within key biomes and forest types
-across a large environmental range, examining to what extent ecosystem
-and forest types impacts these relationships and thus providing insight
-into the applicability of these results globally.
+approaches has stronger or weaker relationships with estimates of
+ecosystem function provides insights into the choice of data used to
+build EBVs. Lastly, we examine how the primary and modelled structure
+variables partition the variance of the DHIs within key biomes and
+forest types across a large environmental range, examining to what
+extent ecosystem and forest types impacts these relationships and thus
+providing insight into the applicability of these results globally.
 
@@
 
-We used a suite of forest structural attributes (canopy height, canopy
-cover, overstory cover, basal area, aboveground biomass, gross stem
-volume, mean elevation, elevation standard deviation, and structural
-complexity \[coefficient of variation in elevation returns\]). This
+We used a suite of forest structure variables (canopy height, canopy
+cover, structural complexity \[coefficient of variation of height
+returns\], basal area, aboveground biomass, and gross stem volume). This
 dataset was created for the year 2015 at a 30 m spatial resolution
@@
 We use an established set of indices of annual productivity shown to be
-related to global biodvierstiy trends: the Dynamic Habitat Indices
+related to global biodiversity trends: the Dynamic Habitat Indices
 [@radeloff2019]. The DHIs are a set of satellite remote sensing derived
@@
 energy throughout a given year [@berry2007; @radeloff2019]. The DHIs
-have previously been produced at a global extent using MODIS imagery ,
-and have been used to assess alpha [@radeloff2019] and beta[@andrew2012]
-diversity, species abundances [@razenkova2023], and construct novel
-ecoregionalizations [@coops2009]; @andrew2013\] . Recent studies have
-began to examine how these indices can be constructed at a finer spatial
-resolution by using multi-annual Landsat imagery to generate a single
-synthetic year of monthly observations [@razenkova2022].
+have previously been produced at a global extent using MODIS imagery,
+and have previously been used to assess alpha [@radeloff2019] and beta
+[@andrew2012] diversity, species abundances [@razenkova2023], and
+construct novel ecoregionalizations [@coops2009; @andrew2013] . Recent
+studies have began to examine how these indices can be constructed at a
+finer spatial resolution by using multi-annual Landsat imagery to
+generate a single synthetic year of monthly observations
+[@razenkova2022].
 
-The DHIs were calculated according to [@razenkova2022; Razenkova et al.,
+The DHIs were calculated according to [@razenkova2022, Razenkova et al.,
 In Press] for all of terrestrial British Columbia. In brief, Google
 Earth Engine [@gorelick2017] was used to obtain all valid Landsat pixels
-for a given study area, filtering out pixels containing shadows, slouds,
-and cloud shadows within each image [@zhu2012], then calculated the NDVI
-for each pixel in each image. They then calculated the median NDVI value
-for each month across the ten year time span (2011-2020) to generate a
-synthetic year of monthly data. The sum, minimum, and coefficient of
-variation across this synthetic year of NDVI values is then calculated.
-More detailed information can be found in Razenkova et al. (In Press).
+for a given study area, filtering out pixels containing shadows, clouds,
+and cloud shadows within each image using the fmask algorithm
+[@zhu2012], then calculated the NDVI for each pixel in each image. They
+then calculated the median NDVI value for each month across a ten year
+time span (2011-2020) to generate a synthetic year of monthly data. The
+sum, minimum, and coefficient of variation across this synthetic year of
+NDVI values was then calculated. More detailed information can be found
+in Razenkova et al. (In Press).
 
-## Sampling
+A summary of remote sensing derived datasets can be found in Table 2.
 
-We conducted model-based sampling across the sixteen ecosystems found
-within British Columbia (@fig-study-map). Samples were randomly selected
-within each BEC zone, in undisturbed pixels. Each sampled pixel had to
-have a forested land cover class (coniferous, deciduous, mixed-wood, or
-wetland-treed), and be surrounded by the same land cover class. The land
-cover mask was generated following [@hermosilla2022] using a
-best-available-pixel composite, and an inverse-distance weighted random
-forest approach across Canada. Additionally, each pixel had to have a
-coefficient of variation less than 0.5 in surrounding pixels in the two
-simplest forest structural attributes, canopy height and canopy cover. A
-maximum of 3000 samples were sampled in each BEC zone with a 1 km
-minimum sampling distance to reduce the effects of spatial
-autocorrelation. All variables were natural-log transformed and
-standardized. Variables containing zeros were natural-log plus one
-transformed. Sampling was conducted in R [@R-base] version 4.2.2 using
-the **sgsR** package [@R-sgsR]. Focal analyses for the land cover
-classes and coefficient of variation were calculated in Python version
-3.9.
-
-# Sampling 2
+# Sampling
 
@@
 forest types -- across the terrestrial ecosystems of British Columbia
-(@fig-study-map). Both have samples randomly selected. For both sampling
-schemes, each individual sample was taken from a forested pixel which
-was surrounded by the same land cover class. The land cover class map
-was generated for the year 2015 following [@hermosilla2022] using a
-best-available-pixel composite, and an inverse-distance weighted random
-forest approach across Canada. Additionally, each pixel had to have a
-coefficient of variation less than 0.5 in surrounding pixels in the two
-simplest forest structural attributes, canopy height and canopy cover. A
-minimum sampling distance of 1 km was implemented to reduce the effects
-of spatial autocorrelation, and samples that had been disturbed in the
-last 30 years were discarded by using a disturbance mask generated for
-the forested ecosystems of Canada by @hermosilla2016.
+(@fig-study-map). For both sampling schemes, each individual sample was
+taken from a forested pixel which was surrounded by the same land cover
+class. The land cover class map was generated for the year 2015
+following [@hermosilla2022] using a best-available-pixel composite, and
+an inverse-distance weighted random forest classification approach
+across Canada. Additionally, each pixel had to have a coefficient of
+variation less than 0.5 in surrounding pixels in the two simplest forest
+structure variables, canopy height and canopy cover. A minimum sampling
+distance of 1 km was implemented to reduce the effects of spatial
+autocorrelation, and samples that had been disturbed in the last 30
+years were discarded by using a disturbance mask generated for the
+forested ecosystems of Canada by @hermosilla2016.
 
@@
 pixels matching the above criteria within each BEC zone of British
-Columbia, up to a maximum of 3000 pixels, regardless of land cover
-class. The forest type sampling scheme was similar in that each BEC zone
-was sampled, however, it implemented stratified sampling on each forest
-type, up to a maximum of 500 pixels per BEC zone and forest type.
+Columbia, up to a maximum of 3000 pixels, regardless of forest class.
+The forest type sampling scheme was similar in that each BEC zone was
+sampled, however, it implemented stratified sampling on each forest
+class, up to a maximum of 500 pixels per BEC zone and forest class.
 
@@
 Redundancy analysis (RDA) and variation partitioning were used to relate
-the primary and modelled forest structure attributes to ecosystem
+the primary and modelled forest structure variables to ecosystem
 function across a broad environment range. Redundancy analysis functions
@@
 assessed by calculating the loadings of both predictor and response
-variables on the RDA axes. Partial redundancy analysis function
+variables on the RDA axes. Partial redundancy analysis functions
 similarly, except also considers co-variates [@legendre2012]. Redundancy
@@
 each axis using the eigenvalues generated from the RDA. We calculate
-axis loadings for both predictor and response variables by calculating
-the correlation between the variables and the RDA axes. Axis loadings
-represent the relationship between a given variable and the RDA axis. We
-only consider and display significant axes. To visualize the RDA for
-both predictor and response variables, we display the results as path
-diagrams, with loadings from each predictor to the RDA axis to the
-response variables. The variance explained by each axis is also
-displayed in the RDA box (@fig-rda-var). All RDA calculations were done
-in R [@R-base] version 4.2.2 using the **vegan** package [@R-vegan].
+axis loadings for both forest structure and DHI variables by calculating
+the correlation between each variable and the RDA axes. Forest strcuture
+variable axis loadings represent the strength of relationship between a
+given variable and the RDA axis, while DHI loadings indicate what is
+being represented by the RDA axes. We only consider and display
+significant axes. To visualize the RDA for both predictor and response
+variables, we display the results as path diagrams, with loadings from
+each predictor to the RDA axis to the response variables. The variance
+explained by each axis is also displayed in the RDA box (@fig-rda-var).
 
@@
 circle, and the overlap between circles represents the overlap in
-variance explained. All variation partition analyses were done in R
-[@R-base] version 4.2.2 using the **vegan** package [@R-vegan].
+variance explained.
 
@@
 ecosystem and forest type. The results were aggregated to BEC zone
-groups (see @sec-study-area).
+groups (see @sec-study-area). All RDA and variation partitioning
+calculations were done in R [@R-base] version 4.2.2 using the **vegan**
+package [@R-vegan].
 
@@
 #| label: fig-rda-var
-#| fig-cap: A) Axis loadings from redundancy analysis of primary and modelled forest structural attributes on the dynamic habitat indices.
-B) Results from variation partitioning of primary and modelled forest structural attributes on the DHIs.
+#| fig-cap: A) Axis loadings from redundancy analysis of primary and modelled forest structure variables on the dynamic habitat indices.
+B) Results from variation partitioning of primary and modelled forest structure variables on the DHIs.
 Both visualized analyses are across all collected samples.
@@
 across a large environmental gradient, we present the results of a
-redundancy analysis of forest structural attributes on the dynamic
-habitat indices across the entire sampled dataset in @fig-rda-var A.
-While there are three RDA axes associated with the full dataset, the
-third axis explains 0.05% of the variance in the DHIs, and as a result
-it is not shown. The first axis is strongly represents all the DHIs
-(loadings \> 0.85 for all DHIs), and has the strongest loadings from
-canopy cover, basal area, aboveground biomass, and gross stem volume.
-The other input attributes (canopy height, structural complexity) have
-smaller loadings. The second axis primarily represents the seasonality
-(Minimum and Variation DHI) of the DHIs, with a very small loading on
-the Cumulative DHI, and has the largest loadings in canopy cover and
-complexity, with the rest of the forest structural attributes having
-negligible loadings (@fig-rda-var A).
+redundancy analysis of forest structure variables on the dynamic habitat
+indices across the entire sampled dataset in @fig-rda-var A. While there
+are three RDA axes associated with the full dataset, the third axis
+explains 0.05% of the variance in the DHIs, and as a result is not
+shown. The first axis strongly represents all the DHIs (loadings \> 0.85
+for all DHIs), with the highest loadings from canopy cover, basal area,
+aboveground biomass, and gross stem volume. The other input variables
+(canopy height, structural complexity) have smaller loadings. The second
+axis primarily represents the seasonality (Minimum and Variation) of the
+DHIs, with no correlation to Cumulative DHI, and has the highest
+loadings from canopy cover and complexity, with the remainder of the
+forest structure variables having negligible influences(@fig-rda-var A).
 
@@
 show that the majority of the variance explained by the input datasets
-is shared across both primary and modelled attributes. In total, only
-13.4% of the variation in the DHIs is explained by the structural
-information, with 9.4% of this being from overlap between the datasets.
-The primary and modelled attributes explain 2.7 and 1.3% of the
-variation on their own, respectively (@fig-rda-var B).
+is shared across both primary and modelled forest structure variables.
+Of the 13.4% of the variation in the DHIs being explained by the
+structural information, 9.4% is common across both sets of structure
+variables. The primary and modelled variables only explain 2.7% and 1.3%
+of the variation uniquely (@fig-rda-var B).
 
@@
 Further, we wanted to examine the strength of the relationship between
-the DHIs and primary structural attributes vs modelled structural
-attributes across the forest types and ecosystems of the province.
-@fig-radar A and B show the axis loadings for structural attributes and
-the DHIs, respectively. Across the BEC zone groups, the loadings are
-generally similar in the structural attributes. Southern ecosystems have
-low or non-significant structural complexity loadings, while the forest
-types have structural complexity as their largest loading. The loadings
-for the DHIs in the first axis are generally larger than the loadings
-for the structural attributes, with the cumulative DHI often being the
-strongest loading overall. In alpine BEC zones, the minimum DHI loading
-is smaller than the variation and cumulative DHIs. In contrast to the
-first RDA axis, the secondary RDA axis is primarily driven by variation
-in the minimum DHI, with medium loadings in the variation DHI, and small
-loadings in the cumulative DHI (@fig-radar A & B). @fig-radar C shows
-boxplots of individual BEC zone loadings for structural attributes and
-the DHIs. Across the BEC zones, canopy cover is generally the strongest
-predictor loading of the first RDA axis, and structural complexity is
-generally the weakest predictor loading. The exception is in the forest
-types, where structural complexity is often the largest driver of the
-first axis. When significant, the second RDA axis always has small
-structural attribute loadings.
+the DHIs and primary structure variables vs modelled structure variables
+across the forest types and ecosystems of the province. @fig-radar A and
+B show the axis loadings for structure variables and the DHIs,
+respectively. Across the BEC zone groups, the loadings are generally
+similar in the structure variables. Variations in the DHIs in the
+southern ecosystems, which are typically warmer and drier, have low or
+non-significant variance explained by the structure variables.
+Conversely, across the forest types (coniferous, broadleaf, mixed wood,
+and wetland-treed) structural complexity is often the highest loading.
+The loadings for the DHIs in the first axis are generally larger than
+the loadings for the structure variables, with the cumulative DHI often
+being the strongest loading overall. In alpine BEC zones, the minimum
+DHI loading is smaller than the variation and cumulative DHIs. In
+contrast to the first RDA axis, the secondary RDA axis is primarily
+driven by variation in the minimum DHI, with medium loadings in the
+variation DHI, and small loadings in the cumulative DHI (@fig-radar A &
+B). @fig-radar C shows boxplots of individual BEC zone loadings for
+structure variables and the DHIs. Across the province, canopy cover
+generally has the highest loadings, with structural complexity generally
+being the weakest.
 
@@
 #| label: fig-fracts-bar
-#| fig-cap: Stacked bar plot proportion of DHI variation explained by extracted, modelled, and the overlap between primary and modelled structural attributes.
+#| fig-cap: Stacked bar plot proportion of DHI variation explained by extracted, modelled, and the overlap between primary and modelled structure variables.
 
@@
 explained in the DHIs by primary, modelled, and the overlap between
-primary and modelled attributes. The forest types sampling scheme
-generally results in higher amounts of variance explained, which is
-especially prevalent in broadleaf and mixed wood forests. The overlap
-between modelled and primary attributes is generally higher than each
-dataset on their own, with the exception being mixed wood and broadleaf
-forests and the Coastal Western Hemlock and Engellmann Spruce - -
-Subalpine Fir BEC zone zones. The Interior Mountain-heather Alpine and
-Montane Spruce ecosystems have larger amounts of DHI variance explained
-than the rest of their ecosystem groups. Notably, the variation
-explained by the primary and modelled attributes is commonly presented
-through a single RDA axis, which generally corresponds to overall
-productivity through the year (@fig-radar). Generally, the DHIs are
-decoupled from the structural attributes, as shown by the overall
-variance explained being under 30%, regardless of whether the attribute
-is directly extracted from the point cloud or is a modelled variable.
+primary and modelled variables. Forest types sampled across the
+environmental gradient generally show higher amounts of variance
+explained, which is especially prevalent in broadleaf and mixed wood
+forests. The overlap between modelled and primary structure variables is
+generally than the primary and modelled attributes individually, with
+the exception being mixed wood and broadleaf forests and colder, wetter
+forests. Higher elevation and colder forest stands have larger amounts
+of DHI variance explained by the forest structure variables than other
+environments in the province. Notably, the variation explained by the
+primary and modelled structure variables is commonly presented through a
+single RDA axis, which generally corresponds to overall productivity
+through the year (@fig-radar). Overall, the DHIs are decoupled from the
+structure variables, as shown by the overall variance explained being
+under 30%, regardless of whether the attribute is directly extracted
+from the point cloud or is a modelled variable.
 
@@
 
-knitr::include_graphics(here::here("outputs", "rgb_norm_edited.png"))
+knitr::include_graphics(here::here("outputs", "rgb_norm_edit.png"))
 ```
@@
 modelled (B & D) loadings across the ecosystems of British Columbia. In
-the first RDA axis there is spatial variation in the primary attributes,
-with interior zones being primarily driven by canopy cover (green), the
-coastal regions having strong structural complexity and canopy height
-loadings, and the boreal in the northwest having the strongest loading
-in canopy height (@fig-fcc A). The modelled attributes generally show
-gray-scale colour, indicating that basal area, total biomass, and gross
-stem volume explain similar variation in the DHIs (@fig-fcc B). Interior
-Cedar -- Hemlock, Boreal White and Black Spruce, Interior Douglas-fir,
-Mountain Hemlock, Montane Spruce and Sub-Boreal Spruce have a secondary
-axis, shown in @fig-radar C & D. All zones with a secondary axis are
-located away from the Pacific coast. Canopy cover has the strongest
-secondary axis loading in the boreal white and black spruce, while the
-secondary axis in the remaining four zones have the highest loadings in
-canopy height (@fig-fcc C). Again, the modeled attributes are in
-grayscale, indicating similar loadings across the three modelled
-attributes (@fig-fcc E).
+the first RDA axis there is spatial variation in the primary structure
+variables, with interior zones being primarily driven by canopy cover
+(green), the coastal regions having strong structural complexity and
+canopy height loadings, and the boreal in the northwest having the
+strongest loading in canopy height (@fig-fcc A). The modelled structure
+variables generally show gray-scale colour, indicating that basal area,
+total biomass, and gross stem volume explain similar variation in the
+DHIs (@fig-fcc B). All ecosystems with a secondary axis are located away
+from the Pacific coast in cooler, drier environments. Canopy cover has
+the strongest secondary axis loading in the northern boreal stands,
+while the secondary axis in the remaining four zones have the highest
+loadings in canopy height (@fig-fcc C). Again, the modeled structure
+variables are in grayscale, indicating similar loadings across the three
+modelled structure variables (@fig-fcc E).
 
@@
 monitoring and understanding biodiversity [@pereira2013; @skidmore2021].
-At the ecosystem scale, the interconnectedness and interdependance of
+At the ecosystem scale, the interconnectedness and interdependence of
 ecosystem structure and function is important to recognize and
@@
 production and forest structure has yet to be shown. In this study, we
-use statistical analyses from community ecology - namely redundancy
-analysis and variation partitioning - to assess the complementarity of
-forest structure and yearly productivity summaries. While RDA is common
-in the ecological literature, this analysis represents one of the first
-times this technique has been applied to assess the complimentary of
-proposed satellite-derived EBVs. We find that the datasets do not
-strongly overlap, with forest structure explaining 13.4% of the
-variation in the DHIs in samples taken across the entirety of British
-Columbia in the ecosystem sampling scheme (@fig-rda-var B). This
-indicates that they are suitable to be used in tandem with one another
-when used as ecosystem EBVs.
+use statistical analyses commonly used in community ecology - namely
+redundancy analysis and variation partitioning - to assess the
+complementarity of forest structure and yearly productivity summaries.
+While RDA is common in the ecological literature, this analysis
+represents one of the first times this technique has been applied to
+assess the complimentary of proposed satellite-derived EBVs. We find
+that the example EBVs do not strongly overlap, with forest structure
+explaining 13.4% of the variation in the DHIs in samples taken across a
+broad environmental gradient (@fig-rda-var B). This indicates that they
+are suitable to be used in tandem with one another when used as
+ecosystem EBVs across wide environmental gradients.
 
-Across most of British Columbia's ecosystem, we identified a single RDA
-axis associated with the DHIs, encompassing the variation in yearly
+Across most of British Columbia's ecosystems, we identified a single RDA
+axis associated with the DHIs, encompassing the variation in annual
 productivity (@fig-rda-var A). Within this first axis, the strongest
-loadings were canopy cover and the modelled attributes. In Alpine
-ecosystems the relationship between the first axis and the Minimum DHI
-was lowest, likely due to the low amount of variation in the Minimum DHI
-in these ecosystems. When a second axis was significant, it consistently
-had strong loadings on the minimum and variation DHIs, which indicates a
-complex internal productivity relationship in certain ecosystems. This
-secondary axis has smaller axis loadings associated with the primary and
-modelled attributes, with the strongest loadings being canopy cover and
-structural complexity across the entire dataset (@fig-rda-var A).
+loadings were canopy cover and the modelled structure variables. In
+Alpine ecosystems the relationship between the first axis and the
+Minimum DHI was lowest, likely due to the low amount of variation in the
+Minimum DHI in these ecosystems. When a second axis was significant, it
+consistently had strong loadings on the minimum and variation DHIs,
+which indicates a complex internal productivity relationship in certain
+ecosystems. This secondary axis has smaller axis loadings associated
+with the primary and modelled structure variables, with the strongest
+loadings being canopy cover and structural complexity across the entire
+dataset (@fig-rda-var A).
 
-Our second question sought to explore whether modelled attributes (basal
-area, gross stem volume, aboveground biomass) add additional explanatory
-information when predicting the DHIs, as compared to primary forest
-structural attributes. We generally found that canopy cover had the
-largest axis loadings in the first RDA axis (@fig-radar). Modelled
-attributes such as basal area, aboveground biomass, and gross stem
-volume shared similar loading magnitudes across the range of studied
-ecosystems, indicating they do not add additional value when utilized
-with one another (@fig-radar C; @fig-fcc B). The loadings between the
-modelled attributes and canopy cover are often similar, and as such, we
-recommend utilizing the attributes derived directly from the point cloud
-in the case of ALS data, or selecting a single modelled attribute.
+Our second question sought to explore whether modelled structure
+variables (basal area, gross stem volume, aboveground biomass) add
+additional explanatory information when predicting the DHIs, as compared
+to primary forest structure variables. We generally found that canopy
+cover had the largest axis loadings in the first RDA axis (@fig-radar).
+Modelled structure variables such as basal area, aboveground biomass,
+and gross stem volume shared similar loading magnitudes across the range
+of studied ecosystems, indicating they do not add additional value when
+utilized with one another (@fig-radar C; @fig-fcc B). The loadings
+between the modelled structure variables and canopy cover are often
+similar, and as such, in this context we recommend utilizing the
+structure variables derived directly from the point cloud in the case of
+ALS data, or selecting a single modelled attribute, such as aboveground
+biomass alone [@duncanson2022].
 
-During the analysis we found a high amount of total variance explained
+During the analysis we found a higher amount of total variance explained
 in deciduous forests (mixed-wood and broadleaf) when compared to the
@@
 where canopy cover is present year-round, there is likely a temporal
-mismatch between forest structural attributes and the DHIs. Further, the
+mismatch between forest structure variables and the DHIs. Further, the
 strongest loadings in these two forest types was vertical structural
@@
 more closely related to forest structure in deciduous forests, however
-even there they are still decoupled from one another.
+even there significant decoupling remains.
 
 Finally, we explored the amount of variation explained by primary vs
-modelled attributes, as well as their overlap (@fig-fracts-bar). We
-generally found that the overlap between the primary and modelled
-attributes explained most of the variation, with some exceptions,
-indicating that using either set of forest structural attributes may be
-suitable when monitoring biodiversity. The exceptions are especially
-prominent in the broadleaf and mixed wood forest classes, and the
-Coastal Western Hemlock ecosystem, where there is no variation explained
-by the overlap between the two datasets. The majority of the variance
-explained in the broadleaf and mixed wood forests was in the forest
-structural attributes derived directly from the point cloud. This was
-expected, as the modelled attributes are calculated based on the derived
-attributes as well as site index and ancillary information [@coops2021].
+modelled structure variables, as well as their overlap
+(@fig-fracts-bar). We generally found that the overlap between the
+primary and modelled structure variables explained most of the
+variation, with some exceptions, indicating that using either set of
+forest structure variables may be suitable when monitoring biodiversity.
+The exceptions are especially prominent in the broadleaf and mixed wood
+forest classes, as well as cool, highly productive forests where there
+is no variation explained by the overlap between the two datasets. The
+majority of the variance explained in the broadleaf and mixed wood
+forests was in the forest structure variables derived directly from the
+point cloud. This was expected, as the modelled structure variables are
+calculated based on the derived structure variables as well as site
+index and ancillary information [@coops2021].
 
@@
 Press), rather than the previously used 1 km DHIs derived from MODIS
-\[@radeloff2019; @razenkova2022\]. This represents a significant
+[@radeloff2019; @razenkova2022]. This represents a significant
 advancement when assessing the utility of EBVs, as the 30 m scale is
@@
 structure and the DHIs. We also separated the forest structure datasets
-into primary and modelled attributes in order to assess the need to
-develop more complex structural attributes, or if data derived directly
-from lidar datasets was suitable. We found that the structural
-attributes are not strongly related to the DHIs, indicating that they
-are suitable to be used together as ecosystem scale EBVs when monitoring
+into primary and modelled structure variables in order to assess the
+need to develop more complex structure variables, or if data derived
+directly from lidar datasets was suitable. We found that the structure
+variables are not strongly related to the DHIs, indicating that they are
+suitable to be used together as ecosystem scale EBVs when monitoring
 forest environments. We also found that variation explained by the
-overlap between primary and modelled attributes was often higher than
-the variation explained by either individually.
+overlap between primary and modelled structure variables was often
+higher than the variation explained by either individually.
 

### 2023-06-05 scripts/get_bec_proportion_vlce.R

@@ -1,0 +1,43 @@
+unique_bec <- bec_normal %>%
+  pull(bec) %>%
+  unique()
+
+bec_prop_vlce_loc <- here::here("data", "bec_prop_vlce.csv")
+
+if (!file.exists(bec_prop_vlce_loc)) {
+  zonal_freq <- function(bec) {
+    single_bec <- subst(bec_rast, bec, 1, others = NA)
+    
+    mask(vlce, single_bec) %>%
+      freq() %>%
+      tibble() %>%
+      mutate(bec = bec)
+  }
+  
+  zonal_prop <- map(unique_bec, zonal_freq, .progress = T) %>%
+    list_rbind()
+  
+  write_csv(zonal_prop, bec_prop_vlce_loc)
+}
+
+zonal_prop <- read_csv(bec_prop_vlce_loc)
+
+per_con_for <- zonal_prop %>%
+  left_join(keys$vlce, by = c("value" = "name_clean")) %>%
+  filter(forest == "Forest") %>%
+  group_by(bec) %>%
+  mutate(per_con = count / sum(count)) %>%
+  filter(value == "Coniferous") %>%
+  select(bec, value, per_con)
+
+per_con_total <- zonal_prop %>%
+  left_join(keys$vlce, by = c("value" = "name_clean")) %>%
+  #filter(forest == "Forest") %>%
+  group_by(bec) %>%
+  mutate(per_tot = count / sum(count)) %>%
+  filter(value == "Coniferous") %>%
+  select(bec, value, per_tot)
+
+per_cons <- left_join(per_con_for, per_con_total)
+
+rm(unique_bec, bec_prop_vlce_loc, zonal_prop, per_con_for, per_con_total)

### 2023-06-06 scripts/manuscript/manuscript.qmd
send to all coauthors

@@
 Key to regular biodiversity monitoring over time and space is the definition and selection of variables capable of accurately monitoring biodiversity trends.
-One framework is the Essential Biodiversity Variables (EBV), which are designed to be analogous to the Essential Climate variables, and capable of monitoring biodiversity globally while also being complementary to one another, within EBV classes.
-Five of six EBV classes have been shown to be well suited to be monitored using satellite remote sensing, with particular focus warranted on the ecosystem structure and function classes.
+One framework is the Essential Biodiversity Variables (EBV), which are designed to be analogous to the Essential Climate variables and capable of monitoring biodiversity globally while also being complementary to one another within EBV classes.
+Five of six EBV classes are well suited to being monitored using satellite remote sensing, with particular focus warranted on the ecosystem structure and function classes.
 In this paper, we seek to address the complementarity of variables describing forest structure, as imputed across the entirety of British Columbia, Canada, with variables representing ecosystem functioning, namely the Dynamic Habitat Indices, a yearly summary of productivity indices.
@@
 
-With biodiversity being in decline, and facing extinction rates above
-the background extinction rate [@thomas2004; @urban2015], as well as the
-homogenization of communities at various scales [@mcgill2015] it is
+With biodiversity in decline, and facing extinction rates above the
+background extinction rate [@thomas2004; @urban2015], as well as the
+homogenization of communities at various scales [@mcgill2015], it is
 integral to be able to monitor how biodiversity is changing across the
 globe. In response, the global biodiversity community is making efforts
-to assess and halt the degradation of biodiversity. The Group for Earth
-Observation Biodiversity Observation Network has developed the Essential
-Biodiversity Variables [EBVs, @pereira2013], designed as an analog to
-the Essential Climate Variables framework [@bojinski2014]. EBVs are
-designed to be global in scope, relevant to biodiversity information,
-feasible to use, and complementary to one another [@skidmore2021]. While
-it can be difficult, time consuming, and expensive to collect data on
-biodiversity across wide swaths of land and varying ecosystems, EBVs,
-which can be correlated to sampled biodiversity information, allow for
-the monitoring and assessment of protected area effectiveness and
-ecosystem health at large spatial scales [@hansen2021]. There are six
-EBV classes, each of which correspond to a different facet of
-biodiversity, including species populations, species traits, community
-composition, ecosystem structure, ecosystem function, and genetic
-composition [@pereira2013].
+to assess and halt the degradation of biodiversity. The Group on Earth
+Observations Biodiversity Observation Network has developed the
+Essential Biodiversity Variables [EBVs, @pereira2013], designed as an
+analog to the Essential Climate Variables framework [@bojinski2014].
+EBVs are designed to be global in scope, relevant to biodiversity
+information, feasible to use, and complementary to one another
+[@skidmore2021]. While it can be difficult, time consuming, and
+expensive to collect data on biodiversity across wide swaths of land and
+varying ecosystems, EBVs, which can be correlated to sampled
+biodiversity information, allow for the monitoring and assessment of
+protected area effectiveness and ecosystem health at large spatial
+scales [@hansen2021]. There are six EBV classes, each of which
+correspond to a different facet of biodiversity: species populations,
+species traits, community composition, ecosystem structure, ecosystem
+function, and genetic composition [@pereira2013].
 
@@
 requires in-situ observation and sampling [@skidmore2021]. Species
-populations - and in turn community composition - can be assessed with
+populations, and in turn community composition, can be assessed with
 very-high-resolution imagery to identify tree species at the tree-crown
-scale, however it is difficult and computationally expensive to extend
-these analyses to broader extents [@fassnacht2016; @graves2016] while
-species traits such as vegetation phenology have been observed at the
-single-tree scale using, for example, PlanetScope imagery and
-drone-based measurements [@wu2021]. However, the spatially limited and
-often *ad-hoc* data collection approaches associated with monitoring
-individuals is not conducive to the global or regional scales required
-for biodiversity trend assessment [@valdez2023].
+scale [@fassnacht2016; @graves2016]. Additionally, species traits such
+as vegetation phenology have been observed at the single-tree scale
+using, for example, PlanetScope imagery and drone-based measurements
+[@wu2021]. However, the spatially limited and often *ad-hoc* data
+collection approaches associated with monitoring individuals is not
+conducive to the global or regional scales required for biodiversity
+trend assessment [@valdez2023], in addition to the difficulties
+associated with extending species-level analyses to broader extents
+[@fassnacht2016; @graves2016].
 
@@
 @darvishzadeh2019] programs. These mid-resolution satellites can monitor
-processes at broader extents but the coarse spatial resolution removes
-the ability to relate these traits to individual organisms. As a result,
-satellite remote sensing data has been shown to be arguably the most
-effective at monitoring ecosystem based EBVs focused on structure and
-function. These EBVs classes can be monitored at regional to global
+processes at broader extents, but their coarse spatial resolution
+removes the ability to relate these traits to individual organisms. As a
+result, satellite remote sensing data has been shown to be arguably the
+most effective at monitoring ecosystem based EBVs focused on structure
+and function. These EBVs classes can be monitored at regional to global
 extents through the use of optical imagery [@cohen2004], as well as
@@
 which has been frequently demonstrated using avian species diversity
-metrics [@macarthur1961]. For example: @herniman2020 used spectral and
+metrics [@macarthur1961]. For example, @herniman2020 used spectral and
 lidar derived forest structure data to model avian habitat suitability;
@@
 @berry2007 first explored this idea by proposing the linkage of
-intra-annual summaries of MODIS derived GPP to dispersive bird species.
+intra-annual summaries of MODIS-derived GPP to dispersive bird species.
 This idea was further refined into the Dynamic Habitat Indices [DHIs,
@@
 and empirical evidence for their relationship at single time points
-(within a single image) [@myneni1994], various relationship directions
-and shapes have been found between forest structure and function metrics
-[@ali2019]. Hypothesized mechanisms such as niche complementary have
-shown that aboveground biomass increases with stand structure
-[@zhang2012], while asymmetric competition for light can reduce forest
-productivity with increased structural complexity [@bourdier2016]. The
-relationship in particular between forest structural diversity metrics -
-which are now more accurately and comprehensively derived from lidar
-data - and temporal variation in functional metrics, specifically the
-metrics of ecosystem productivity via the DHI framework, have yet to be
-fully examined.
+(e.g. within a single image) [@myneni1994], various relationship
+directions and shapes have been found between forest structure and
+function metrics [@ali2019]. Hypothesized mechanisms such as niche
+complementarity have shown that aboveground biomass increases with stand
+structure [@zhang2012], while asymmetric competition for light can
+reduce forest productivity with increased structural complexity
+[@bourdier2016]. The relationship in particular between forest
+structural diversity metrics - which are now more accurately and
+comprehensively derived from lidar data - and temporal variation in
+functional metrics, specifically the metrics of ecosystem productivity
+via the DHI framework, have yet to be fully examined.
 
@@
 and vertical complexity, as well as modelled forest structure attributes
-including volume, and aboveground biomass, with a well-established
-remote sensing derived index on ecosystem function. First, we examine
-how ecosystem structure and function complement one another across a
-large environmental gradient and then compare the simple and modelled
+including volume and aboveground biomass, with a well-established remote
+sensing derived index on ecosystem function. First, we examine how
+ecosystem structure and function complement one another across a large
+environmental gradient and then compare the simple and modelled
 representations of forest structure to different levels of ecosystem
@@
 Our second question examines the independent and shared relationships of
-ecosystem structure height and cover, with modelled forest structure, on
-ecosystem function. This provides insight into the choice of remote
-sensing attributes to use when developing EBVs within a single EBV
-class. Remote sensing datasets can comprise relatively unprocessed
-observations, in this case ALS measures of height and cover which are
-derived from the raw 3D point cloud vs modelled attributes, such as
-biomass and volume, which involve the use the statistical relationships
-with field data to transform the observations into more refined data
-products. Assessing which of these two (or combination of the two)
-approaches has stronger or weaker relationships with estimates of
-ecosystem function provides insights into the choice of data used to
-build EBVs. Lastly, we examine how the primary and modelled structure
-variables partition the variance of the DHIs within key biomes and
-forest types across a large environmental range, examining to what
-extent ecosystem and forest types impacts these relationships and thus
-providing insight into the applicability of these results globally.
+lidar derived forest structural attributes, with modelled forest
+structure, on ecosystem function. This provides insight into the
+selection of remote sensing attributes to use when developing EBVs
+within a single EBV class. Remote sensing datasets can comprise
+relatively unprocessed observations, in this case lidar-imputed measures
+of height and cover vs modelled attributes, such as biomass and volume,
+which involve the use the statistical relationships with field data to
+transform the observations into more refined data products. Assessing
+which of these two (or combination of the two) approaches has stronger
+or weaker relationships with estimates of ecosystem function provides
+insights into the choice of data used to build EBVs. Lastly, we examine
+how the primary and modelled structure variables partition the variance
+of the DHIs within key biomes and forest types across a large
+environmental range, examining to what extent ecosystem and forest types
+impacts these relationships and thus providing insight into the
+applicability of these results globally.
 
@@
 examine trends across the large environmental gradients, we group the
-BEC zones into five broad biomes, specifically, the southern interior,
-northern interior, montane, alpine, and coastal groups similar to
-@hamann2006. We also report each BEC zone's average climate data,
-according to @wang2020 (@tbl-bec-group).
+BEC zones into five broad biomes: the southern interior, northern
+interior, montane, alpine, and coastal groups similar to @hamann2006. We
+also report each BEC zone's average climate data, according to @wang2020
+(@tbl-bec-group).
 
@@
 [@andrew2012] diversity, species abundances [@razenkova2023], and
-construct novel ecoregionalizations [@coops2009; @andrew2013] . Recent
+construct novel ecoregionalizations [@coops2009; @andrew2013]. Recent
 studies have began to examine how these indices can be constructed at a

### 2023-07-24 scripts/0_whole_game.qmd
return from australia
secondary sampling scheme, revised manuscript for ncc
sent to meg and tx 2023-07-21b

@@
 
-Needed things
+Needed rasters for 1 2 and 3
 
 ```{r loads}
+vlce_loc <- "F:/mosaiced/VLCE2.0/LC_Class_terr_bc.dat"
+
+if (!file.exists(vlce_loc)) {
   vlce <- rast("F:/mosaiced/VLCE2.0/LC_Class_HMM_2015_v20_v20.dat")
@@
     mask(bcb_rast)
+
+  writeRaster(vlce, vlce_loc, filetype = "envi")
+}
+vlce <- rast(vlce_loc)
+bcb <- bcmaps::bc_bound() %>%
+  vect()
+
+bcb_rast <- bcb %>%
+  rasterize(y = vlce) %>%
+  crop(bcb)
 ```
@@
 
+dir.create(sample_loc, showWarnings = F)
+
 list.files(sample_loc, pattern = ".shp$")

### 2023-07-24 scripts/4_vegan.qmd
sent to meg and tx 2023-07-21b

@@
 ```{r data-prep}
-bec_sample_loc <- here::here("data", "samples", "bec_extracted.shp")
+bec_sample_loc <- here::here("data", "samples", "vlce_extracted.shp")
 
@@
   tibble() %>%
-  filter(elev_cv <= 1000) %>%
-  filter(elev_p95 >= 2000)
+  filter(elev_cv <= 1000) %>% #erroneous pixels
+  filter(elev_p95 >= 2000) # non-treed pixels
 
@@
 
-extracted <- to_rda %>%
+primary <- to_rda %>%
   select(elev_p95,
@@
 rda_out <- 
-  rda(dhi_tab ~ elev_p95 + percentage_first_returns_above_2m + elev_cv + # simple/raw/extracted
+  rda(dhi_tab ~ elev_p95 + percentage_first_returns_above_2m + elev_cv + # simple/raw/primary
       basal_area + gross_stem_volume + total_biomass, # modelled
@@
   pivot_wider(names_from = var, values_from = axis_loading) %>%
-  mutate(cluster = paste(axis, "- All Data")) %>%
+  mutate(cluster = paste(axis, "- British Columbia")) %>%
   select(-axis) %>%
@@
 
-var_out <- varpart(dhi_tab, extracted, modelled)
+var_out <- varpart(dhi_tab, primary, modelled)
 
 var_out %>% 
-  plot(Xnames = c("Derived", "Modelled"))
+  plot(Xnames = c("Primary", "Modelled"))
 
 path_r2 <- var_out$part$indfract %>%
-      mutate(type = c("extracted", "modelled", "overlap", "residual")) %>%
+      mutate(type = c("primary", "modelled", "overlap", "residual")) %>%
       select(type, Adj.R.squared) %>%
@@
 
-# vlce sampling regime to get the samples up
-vlce_samples <- read_sf(here::here("data", "samples", "vlce_extracted.shp"))
-
-vlce_df <- vlce_samples %>%
-  st_drop_geometry() %>%
-  select(-category_1, -category_2) %>%
-  rename(category = strata) %>%
-  relocate(bec) %>%
-  drop_na() %>%
-  ungroup()
-
-names(vlce_df) <- names_manual
-
-vlce_normal <- vlce_df %>%
-  mutate(across(basal_area:VarDHI,  .fns = ~ normal(.x)))
-
-to_nest <- vlce_normal %>%
-  select(-percentage_first_returns_above_mean,
-         -loreys_height,
-         -elev_stddev,
-         -elev_mean)
-
 class_nested <- to_nest %>%
   group_by(category) %>%
-  slice_sample(n = 3000) %>%
   nest() %>%
@@
   rename(strata = category)
+
+# # vlce sampling regime to get the samples up
+# switched away from this to a single double stratified sample
+# vlce_samples <- read_sf(here::here("data", "samples", "vlce_extracted.shp"))
+# 
+# vlce_df <- vlce_samples %>%
+#   st_drop_geometry() %>%
+#   select(-category_1, -category_2) %>%
+#   rename(category = strata) %>%
+#   relocate(bec) %>%
+#   drop_na() %>%
+#   ungroup()
+# 
+# names(vlce_df) <- names_manual
+# 
+# vlce_normal <- vlce_df %>%
+#   mutate(across(basal_area:VarDHI,  .fns = ~ normal(.x))) 
+# 
+# to_nest <- vlce_normal %>%
+#   select(-percentage_first_returns_above_mean,
+#          -loreys_height,
+#          -elev_stddev,
+#          -elev_mean)
+# 
+# 
+# 
+# 
+# class_nested <- to_nest %>%
+#   group_by(category) %>%
+#   slice_sample(n = 3000) %>%
+#   nest() %>% 
+#   ungroup() %>%
+#   rename(strata = category)
 
@@
 
-  rda(dhi_tab ~ elev_p95 + percentage_first_returns_above_2m + elev_cv + # simple/raw/extracted
+  rda(dhi_tab ~ elev_p95 + percentage_first_returns_above_2m + elev_cv + # simple/raw/primary
       basal_area + gross_stem_volume + total_biomass, # modelled, put condition here for a conditional rda
@@
 
-  extracted <- tibble %>%
+  primary <- tibble %>%
     select(elev_p95,
@@
 
-  varpart(dhi_tab, extracted, modelled)
+  varpart(dhi_tab, primary, modelled)
 }
@@
     x$part$indfract %>%
-      mutate(type = c("extracted", "modelled", "overlap", "residual")) %>%
+      mutate(type = c("primary", "modelled", "overlap", "residual")) %>%
       select(type, Adj.R.squared) %>%
@@
   arrange(desc(residual)) %>%
-  select(strata, extracted, modelled, overlap) %>%
-  pivot_longer(c(extracted, modelled, overlap)) %>%
+  select(strata, primary, modelled, overlap) %>%
+  pivot_longer(c(primary, modelled, overlap)) %>%
   mutate(value = ifelse(value > 0, value, 0)) %>%
@@
 fracts_out %>% 
-  mutate(total = extracted + modelled + overlap) %>%
+  mutate(total = primary + modelled + overlap) %>%
   arrange(strata %in% forest_types,
@@
   select(-residual) %>%
-  mutate(strata = "All Data") %>% 
+  mutate(strata = "British Columbia") %>% 
   relocate(strata)
@@
 fracts_tot <- fracts_all %>% 
-  mutate(total = extracted + modelled + overlap) %>%
+  mutate(total = primary + modelled + overlap) %>%
   select(strata, total)
@@
   mutate(group = ifelse(is.na(group), "Forest Types", group),
-         group = ifelse(strata == "All Data", "All", group)) %>%
+         group = ifelse(strata == "British Columbia", "BC", group)) %>%
   left_join(fracts_tot) %>%
@@
 fracts_bar_plot <- fracts_bar_tab %>%
-  ggplot(aes(x = value, y = fct_reorder(zone_nm, total), fill = fct_rev(fct_relevel(str_to_title(name), "Overlap", after = 1)))) +
+  ggplot(aes(x = value, y = fct_reorder(zone_nm, total), fill = str_to_title(name))) +
   geom_col(position = "stack") +
@@
 fracts_bar_plot
-ggsave(here::here("outputs", "fracts_bar_plot.png"), plot = fracts_bar_plot)
+ggsave(here::here("outputs", "fracts_bar_plot.png"), plot = fracts_bar_plot,
+       height = 6, width = 8)
 ```
@@
 fracts_out %>%
-  mutate(total = extracted + modelled + overlap) %>%
+  mutate(total = primary + modelled + overlap) %>%
   #left_join(keys$bec, by = c("bec" = "zone")) %>%
@@
   # geom_text(aes(x = 0, y = -1.15), label = "Modelled\nAttributes") +
-  # geom_text(aes(x = 0, y = 1.15), label = "Extracted\nAttributes") +
+  # geom_text(aes(x = 0, y = 1.15), label = "primary\nAttributes") +
   # geom_text(aes(x = .65, y = 0), label = "Overlap") +
@@
 
-path_r2_plot <- tibble(type = c("extracted", "modelled", "overlap"),
+path_r2_plot <- tibble(type = c("primary", "modelled", "overlap"),
        x1 = c(0.85, 0.85, 0.85),
@@
                      select(any_of(var_oi)),
-                   extracted, modelled)
-
-# plot(var_out, Xnames = c("Extracted", "Modelled"))
+                   primary, modelled)
+
+# plot(var_out, Xnames = c("Primary", "Modelled"))
 
 path_r2 <- var_out$part$indfract %>%
-      mutate(type = c("extracted", "modelled", "overlap", "residual")) %>%
+      mutate(type = c("primary", "modelled", "overlap", "residual")) %>%
       select(type, Adj.R.squared) %>%
@@
 
-clean_loads <- sig_loads %>%
+clean_loads_raw <- sig_loads %>%
   filter(anova_sig != "NS",
@@
   bind_rows(all_data_loadings %>%
-              mutate(strata = "All Data") %>%
+              mutate(strata = "British Columbia") %>%
               rename(loading_sig = sig)) %>%
-  select(strata:axis_loading) %>%
+  select(strata:axis_loading)
+
+clean_loads_abs <- clean_loads_raw %>%
   mutate(axis_loading = abs(axis_loading))
@@
 ##### run to here for analysis to be done
+
 radar chart based on groups, not hierarchical clustering
@@
   mutate(group = ifelse(is.na(group), "Forest Types", group),
-         group = ifelse(strata == "All Data", "All Data", group),
+         group = ifelse(strata == "British Columbia", "British Columbia", group),
          group = ifelse(axis == "RDA2", "RDA2", group)) %>%
@@
 radar_labels <- radar_loads %>%
-  select(percentage_first_returns_above_2m:elev_p95) %>%
+  select(-group, -ends_with("DHI")) %>%
   colnames() %>%
@@
          group = ifelse(axis == "RDA2", "RDA2", group),
-         group = ifelse(strata == "All Data", "All Data", group),
+         group = ifelse(strata == "British Columbia", "British Columbia", group),
          inout = ifelse(endsWith(var, "DHI"), "Response", "Predictor")) %>%
@@
        plot = radar_boxplot,
-       device = "png")
-```
+       device = "png",
+       height = 6,
+       width = 8)
+```
+
+# drawio
 
@@
 # list to iterate
-stratum <- clean_loads %>%
+stratum <- clean_loads_raw %>%
   pull(strata) %>%
@@
 # make sure everything gets change in rda fig
-possibles <- clean_loads %>%
+possibles <- clean_loads_raw %>%
   select(var, axis) %>%
@@
   filter(name != "RDA3") %>%
-  mutate(strata = "All Data") %>%
+  mutate(strata = "British Columbia") %>%
   rename(axis = name) %>%
@@
 for (s in stratum) {
-  strata_tp <- clean_loads %>%
+  print(s)
+  strata_tp <- clean_loads_raw %>%
     filter(strata == s) %>%
@@
 
-  for(p in possibles) {
+  for (p in possibles) {
     strata_drawio <- str_replace(strata_drawio, p, "NS")
@@
     filter(strata == s) %>%
-    pivot_longer(extracted:overlap) %>%
+    pivot_longer(primary:overlap) %>%
     mutate(value = (value * 100) %>%
@@
 
-  for(i in 1:length(fract_names)) {
-    strata_drawio <- str_replace(strata_drawio, fract_names[[i]], fract_value[[i]])
+  for (i in 1:length(fract_names)) {
+    strata_drawio <-
+      str_replace(strata_drawio, fract_names[[i]], fract_value[[i]])
   }
@@
 
-  for(i in 1:length(axis_names)) {
-    strata_drawio <- str_replace(strata_drawio, axis_names[[i]], axis_vals[[i]])
+  for (i in 1:length(axis_names)) {
+    strata_drawio <-
+      str_replace(strata_drawio, axis_names[[i]], axis_vals[[i]])
   }
@@
   theme(axis.text.x = element_text(angle = 45, hjust = 1))
-```
-
-# rgb map
-
-```{r}
-bec_dissolve <- read_sf(here::here("data", "shapefiles", "bec_dissolve.shp")) %>%
-  janitor::clean_names() %>%
-  select(zone)
-
-# non normalized
-loadings_rgb <- clean_loads %>% 
-  group_by(axis) %>%
-  pivot_wider(names_from = var,
-              values_from = axis_loading) %>%
-  ungroup() %>%
-  replace(is.na(.), 0) %>%
-  mutate(across(elev_cv:elev_p95, \(x) ifelse(x > 0.5, 0.5, x))) %>%
-  mutate(dhi_rgb = rgb(CumDHI, VarDHI, MinDHI),
-         pri_rgb = rgb(elev_p95, percentage_first_returns_above_2m, elev_cv, maxColorValue = 1),
-         mod_rgb = rgb(basal_area, total_biomass, gross_stem_volume, maxColorValue = 1)) %>%
-  select(strata, axis, percentage_first_returns_above_2m, elev_p95, elev_cv, ends_with("rgb")) %>%
-  pivot_longer(ends_with("rgb"))
-
-# normalized to max
-loadings_rgb <- clean_loads %>% 
-  group_by(var) %>%
-  mutate(axis_loading = axis_loading / max(axis_loading)) %>%
-  group_by(axis) %>%
-  pivot_wider(names_from = var,
-              values_from = axis_loading) %>%
-  ungroup() %>%
-  replace(is.na(.), 0) %>%
-  #mutate(across(elev_cv:elev_p95, \(x) ifelse(x > 0.5, 0.5, x))) %>%
-  mutate(dhi_rgb = rgb(CumDHI, VarDHI, MinDHI),
-         pri_rgb = rgb(elev_p95, percentage_first_returns_above_2m, elev_cv, maxColorValue = 1),
-         mod_rgb = rgb(basal_area, total_biomass, gross_stem_volume, maxColorValue = 1)) %>%
-  #select(strata, axis, percentage_first_returns_above_2m, elev_p95, elev_cv, ends_with("rgb")) %>%
-  pivot_longer(ends_with("rgb"))
-
-becs <- bec_normal %>%
-  pull(bec) %>%
-  unique()
-
-rgb_loads_split <- loadings_rgb %>%
-  filter(strata %in% becs) %>%
-  group_by(axis, name) %>%
-  group_split()
-```
-
-```{r}
-working <- rgb_loads_split[[1]]
-
-make_rgb_map <- function(tibble) {
-
-  rgb_cols = tibble %>%
-    pull(value) %>%
-    sort()
-
-  bec_dissolve %>%
-    left_join(tibble, by = c("zone" = "strata")) %>%
-    mutate(ifelse(is.na(value), "grey", value)) %>%
-    ggplot() +
-    geom_sf(aes(geometry = geometry, fill = value), colour = NA) +
-    scale_fill_manual(values = rgb_cols) +
-    theme_void() +
-    theme(legend.position = "none")
-}
-
-rgb_maps <- map(rgb_loads_split, make_rgb_map)
-
-patchwork_rgbplots <- (rgb_maps[[3]] + rgb_maps[[2]]) / 
-                      (rgb_maps[[6]] + rgb_maps[[5]]) &
-  plot_annotation(tag_levels = "A")
-
-ggsave(here::here("outputs", "rgbplots_norm.png"), patchwork_rgbplots,
-       height = 6,
-       width = 8)
-
-print("png saved")
-
-ggsave(here::here("outputs", "rgbplots_norm.pdf"), patchwork_rgbplots,
-       height = 6,
-       width = 8)
-
-print("pdf saved")
 ```
@@
 ```
+
+# new radar loads
+
+```{r}
+clean_loads %>%
+  pivot_wider(names_from = var, 
+              values_from = axis_loading) %>%
+  left_join(keys$bec, by = c("strata" = "zone")) %>%
+  mutate(group = ifelse(is.na(group), strata, group),
+         group = ifelse(strata == "British Columbia", "British Columbia", group)) %>%
+  group_by(group, axis) %>%
+  summarize(across(CumDHI:elev_p95, \(x) mean(x, na.rm = T))) %>%
+  pivot_longer(CumDHI:elev_p95) %>%
+  mutate(value = ifelse(is.na(value), 0, value)) %>%
+  left_join(keys$continuous, by = c("name" = "variable")) %>%
+  mutate(var_long = ifelse(is.na(var_long), name, var_long),
+         order = endsWith(var_long, "DHI")) %>%
+  ggplot(aes(x = fct_reorder(var_long, value, .desc = T), y = value)) +
+  geom_col() +
+  #coord_polar() +
+  coord_flip() +
+  facet_grid(cols = vars(group), rows = vars(axis)) +
+  labs(x = "Variable",
+       y = "Axis Loading (absolute)")
+
+clean_loads %>%
+  left_join(keys$bec, by = c("strata" = "zone")) %>%
+  mutate(group = ifelse(is.na(group), strata, group),
+         group = ifelse(strata == "British Columbia", "British Columbia", group)) %>%
+  ggplot(aes(x = var, y = axis_loading, col = group)) +
+  geom_line()
+```
+
+```{r}
+clean_loads %>%
+  left_join(keys$bec, by = c("strata" = "zone")) %>%
+  mutate(group = ifelse(is.na(group), strata, group),
+         group = ifelse(strata == "British Columbia", "British Columbia", group)) %>%
+  pivot_wider(names_from = var, values_from = axis_loading) %>%
+  select(-zone_nm, -order) %>%
+  replace(is.na(.), 0) %>%
+  GGally::ggparcoord(columns = 4:12,
+                     groupColumn = "group",
+                     scale = "globalminmax") +
+  facet_wrap(~axis) +
+  coord_flip()
+```
+
+# parallel coordinate plots
+
+```{r}
+data <- clean_loads_abs %>%
+  left_join(keys$bec %>%
+              select(strata = zone, group)) %>%
+  left_join(keys$continuous %>%
+              select(var = variable, var_long)) %>%
+  mutate(group = ifelse(is.na(group), strata, group),
+         var_long = ifelse(is.na(var_long), var, var_long)) %>%
+  group_by(group, var_long, axis) %>%
+  summarize(axis_loading = mean(axis_loading)) %>%
+  ungroup() %>%
+  complete(group, var_long, axis, fill = list(axis_loading = NA)) %>%
+  mutate(inout = ifelse(endsWith(var_long, "DHI"), "Response", "Predictor"), 
+         group = fct_relevel(group, "British Columbia", 
+                             "Broadleaf", 
+                             "Coniferous", 
+                             "Mixed Wood", 
+                             "Wetland-Treed", 
+                             after = 0),
+         var_long = fct_relevel(var_long, 
+                                c("Canopy Height", "Canopy Cover", "Structural Complexity",
+                                "Basal Area", "Aboveground Biomass", "Gross Stem Volume",
+                                "Cumulative DHI", "Variation DHI", "Minimum DHI")))
+
+# nojitter/lines
+parcoord_noflip <- data %>%
+  ggplot(aes(x = var_long, y = axis_loading)) +
+  #geom_boxplot(col = "#00000025", outlier.alpha = 0) +
+
+  geom_line(aes(col = group, group = group), lty = "dashed") +
+  geom_point(aes(col = group, group = group, shape = group, fill = group)) +
+  # geom_hline(yintercept = 0, lty = "dashed", col = "red") +
+  facet_wrap(~paste(axis, inout),
+             scale = "free") +
+  #theme_bw() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
+  labs(x = NULL,
+       y = "Axis Loading (absolute)",
+       col = NULL,
+       shape = NULL,
+       fill = NULL) +
+  # coord_flip() +
+  scale_colour_discrete_c4a_cat(palette = "bold") +
+  scale_fill_discrete_c4a_cat(palette = "bold") +
+  scale_shape_manual(values = c(21, 22, 22, 22, 22, 25, 25, 25, 25, 25))
+
+parcoord_noflip
+
+ggsave(here::here("outputs", "parcoord_noflip.png"), plot = parcoord_noflip,
+       height = 6, width = 8)
+
+ggsave(here::here("outputs", "parcoord_noflip.svg"), plot = parcoord_noflip,
+       height = 6, width = 8)
+```
+
+# rgb map
+
+```{r}
+bec_dissolve <- read_sf(here::here("data", "shapefiles", "bec_dissolve.shp")) %>%
+  janitor::clean_names() %>%
+  select(zone)
+
+# non normalized
+loadings_rgb <- clean_loads %>% 
+  group_by(axis) %>%
+  pivot_wider(names_from = var,
+              values_from = axis_loading) %>%
+  ungroup() %>%
+  replace(is.na(.), 0) %>%
+  mutate(across(elev_cv:elev_p95, \(x) ifelse(x > 0.5, 0.5, x))) %>%
+  mutate(dhi_rgb = rgb(CumDHI, VarDHI, MinDHI),
+         pri_rgb = rgb(elev_p95, percentage_first_returns_above_2m, elev_cv, maxColorValue = 1),
+         mod_rgb = rgb(basal_area, total_biomass, gross_stem_volume, maxColorValue = 1)) %>%
+  select(strata, axis, percentage_first_returns_above_2m, elev_p95, elev_cv, ends_with("rgb")) %>%
+  pivot_longer(ends_with("rgb"))
+
+# normalized to max
+loadings_rgb <- clean_loads %>% 
+  group_by(var) %>%
+  mutate(axis_loading = axis_loading / max(axis_loading)) %>%
+  group_by(axis) %>%
+  pivot_wider(names_from = var,
+              values_from = axis_loading) %>%
+  ungroup() %>%
+  replace(is.na(.), 0) %>%
+  #mutate(across(elev_cv:elev_p95, \(x) ifelse(x > 0.5, 0.5, x))) %>%
+  mutate(dhi_rgb = rgb(CumDHI, VarDHI, MinDHI),
+         pri_rgb = rgb(elev_p95, percentage_first_returns_above_2m, elev_cv, maxColorValue = 1),
+         mod_rgb = rgb(basal_area, total_biomass, gross_stem_volume, maxColorValue = 1)) %>%
+  #select(strata, axis, percentage_first_returns_above_2m, elev_p95, elev_cv, ends_with("rgb")) %>%
+  pivot_longer(ends_with("rgb"))
+
+becs <- bec_normal %>%
+  pull(bec) %>%
+  unique()
+
+rgb_loads_split <- loadings_rgb %>%
+  filter(strata %in% becs) %>%
+  group_by(axis, name) %>%
+  group_split()
+```
+
+```{r}
+make_rgb_map <- function(tibble) {
+
+  input <- bec_dissolve %>%
+    left_join(tibble, by = c("zone" = "strata")) %>%
+    mutate(value = ifelse(is.na(value), "#000000", value))
+
+  rgb_cols = input %>%
+    st_drop_geometry() %>%
+    pull(value) %>%
+    unique() %>%
+    sort()
+
+  input %>%
+    ggplot(aes(x = zone, y = 2, fill = value)) +
+    geom_col() +
+    scale_fill_manual(values = rgb_cols)
+
+   input %>%
+    ggplot() +
+    geom_sf(aes(geometry = geometry, fill = value, col = value), linewidth = 0.1) +
+    scale_fill_manual(values = rgb_cols) +
+    scale_color_manual(values = rgb_cols) +
+    theme_void() +
+    theme(legend.position = "none")
+}
+
+rgb_maps <- map(rgb_loads_split, make_rgb_map)
+
+labels = c("DHI", "B - Modelled RDA1", "A - Primary RDA1", "DHI", "D - Modelled RDA2", "C - Primary RDA2")
+
+# for (i in 1:length(rgb_maps)) {
+#   print(i)
+#   gg <- rgb_maps[[i]] + labs(title = labels[[i]])
+#   
+#   ggsave(here::here("outputs", "black_plots", paste(i, ".png")),
+#          plot = gg,
+#          height = 3,
+#          width = 3)
+# }
+
+ggsave(here::here("outputs", "test_black.png"), rgb_maps[[6]])
+
+patchwork_rgbplots <- (rgb_maps[[3]] + rgb_maps[[2]]) / 
+                      (rgb_maps[[6]] + rgb_maps[[5]]) &
+  plot_annotation(tag_levels = "A")
+
+cowplot_rgb <- cowplot::plot_grid(rgb_maps[[3]], rgb_maps[[2]],
+                                  rgb_maps[[6]], rgb_maps[[5]],
+                                  labels = c("A - Primary RDA1", "B - Modelled RDA1",
+                                             "C - Primary RDA2", "D - Modelled RDA2"),
+                                  vjust = 1.25)
+
+ggsave(here::here("outputs", "rgbplots_cow.png"), cowplot_rgb, height = 7, width = 8, bg = "white")
+
+# ggsave(here::here("outputs", "rgbplots_norm.png"), patchwork_rgbplots,
+#        height = 6,
+#        width = 8)
+# 
+# print("png saved")
+
+# ggsave(here::here("outputs", "rgbplots_norm.pdf"), patchwork_rgbplots,
+#        height = 6,
+#        width = 8)
+# 
+# print("pdf saved")
+```

### 2023-07-24 scripts/manuscript/manuscript.qmd
sent to meg and tx 2023-07-21b

@@
 ---
-title: Disentangling linkages between satellite derived forest structure and productivity essential biodiversity variables.
+title: Disentangling linkages between satellite-derived forest structure and productivity essential biodiversity variables.
 author:
@@
           postal-code: V8V 1T7
-abstract: Under a changing climate and in light of the biodiversity crises, monitoring change in biodiversity at broad scales has become ever more imperative.
-Key to regular biodiversity monitoring over time and space is the definition and selection of variables capable of accurately monitoring biodiversity trends.
-One framework is the Essential Biodiversity Variables (EBV), which are designed to be analogous to the Essential Climate variables and capable of monitoring biodiversity globally while also being complementary to one another within EBV classes.
-Five of six EBV classes are well suited to being monitored using satellite remote sensing, with particular focus warranted on the ecosystem structure and function classes.
-In this paper, we seek to address the complementarity of variables describing forest structure, as imputed across the entirety of British Columbia, Canada, with variables representing ecosystem functioning, namely the Dynamic Habitat Indices, a yearly summary of productivity indices.
-Both sets of variables have previously been linked with biodiversity metrics across a range of scales.
-Using redundancy analysis, we find that forest structural attributes and the DHIs are essentially decoupled from each other, with the forest structure variables explaining 14% of the variation in the DHIs.
-Further, we explore how the proportion of variance explained in the DHIs varies with either primary structure variables, such as canopy height and cover, vs more complex modelled forest structural attributes, such as aboveground biomass and volume.
-Results suggest there is large overlap between the two types of forest structure variables suggesting little added benefit of derived attributes beyond height and cover, in this context.
-Overall, we find that forest structure as an ecosystem structure EBV and the DHIs as an ecosystem function EBV represent markedly different processes as observed by remote sensing and they are highly complementary across large environmental gradients, suggesting that in an earth-observation based monitoring program both are needed.
 keywords: 
@@
   - forest productivity
-  - dynamic habitat indices
+  - dynamic habitat indices (DHIs)
   - essential biodiversity variables
-  - redundancy analysis
+  - redundancy analysis (RDA)
 date: last-modified
@@
     journal:
-      name: Remote Sensing of Environment
+      name: Nature Scientific Reports
       formatting: review
@@
 fig-width: 300
+number-sections: true
 ---
@@
 
+\newpage
+
+**Abstract**
+
+Key to regular biodiversity monitoring is the definition and selection
+of variables capable of accurately monitoring biodiversity trends. One
+framework is the Essential Biodiversity Variables (EBV), which are
+standardized, comparable data that represents a minimum set of
+biological information required for monitoring biodiversity change at
+large spatial extents, with each class being complementary to one
+another -- i.e., capturing unique biological information. The
+interrelationships between forest ecosystem structure and function and
+the complementarity between these two classes of EBVs across a range of
+ecosystems encompassing significant environmental gradients are examined
+using redundancy analysis (RDA) and variation partitioning. We found
+that the primary and modelled forest structure variables are not
+strongly related to the Dynamic Habitat Indices (DHIs), with the
+structural information only explaining 16% of the variation in the DHIs,
+indicating that structure and function EBVs are suitable to be used
+together for monitoring forest ecosystems. We also found that forest
+types and ecosystems dominated by coniferous trees had less variation
+explained, potentially due to the temporal decoupling of canopy cover
+and the DHIs. We suggest that biodiversity researchers focus on using
+forest structural attributes derived directly from lidar data alongside
+the DHIs as complementary EBVs for monitoring forest ecological
+integrity.
+
+\newpage
+
 # Introduction
 
-With biodiversity in decline, and facing extinction rates above the
-background extinction rate [@thomas2004; @urban2015], as well as the
-homogenization of communities at various scales [@mcgill2015], it is
-integral to be able to monitor how biodiversity is changing across the
-globe. In response, the global biodiversity community is making efforts
-to assess and halt the degradation of biodiversity. The Group on Earth
-Observations Biodiversity Observation Network has developed the
-Essential Biodiversity Variables [EBVs, @pereira2013], designed as an
-analog to the Essential Climate Variables framework [@bojinski2014].
-EBVs are designed to be global in scope, relevant to biodiversity
-information, feasible to use, and complementary to one another
-[@skidmore2021]. While it can be difficult, time consuming, and
-expensive to collect data on biodiversity across wide swaths of land and
-varying ecosystems, EBVs, which can be correlated to sampled
-biodiversity information, allow for the monitoring and assessment of
-protected area effectiveness and ecosystem health at large spatial
-scales [@hansen2021]. There are six EBV classes, each of which
-correspond to a different facet of biodiversity: species populations,
-species traits, community composition, ecosystem structure, ecosystem
-function, and genetic composition [@pereira2013].
+Monitoring the changing state of biodiversity globally is key to
+understanding and mitigating the accelerated extinction risk for many
+species [@thomas2004; @urban2015] and the homogenization of biotic
+communities at various spatial and temporal scales
+[@mcgillFifteenFormsBiodiversity2015]. The Group on Earth Observations
+Biodiversity Observation Network has developed the Essential
+Biodiversity Variables [EBVs, @pereira2013], which are standardized,
+comparable data that represents a minimum set of biological information
+required for monitoring biodiversity change at large spatial extents
+[@skidmore2021]. The EBVs are designed as an analog to the Essential
+Climate Variables framework [@bojinski2014] to be global in scope,
+relevant to biodiversity information, feasible to implement and use, and
+complementary to one another [@skidmore2021]. While it can be difficult,
+time consuming, and expensive to collect data on biodiversity across
+wide swaths of land, EBVs can be combined with sampled biodiversity
+information to allow for the spatial extrapolation of more sparsely
+collected in situ data [@cavender-bares2022]. The standardization and
+global availability of the EBVs allows for the assessment and monitoring
+of protected area effectiveness and ecosystem health at large spatial
+scales [@hansen2021], which can then be integrated into policy
+development [@geijzendorffer2016]. There are six EBV classes, each of
+which provide information on a different facet of biodiversity: species
+populations, species traits, community composition, ecosystem structure,
+ecosystem function, and genetic composition [@pereira2013]. While these
+EBV classes can be related to one another under a variety of ecological
+and evolutionary mechanisms, they provide distinct information; multiple
+in concert can provide a more complete picture on the status and trends
+of biodiversity in a region.
 
-Satellite remote sensing has proven to be capable of measuring five of
-the six EBV classes, the exception being genetic composition, which
-requires in-situ observation and sampling [@skidmore2021]. Species
-populations, and in turn community composition, can be assessed with
-very-high-resolution imagery to identify tree species at the tree-crown
-scale [@fassnacht2016; @graves2016]. Additionally, species traits such
-as vegetation phenology have been observed at the single-tree scale
-using, for example, PlanetScope imagery and drone-based measurements
-[@wu2021]. However, the spatially limited and often *ad-hoc* data
-collection approaches associated with monitoring individuals is not
-conducive to the global or regional scales required for biodiversity
-trend assessment [@valdez2023], in addition to the difficulties
-associated with extending species-level analyses to broader extents
-[@fassnacht2016; @graves2016].
+The global coverage and moderate spatial resolutions of many optical
+satellite remote sensing missions [@wulder2022], combined with open-data
+policies [@wulder2012], make these datasets exceptionally well suited
+for contributing to some EBV classes [@skidmore2021]. The six EBV
+classes can be divided into two general approaches: species-focused and
+ecosystem-focused [@fernández2020]. Genetic composition, species
+populations, and species traits are species-focused EBVs, with community
+composition being directly related to species populations. While
+species-focused and community composition EBVs can be assessed from
+measurements at the tree-crown scale using very-high spatial resolution
+imagery [@fassnacht2016; @graves2016; @wu2021], the spatially
+limited--and often *ad-hoc*--data collection associated with monitoring
+individuals is not conducive to the global or regional scales often
+required for biodiversity trend assessments [@valdez2023]. In contrast,
+two ecosystem-based EBV classes (ecosystem structure and ecosystem
+function) are well suited to be monitored across large extents using
+moderate spatial resolution optical imagery, such as the Moderate
+Resolution Imaging Spectroradiometer [MODIS, @zhang2003], Landsat
+imaging systems [@wulder2022] or Sentinel-2 [@helfenstein2022;
+@drusch2012] programs, especially in forested environments
+[@pettorelli2018; @coops2016].
 
-The two landscape-level EBVs (ecosystem structure and function) are well
-suited to be examined at large spatial scales using coarser spatial
-measurements, such as those taken from satellites by the Moderate
-Resolution Imaging Spectroradiometer [MODIS, @zhang2003], the Landsat
-imaging systems [@fisher2006], or Sentinel-2 [@helfenstein2022;
-@darvishzadeh2019] programs. These mid-resolution satellites can monitor
-processes at broader extents, but their coarse spatial resolution
-removes the ability to relate these traits to individual organisms. As a
-result, satellite remote sensing data has been shown to be arguably the
-most effective at monitoring ecosystem based EBVs focused on structure
-and function. These EBVs classes can be monitored at regional to global
-extents through the use of optical imagery [@cohen2004], as well as
-active sensors such as lidar (light detection and ranging) and radar
-[@guo2017; @lefsky2002; @lang2021; @neuenschwander2019; @coops2016].
+Ecosystem structure is the composition, abundance, and spatial
+arrangement of different ecosystem components, and is indicative of
+habitat quality and connectivity [@noss1990; @valbuena2020]. There are
+both horizontal and vertical components of ecosystem structure. In
+forest environments, forest structural diversity has been linked to
+species richness at various spatial scales, ranging from individual
+plots to landscapes [@bergen2009; @gao2014]. Structural attributes range
+in complexity from simple (e.g., canopy cover, canopy height), to more
+complex (e.g., foliage height diversity, leaf area index) to modelled
+(e.g., aboveground biomass, basal area), which can be derived using
+lidar (light detection and ranging) data [@coops2021]. A suite of these,
+and other, structural attributes has been used as indicators of
+biodiversity at local to landscape scales [@lefsky1999; @guo2017;
+@coops2016; @bergen2009]. Increased forest structural complexity has
+been hypothesized to create additional niches, leading to increased
+species diversity [@bergen2009], which has been frequently demonstrated
+using avian species diversity metrics [@macarthur1961], with lidar
+[@clawges2008], a combination of lidar and satellite imagery
+[@herniman2020], with some research indicating that canopy vertical
+distribution was the strongest predictor of species richness
+[@goetz2007]. Although lidar datasets have traditionally been local in
+extent, advances in satellite remote sensing processing have allowed the
+extension of three-dimensional forest structure attributes nationally in
+a wall-to-wall fashion. The wall-to-wall data often uses data fusion
+approaches involving lidar data and medium spatial resolution optical
+imagery [@matasci2018; @matasci2018b; @coops2021].
 
-## EBV - Ecosystem Structure
-
-Forest structural diversity has been linked to biodiversity at various
-scales [@guo2017; @bergen2009; @gao2014]. Structural attributes range in
-complexity from simple (canopy cover; canopy height), to more complex
-(vertical and horizontal structural complexity) to modelled (aboveground
-biomass; basal area), all of which can be assessed using lidar data
-[@coops2021]. A suite of these lidar-derived attributes have been used
-as local indicators of biodiversity, including simple metrics such as
-canopy cover and canopy height as well as derived metrics including
-vertical profiles, aboveground biomass, or lidar derived indices
-[@lefsky1999; @guo2017; @coops2016]. Other second order derived metrics
-such as canopy texture, height class distribution, edges, and patch
-metrics have also been used to examine habitat and biodiversity at
-landscape scales [@bergen2009]. Advances in satellite remote sensing
-processing have now allowed 3D forest structure data to be imputed
-across wide spatial scales [@matasci2018; @coops2021] using data fusion
-approaches involving collected lidar data and optical/radar data.
-
-Increased forest structural complexity has been hypothesized to create
-additional niches, leading to increased species diversity [@bergen2009],
-which has been frequently demonstrated using avian species diversity
-metrics [@macarthur1961]. For example, @herniman2020 used spectral and
-lidar derived forest structure data to model avian habitat suitability;
-@clawges2008 found that lidar derived forest structural attributes are
-capable of identifying habitat types associated with avian species in
-pine/aspen forests; @goetz2007 used canopy structural diversity to
-predict bird species richness, finding that canopy vertical distribution
-was the strongest predictor of species richness. Forest structural
-metrics have also been used to study biodiversity in other clades as
-well [@davies2014; @nelson2005].
-
-## EBV - Ecosystem Function
-
-With respect to ecosystem function, energy availability in an ecosystem
+Ecosystem functions are measures of ecosystem performance that are the
+consequence of one or multiple ecosystem processes [@pettorelli2018;
+@ecosyste2005]. Within the EBV framework, vegetation phenology and
+metrics of primary productivity are commonly used as potential EBVs
+[@skidmore2021]. Energy availability is the amount of accessible and
+utilizable energy available for growth in an ecosystem, and is often
+measured as net primary productivity [@bonn2004]. Energy availability
 has shown to be a predictor of species richness and abundances at
 various scales [@chase2002; @radeloff2019; @coops2019; @razenkova2023],
-and is measurable using satellite remote sensing via the use of various
-vegetation indices [@huete2002; @radeloff2019]. Vegetation indices,
-which are indicative of photosynthetic activity, are commonly used as
-proxies of gross primary productivity [@huang2019]. These vegetation
-indices have also been used to assess patterns in biodiversity at single
-time points [@bonn2004], and more recently, through yearly summaries of
-productivity [@berry2007; @radeloff2019]. The relationship between
-energy availability and biodiversity occurs via various hypothesized
-mechanisms, such as the available energy hypothesis [@currie2004;
-@wright1983], the environmental stress hypothesis [@currie2004], and the
-environmental stability hypothesis [@williams2008]. These three
-hypotheses have in turn been linked to patterns of annual surface
-reflectance in remote sensing data [@berry2007; @radeloff2019].
+and is measurable using satellite remote sensing via various vegetation
+indices [@huete2002; @radeloff2019]. Vegetation indices are indicative
+of photosynthetic activity, and are commonly used as proxies of gross
+primary productivity [@huang2019]. These vegetation indices enable the
+assessment of biodiversity patterns at single time points [@bonn2004],
+or through time when using time series data [@berry2007; @radeloff2019].
 
-@berry2007 first explored this idea by proposing the linkage of
-intra-annual summaries of MODIS-derived GPP to dispersive bird species.
-This idea was further refined into the Dynamic Habitat Indices [DHIs,
-@coops2008], which have now been shown to be well suited to assess the
-three aforementioned hypotheses at global scales [@radeloff2019]. The
-cumulative DHI calculates the total amount of energy available in a
-given pixel over the course of a year. Cumulative DHI is strongly linked
-to the available energy hypothesis, which suggests that with greater
-available energy species richness will increase [@wright1983]. The
-minimum DHI, which calculates the lowest productivity over the course of
-a year can be matched to the environmental stress hypothesis, which
-proposes that higher levels of minimum available energy will lead to
-higher species richness [@currie2004]. Finally, the variation DHI, which
-calculates the coefficient of variance in a vegetation index through the
-course of a year, corresponds to the environmental stability hypothesis
-which states that lower energy variation throughout a year will lead to
-increased species richness [@williams2008].
+The relationship between energy availability and biodiversity can be
+explained through various hypothesized mechanisms: including available
+energy, environmental stress, and environmental stability. These three
+hypotheses have been incorporated into the Dynamic Habitat Indices
+(DHIs), which summarizes intra-annual dynamics in remotely-sensed
+productivity metrics and have been evaluated as broad-scale biodiversity
+indicators [@berry2007; @coops2008; @radeloff2019]. According to the
+available energy hypothesis [@wright1983] an increase in available
+energy results in higher species richness. In the context of the DHIs,
+the total energy available within a specific area is quantified using
+the cumulative DHI, which sums each productivity observation throughout
+the course of a year. The environmental stress hypothesis [@currie2004]
+suggests that higher levels of minimum available energy contribute to
+greater species richness. The minimum DHI, derived from the lowest
+productivity observed over the course of a year, indicates the level of
+available energy during stressful periods. The environmental stability
+hypothesis [@williams2008] proposes that lower energy variation
+throughout the year leads to increased species richness. The variation
+DHI captures the coefficient of variation in a vegetation index over the
+course of a year, reflecting the stability of energy availability. The
+DHIs have previously been produced at a global extent using MODIS
+imagery and have been used to assess alpha [@radeloff2019] and beta
+[@andrew2012] diversity, species abundances [@razenkova2023], and
+construct novel ecoregionalizations [@coops2009; @andrew2013]. Recent
+studies have increased the spatial resolution of these indices by
+calculating the DHIs from multi-annual Landsat imagery
+[@razenkovaMediumresolutionDynamicHabitatInPress].
 
-## Biodiversity Monitoring with EBVs
-
-Biodiversity monitoring programs often require a range of information in
-order to accurately assess changes in ecological integrity
-[@lindenmayer2010]. Choosing datasets that are most closely related to
-the phenomenon of interest in a given application allows for direct
-connections to monitoring results and management actions [@pressey2021].
-With the advent of large-extent monitoring methods like satellite remote
-sensing, and a proliferation of potential EBVs datasets, it becomes
-important to assess the interrelationships between these datasets, and
-assess their complementary of the information to reduce the duplication
-of efforts [@pereira2013; @skidmore2021]. When strong relationships are
-present between EBVs, it is possible to assess the ecological
-relationships between potential EBVs. On the other hand, when datasets
-do not appear related, they may be well suited to be used in monitoring
-programs together, as complementary EBVs.
+Choosing datasets that are most closely related to the phenomenon of
+interest in a given application allows for direct integration of
+monitoring results with management actions [@pressey2021]. With the
+advent of large-extent monitoring methods like satellite remote sensing
+which have led to a proliferation of potential EBV datasets, it becomes
+important to assess the interrelationships between these datasets and
+the complementary of the information they provide to reduce redundancy
+in future EBV development efforts [@pereira2013; @skidmore2021].
+Identifying and exploring strong relationships between EBVs may help
+elucidate the ecological relationships between these facets of
+biodiversity, such as ecosystem structure and function, at local to
+global scales. On the other hand, when variables are not related, they
+may be well suited to be used in monitoring programs together, as
+complementary EBVs.
 
 Linkages between forest ecosystem structure and function have been
-examined within a remote sensing context for over 20 years [@huete2002,
+examined within a remote sensing context for decades [@huete2002,
 @knyazikhin1998; @myneni1994]. While there is significant theoretical
-and empirical evidence for their relationship at single time points
-(e.g. within a single image) [@myneni1994], various relationship
-directions and shapes have been found between forest structure and
-function metrics [@ali2019]. Hypothesized mechanisms such as niche
-complementarity have shown that aboveground biomass increases with stand
-structure [@zhang2012], while asymmetric competition for light can
-reduce forest productivity with increased structural complexity
-[@bourdier2016]. The relationship in particular between forest
-structural diversity metrics - which are now more accurately and
-comprehensively derived from lidar data - and temporal variation in
-functional metrics, specifically the metrics of ecosystem productivity
-via the DHI framework, have yet to be fully examined.
+and empirical evidence for their relationship [@myneni1994], the
+direction and significance has not been consistent, with positive,
+negative, and non-significant relationships being found [@ali2019].
+Hypothesized mechanisms such as niche complementarity suggest that
+productivity increases with stand structural complexity [@zhang2012],
+while asymmetric competition for light can reduce forest productivity
+with increased structural complexity [@bourdier2016]. Some aspects of
+forest structure, such as canopy cover and leaf area index, also
+directly influence vegetation indices, which are used to estimate
+productivity, especially in very dense or very sparse forests
+[@pettorelli2005]. Further, the relationship between forest structural
+diversity metrics--which are now more accurately and comprehensively
+derived from lidar data--and the temporal variation in functional
+metrics, specifically the metrics of ecosystem productivity via the DHIs
+framework, have yet to be fully examined.
 
-The overall goal of this paper is to assess patterns of forest ecosystem
-structure and function, as observed using remote sensing technologies,
-and their complementarity across a wide range of ecosystems encompassing
-significant environmental gradients. To do so, we synthesize data from
-moderate-scale remote-sensing derived metrics of forest structure,
-represented as both simple ALS-extracted metrics of canopy height, cover
-and vertical complexity, as well as modelled forest structure attributes
-including volume and aboveground biomass, with a well-established remote
-sensing derived index on ecosystem function. First, we examine how
-ecosystem structure and function complement one another across a large
-environmental gradient and then compare the simple and modelled
-representations of forest structure to different levels of ecosystem
-function. This question is important as it provides insights to the EBV
-community around complementarity of remote sensing metrics when
-describing the structure and function of ecosystems and proposes a
-method to examine potential overlap when generating remote sensing EBVs.
+The goal of this study is to examine relationships between forest
+ecosystem structure and function, and the complementarity between these
+two classes of EBVs across significant environmental gradients. We
+synthesize data from moderate-scale remote sensing-derived metrics of
+forest structure (simple lidar-extracted metrics of canopy height,
+canopy cover, and structural complexity), as well as modelled forest
+structure attributes (gross stem volume, basal area, and aboveground
+biomass), with a well established remote sensing-derived index for
+ecosystem function (DHIs). First, we analyze remote sensing metric
+complementarity by examining the multivariate relationship between
+ecosystem structure and function and propose a method to examine
+potential overlap in remote sensing EBVs. Second, we propose a method
+for the selection of remote sensing attributes for developing and
+applying EBVs within a single EBV using the independent and shared
+relationships of lidar-derived forest structural attributes and modelled
+forest structure on ecosystem function. Third, we examine the global
+applicability of these results using the relationship between simple and
+modelled forest structural attributes with DHIs across a range of
+environment gradients and forest types.
 
-Our second question examines the independent and shared relationships of
-lidar derived forest structural attributes, with modelled forest
-structure, on ecosystem function. This provides insight into the
-selection of remote sensing attributes to use when developing EBVs
-within a single EBV class. Remote sensing datasets can comprise
-relatively unprocessed observations, in this case lidar-imputed measures
-of height and cover vs modelled attributes, such as biomass and volume,
-which involve the use the statistical relationships with field data to
-transform the observations into more refined data products. Assessing
-which of these two (or combination of the two) approaches has stronger
-or weaker relationships with estimates of ecosystem function provides
-insights into the choice of data used to build EBVs. Lastly, we examine
-how the primary and modelled structure variables partition the variance
-of the DHIs within key biomes and forest types across a large
-environmental range, examining to what extent ecosystem and forest types
-impacts these relationships and thus providing insight into the
-applicability of these results globally.
-
-# Methods
+# Methods and Materials
 
@@
 
-British Columbia is the westernmost province of Canada, and is home to a
-variety of terrestrial ecosystems. Approximately 64% of the province is
+British Columbia (BC) is the westernmost province of Canada, and is home
+to a variety of terrestrial ecosystems. Located between the Pacific
+Ocean and the Rocky Mountains, approximately 64% of the province is
 forested, with large environmental and topographic gradients
-[@pojar1987; @bcministryofforests2003]. The Biogeoclimatic Ecosystem
-Classification (BEC) system identifies 16 zones based on the dominant
-tree species and the ecosystems general climate (@fig-study-map).
+[@pojar1987; @bcministryofforests2003]. The majority of the forests of
+BC are coniferous forests (87.4%), with the remaining forests being
+broadleaf (8.9%), wetland-treed (2.7%) and mixed-wood (0.9%)
+[@hermosilla2022]. There is a large topographic gradient in the
+province, with elevations ranging from sea level to over 4000 m, with
+much of the province being mountainous (elevations above 1000 m)
+[@valentine1978soil]. Wildfires are a regular disturbance in the
+province, with the historical fire regime being disrupted due to
+management actions, leading to larger, higher intensity fires becoming
+more common in recent years [@brookes2021]. In addition, forestry is a
+large industry in the province, with approximately 136000 ha of the
+province harvested accounting for \~54 million cubic metres of volume of
+wood in 2021 [@canadiancouncilofforestministers2023].
+
+BC is stratified into 16 zones based on the dominant tree species and
+climate by the Biogeoclimatic Ecosystem Classification (BEC) system
+(@fig-study-map). To examine trends across the large environmental
+gradients, we group the BEC zones into five broad biomes: the Nouthern
+interior, Sorthern interior, Montane, Alpine, and Coastal groups,
+similar to @hamann2006. We also report the average climate data for BEC
+zones from 1991-2020, according to @wang2016 (Table 1).
 
@@
 #| label: fig-study-map
-#| fig-cap: Location of biogeoclimatic ecosystem classification (BEC) zones in British Columbia.
+#| fig-cap: Location of Biogeoclimatic Ecosystem Classification (BEC) zones in British Columbia.
 
@@
 
-These zones can be further split into subzones, variants, and phases
-based on microclimate, precipitation, and topography [@pojar1987]. To
-examine trends across the large environmental gradients, we group the
-BEC zones into five broad biomes: the southern interior, northern
-interior, montane, alpine, and coastal groups similar to @hamann2006. We
-also report each BEC zone's average climate data, according to @wang2020
-(@tbl-bec-group).
+```{r}
+#| warning: false
+#| echo: false
 
-```{r}
-#| label: tbl-bec-group
-#| tbl-cap: BEC Zones, their aggregate groups, and their average climate values for precipitation, maximum temperature, and minimum temperature. Climate data from Wang et al. 2016.
-
-# bec_tab %>%
-#   gt() %>%
-#   cols_label(
-#     group = "BEC Group",
-#     zone = "BEC Zone",
-#     zone_nm = "Full Name",
-#     ppt = "Precipitation (mm/yr)",
-#     tmax = "Max Temperature (°C)",
-#     tmin = "Min Temperature (°C)"
-#   ) %>%
-#   as_word()
-
-knitr::kable(
-  bec_tab,
-  col.names = c(
-    "BEC Group",
-    "BEC Zone",
-    "Full Name",
-    "Precipitation (mm/yr)",
-    "Max Temperature (°C)",
-    "Min Temperature (°C)"
-  )
-)
+bec_tab %>%
+  gt() %>%
+  cols_label(
+    group = "BEC Group",
+    zone = "BEC Zone",
+    zone_nm = "Full Name",
+    ppt = "Precipitation (mm/yr)",
+    tmax = "Max Temperature (°C)",
+    tmin = "Min Temperature (°C)"
+  ) %>%
+  tab_header(title = "Biogeoclimatic Ecosystem Classification (BEC) Zones, their aggregated biomes, and their average climate values for precipitation, maximum temperature, and minimum temperature.
+Climate data from Wang et al.
+2016 averaged for 1991-2020.")
 ```
@@
 
-### Forest Structure
+### National Terrestrial Ecosystem Monitoring System (NTEMS)
 
-We used a suite of forest structure variables (canopy height, canopy
-cover, structural complexity \[coefficient of variation of height
-returns\], basal area, aboveground biomass, and gross stem volume). This
-dataset was created for the year 2015 at a 30 m spatial resolution
-according to @matasci2018. In brief, the method used a set of lidar
-collections and field plots across Canada, and imputed the remaining
-pixels using a random forest k-Nearest Neighbour approach on
-Landsat-derived surface reflectance and auxiliary data such as
-topography. Detailed information on the creation of this dataset can be
-found in @matasci2018.
+#### Best-Available-Pixel
+
+The forest structure, forest disturbance, and land cover layers are
+based on a 30-m best-available-pixel (BAP) composite generated by
+@hermosilla2016 using the Composite2Change approach. The composites were
+developed for each year from 1984-2019 by selecting the best available
+imagery from the available growing season Landsat observations, removing
+pixels with clouds, cloud shadows, and haze. The methodology uses the
+scoring method from @white2014 to select the best pixel for a given year
+from Landsat-5 Thematic Mapper, Landsat-7 Enhanced Thematic Mapper Plus,
+and Landsat-8 Operational Land Imager imagery. To reduce noise in the
+BAP between years, a spectral trend analysis on the Normalized Burn
+Ratio was conducted on each pixel, which removes unscreened noise,
+detects changes, and temporally interpolates data gaps. This resulted in
+a gap-free surface-reflectance composite across the entirety of Canada
+[@hermosilla2015]. We utilize data from 2015 for all NTEMS data, except
+forest disturbances, where we use all available years of data in the
+terrestrial area of BC, Canada.
+
+#### Forest Disturbances
+
+We used a forest disturbance layer generated by @hermosilla2015a to
+restrict the analysis to stands older than 35 years as productivity and
+structure metrics have been shown to be strongly decoupled in early
+regenerative stands [@bolton2017], and also excluding recently
+anthropogenically disturbed stands. Changes were detected using spectral
+trend analysis and were attributed using an object-based analysis
+approach to a disturbance agent (i.e., wildfire, harvest) using a Random
+Forests classification model [@hermosilla2015a]. Changes were detected
+with an overall accuracy of 89%, attributed with an accuracy of 92%, and
+assigned to the correct year with an accuracy of 89% [@hermosilla2016].
+
+#### Land Cover {#sec-lcc}
+
+We used a land cover map generated by @hermosilla2022 to stratify our
+samples alongside the BEC zones. The 30-m spatial resolution land cover
+map for Canada's forested ecosystems was generated using the Virtual
+Land Cover Engine framework [@hermosilla2022]. This framework uses a
+Random Forests based classification approach, and integrates logical
+land cover transitions using a Hidden Markov Model to generate 12 land
+cover classes. Training samples were locally selected using an inverse
+distance weighting approach [@hermosilla2022]. The treed land cover
+classes are broadleaf, coniferous, mixed wood, and wetland-treed, which
+we use as a stratification. The land cover classification reached an
+overall accuracy of 77.9% ± 4% [@hermosilla2022].
+
+#### Forest Structure
+
+Forest structural attributes (canopy height, canopy cover, structural
+complexity \[coefficient of variation of height returns\], basal area,
+aboveground biomass, and gross stem volume) for 2015 were extracted for
+BC from the 30-m spatial resolution dataset generated by Matasci et al.
+(2018a, 2018b). In brief, the method extends structural attributes
+derived from a set of lidar acquisitions and field plots across Canada.
+These data were generated for the treed pixels of Canada using the BAP
+Landsat surface reflectance composite and auxiliary data (i.e.,
+topography, geography) using a k-Nearest Neighbour imputation
+[@matasci2018; @matasci2018b]. Accuracy metrics for imputed forest
+structural attributes range from an RMSE of 24.5% (structural
+complexity) to 82.3% (gross stem volume) and *R^2^* ranging from 0.125
+(structural complexity) to 0.712 (gross stem volume) [@matasci2018;
+@matasci2018b].
 
@@
 
-We use an established set of indices of annual productivity shown to be
-related to global biodiversity trends: the Dynamic Habitat Indices
-[@radeloff2019]. The DHIs are a set of satellite remote sensing derived
-productivity variables that summarize the cumulative amount of available
-energy, the minimum available energy, and the variation in available
-energy throughout a given year [@berry2007; @radeloff2019]. The DHIs
-have previously been produced at a global extent using MODIS imagery,
-and have previously been used to assess alpha [@radeloff2019] and beta
-[@andrew2012] diversity, species abundances [@razenkova2023], and
-construct novel ecoregionalizations [@coops2009; @andrew2013]. Recent
-studies have began to examine how these indices can be constructed at a
-finer spatial resolution by using multi-annual Landsat imagery to
-generate a single synthetic year of monthly observations
-[@razenkova2022].
-
-The DHIs were calculated according to [@razenkova2022, Razenkova et al.,
-In Press] for all of terrestrial British Columbia. In brief, Google
-Earth Engine [@gorelick2017] was used to obtain all valid Landsat pixels
-for a given study area, filtering out pixels containing shadows, clouds,
-and cloud shadows within each image using the fmask algorithm
-[@zhu2012], then calculated the NDVI for each pixel in each image. They
-then calculated the median NDVI value for each month across a ten year
-time span (2011-2020) to generate a synthetic year of monthly data. The
-sum, minimum, and coefficient of variation across this synthetic year of
-NDVI values was then calculated. More detailed information can be found
-in Razenkova et al. (In Press).
+The DHIs were calculated following
+@razenkovaMediumresolutionDynamicHabitatInPress for the terrestrial area
+of BC. In brief, we used Google Earth Engine [@gorelick2017] to obtain
+valid Landsat pixels for the study area across a ten-year time span,
+centered on 2015 (2011-2020), filtering out pixels containing shadows,
+clouds, and cloud shadows using the QA band derived from the fmask
+algorithm [@zhu2012], and then calculated the NDVI index. These
+estimates were composited into a synthetic year of monthly data by
+calculating the monthly median NDVI value. Finally, images of the sum,
+minimum, and coefficient of variation across this synthetic year of NDVI
+values were then calculated to represent the cumulative, minimum, and
+variation DHIs, respectively. More detailed information can be found in
+@razenkovaMediumresolutionDynamicHabitatInPress.
 
@@
 
-# Sampling
+## Sampling
 
-We implement two sampling schemes -- one for ecosystems, and one for
-forest types -- across the terrestrial ecosystems of British Columbia
-(@fig-study-map). For both sampling schemes, each individual sample was
-taken from a forested pixel which was surrounded by the same land cover
-class. The land cover class map was generated for the year 2015
-following [@hermosilla2022] using a best-available-pixel composite, and
-an inverse-distance weighted random forest classification approach
-across Canada. Additionally, each pixel had to have a coefficient of
-variation less than 0.5 in surrounding pixels in the two simplest forest
-structure variables, canopy height and canopy cover. A minimum sampling
-distance of 1 km was implemented to reduce the effects of spatial
-autocorrelation, and samples that had been disturbed in the last 30
-years were discarded by using a disturbance mask generated for the
-forested ecosystems of Canada by @hermosilla2016.
+To obtain representative sample units of the EBV metrics across the
+study area, we implemented stratified random sampling across the union
+of BEC zones (Table 1) and forest types (i.e., coniferous, broadleaf,
+mixed wood, and wetland treed; see @sec-study-area) of BC. A maximum of
+500 sample units were selected from each stratum. Sample units were
+restricted to forested pixels that were surrounded by the same forest
+type to reduce the chance of sampling mixed pixels. To ensure that
+sample units were selected from a homogeneous area, reducing uncertainty
+associated with the input datasets [@shang2020], we required sample
+units to have a coefficient of variation for canopy cover and canopy
+height to be lower than 0.5 in a 3 x 3-pixel window. A minimum sampling
+distance of 1-km was implemented to reduce the effects of spatial
+autocorrelation, and sample units that had been disturbed in the last 30
+years were discarded by using the disturbance mask generated for the
+forested ecosystems of Canada as early regenerative forests have shown
+to have strongly decoupled productivity and canopy cover [@bolton2017].
 
-The ecosystem sampling scheme was implemented by randomly sampling
-pixels matching the above criteria within each BEC zone of British
-Columbia, up to a maximum of 3000 pixels, regardless of forest class.
-The forest type sampling scheme was similar in that each BEC zone was
-sampled, however, it implemented stratified sampling on each forest
-class, up to a maximum of 500 pixels per BEC zone and forest class.
-
-Each sampling scheme was natural-log transformed and standardized to
-Z-scores. Variables containing zeros were natural-log plus one
-transformed. Sampling was conducted in R [@R-base] version 4.2.2 using
-the **sgsR** package [@R-sgsR]. Focal analyses for the land cover
-classes and coefficient of variations of canopy height and cover were
-calculated in Python version 3.9.
+Sampling was conducted in R [@R-base] version 4.2.2 using the **sgsR**
+package [@R-sgsR]. Neighbourhood analyses for the land cover classes and
+coefficient of variations of canopy height and cover were calculated in
+Python version 3.9.
 
@@
 
-### Redundancy Analysis and Variation Partitioning
-
 Redundancy analysis (RDA) and variation partitioning were used to relate
 the primary and modelled forest structure variables to ecosystem
-function across a broad environment range. Redundancy analysis functions
-similarly to a multiple linear regression, except it is capable of
-predicting multiple response variables. It accomplishes this by first
-running a multiple linear regression of each predictor variable on each
-response variable, then running a principle component analysis on the
-residuals from each multiple linear regression. This reduces the
-dimensionality of the output, and allows the relationship strength to be
-assessed by calculating the loadings of both predictor and response
-variables on the RDA axes. Partial redundancy analysis functions
-similarly, except also considers co-variates [@legendre2012]. Redundancy
-analysis has widely been used in community ecology where environmental
-variables of interest are compared to species composition
-[@blanchet2014; @kleyer2012] and has similarities to partial least
-squares regression -- namely, their multivariate approach, usage of
-dimensionality reduction, and linearity assumptions -- which is commonly
-used in remote sensing literature [@roelofsen2014; @burnett2021].
+function across a broad environment range. RDA has widely been used in
+community ecology where environmental variables of interest are compared
+to species composition [@blanchet2014; @kleyer2012]. RDA predicts
+multiple response variables by first running a multiple linear
+regression of the predictor variables on each response variable, and
+then a principal component analysis on the fitted values from each
+multiple linear regression. This reduces the dimensionality of the
+fitted values by transforming them into a set of independent RDA axes,
+and allows the strength of the multivariate relationship between
+predictors and responses to be assessed by calculating eigenvalues of
+the RDA. Relationships between variables are revealed by the loadings of
+both predictor and response variables on the RDA axes. RDA axes are
+labelled as RDA1, RDA2, etc. to indicate they are derived from
+redundancy analysis, rather than other ordination techniques such as
+principle component analysis or constrained correspondence analysis
+[@legendre2012]. Variation partitioning is an extension of partial RDA
+-- which allows predictor variables to be considered within pre-defined
+groups of co-variates -- that can assess the overlap between the
+explanatory power of two datasets by utilizing multiple partial RDAs and
+exchanging which datasets are considered the predictor, and which are
+considered the co-variates [@legendre2012].
 
-Following the RDA, we employ ANOVAs to determine which axes are
-significant, and calculate the proportion of variance attributable to
-each axis using the eigenvalues generated from the RDA. We calculate
-axis loadings for both forest structure and DHI variables by calculating
-the correlation between each variable and the RDA axes. Forest strcuture
-variable axis loadings represent the strength of relationship between a
-given variable and the RDA axis, while DHI loadings indicate what is
-being represented by the RDA axes. We only consider and display
-significant axes. To visualize the RDA for both predictor and response
-variables, we display the results as path diagrams, with loadings from
-each predictor to the RDA axis to the response variables. The variance
-explained by each axis is also displayed in the RDA box (@fig-rda-var).
+We performed RDAs for our full sample, pooling BEC zones and forest
+types, as well as individually by BEC zone and forest type, with results
+aggregated to BEC zone groups (see @sec-study-area). All measured values
+were natural-log transformed and standardized to Z-scores based on the
+summary statistics of the full sample prior to analyses. Forest
+structure variables were treated as predictor variables and were grouped
+based on whether they were primary estimates from the lidar data or
+modelled from lidar and ancillary information (Table 2) for variance
+partitioning. The three DHIs were used as the response variables.
+Analysis of variance tests were used to determine which RDA axes were
+significant, with p-values below 0.1 using an F-test. The proportion of
+variance attributable to each axis was calculated from the eigenvalues
+generated by the RDA, and axis loadings were calculated as the Pearson
+correlation between each forest structure or DHI variable and the RDA
+axes. Forest structure variable axis loadings represent the strength of
+relationship between a given variable and the RDA axis, while the DHIs
+loadings indicate what is being represented by the RDA axes. To
+visualize the RDA results for both predictor and response variables, we
+display the results as path diagrams, with paths annotated by the
+loadings and RDA axes annotated by the variance in the DHIs explained by
+each axis (example shown in @fig-rda-var-template). We only consider and
+display significant RDA axes.
 
-Variation partitioning is an extension of partial RDA which can assess
-the overlap between the explanatory power of two datasets by utilizing
-multiple partial RDAs and exchanging which datasets are considered the
-predictor, and which is considered the co-variate [@legendre2012].
-Variation partitioning is traditionally displayed using a Venn diagram,
-in which the percentage of variance explained by each dataset is in a
-circle, and the overlap between circles represents the overlap in
-variance explained.
+Variation partitioning is displayed using a Venn diagram in which the
+percentage of variance explained by each dataset is in a circle, and the
+overlap between circles represents the overlap in variance explained. We
+add paths to the Venn diagram to indicate how the structural attributes
+contribute to the variation in the DHIs. Because the variation
+partitioning results summarize the full multivariate solution, rather
+than individual RDA axes, we present variation partitioning results
+alongside the path diagrams illustrating the details of the RDA results
+(@fig-rda-var-template).
 
-RDA and variation partitioning analyses were conducted for all samples
-in the ecosystem sampling scheme, as well as individually across each
-ecosystem and forest type. The results were aggregated to BEC zone
-groups (see @sec-study-area). All RDA and variation partitioning
-calculations were done in R [@R-base] version 4.2.2 using the **vegan**
-package [@R-vegan].
+```{r}
+#| label: fig-rda-var-template
+#| fig-cap: Template for visualization of RDA and variation partitioning results.
+Paths between RDA axes and predictor/response variables will be labelled with axis loadings, caluclated as the correlation between the variable and the RDA axis.
+If loadings are non-significant, NS will be shown on the path.
 
-All code associated with the processing and analysis is available at
+knitr::include_graphics(here::here("outputs", "template.drawio.png"))
+```
+
+RDA and variation partitioning calculations were done in R [@R-base]
+version 4.2.2 using the **vegan** package [@R-vegan]. The code
+associated with the processing and analysis is available at
 https://github.com/emuise/code-structProdSem.
@@
 
+Overall in BC, the intra-annual patterns of productivity as represented
+by the DHIs variables were weakly related to forest structure
+(@fig-rda-var). Less than 16% of the overall variation in the DHIs was
+explained by the forest structural attributes. Due to strong
+correlations between the axes of the DHIs, the majority of the variation
+was held within the first RDA axis, which had strong loadings for all
+DHIs, with positive loadings for the cumulative and minimum DHIs, and
+negative loadings for the variation DHI. As such, we identified the RDA1
+axis as representing overall productivity. The overall productivity axis
+had similar loadings (between 0.24-0.27) for all modelled predictors and
+canopy cover. Canopy height had a smaller loading (0.1), while
+structural complexity was negatively correlated with the overall
+productivity axis. RDA2 explained much less of the variation in the
+DHIs. RDA2 had a large positive loading on the minimum DHI (0.55), with
+negative loadings on the cumulative (-0.19) and variation (-0.39) DHIs.
+Due to the minimum and variation having the most influence on RDA2, we
+identified it as a seasonality axis. Canopy cover and structural
+complexity had the strongest (albeit negative; -0.17 and -0.08,
+respectively) loadings on the seasonality axis (@fig-rda-var A).
+Variation partitioning highlighted that the majority of the explained
+variation in the DHIs was due to the primary attributes (canopy cover,
+canopy height, and structural complexity; 9% variation explained). The
+overlap between primary and modelled attributes was 3.5%, and the
+modelled attributes explained 3.2% of the variation on their own
+(@fig-rda-var B).
+
 ```{r}
 #| label: fig-rda-var
-#| fig-cap: A) Axis loadings from redundancy analysis of primary and modelled forest structure variables on the dynamic habitat indices.
+#| fig-cap: A) Axis loadings from redundancy analysis (RDA) of primary and modelled forest structure variables on the dynamic habitat indices (DHIs).
 B) Results from variation partitioning of primary and modelled forest structure variables on the DHIs.
-Both visualized analyses are across all collected samples.
+Both visualized analyses are across the entire dataset.
+Non-significant paths are labelled as "NS".
 See supplementary information for results from each BEC zone and forest type.
 
-knitr::include_graphics(here::here("outputs", "all_data_figure.drawio.png"))
+knitr::include_graphics(here::here("outputs", "drawio", "British Columbia.drawio.png"))
 ```
 
-To examine the relationship between ecosystem structure and function
-across a large environmental gradient, we present the results of a
-redundancy analysis of forest structure variables on the dynamic habitat
-indices across the entire sampled dataset in @fig-rda-var A. While there
-are three RDA axes associated with the full dataset, the third axis
-explains 0.05% of the variance in the DHIs, and as a result is not
-shown. The first axis strongly represents all the DHIs (loadings \> 0.85
-for all DHIs), with the highest loadings from canopy cover, basal area,
-aboveground biomass, and gross stem volume. The other input variables
-(canopy height, structural complexity) have smaller loadings. The second
-axis primarily represents the seasonality (Minimum and Variation) of the
-DHIs, with no correlation to Cumulative DHI, and has the highest
-loadings from canopy cover and complexity, with the remainder of the
-forest structure variables having negligible influences(@fig-rda-var A).
+Subdividing the sample by forest type led to similar or increased
+amounts of variance explained when compared to the overall dataset
+(@fig-fracts-bar). Subdividing by BEC zones generally led to smaller
+amounts of variation explained than BC or stratifying by forest type,
+which may be due to the length of the environmental gradients included
+in the resulting subsamples (@fig-fracts-bar). The overlap between the
+primary and modelled datasets often explained the majority of the
+variation in the DHIs, except when stratifying by forest types, where
+this is only true for the coniferous subsample. The broadleaf, mixed
+wood, and wetland-treed forest types contained most of the DHIs
+explained variation in the primary structural attributes. The BEC zones
+followed a similar pattern, with most zones' DHI variation principally
+explained by the overlap, with the exception being CWH, which has no
+variation explained by the overlap between primary and modelled
+datasets. The Coastal BEC group had the lowest amount of variation
+explained, with Coastal Douglas-fir having \<1% of variation explained
+by the structural data. Overall, the DHIs are decoupled from the forest
+structural attributes, as shown by the overall variation explained being
+under 30% regardless of data stratification (@fig-fracts-bar).
 
-The results from the variation partitioning analysis (@fig-rda-var B)
-show that the majority of the variance explained by the input datasets
-is shared across both primary and modelled forest structure variables.
-Of the 13.4% of the variation in the DHIs being explained by the
-structural information, 9.4% is common across both sets of structure
-variables. The primary and modelled variables only explain 2.7% and 1.3%
-of the variation uniquely (@fig-rda-var B).
+```{r}
+#| label: fig-fracts-bar 
+#| fig-cap: Stacked bar plot proportion of the variation in the Dynamic Habitat Indices (DHIs) variation explained by primary, modelled, and the overlap between primary and modelled structure variables across the stratified and overall datasets.
+knitr::include_graphics(here::here("outputs", "fracts_bar_plot.png"))
+```
+
+Stratifying the dataset by forest type and ecosystem led to varying
+loadings between the RDA axes and predictor/response variables
+(@fig-radar). The DHIs' loadings on RDA1 are large (similar to the
+overall dataset), except in colder BEC groups (Montane, North, and
+Alpine), where the minimum DHI loading half as large as the cumulative
+or variation DHI loadings. As such, we do not identify the first axis as
+being consistently associated with overall productivity. Non-coniferous
+forest types (broadleaf, mixed wood and wetland-treed) typically had
+large loadings between RDA1 and structural complexity. The mixed wood
+forest and Coastal BEC group had no significant loadings from modelled
+forest structural attributes for RDA1. Canopy cover generally had the
+highest significant predictor loading on average in RDA1. Examining the
+predictor loadings for RDA1 spatially across the province shows a
+geographic pattern in the loading strengths of primary structural
+attributes (@fig-fcc A). The interior (Northern and Southern BEC groups)
+of the province generally showed high canopy cover loadings, while the
+Boreal zones in the northwest had equal loadings in canopy cover and
+canopy height. Coastal were driven similarly by canopy height and
+structural complexity, with very low or non-significant loadings from
+canopy cover (@fig-fcc A; @fig-radar). Modelled attributes for RDA1
+generally show similar loadings across the province (@fig-fcc B).
 
@@
 #| label: fig-radar
-#| fig-cap: Radar plots of average loadings strength by group.
-A and B show input and response loadings, respectively.
-C) Boxplots of BEC zone, forest types, and all data loadings for predictor and response variables.
+#| fig-cap: Parallel coordinate plots of average absolute significant loading strength by Biogeoclimatic Ecosystem Classification (BEC) group, forest type, and British Columbia.
+Note the varying y axes.
 
-knitr::include_graphics(here::here("outputs", "radar_boxplot.png"))
+knitr::include_graphics(here::here("outputs", "parcoord_noflip_edit.png"))
 ```
 
-Further, we wanted to examine the strength of the relationship between
-the DHIs and primary structure variables vs modelled structure variables
-across the forest types and ecosystems of the province. @fig-radar A and
-B show the axis loadings for structure variables and the DHIs,
-respectively. Across the BEC zone groups, the loadings are generally
-similar in the structure variables. Variations in the DHIs in the
-southern ecosystems, which are typically warmer and drier, have low or
-non-significant variance explained by the structure variables.
-Conversely, across the forest types (coniferous, broadleaf, mixed wood,
-and wetland-treed) structural complexity is often the highest loading.
-The loadings for the DHIs in the first axis are generally larger than
-the loadings for the structure variables, with the cumulative DHI often
-being the strongest loading overall. In alpine BEC zones, the minimum
-DHI loading is smaller than the variation and cumulative DHIs. In
-contrast to the first RDA axis, the secondary RDA axis is primarily
-driven by variation in the minimum DHI, with medium loadings in the
-variation DHI, and small loadings in the cumulative DHI (@fig-radar A &
-B). @fig-radar C shows boxplots of individual BEC zone loadings for
-structure variables and the DHIs. Across the province, canopy cover
-generally has the highest loadings, with structural complexity generally
-being the weakest.
-
-```{r}
-#| label: fig-fracts-bar
-#| fig-cap: Stacked bar plot proportion of DHI variation explained by extracted, modelled, and the overlap between primary and modelled structure variables.
-
-knitr::include_graphics(here::here("outputs", "fracts_bar_plot.png"))
-```
-
-@fig-fracts-bar shows stacked bar plots of the proportion of variation
-explained in the DHIs by primary, modelled, and the overlap between
-primary and modelled variables. Forest types sampled across the
-environmental gradient generally show higher amounts of variance
-explained, which is especially prevalent in broadleaf and mixed wood
-forests. The overlap between modelled and primary structure variables is
-generally than the primary and modelled attributes individually, with
-the exception being mixed wood and broadleaf forests and colder, wetter
-forests. Higher elevation and colder forest stands have larger amounts
-of DHI variance explained by the forest structure variables than other
-environments in the province. Notably, the variation explained by the
-primary and modelled structure variables is commonly presented through a
-single RDA axis, which generally corresponds to overall productivity
-through the year (@fig-radar). Overall, the DHIs are decoupled from the
-structure variables, as shown by the overall variance explained being
-under 30%, regardless of whether the attribute is directly extracted
-from the point cloud or is a modelled variable.
+Only seven of sixteen BEC zones had a second RDA axis, with none being
+found in Coastal or Alpine BEC zones (@fig-radar; @fig-fcc) . All of the
+Montane BEC zones had a secondary axis, while some in the Northern and
+Southern interior had a secondary axis. RDA2 often showed the largest
+absolute loadings in the minimum DHI, which was frequently negative
+(@fig-radar; see supplementary material). The secondary axis had smaller
+loadings in the variation and cumulative DHIs, with only the Southern
+BEC group having a lower variation DHI loading than cumulative DHI. RDA2
+has lower predictor loadings (maximum of \~0.25) when compared to RDA1
+(maximum of \~0.5), with infrequently significant loadings for canopy
+cover and structural complexity (@fig-radar). Canopy height had much
+larger primary attribute loadings in the Southern group, while the
+Northern group had slightly larger structural complexity loadings
+(@fig-radar; @fig-fcc C) Similar to RDA1, the modelled structural
+attribute loadings were generally similar to one another, although gross
+stem volume was slightly higher in some southern BEC zones (@fig-radar;
+@fig-fcc D).
 
@@
 #| label: fig-fcc
-#| fig-cap: False colour maps of axis loadings for the first RDA axis (top) and second RDA axis (bottom).
-All colour values are normalized to the maximum loading of each variable.
-A and D show axis loadings for canopy height (r), canopy cover (g) and structural complexity (b).
-B and E show axis loadings for basal area(r), total biomass (g) and gross stem volume (b).
-C and F show axis loadings for the cumulative DHI (r), variation DHI (g) and minimum DHI (b).
+#| fig-cap: False colour maps by Biogeoclimatic Ecosystem Classification (BEC) zone of axis loadings for the first redundancy analysis (RDA) axis (A and B) and second RDA axis (C and D).
+Colour values are normalized to the maximum loading of each variable.
+A and C show axis loadings for canopy height (CH; red), canopy cover (CC; green) and structural complexity (SC; blue).
+B and D show axis loadings for basal area (BA; red), total biomass (TB; green) and gross stem volume (GSV; blue).
+Black indicates no significant loadings, while greyscale colours indicate similar loadings across the three variables.
+Annotated colour wheels for each set of variables also shown.
 
-knitr::include_graphics(here::here("outputs", "rgb_norm_edit.png"))
+knitr::include_graphics(here::here("outputs", "maps_colourwheel.png"))
 ```
-
-@fig-fcc shows false colour composites of the primary (A & C) and
-modelled (B & D) loadings across the ecosystems of British Columbia. In
-the first RDA axis there is spatial variation in the primary structure
-variables, with interior zones being primarily driven by canopy cover
-(green), the coastal regions having strong structural complexity and
-canopy height loadings, and the boreal in the northwest having the
-strongest loading in canopy height (@fig-fcc A). The modelled structure
-variables generally show gray-scale colour, indicating that basal area,
-total biomass, and gross stem volume explain similar variation in the
-DHIs (@fig-fcc B). All ecosystems with a secondary axis are located away
-from the Pacific coast in cooler, drier environments. Canopy cover has
-the strongest secondary axis loading in the northern boreal stands,
-while the secondary axis in the remaining four zones have the highest
-loadings in canopy height (@fig-fcc C). Again, the modeled structure
-variables are in grayscale, indicating similar loadings across the three
-modelled structure variables (@fig-fcc E).
 
@@
 
-Complementarity plays a crucial role in the context of Essential
-Biodiversity Variables (EBVs) and holds significant importance for
-monitoring and understanding biodiversity [@pereira2013; @skidmore2021].
-At the ecosystem scale, the interconnectedness and interdependence of
-ecosystem structure and function is important to recognize and
-disentangle. In forested environments, forest structure and productivity
-have been shown to be linked in multiple studies spanning three decades
-[@ali2019; @myneni1994], however, a linkage between intra-annual
-production and forest structure has yet to be shown. In this study, we
-use statistical analyses commonly used in community ecology - namely
-redundancy analysis and variation partitioning - to assess the
-complementarity of forest structure and yearly productivity summaries.
-While RDA is common in the ecological literature, this analysis
-represents one of the first times this technique has been applied to
-assess the complimentary of proposed satellite-derived EBVs. We find
-that the example EBVs do not strongly overlap, with forest structure
-explaining 13.4% of the variation in the DHIs in samples taken across a
-broad environmental gradient (@fig-rda-var B). This indicates that they
-are suitable to be used in tandem with one another when used as
-ecosystem EBVs across wide environmental gradients.
+Relationships between forest ecosystem structure and function are
+increasingly being examined due to advances in data collection methods
+[@ali2019; @radeloff2019; @ali2016]. Prior studies have shown both
+theoretical [@knyazikhin1998] and empirical [@atkins2018] relationships
+between forest ecosystem structure and function. We examined these
+relationships across the province of BC, Canada, which has significant
+environmental gradients [@valentine1978soil; @pojar1987] by using
+wall-to-wall lidar-derived forest structural attributes [@matasci2018;
+@matasci2018b] and intra-annual satellite-derived summaries of
+vegetation productivity [@radeloff2019;
+@razenkovaMediumresolutionDynamicHabitatInPress]. We found a weak
+relationship between in forest ecosystem structure and function overall
+across BC (15.7% of variation explained in the DHIs by forest structural
+attributes), as well as within individual BEC zones and forest types
+found in BC (@fig-fracts-bar).
 
-Across most of British Columbia's ecosystems, we identified a single RDA
-axis associated with the DHIs, encompassing the variation in annual
-productivity (@fig-rda-var A). Within this first axis, the strongest
-loadings were canopy cover and the modelled structure variables. In
-Alpine ecosystems the relationship between the first axis and the
-Minimum DHI was lowest, likely due to the low amount of variation in the
-Minimum DHI in these ecosystems. When a second axis was significant, it
-consistently had strong loadings on the minimum and variation DHIs,
-which indicates a complex internal productivity relationship in certain
-ecosystems. This secondary axis has smaller axis loadings associated
-with the primary and modelled structure variables, with the strongest
-loadings being canopy cover and structural complexity across the entire
-dataset (@fig-rda-var A).
+Ecosystem structure and function are key EBV classes which are capable
+of being measured using satellite remote sensing at global extents and
+with large temporal depth [@pereira2013; @skidmore2021]. Both EBV
+classes have been related to biodiversity metrics [@radeloff2019;
+@bergen2009]. One of the key attributes of the EBV framework is the
+complementarity of the proposed datasets, therefore it is necessary to
+determine if potential EBVs comprise unique or duplicate information
+when compared with one another [@skidmore2021]. This can be especially
+important when both datasets being examined were primarily derived using
+data from the same series of satellites, in our case, the Landsat series
+[@matasci2018; @matasci2018b;
+@razenkovaMediumresolutionDynamicHabitatInPress]. Our findings show weak
+relationships between these two classes, indicating that these datasets
+are very suitable to be used in a fashion complementary to one another,
+especially when examining ecosystem integrity, even with the same data
+provenance [@hansen2021].
 
-Our second question sought to explore whether modelled structure
-variables (basal area, gross stem volume, aboveground biomass) add
-additional explanatory information when predicting the DHIs, as compared
-to primary forest structure variables. We generally found that canopy
-cover had the largest axis loadings in the first RDA axis (@fig-radar).
-Modelled structure variables such as basal area, aboveground biomass,
-and gross stem volume shared similar loading magnitudes across the range
-of studied ecosystems, indicating they do not add additional value when
-utilized with one another (@fig-radar C; @fig-fcc B). The loadings
-between the modelled structure variables and canopy cover are often
-similar, and as such, in this context we recommend utilizing the
-structure variables derived directly from the point cloud in the case of
-ALS data, or selecting a single modelled attribute, such as aboveground
-biomass alone [@duncanson2022].
+While we assess the complementarity of ecosystem structure and function,
+a single pair of EBVs easily measurable at similar scales, other EBV
+class combinations should also be assessed for complementarity. This is
+especially relevant in cases such as ours, where remote sensing datasets
+can proliferate to a number of EBVs in different classes
+[@skidmore2021]. Other EBVs, such as those found in the community
+composition and species population classes, will be inherently linked
+due to being based on the same data. In addition, @skidmore2021
+highlighted that many satellite-derived biodiversity products can fall
+into multiple EBV classes simultaneously -- e.g., leaf area index being
+in the ecosystem structure, ecosystem function, and species trait EBV
+classes -- depending on the spatial scale of analysis and goal of the
+monitoring system. As such, it is relevant to consider which classes
+each biodiversity product should belong to, for given extents and
+temporal depths.
 
-During the analysis we found a higher amount of total variance explained
-in deciduous forests (mixed-wood and broadleaf) when compared to the
-other two forest types and most BEC zones (@fig-fracts-bar). This could
-potentially be due to the temporal linkages between canopy cover in
-deciduous forests and productivity. In evergreen coniferous forests
-where canopy cover is present year-round, there is likely a temporal
-mismatch between forest structure variables and the DHIs. Further, the
-strongest loadings in these two forest types was vertical structural
-complexity, rather than canopy cover. This indicates that the DHIs are
-more closely related to forest structure in deciduous forests, however
-even there significant decoupling remains.
+Further, we assessed the utility of primary forest structural attributes
+when compared to modelled forest structural attributes to examine
+duplicated information within a single EBV class. We found that standard
+lidar-derived measurements, such as canopy height, canopy cover, and
+structural complexity (as proposed by Valbuena et al., 2020) comprise
+the majority of the explained variation in the DHIs, either
+independently or in their overlap with the modelled attributes
+(@fig-fracts-bar). This could be expected as modelled forest structural
+attributes are generally directly based on the primary forest structural
+attributes [@coops2021], however, the creation of modelled forest
+structural attributes involves modelling expertise and processing time
+overhead [@matasci2018; @matasci2018b; @duncanson2022]. One advantage of
+these interpreted products is potential uptake by a broader set of
+users, and their ability to be more closely linked to management actions
+and monitoring results, as is the case with forest biomass for climate
+change mitigation [@duncanson2022].
 
-Finally, we explored the amount of variation explained by primary vs
-modelled structure variables, as well as their overlap
-(@fig-fracts-bar). We generally found that the overlap between the
-primary and modelled structure variables explained most of the
-variation, with some exceptions, indicating that using either set of
-forest structure variables may be suitable when monitoring biodiversity.
-The exceptions are especially prominent in the broadleaf and mixed wood
-forest classes, as well as cool, highly productive forests where there
-is no variation explained by the overlap between the two datasets. The
-majority of the variance explained in the broadleaf and mixed wood
-forests was in the forest structure variables derived directly from the
-point cloud. This was expected, as the modelled structure variables are
-calculated based on the derived structure variables as well as site
-index and ancillary information [@coops2021].
+In the axis loadings we found a strong linkage between canopy cover,
+structural complexity, and the DHIs, which was expected as these metrics
+are more directly related to photosynthesis than canopy height
+[@myneni1994]. We also found that axis loadings for modelled forest
+structural attributes were more similar to canopy cover and structural
+complexity than canopy height (@fig-rda-var). This was unexpected as
+while the modelled forest structural attributes are generally correlated
+due to their creation methods, these modelled attributes are primarily
+generated using canopy height information [@duncanson2022], with
+relationships often based on tree allometry rather than leaf volume or
+placement [@chave2005; @lutz2018].
 
-Recent advances in creating synthetic yearly observations have allowed
-the DHIs to be generated at a finer scale (30 m; Razenkova et al., In
-Press), rather than the previously used 1 km DHIs derived from MODIS
-[@radeloff2019; @razenkova2022]. This represents a significant
-advancement when assessing the utility of EBVs, as the 30 m scale is
-well suited to examine a range of ecological applications, including
-forest structure and productivity, and more closely approximates the
-scale at which species perceive habitat [@cohen2004; @kennedy2014]. In
-addition to more closely matching the scale of ecological phenomenon,
-this allows these datasets to be matched and analyzed with other
-datasets generated from the Landsat archive.
+Forest types with deciduous trees (namely broadleaf and mixed wood) had
+large, positive structural complexity loadings when compared to the
+other forest structural attributes (@fig-radar), with mixed wood forests
+having no modelled structural attributes associated with the DHIs. This
+structural complexity-DHIs linkage may be due to understory cover
+providing additional photosynthesis in these forest types
+[@nilsson2005]. Coniferous forests DHIs, on the other hand, show a
+weaker relationship with structural complexity than all other forest
+structural attributes. Notably, the Alpine, Montane, and Northern BEC
+groups show decoupling of the three DHIs (@fig-radar), with lower
+loadings between the Minimum DHI and RDA1. This decoupling in the DHIs
+is potentially due to the the near certainty of a pixel containing snow
+cover over the course of a year leading to consistently low minimum DHI
+values across the ecosystem and thus to weaker minimum DHI loadings
+[@razenkova2020; @beck2006]. These ecosystems with DHI decoupling also
+showed relatively strong relationships to canopy cover and weak
+relationships with structural complexity, potentially due to the
+dominance of coniferous forests in these ecosystems.
 
-In conclusion, we used redundancy analysis and variation partitioning to
-assess the complementarity of two potential EBV datasets - forest
-structure and the DHIs. We also separated the forest structure datasets
-into primary and modelled structure variables in order to assess the
-need to develop more complex structure variables, or if data derived
-directly from lidar datasets was suitable. We found that the structure
-variables are not strongly related to the DHIs, indicating that they are
-suitable to be used together as ecosystem scale EBVs when monitoring
-forest environments. We also found that variation explained by the
-overlap between primary and modelled structure variables was often
-higher than the variation explained by either individually.
+BC shows varying forest structural drivers of the DHIs depending on the
+geographic location [@fig-fcc]. The interior of BC is drier and warmer,
+and also has productivity being strongly driven by canopy cover, while
+the Coastal zones are the only BEC zone with strong structural
+complexity loadings. The climate of these regions could be influencing
+the DHIs [@perez2016], leading to varying loadings strengths depending
+on the distance to the coastline. Further, the extreme environmental and
+topographic gradients within BC's ecosystems [@pojar1987;
+@valentine1978soil] may be leading to highly variable data within the
+BEC zones we used for this study, and a finer ecoregionalization may
+reveal a different relationship between ecosystem structure and
+function.
+
+Across British Columbia, the DHIs have been shown to be strongly
+correlated with one another (@fig-rda-var). While this does imply that
+utilizing a single one of the three DHI axis may be suitable within this
+geographic region, the the DHIs do provide unique information to one
+another, which can be especially relevant when stratifying the study
+area. High levels of minimum DHI, for example, may indicate winter
+forage availability, which is difficult to capture in the variation or
+cumulative DHIs [@razenkova2020]. The variation DHI has been shown to be
+strongly related to seasonality, a key predictor of avian species
+richness, especially in migratory species [@rowhani2008]. In Coastal
+ecosystems, which are characterized by mild winters and high
+precipitation, we found that forest structure and the DHIs are strongly
+decoupled, with very little variance explained in the DHIs by the forest
+structural attributes (@fig-fracts-bar). This indicates that the
+complementarity of the DHIs and forest structural attributes should
+remain, even when expanding the analysis outside of ecosystems with
+little to no snow cover. Future research should continue to examine the
+complementarity of various EBV pairs, across varying ecosystems and
+geographic locations.
+
+# Conclusion
+
+We used redundancy analysis and variation partitioning to assess the
+complementarity of two potential EBV datasets - forest structure and the
+DHIs. We also separated the forest structure datasets into primary and
+modelled structure variables in order to assess the need to develop more
+complex structure variables, or if data derived directly from lidar
+datasets are suitable. We found that the structure variables were not
+strongly related to the DHIs, indicating that they are suitable to be
+used together as complementary ecosystem-scale EBVs when monitoring
+forest environments. It was also found that variation in the DHIs
+explained by the overlap between primary and modelled structure
+variables was often higher than the variation explained by either
+individually, both in the BC-wide dataset and the stratified datasets.
+These ecosystem structure and function variables are often attainable at
+global scales using satellite remote sensing, and recent advances are
+allowing them both to be generated at medium spatial resolutions. We
+suggest that biodiversity researchers focus on using forest structural
+attributes derived directly from the lidar data, and if needed, use a
+single modelled forest structural attribute, such as total aboveground
+biomass. In addition, we highlight that the usage of intra-annual
+summarizations of productivity provide novel information for
+biodiversity monitoring when used with forest structural attributes, and
+utilizing them as complementary EBVs for a holistic monitoring system
+for forest ecological integrity.
+
+# Acknowledgements
+
+This research was funded by NSERC support of Coops (RGPIN-2018-03851).
+Remote sensing data products utilized in this research are free and open
+and available for download at <https://ca.nfis.org/maps_eng.html>. We
+thank Dr. Michael Wulder and Dr. Joanne White for development and early
+access to these National Terrestrial Ecosystem Mapping System (NTEMS)
+products. We thank Dr. Elena Razenkova for early access to the
+Landsat-derived Dynamic Habitat Indices.
 

### 2023-11-08 scripts/0_whole_game.qmd
final comments

@@
 
+bcb <- bcmaps::bc_bound_hres() %>%
+    vect()
+
 if (!file.exists(vlce_loc)) {
   vlce <- rast("F:/mosaiced/VLCE2.0/LC_Class_HMM_2015_v20_v20.dat")
-  bcb <- bcmaps::bc_bound() %>%
-    vect()
 
   bcb_rast <- bcb %>%
-    rasterize(y = vlce) %>%
-    crop(bcb)
+    rasterize(y = vlce)
 
@@
 vlce <- rast(vlce_loc)
-bcb <- bcmaps::bc_bound() %>%
-  vect()
 

### 2023-11-08 scripts/4_vegan.qmd
final comments

@@
 library(patchwork)
-library(ggradar)
+# library(ggradar)
 source(here::here("scripts", "get_keys.R"))
@@
 
+```{r}
+# linear relationships * homoschedasticity check
+
+bec_normal %>%
+  pivot_longer(cols = ends_with("DHI"),
+               names_to = "dhi",
+               values_to = "dhi_val") %>%
+  pivot_longer(
+    cols = basal_area:total_biomass,
+    names_to = "struct",
+    values_to = "struct_val"
+  ) %>%
+  ggplot(aes(x = struct_val, y = dhi_val)) +
+  geom_hex() +
+  facet_grid(rows = vars(dhi),
+             cols = vars(struct),
+             scales = "free") +
+  labs(x = "Structure Value (Normalized)",
+       y = "DHI Value (Normalized)")
+
+strata_nested_long <- strata_nested %>%
+  unnest(cols = data) %>%
+  pivot_longer(cols = ends_with("DHI"),
+               names_to = "dhi",
+               values_to = "dhi_val") %>%
+  pivot_longer(
+    cols = basal_area:total_biomass,
+    names_to = "struct",
+    values_to = "struct_val"
+  ) %>%
+  group_by(strata) %>%
+  group_split()
+
+test <- function(df) {
+  strata <- df %>%
+    pull(strata) %>%
+    unique()
+
+  plot <- df %>% ggplot(aes(x = struct_val, y = dhi_val)) +
+    geom_hex() +
+    facet_grid(rows = vars(dhi),
+               cols = vars(struct),
+               scales = "free") +
+    labs(x = "Structure Value (Normalized)",
+         y = "DHI Value (Normalized)")
+
+  ggsave(here::here("outputs", "linear", glue::glue("{strata}.png")))
+}
+
+map(strata_nested_long, test)
+```
+
 ```{r rda-var-funs}
@@
   ggplot(aes(x = value, y = fct_reorder(zone_nm, total), fill = str_to_title(name))) +
-  geom_col(position = "stack") +
+  geom_col(position = "stack", col = "white") +
   facet_grid(rows = vars(fct_reorder(group, total, .desc = T)),
@@
   guides(fill = guide_legend(reverse = T)) +
-  scale_x_continuous(labels = scales::label_percent())
+  scale_x_continuous(labels = scales::label_percent()) +
+  scale_fill_manual(values = c("#6666ff", "#c22966", "#ff6666"))
 
@@
 
-##### run to here for analysis to be done
+# run to here for analysis to be done
 
@@
 ```
+
+# parallel coordinate pos neg
+
+```{r}
+neg_pos <- clean_loads_raw %>%
+  left_join(keys$bec %>%
+              select(strata = zone, group)) %>%
+  left_join(keys$continuous %>%
+              select(var = variable, var_long)) %>%
+  mutate(group = ifelse(is.na(group), strata, group),
+         var_long = ifelse(is.na(var_long), var, var_long)) %>%
+  group_by(group, var_long, axis) %>%
+  summarize(n_pos = sum(ifelse(axis_loading > 0, 1, 0)),
+            n_neg = sum(ifelse(axis_loading < 0, 1, 0)),
+            axis_loading = mean(axis_loading)
+            ) %>%
+  ungroup() %>%
+  complete(group, var_long, axis, fill = list(axis_loading = NA,
+                                              n_pos = NA,
+                                              n_neg = NA)) %>%
+  mutate(pos_neg = case_when(n_pos > 0 & n_neg > 0 ~ "Mixed",
+                             n_pos > 0 ~ "Positive",
+                             n_neg > 0 ~ "Negative")) %>%
+  select(-n_pos, -n_neg, -axis_loading)
+
+data <- clean_loads_abs %>%
+  left_join(keys$bec %>%
+              select(strata = zone, group)) %>%
+  left_join(keys$continuous %>%
+              select(var = variable, var_long)) %>%
+  mutate(group = ifelse(is.na(group), strata, group),
+         var_long = ifelse(is.na(var_long), var, var_long)) %>%
+  group_by(group, var_long, axis) %>%
+  summarize(axis_loading = mean(axis_loading)) %>%
+  ungroup() %>%
+  complete(group, var_long, axis, fill = list(axis_loading = NA)) %>%
+  left_join(neg_pos) %>%
+  mutate(inout = ifelse(endsWith(var_long, "DHI"), "Response", "Predictor"), 
+         group = fct_relevel(group, "British Columbia", 
+                             "Broadleaf", 
+                             "Coniferous", 
+                             "Mixed Wood", 
+                             "Wetland-Treed", 
+                             after = 0),
+         var_long = fct_relevel(var_long, 
+                                c("Canopy Height", "Canopy Cover", "Structural Complexity",
+                                "Basal Area", "Aboveground Biomass", "Gross Stem Volume",
+                                "Cumulative DHI", "Variation DHI", "Minimum DHI")))
+
+# nojitter/lines
+parcoord_noflip <- data %>%
+  ggplot(aes(x = var_long, y = axis_loading)) +
+  #geom_boxplot(col = "#00000025", outlier.alpha = 0) +
+
+  geom_line(aes(col = group, group = group), lty = "dashed") +
+  geom_point(data = data %>% 
+               filter(pos_neg == "Mixed"), aes(group = group, shape = group), size = 2.5, col = "black", fill = "black") +
+  geom_point(data = data %>% 
+               filter(pos_neg == "Positive"), aes(group = group, shape = group), size = 2.5, col = "red", fill = "red") +
+  geom_point(data = data %>% 
+               filter(pos_neg == "Negative"), aes(group = group, shape = group), size = 2.5, col = "grey", fill = "grey") +
+  geom_point(aes(col = group, group = group, shape = group, fill = group)) +
+  # geom_hline(yintercept = 0, lty = "dashed", col = "red") +
+  facet_wrap(~paste(axis, inout),
+             scale = "free") +
+  #theme_bw() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
+  labs(x = NULL,
+       y = "Axis Loading (absolute)",
+       col = NULL,
+       shape = NULL,
+       fill = NULL) +
+  # coord_flip() +
+  scale_colour_discrete_c4a_cat(palette = "bold") +
+  scale_fill_discrete_c4a_cat(palette = "bold") +
+  scale_shape_manual(values = c(21, 22, 22, 22, 22, 25, 25, 25, 25, 25))
+
+parcoord_noflip
+
+ggsave(here::here("outputs", "parcoord_test.png"), plot = parcoord_noflip,
+       height = 6, width = 8)
+
+ggsave(here::here("outputs", "parcoord_noflip.png"), plot = parcoord_noflip,
+       height = 6, width = 8)
+
+ggsave(here::here("outputs", "parcoord_noflip.svg"), plot = parcoord_noflip,
+       height = 6, width = 8)
+
+```
+
+```{r}
+data %>%
+  ggplot(aes(x = var_long, y = axis_loading)) +
+  geom_point(aes(shape = group, col = pos_neg, fill = group)) +
+  geom_line(aes(col = group)) +
+  facet_wrap(~paste(axis, inout), scales = "free") +
+  scale_shape_manual(values = c(21, 22, 22, 22, 22, 25, 25, 25, 25, 25)) +
+  scale_fill_discrete_c4a_cat(palette = "bold") +
+  scale_colour_manual(values = c("black", "grey", "red"))
+
+ggsave(here::here("outputs", "parcoord_test.png"), height = 6, width = 8)
+```
+
+```{r}
+clean_loads_raw %>%
+  mutate(pos_neg = ifelse(axis_loading > 0, "Positive", "Negative")) %>%
+  count(var, axis, pos_neg) %>%
+  ggplot(aes(y = var, x = n, fill = pos_neg)) +
+  geom_col() +
+  facet_wrap( ~ axis)
+
+clean_loads_raw %>%
+  left_join(keys$bec %>%
+              select(strata = zone, group)) %>%
+  mutate(pos_neg = ifelse(axis_loading > 0, "Positive", "Negative")) %>%
+  count(var, group, axis, pos_neg) %>%
+  mutate(n = ifelse(pos_neg == "Positive", n, -n)) %>%
+  ggplot(aes(y = var, x = n, fill = pos_neg)) +
+  geom_col() +
+  facet_grid(cols = vars(axis),
+             rows = vars(group))
+```
+
+```{r}
+groups <- keys$bec %>%
+  pull(group) %>% unique()
+
+lollipop_data <- clean_loads_raw %>%
+  left_join(keys$bec %>%
+              select(strata = zone, group)) %>%
+  left_join(keys$continuous %>%
+              select(var = variable, var_long)) %>%
+  mutate(
+    group = ifelse(is.na(group), strata, group),
+    var_long = ifelse(is.na(var_long), var, var_long),
+    pos_neg = ifelse(axis_loading > 0, "Positive", "Negative")
+  ) %>%
+  group_by(var_long, axis, group, pos_neg) %>%
+  summarize(n = n(),
+            avg_loading = mean(axis_loading)) %>%
+  ungroup() %>%
+  mutate(
+    inout = ifelse(endsWith(var_long, "DHI"), "Response", "Predictor"),
+    group = fct_relevel(
+      group,
+      "British Columbia",
+      "Broadleaf",
+      "Coniferous",
+      "Mixed Wood",
+      "Wetland-Treed",
+      after = 0
+    ),
+    var_long = fct_relevel(
+      var_long,
+      c(
+        "Canopy Height",
+        "Canopy Cover",
+        "Structural Complexity",
+        "Basal Area",
+        "Aboveground Biomass",
+        "Gross Stem Volume",
+        "Cumulative DHI",
+        "Variation DHI",
+        "Minimum DHI"
+      )
+    )
+  )
+
+lollipops <- lollipop_data %>%
+  group_by(group %in% groups) %>%
+  group_split()
+
+left <- lollipops[[1]] %>%
+  ggplot(aes(x = avg_loading,
+             y = var_long,
+             col = pos_neg)) +
+  geom_point() +
+  geom_segment(aes(
+    x = avg_loading,
+    xend = 0,
+    y = var_long,
+    yend = var_long
+  )) +
+  geom_vline(xintercept = 0,
+             alpha = 0.5,
+             lty = "dotted") +
+  facet_grid(cols = vars(axis),
+             rows = vars(group)) +
+  theme(legend.position = "bottom") +
+  labs(x = "Average Loading",
+       y = NULL)
+
+right = lollipops[[2]] %>%
+  ggplot(aes(x = avg_loading,
+             y = var_long,
+             col = pos_neg)) +
+  geom_point(aes(size = n)) +
+  geom_segment(aes(
+    x = avg_loading,
+    xend = 0,
+    y = var_long,
+    yend = var_long
+  )) +
+  geom_vline(xintercept = 0,
+             alpha = 0.5,
+             lty = "dotted") +
+  facet_grid(cols = vars(axis),
+             rows = vars(group)) +
+  theme(legend.position = "bottom",
+        axis.text.y = element_blank()) +
+  labs(x = "Average Loading",
+       y = NULL)
+
+left + right
+
+ggsave(here::here("outputs", "chris.png"))
+```
+
+```{r}
+lollipop_data %>%
+  ggplot(aes(x = var_long, y = avg_loading)) +
+  #geom_boxplot(col = "#00000025", outlier.alpha = 0) +
+  geom_line(data = lollipop_data %>%
+              filter(pos_neg == "Positive"), aes(col = group, group = group), lty = "dashed") +
+  geom_line(data = lollipop_data %>%
+              filter(pos_neg == "Negative"), aes(col = group, group = group), lty = "dashed") +
+  geom_point(aes(col = group, group = group, shape = group, fill = group)) +
+  # geom_hline(yintercept = 0, lty = "dashed", col = "red") +
+  facet_wrap(~paste(axis, inout),
+             scale = "free") +
+  #theme_bw() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
+  labs(x = NULL,
+       y = "Axis Loading (absolute)",
+       col = NULL,
+       shape = NULL,
+       fill = NULL) +
+  # coord_flip() +
+  scale_colour_discrete_c4a_cat(palette = "bold") +
+  scale_fill_discrete_c4a_cat(palette = "bold") +
+  scale_shape_manual(values = c(21, 22, 22, 22, 22, 25, 25, 25, 25, 25))
+```
+
+# flip cum to positive
+
+```{r}
+zones <- keys$bec %>% pull(zone)
+
+flip_cum <- function(data) {
+  if(data %>% filter(var == "CumDHI") %>% nrow() == 0) {
+    return(data)
+  }
+  flip <- data %>%
+    filter(var == "CumDHI") %>%
+    pull(axis_loading) %>%
+    sign()
+
+  data %>% mutate(axis_loading = axis_loading * flip)
+}
+
+clean_loads_cum_pos <- clean_loads_raw %>%
+  group_by(strata, axis) %>%
+  group_split() %>%
+  map_dfr(., .f = flip_cum)
+
+clean_loads_cum_pos %>%
+  filter(strata %in% zones) %>%
+  count(var, axis, sign = sign(axis_loading)) %>%
+  pivot_wider(names_from = sign, values_from = n) %>%
+  arrange(axis)
+
+```
+
+```{r}
+groups <- keys$bec %>% pull(group) %>% unique()
+
+clean_loads_joined <- clean_loads_cum_pos %>%
+  left_join(keys$bec %>%
+              select(strata = zone, group)) %>%
+  left_join(keys$continuous %>%
+              select(var = variable, var_long)) %>%
+  mutate(
+    group = ifelse(is.na(group), strata, group),
+    var_long = ifelse(is.na(var_long), var, var_long),
+    pos_neg = ifelse(axis_loading > 0, "Positive", "Negative")
+  )
+
+lollipop_data <- clean_loads_joined %>%
+  group_by(var_long, axis, group, pos_neg) %>%
+  summarize(n = n(),
+            avg_loading = mean(axis_loading)) %>%
+  ungroup() %>%
+  mutate(
+    inout = ifelse(endsWith(var_long, "DHI"), "Response", "Predictor"),
+    group = fct_relevel(
+      group,
+      "British Columbia",
+      "Broadleaf",
+      "Coniferous",
+      "Mixed Wood",
+      "Wetland-Treed",
+      after = 0
+    ),
+    var_long = fct_relevel(
+      var_long,
+      c(
+        "Canopy Height",
+        "Canopy Cover",
+        "Structural Complexity",
+        "Basal Area",
+        "Aboveground Biomass",
+        "Gross Stem Volume",
+        "Minimum DHI",
+        "Variation DHI",
+        "Cumulative DHI"
+      )
+    )
+  )
+
+lollipops <- lollipop_data %>%
+  group_by(group %in% groups) %>%
+  group_split()
+
+left <- lollipops[[1]] %>%
+  ggplot(aes(x = avg_loading,
+             y = var_long,
+             col = pos_neg)) +
+  geom_point() +
+  geom_segment(aes(
+    x = avg_loading,
+    xend = 0,
+    y = var_long,
+    yend = var_long
+  )) +
+  geom_vline(xintercept = 0,
+             alpha = 0.5,
+             lty = "dotted") +
+  facet_grid(cols = vars(axis),
+             rows = vars(group)) +
+  theme(legend.position = "bottom") +
+  labs(x = "Average Loading",
+       y = NULL)
+
+right = lollipops[[2]] %>%
+  ggplot(aes(x = avg_loading,
+             y = var_long,
+             col = pos_neg)) +
+  geom_point(aes(size = n)) +
+  geom_segment(aes(
+    x = avg_loading,
+    xend = 0,
+    y = var_long,
+    yend = var_long
+  )) +
+  geom_vline(xintercept = 0,
+             alpha = 0.5,
+             lty = "dotted") +
+  facet_grid(cols = vars(axis),
+             rows = vars(group)) +
+  theme(legend.position = "bottom",
+        axis.text.y = element_blank()) +
+  labs(x = "Average Loading",
+       y = NULL)
+
+left + right
+
+ggsave(here::here("outputs", "chris.png"))
+```
+
+```{r}
+data <- clean_loads_joined %>%
+  # mutate(axis_loading = abs(axis_loading)) %>%
+  group_by(group, var_long, axis) %>%
+  summarize(axis_loading = mean(axis_loading)) %>%
+  ungroup() %>%
+  complete(group, var_long, axis, fill = list(axis_loading = NA)) %>%
+  mutate(inout = ifelse(endsWith(var_long, "DHI"), "Response", "Predictor"), 
+         group = fct_relevel(group, "British Columbia", 
+                             "Broadleaf", 
+                             "Coniferous", 
+                             "Mixed Wood", 
+                             "Wetland-Treed", 
+                             after = 0),
+         var_long = ifelse(endsWith(var_long, "DHI"), var_long, str_replace_all(var_long, " ", "\n")),
+         var_long = fct_relevel(var_long, 
+                                c("Structural\nComplexity", "Canopy\nHeight", "Canopy\nCover",
+                                "Basal\nArea", "Aboveground\nBiomass", "Gross\nStem\nVolume",
+                                "Cumulative DHI", "Variation DHI", "Minimum DHI")))
+
+# nojitter/lines
+parcoord_noflip <- data %>%
+  ggplot(aes(x = var_long, y = axis_loading)) +
+  #geom_boxplot(col = "#00000025", outlier.alpha = 0) +
+
+  geom_line(aes(col = group, group = group), lty = "dashed") +
+  geom_point(aes(col = group, group = group, shape = group, fill = group)) +
+  geom_hline(yintercept = 0, lty = "dotted", col = "grey", linewidth = 0.5) +
+  facet_wrap(~paste(axis, inout),
+             scale = "free") +
+  #theme_bw() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
+  labs(x = NULL,
+       y = "Axis Loading",
+       col = NULL,
+       shape = NULL,
+       fill = NULL) +
+  # coord_flip() +
+  scale_colour_discrete_c4a_cat(palette = "bold") +
+  scale_fill_discrete_c4a_cat(palette = "bold") +
+  scale_shape_manual(values = c(21, 22, 22, 22, 22, 25, 25, 25, 25, 25))
+
+parcoord_noflip
+
+ggsave(here::here("outputs", "parcoord_noflip.png"), plot = parcoord_noflip,
+       height = 6, width = 8)
+
+# ggsave(here::here("outputs", "parcoord_noflip.svg"), plot = parcoord_noflip,
+#        height = 6, width = 8)
+```
+
+# meg scatterplots
+```{r}
+scatter <- bec_normal %>%
+  mutate(label = case_when(bec %in% c("CWH", "MS") ~ bec,
+                           T ~ "Other"))
+
+ggplot(data = scatter, aes(col = label), alpha = 0.25) +
+  geom_point(aes(x = percentage_first_returns_above_2m, y = CumDHI)) +
+  geom_density(aes(x = percentage_first_returns_above_2m)) +
+  geom_density(aes(y = CumDHI)) +
+  scale_fill_manual(values = c("grey", "red", "blue")) +
+  scale_colour_manual(values = c("grey", "red", "blue"))
+
+ggplot() +
+  geom_point(data = bec_df, aes(x = percentage_first_returns_above_2m, y = CumDHI), alpha = 0.5) +
+  geom_point(data = bec_df %>%
+               filter(bec == "CWH"), aes(x = percentage_first_returns_above_2m, y = CumDHI), col = "orange") +
+  geom_point(data = bec_df %>%
+               filter(bec == "MS"), aes(x = percentage_first_returns_above_2m, y = CumDHI), col = "red") +
+  geom_histogram(data = bec_df, aes(x = percentage_first_returns_above_2m)) +
+  geom_histogram(data = bec_df, aes(y = CumDHI))
+```
+
+```{r}
+fracts_out
+
+strata_nested %>% 
+  unnest(cols = data) %>%
+  select(-bec, -category) %>%
+  left_join(keys$bec, by = c("strata" = "zone")) %>%
+  mutate(group = ifelse(is.na(group), "Forest Type", group)) %>%
+  ggplot() +
+  geom_boxplot(aes(x = fct_reorder(.f = strata, .x = MinDHI, .fun = mean), y = MinDHI, fill = group))
+```
+
+```{r}
+strata_nested %>% 
+  unnest(cols = data) %>%
+  select(-bec, -category) %>%
+  group_by(strata) %>%
+  summarize(across(basal_area:VarDHI, .fns = sd)) %>%
+  left_join(keys$bec, by = c("strata" = "zone")) %>%
+  mutate(group = ifelse(is.na(group), "Forest Type", group)) %>%
+  select(strata, group, basal_area:VarDHI) %>%
+  pivot_longer(-c(strata, group)) %>%
+  left_join(fracts_out) %>%
+  mutate(varexp = primary + modelled + overlap) %>%
+  ggplot(aes(x = value, y = varexp, col = group)) +
+  geom_point() +
+  facet_wrap(~name, strip.position = "bottom") +
+  labs(x = NULL,
+       y = "Variation Explained",
+       title = "x axis is sd of variable") +
+  theme(strip.placement = "outside",
+        strip.background = element_blank())
+```
+
+```{r}
+bec_df %>%
+  left_join(fracts_out %>%
+              mutate(all = (primary + modelled + overlap) * 100) %>%
+              select(strata, all), by = c("bec" = "strata")) %>%
+  mutate(facet = paste0(bec, " (", round(all, 2), "%)")) %>%
+  ggplot(aes(x = percentage_first_returns_above_2m / 100, y = CumDHI)) +
+  geom_hex(aes(fill = after_stat(log(count)))) +
+  labs(x = "Canopy Cover (%)") +
+  facet_wrap(~fct_reorder(facet, all)) +
+  theme(legend.position = "bottom")
+
+ggsave(here::here("outputs", "supplement_hex.png"),
+       height = 6, width = 8)
+```

### 2023-11-08 scripts/manuscript/manuscript.qmd
final comments

@@
 ---
-title: Disentangling linkages between satellite-derived forest structure and productivity essential biodiversity variables.
+title: Disentangling linkages between satellite-derived indicators of forest structure and productivity for ecosystem monitoring.
 author:
@@
           department: Environmental and Conservation Sciences and Harry Butler Institute
+          address: 90 South St
           city: Murdoch, WA, Australia
@@
           city: Vancouver, BC, Canada
-          postal-code: V6T 1Z4
   - name: Txomin Hermosilla
@@
 bibliography: [bibliography.bib, packages.bib]
-csl: remote-sensing-of-environment.csl
+csl: scientific-reports.csl
 format:
@@
 number-sections: true
+nocite: |
+  @matasci2018, @matasci2018b
 ---
@@
 
-Key to regular biodiversity monitoring is the definition and selection
-of variables capable of accurately monitoring biodiversity trends. One
-framework is the Essential Biodiversity Variables (EBV), which are
-standardized, comparable data that represents a minimum set of
-biological information required for monitoring biodiversity change at
-large spatial extents, with each class being complementary to one
-another -- i.e., capturing unique biological information. The
-interrelationships between forest ecosystem structure and function and
-the complementarity between these two classes of EBVs across a range of
-ecosystems encompassing significant environmental gradients are examined
-using redundancy analysis (RDA) and variation partitioning. We found
-that the primary and modelled forest structure variables are not
-strongly related to the Dynamic Habitat Indices (DHIs), with the
-structural information only explaining 16% of the variation in the DHIs,
-indicating that structure and function EBVs are suitable to be used
-together for monitoring forest ecosystems. We also found that forest
-types and ecosystems dominated by coniferous trees had less variation
-explained, potentially due to the temporal decoupling of canopy cover
-and the DHIs. We suggest that biodiversity researchers focus on using
-forest structural attributes derived directly from lidar data alongside
-the DHIs as complementary EBVs for monitoring forest ecological
-integrity.
+The Essential Biodiversity Variables (EBV) framework has been proposed
+as a monitoring system of standardized, comparable variables that
+represents a minimum set of biological information to monitor
+biodiversity change at large spatial extents. Six classes of EBVs
+(genetic composition, species populations, species traits, community
+composition, ecosystem structure and ecosystem function) are defined, a
+number of which are ideally suited to observation and monitoring by
+remote sensing systems. We used moderate-resolution remotely-sensed
+indicators representing two ecosystem level EBV classes (ecosystem
+structure and function) to assess their complementarity and redundancy
+across a range of ecosystems encompassing significant environmental
+gradients. Redundancy analyses found that remote sensing indicators of
+forest structure were not strongly related to indicators of ecosystem
+productivity (represented by the Dynamic Habitat Indices, DHIs), with
+the structural information only explaining 16% of the variation in the
+DHIs. Complex metrics of forest structure, such as above ground biomass,
+did not contribute additional information over simpler height-based
+attributes that can be directly estimated with light detection and
+ranging (LIDAR) observations. With respect to ecosystem conditions, we
+found that forest types and ecosystems dominated by coniferous trees had
+less redundancy between the remote sensing indicators when compared to
+broadleaf or mixed forest types. Likewise, higher productivity
+environments exhibited the least redundancy between indicators, in
+contrast to more environmentally stressed regions. We suggest that
+biodiversity researchers continue to exploit multiple dimensions of
+remote sensing data given the complementary information they provide on
+structure and function focused EBVs, which makes them jointly suitable
+for monitoring forest ecosystems.
 
@@
 Monitoring the changing state of biodiversity globally is key to
-understanding and mitigating the accelerated extinction risk for many
+understanding and mitigating the increased extinction risk for many
 species [@thomas2004; @urban2015] and the homogenization of biotic
@@
 Biodiversity Observation Network has developed the Essential
-Biodiversity Variables [EBVs, @pereira2013], which are standardized,
-comparable data that represents a minimum set of biological information
-required for monitoring biodiversity change at large spatial extents
-[@skidmore2021]. The EBVs are designed as an analog to the Essential
-Climate Variables framework [@bojinski2014] to be global in scope,
-relevant to biodiversity information, feasible to implement and use, and
-complementary to one another [@skidmore2021]. While it can be difficult,
-time consuming, and expensive to collect data on biodiversity across
-wide swaths of land, EBVs can be combined with sampled biodiversity
-information to allow for the spatial extrapolation of more sparsely
-collected in situ data [@cavender-bares2022]. The standardization and
-global availability of the EBVs allows for the assessment and monitoring
-of protected area effectiveness and ecosystem health at large spatial
-scales [@hansen2021], which can then be integrated into policy
-development [@geijzendorffer2016]. There are six EBV classes, each of
-which provide information on a different facet of biodiversity: species
-populations, species traits, community composition, ecosystem structure,
-ecosystem function, and genetic composition [@pereira2013]. While these
-EBV classes can be related to one another under a variety of ecological
-and evolutionary mechanisms, they provide distinct information; multiple
-in concert can provide a more complete picture on the status and trends
-of biodiversity in a region.
-
-The global coverage and moderate spatial resolutions of many optical
-satellite remote sensing missions [@wulder2022], combined with open-data
-policies [@wulder2012], make these datasets exceptionally well suited
-for contributing to some EBV classes [@skidmore2021]. The six EBV
-classes can be divided into two general approaches: species-focused and
-ecosystem-focused [@fernández2020]. Genetic composition, species
-populations, and species traits are species-focused EBVs, with community
-composition being directly related to species populations. While
-species-focused and community composition EBVs can be assessed from
-measurements at the tree-crown scale using very-high spatial resolution
-imagery [@fassnacht2016; @graves2016; @wu2021], the spatially
-limited--and often *ad-hoc*--data collection associated with monitoring
-individuals is not conducive to the global or regional scales often
-required for biodiversity trend assessments [@valdez2023]. In contrast,
-two ecosystem-based EBV classes (ecosystem structure and ecosystem
-function) are well suited to be monitored across large extents using
-moderate spatial resolution optical imagery, such as the Moderate
-Resolution Imaging Spectroradiometer [MODIS, @zhang2003], Landsat
-imaging systems [@wulder2022] or Sentinel-2 [@helfenstein2022;
-@drusch2012] programs, especially in forested environments
-[@pettorelli2018; @coops2016].
+Biodiversity Variables (EBVs) [@pereira2013], which are standardized,
+comparable variables that represents a minimum set of biological
+information required for monitoring biodiversity change at large spatial
+extents [@skidmore2021]. The EBVs are designed as an analog to the
+Essential Climate Variables framework [@bojinski2014] to be global in
+scope, relevant to biodiversity information, feasible to implement and
+use, and complementary to one another [@skidmore2021]. There are six
+classes of EBVs collected at various scales. The ecosystem EBV classes
+(ecosystem structure and ecosystem function) are well suited to be
+monitored across large areas using remote sensing technologies
+[@skidmore2021]. This has led to a proliferation of potential EBV
+datasets that have been generated to describe these two ecosystem EBV
+classes, especially in forested environments.
 
@@
 arrangement of different ecosystem components, and is indicative of
-habitat quality and connectivity [@noss1990; @valbuena2020]. There are
-both horizontal and vertical components of ecosystem structure. In
-forest environments, forest structural diversity has been linked to
-species richness at various spatial scales, ranging from individual
-plots to landscapes [@bergen2009; @gao2014]. Structural attributes range
-in complexity from simple (e.g., canopy cover, canopy height), to more
-complex (e.g., foliage height diversity, leaf area index) to modelled
-(e.g., aboveground biomass, basal area), which can be derived using
-lidar (light detection and ranging) data [@coops2021]. A suite of these,
-and other, structural attributes has been used as indicators of
-biodiversity at local to landscape scales [@lefsky1999; @guo2017;
-@coops2016; @bergen2009]. Increased forest structural complexity has
-been hypothesized to create additional niches, leading to increased
-species diversity [@bergen2009], which has been frequently demonstrated
-using avian species diversity metrics [@macarthur1961], with lidar
-[@clawges2008], a combination of lidar and satellite imagery
+habitat quality and connectivity [@noss1990]. Forest ecosystem structure
+is defined as the geographic distribution of trees or biomass, with
+forest structural diversity encompassing the measurable morphological
+components of the forests (e.g., canopy cover, canopy height, and
+structural complexity) [@atkins2023; @valbuena2020]. Forest structural
+diversity has been linked to species richness at various spatial scales,
+ranging from individual plots to landscapes [@bergen2009; @gao2014]. A
+variety of forest structural attributes can be extracted directly from
+lidar (light detection and ranging) data, ranging in complexity from
+simple (e.g., canopy cover, canopy height) to more complex (e.g.,
+foliage height diversity, leaf area index), while others can be
+indirectly modelled from lidar and auxiliary data (e.g., aboveground
+biomass, basal area) [@coops2021]. A suite of these, and other,
+structural attributes have been used as indicators of biodiversity at
+local to landscape scales [@lefsky1999; @guo2017; @coops2016;
+@bergen2009]. Increased forest structural complexity has been
+hypothesized to create additional niches, leading to increased species
+diversity [@bergen2009; @heidrich2020], which has been frequently
+demonstrated using avian species diversity metrics [@macarthur1961] and
+lidar [@clawges2008], or a combination of lidar and satellite imagery
 [@herniman2020], with some research indicating that canopy vertical
@@
 extension of three-dimensional forest structure attributes nationally in
-a wall-to-wall fashion. The wall-to-wall data often uses data fusion
-approaches involving lidar data and medium spatial resolution optical
-imagery [@matasci2018; @matasci2018b; @coops2021].
+a wall-to-wall fashion. The computation of these attributes often
+requires data fusion approaches involving lidar data and medium spatial
+resolution optical imagery [@coops2021; @matasci2018; @matasci2018b].
 
@@
 consequence of one or multiple ecosystem processes [@pettorelli2018;
-@ecosyste2005]. Within the EBV framework, vegetation phenology and
-metrics of primary productivity are commonly used as potential EBVs
-[@skidmore2021]. Energy availability is the amount of accessible and
-utilizable energy available for growth in an ecosystem, and is often
-measured as net primary productivity [@bonn2004]. Energy availability
-has shown to be a predictor of species richness and abundances at
-various scales [@chase2002; @radeloff2019; @coops2019; @razenkova2023],
-and is measurable using satellite remote sensing via various vegetation
-indices [@huete2002; @radeloff2019]. Vegetation indices are indicative
-of photosynthetic activity, and are commonly used as proxies of gross
-primary productivity [@huang2019]. These vegetation indices enable the
-assessment of biodiversity patterns at single time points [@bonn2004],
-or through time when using time series data [@berry2007; @radeloff2019].
+@ecosyste2005]. Within the EBV framework, vegetation phenology --- the
+timing and duration of leaf-on conditions --- and ecologically available
+energy --- the amount of accessible and useable energy available for
+growth in an ecosystem --- are useful predictors of species richness and
+abundances at various scales [@chase2002; @radeloff2019; @coops2019;
+@razenkova2023]. Ecologically available energy is the amount of energy
+produced in an ecosystem, which is sensitive to limitations such as
+recent disturbances and the water and nutrient supplies, and is
+measurable via productivity [@bonn2004; @nguyen2022]. This differs from
+ambient energy in that it goes beyond the climatic conditions of an
+area, and concerns the amount of resources available for consumption in
+an ecosystem [@hawkins2003]. Clear links have been developed between
+remote sensing-derived spectral reflectance, particularly in the
+near-infrared and the red parts of the electromagnetic spectrum
+[@rouse1974], which relate to pigment concentrations and leaf structure,
+respectively, as well as productivity [@yoder1994]. Consequently, the
+normalized difference vegetation index (NDVI) and other vegetation
+indices have been correlated with productivity measures [@huang2021;
+@sims2006]. The relationship however, is typically non-linear,
+principally due to saturation in the spectral response at moderate to
+high leaf area index values [@myneni1994]. As a result, the strength of
+the relationship between spectral indices and productivity depends on
+the ecosystem type, land cover type, amount of canopy cover, and
+productivity in the region [@shi2017]. Likewise, relationships between
+productivity and biodiversity also contain uncertainty, recognizing that
+biodiversity is governed by a variety of factors of which productivity
+is but one [@skidmore2021]. Despite these uncertainties, remotely sensed
+vegetation indices have enabled the assessment of biodiversity patterns
+at single time points [@bonn2004; @nagendra2010; @phillips2008], or
+through time when using time series data [@berry2007; @radeloff2019;
+@coops2019]
 
-The relationship between energy availability and biodiversity can be
-explained through various hypothesized mechanisms: including available
-energy, environmental stress, and environmental stability. These three
-hypotheses have been incorporated into the Dynamic Habitat Indices
-(DHIs), which summarizes intra-annual dynamics in remotely-sensed
-productivity metrics and have been evaluated as broad-scale biodiversity
-indicators [@berry2007; @coops2008; @radeloff2019]. According to the
-available energy hypothesis [@wright1983] an increase in available
-energy results in higher species richness. In the context of the DHIs,
-the total energy available within a specific area is quantified using
-the cumulative DHI, which sums each productivity observation throughout
-the course of a year. The environmental stress hypothesis [@currie2004]
-suggests that higher levels of minimum available energy contribute to
-greater species richness. The minimum DHI, derived from the lowest
-productivity observed over the course of a year, indicates the level of
-available energy during stressful periods. The environmental stability
-hypothesis [@williams2008] proposes that lower energy variation
-throughout the year leads to increased species richness. The variation
-DHI captures the coefficient of variation in a vegetation index over the
-course of a year, reflecting the stability of energy availability. The
-DHIs have previously been produced at a global extent using MODIS
-imagery and have been used to assess alpha [@radeloff2019] and beta
-[@andrew2012] diversity, species abundances [@razenkova2023], and
-construct novel ecoregionalizations [@coops2009; @andrew2013]. Recent
-studies have increased the spatial resolution of these indices by
-calculating the DHIs from multi-annual Landsat imagery
+While biodiversity is not solely driven by productivity, the
+relationship between productivity and biodiversity can be explained
+through various hypothesized mechanisms: including available energy,
+environmental stress, and environmental stability. These three
+hypotheses have been used to guide the development of a set of remote
+sensing indicators, the Dynamic Habitat Indices (DHIs), which summarize
+intra-annual dynamics in productivity metrics and have been evaluated as
+broad-scale biodiversity indicators globally for a variety of taxa
+[@berry2007; @coops2008; @radeloff2019]. According to the available
+energy hypothesis [@wright1983; @bonn2004] an increase in ecologically
+available energy results in higher species richness. In the context of
+the DHIs, the total energy available within a specific area is
+quantified using the cumulative DHI, which sums each productivity
+observation throughout the course of a year. The environmental stress
+hypothesis [@currie2004] suggests that higher levels of minimum
+available energy contribute to greater species richness. The minimum
+DHI, derived from the lowest productivity observed over the course of a
+year, indicates the level of available energy during stressful periods.
+The environmental stability hypothesis [@williams2008] proposes that
+lower energy variation throughout the year leads to increased species
+richness due to reduced resource bottlenecks. The variation DHI captures
+the coefficient of variation in a vegetation index over the course of a
+year, reflecting the prevalence of resource bottlenecks. The DHIs have
+previously been produced at continental to global extents using Moderate
+Resolution Imaging Spectroradiometer (MODIS) imagery, while recent
+research has focused on producing the DHIs at finer spatial resolutions
+using multi-annual time series of Landsat imagery
 [@razenkovaMediumresolutionDynamicHabitatInPress].
 
-Choosing datasets that are most closely related to the phenomenon of
-interest in a given application allows for direct integration of
-monitoring results with management actions [@pressey2021]. With the
-advent of large-extent monitoring methods like satellite remote sensing
-which have led to a proliferation of potential EBV datasets, it becomes
-important to assess the interrelationships between these datasets and
-the complementary of the information they provide to reduce redundancy
-in future EBV development efforts [@pereira2013; @skidmore2021].
-Identifying and exploring strong relationships between EBVs may help
+The DHIs have been shown to be correlated with aspects of biodiversity
+across a variety of spatial scales and locations, including Moose
+occurrence and abundance in Ontario, Canada [@WOS:000343392200016] and
+across Russia [@razenkova2020], beta diversity in butterflies in Canada
+[@andrew2012], avian diversity in the coterminous U.S.
+[@WOS:000659187700008], Ontario [@coops2009bird] and France
+[@bonthoux2018], and global alpha diversity [@radeloff2019]. The DHIs
+have also been used to construct novel ecoregionalizations [@andrew2013;
+@coops2009]. DHIs, however, are not able to capture all components of
+biodiversity due to the many environmental and landscape processes not
+observable from satellite data [@coops2009bird].
+
+The suite of remote sensing indicators used in biodiversity monitoring
+should be carefully considered. Ideally, indicators should be selected
+that are closely related to the phenomenon of interest to allow for
+direct integration of monitoring results with management actions
+[@pressey2021]. With the advent of large-extent monitoring methods like
+satellite remote sensing which have led to a proliferation of potential
+EBV datasets, it becomes important to assess the interrelationships
+between these datasets and the complementary of the information they
+provide to reduce redundancy in future EBV development efforts
+[@pereira2013; @skidmore2021]. Identifying and exploring the strong
+relationships that may exist between indicators of EBVs may help
 elucidate the ecological relationships between these facets of
 biodiversity, such as ecosystem structure and function, at local to
-global scales. On the other hand, when variables are not related, they
-may be well suited to be used in monitoring programs together, as
-complementary EBVs.
+global scales. On the other hand, when remote sensing-derived EBVs are
+not related, they may be well suited to be used in monitoring programs
+together, as complementary sets of observations.
 
-Linkages between forest ecosystem structure and function have been
-examined within a remote sensing context for decades [@huete2002,
-@knyazikhin1998; @myneni1994]. While there is significant theoretical
-and empirical evidence for their relationship [@myneni1994], the
-direction and significance has not been consistent, with positive,
-negative, and non-significant relationships being found [@ali2019].
-Hypothesized mechanisms such as niche complementarity suggest that
-productivity increases with stand structural complexity [@zhang2012],
-while asymmetric competition for light can reduce forest productivity
-with increased structural complexity [@bourdier2016]. Some aspects of
-forest structure, such as canopy cover and leaf area index, also
-directly influence vegetation indices, which are used to estimate
-productivity, especially in very dense or very sparse forests
-[@pettorelli2005]. Further, the relationship between forest structural
-diversity metrics--which are now more accurately and comprehensively
-derived from lidar data--and the temporal variation in functional
-metrics, specifically the metrics of ecosystem productivity via the DHIs
-framework, have yet to be fully examined.
+The relationship between ecosystem structure and function have been
+examined within a remote sensing context for decades [@huete2002;
+@knyazikhin1998; @myneni1994]. Hypothesized mechanisms such as niche
+complementarity suggest that productivity increases with stand
+structural complexity [@zhang2012], while asymmetric competition for
+light can reduce forest productivity with increased structural
+complexity [@bourdier2016]. Some aspects of forest structure, such as
+canopy cover and leaf area index, also directly influence vegetation
+indices, which are used to estimate productivity [@pettorelli2005].
+Often, these large-extent estimates of structure and function metrics
+are generated from similar data sources, such as the Landsat series of
+satellites [@razenkovaMediumresolutionDynamicHabitatInPress ;
+@radeloff2019; @matasci2018; @matasci2018b]. Due to this shared data
+provenance, and the linkages between biodiversity and both structure and
+function, it is important to assess how much unique information these
+structure and function metrics represent when used as indicators of
+biodiversity.
 
-The goal of this study is to examine relationships between forest
-ecosystem structure and function, and the complementarity between these
-two classes of EBVs across significant environmental gradients. We
-synthesize data from moderate-scale remote sensing-derived metrics of
-forest structure (simple lidar-extracted metrics of canopy height,
-canopy cover, and structural complexity), as well as modelled forest
-structure attributes (gross stem volume, basal area, and aboveground
-biomass), with a well established remote sensing-derived index for
-ecosystem function (DHIs). First, we analyze remote sensing metric
-complementarity by examining the multivariate relationship between
-ecosystem structure and function and propose a method to examine
-potential overlap in remote sensing EBVs. Second, we propose a method
-for the selection of remote sensing attributes for developing and
-applying EBVs within a single EBV using the independent and shared
-relationships of lidar-derived forest structural attributes and modelled
-forest structure on ecosystem function. Third, we examine the global
-applicability of these results using the relationship between simple and
-modelled forest structural attributes with DHIs across a range of
-environment gradients and forest types.
+The goal of this study is to assess the complementarity of two remote
+sensing datasets often proposed as indicators of ecosystem structure and
+function EBVs. We interpret weak or absent relationships between metrics
+to constitute complementary information. To do so, we synthesize
+wall-to-wall forest structure data previously generated by modelling
+lidar-derived forest structure metrics with medium spatial resolution
+satellite imagery using a k-NN imputation method [@matasci2018;
+@matasci2018b], alongside a well-established remote sensing-derived
+index of ecosystem function (DHIs) [@coops2008; @berry2007;
+@radeloff2019; @razenkovaMediumresolutionDynamicHabitatInPress]. First,
+we analyze remote sensing indicator complementarity by examining the
+multivariate relationship between ecosystem structure and function.
+Second, we propose a method for selecting which remote sensing
+attributes to develop and apply within a single EBV class by determining
+the independent and shared contributions of lidar-derived forest
+structural attributes and modelled forest structure to explaining
+ecosystem function. Third, we examine the global applicability of these
+results by evaluating the consistency of the observed forest
+structure-function relationships across a broad range of environment
+gradients and forest types.
 
@@
 [@valentine1978soil]. Wildfires are a regular disturbance in the
-province, with the historical fire regime being disrupted due to
-management actions, leading to larger, higher intensity fires becoming
-more common in recent years [@brookes2021]. In addition, forestry is a
-large industry in the province, with approximately 136000 ha of the
-province harvested accounting for \~54 million cubic metres of volume of
-wood in 2021 [@canadiancouncilofforestministers2023].
+province, although larger, higher intensity fires have become more
+common in recent years due to disruption of the historical fire regime
+[@brookes2021]. In addition, forestry is a large industry in the
+province, with approximately 136,000 ha of the province harvested
+accounting for \~54 million cubic metres of volume of wood in 2021
+[@canadiancouncilofforestministers2023].
 
@@
 (@fig-study-map). To examine trends across the large environmental
-gradients, we group the BEC zones into five broad biomes: the Nouthern
-interior, Sorthern interior, Montane, Alpine, and Coastal groups,
-similar to @hamann2006. We also report the average climate data for BEC
-zones from 1991-2020, according to @wang2016 (Table 1).
+gradients, we group the BEC zones into five broad biomes: the Northern
+interior, Southern interior, Montane, Alpine, and Coastal groups,
+similar to Hamann et al. [@hamann2006]. We also report the average
+climate data for BEC zones from 1991-2020, according to Wang et al.
+[@wang2016] (Table 1).
 
@@
     zone_nm = "Full Name",
-    ppt = "Precipitation (mm/yr)",
-    tmax = "Max Temperature (°C)",
-    tmin = "Min Temperature (°C)"
+    ppt = "Average Monthly Precipitation (mm/yr)",
+    tmax = "Average Monthly Max Temperature (°C)",
+    tmin = "Average Monthly Min Temperature (°C)"
   ) %>%
-  tab_header(title = "Biogeoclimatic Ecosystem Classification (BEC) Zones, their aggregated biomes, and their average climate values for precipitation, maximum temperature, and minimum temperature.
+  tab_header(title = "Biogeoclimatic Ecosystem Classification (BEC) Zones, their aggregated biomes, and their monthly average climate values for precipitation, maximum temperature, and minimum temperature.
 Climate data from Wang et al.
@@
 
-### National Terrestrial Ecosystem Monitoring System (NTEMS)
+### National Terrestrial Ecosystem Monitoring System data products
 
-#### Best-Available-Pixel
+#### Surface-reflectance image composites
 
@@
 based on a 30-m best-available-pixel (BAP) composite generated by
-@hermosilla2016 using the Composite2Change approach. The composites were
-developed for each year from 1984-2019 by selecting the best available
-imagery from the available growing season Landsat observations, removing
-pixels with clouds, cloud shadows, and haze. The methodology uses the
-scoring method from @white2014 to select the best pixel for a given year
-from Landsat-5 Thematic Mapper, Landsat-7 Enhanced Thematic Mapper Plus,
-and Landsat-8 Operational Land Imager imagery. To reduce noise in the
-BAP between years, a spectral trend analysis on the Normalized Burn
-Ratio was conducted on each pixel, which removes unscreened noise,
-detects changes, and temporally interpolates data gaps. This resulted in
-a gap-free surface-reflectance composite across the entirety of Canada
-[@hermosilla2015]. We utilize data from 2015 for all NTEMS data, except
-forest disturbances, where we use all available years of data in the
-terrestrial area of BC, Canada.
+Hermosilla et al. @hermosilla2016 using the Composite2Change approach.
+The composites were developed for each year from 1984-2019 by selecting
+the best available imagery from the available growing season Landsat
+observations, removing pixels with clouds, cloud shadows, and haze. The
+methodology uses the scoring method from White et al. @white2014 to
+select the best pixel for a given year from Landsat-5 Thematic Mapper,
+Landsat-7 Enhanced Thematic Mapper Plus, and Landsat-8 Operational Land
+Imager imagery. A spectral trend analysis was conducted over these
+initial BAP composites using the Normalized Burn Ratio (a vegetation
+index) on each pixel, to remove unscreened noise, detect changes, and
+temporally interpolate data gaps. This resulted in a gap-free
+surface-reflectance composite across Canada [@hermosilla2015]. We use
+data from 2015 for all National Terrestrial Ecosystem Monitoring System
+products, except forest disturbances, which comprise the period
+1985-2020.
 
@@
 
-We used a forest disturbance layer generated by @hermosilla2015a to
-restrict the analysis to stands older than 35 years as productivity and
-structure metrics have been shown to be strongly decoupled in early
-regenerative stands [@bolton2017], and also excluding recently
-anthropogenically disturbed stands. Changes were detected using spectral
-trend analysis and were attributed using an object-based analysis
-approach to a disturbance agent (i.e., wildfire, harvest) using a Random
-Forests classification model [@hermosilla2015a]. Changes were detected
-with an overall accuracy of 89%, attributed with an accuracy of 92%, and
-assigned to the correct year with an accuracy of 89% [@hermosilla2016].
+We used the forest disturbance layer generated by Hermosilla et al.
+@hermosilla2016 to restrict the analysis to stands older than 35 years
+as productivity and structure metrics have been shown to be strongly
+decoupled in early regenerative stands [@bolton2017], and to exclude
+recently anthropogenically disturbed stands. The disturbances detected
+using spectral trend analysis were attributed to a disturbance agent
+(i.e., wildfire, harvest) within an object-based analysis approach using
+a Random Forests classification model [@hermosilla2015a]. Changes were
+detected with an overall accuracy of 89%, attributed with an accuracy of
+92%, and assigned to the correct year with an accuracy of 89%
+[@hermosilla2016].
 
@@
 
-We used a land cover map generated by @hermosilla2022 to stratify our
-samples alongside the BEC zones. The 30-m spatial resolution land cover
-map for Canada's forested ecosystems was generated using the Virtual
-Land Cover Engine framework [@hermosilla2022]. This framework uses a
-Random Forests based classification approach, and integrates logical
-land cover transitions using a Hidden Markov Model to generate 12 land
-cover classes. Training samples were locally selected using an inverse
-distance weighting approach [@hermosilla2022]. The treed land cover
-classes are broadleaf, coniferous, mixed wood, and wetland-treed, which
-we use as a stratification. The land cover classification reached an
-overall accuracy of 77.9% ± 4% [@hermosilla2022].
+We used the land cover map for 2015 generated by Hermosilla et al.
+@hermosilla2022 to stratify our samples alongside the BEC zones. The
+30-m spatial resolution land cover map for Canada's forested ecosystems
+was generated using the Virtual Land Cover Engine framework
+[@hermosilla2018]. This framework uses a Random Forests based
+classification approach, and integrates logical land cover transitions
+using a Hidden Markov Model to generate 12 land cover classes. The land
+cover classification reached an overall accuracy of 77.9% ± 4%
+[@hermosilla2022]. We restricted our sample to the four treed land cover
+classes: broadleaf, coniferous, mixed wood, and wetland-treed.
 
@@
 
-Forest structural attributes (canopy height, canopy cover, structural
-complexity \[coefficient of variation of height returns\], basal area,
-aboveground biomass, and gross stem volume) for 2015 were extracted for
-BC from the 30-m spatial resolution dataset generated by Matasci et al.
-(2018a, 2018b). In brief, the method extends structural attributes
-derived from a set of lidar acquisitions and field plots across Canada.
-These data were generated for the treed pixels of Canada using the BAP
-Landsat surface reflectance composite and auxiliary data (i.e.,
-topography, geography) using a k-Nearest Neighbour imputation
+The forest structural attributes (canopy height, canopy cover,
+structural complexity \[coefficient of variation of height returns\],
+basal area, aboveground biomass, and gross stem volume) were derived
+from the 30-m spatial resolution dataset generated by Matasci et al.
+[@matasci2018; @matasci2018b]. In brief, the method extends structural
+attributes derived from a set of lidar acquisitions and field plots
+across Canada. These data were generated for the treed pixels of Canada
+using the BAP Landsat surface reflectance composite and auxiliary data
+(i.e., topography, geography) using a k-Nearest Neighbour imputation
 [@matasci2018; @matasci2018b]. Accuracy metrics for imputed forest
@@
 
-The DHIs were calculated following
+The DHIs were calculated following Razenkova et al.
 @razenkovaMediumresolutionDynamicHabitatInPress for the terrestrial area
@@
 values were then calculated to represent the cumulative, minimum, and
-variation DHIs, respectively. More detailed information can be found in
-@razenkovaMediumresolutionDynamicHabitatInPress.
+variation DHIs, respectively.
 
-A summary of remote sensing derived datasets can be found in Table 2.
+A summary of the datasets used in this study is presented in Table 2.
 
@@
 of BEC zones (Table 1) and forest types (i.e., coniferous, broadleaf,
-mixed wood, and wetland treed; see @sec-study-area) of BC. A maximum of
-500 sample units were selected from each stratum. Sample units were
-restricted to forested pixels that were surrounded by the same forest
-type to reduce the chance of sampling mixed pixels. To ensure that
-sample units were selected from a homogeneous area, reducing uncertainty
-associated with the input datasets [@shang2020], we required sample
-units to have a coefficient of variation for canopy cover and canopy
-height to be lower than 0.5 in a 3 x 3-pixel window. A minimum sampling
-distance of 1-km was implemented to reduce the effects of spatial
-autocorrelation, and sample units that had been disturbed in the last 30
-years were discarded by using the disturbance mask generated for the
-forested ecosystems of Canada as early regenerative forests have shown
-to have strongly decoupled productivity and canopy cover [@bolton2017].
-
-Sampling was conducted in R [@R-base] version 4.2.2 using the **sgsR**
-package [@R-sgsR]. Neighbourhood analyses for the land cover classes and
-coefficient of variations of canopy height and cover were calculated in
-Python version 3.9.
+mixed wood, and wetland treed; see @sec-study-area) of BC. Sample units
+were restricted to forested pixels that were surrounded by the same
+forest type to reduce the chance of sampling mixed pixels. We also
+required that sample units were selected from a homogeneous area, to
+reduce uncertainty associated with the input datasets [@shang2020];
+homogeneous areas were defined as 3x3-pixel windows with a coefficient
+of variation for canopy cover and canopy height below 0.5. A minimum
+sampling distance of 1-km was implemented to reduce the effects of
+spatial autocorrelation. Sample units that had been disturbed in the
+last 35 years were discarded by using the disturbance mask generated for
+the forested ecosystems of Canada [@hermosilla2016 ] . A maximum of 500
+sample units were selected from each stratum; in the many strata where
+this was not achieved, all sampling units meeting the above restrictions
+were selected. Sampling was conducted in R version 4.2.2 \[@R-base\]
+using the **sgsR** package [@R-sgsR]. Neighbourhood analyses for the
+land cover classes and coefficient of variations of canopy height and
+cover were calculated in Python version 3.9.
 
@@
 redundancy analysis, rather than other ordination techniques such as
-principle component analysis or constrained correspondence analysis
-[@legendre2012]. Variation partitioning is an extension of partial RDA
--- which allows predictor variables to be considered within pre-defined
-groups of co-variates -- that can assess the overlap between the
-explanatory power of two datasets by utilizing multiple partial RDAs and
-exchanging which datasets are considered the predictor, and which are
-considered the co-variates [@legendre2012].
+principal component analysis or constrained correspondence analysis
+[@legendre2012]. Partial RDA extends RDA by allowing groups of predictor
+variables to be considered as co-variates. Variation partitioning
+leverages multiple partial RDAs to assess the overlap between the
+explanatory power of the groups of predictor variables [@legendre2012].
 
@@
 types, as well as individually by BEC zone and forest type, with results
-aggregated to BEC zone groups (see @sec-study-area). All measured values
-were natural-log transformed and standardized to Z-scores based on the
-summary statistics of the full sample prior to analyses. Forest
-structure variables were treated as predictor variables and were grouped
-based on whether they were primary estimates from the lidar data or
-modelled from lidar and ancillary information (Table 2) for variance
-partitioning. The three DHIs were used as the response variables.
-Analysis of variance tests were used to determine which RDA axes were
-significant, with p-values below 0.1 using an F-test. The proportion of
-variance attributable to each axis was calculated from the eigenvalues
-generated by the RDA, and axis loadings were calculated as the Pearson
-correlation between each forest structure or DHI variable and the RDA
-axes. Forest structure variable axis loadings represent the strength of
-relationship between a given variable and the RDA axis, while the DHIs
-loadings indicate what is being represented by the RDA axes. To
-visualize the RDA results for both predictor and response variables, we
-display the results as path diagrams, with paths annotated by the
-loadings and RDA axes annotated by the variance in the DHIs explained by
-each axis (example shown in @fig-rda-var-template). We only consider and
-display significant RDA axes.
+aggregated to BEC zone groups (Table 1) . All measured values were
+natural-log transformed and standardized to Z-scores based on the
+summary statistics of the full sample prior to analyses. Following the
+data transformation, we visually assessed the pairwise relationships
+between the overall and stratified datasets for the assumptions of
+linearity and homoskedasticity. Forest structure variables were treated
+as predictor variables and were grouped based on whether they were
+primary estimates from the lidar data or modelled from lidar and
+ancillary information (Table 2) for variance partitioning. The three
+DHIs were used as the response variables. Analysis of variance tests
+were used to determine which RDA axes were significant, with p-values
+below 0.1 using an F-test. The proportion of variance attributable to
+each axis was calculated from the eigenvalues generated by the RDA, and
+axis loadings were calculated as the Pearson correlation between each
+forest structure or DHI variable and the RDA axes. To ensure similar
+directionality (+ or -) in the loadings across each variable in every
+strata, we transformed the data such that the cumulative DHI loading was
+always positive. To do so, if the cumulative DHI loading was negative we
+multiplied each loading in that strata and axis by -1. Forest structural
+attribute axis loadings represent the strength of relationship between a
+given variable and the RDA axis, while the DHI loadings indicate what is
+being represented by the RDA axes. To visualize the RDA results, we plot
+the significant loadings as annotated arrows between the
+predictor/response variables and each RDA axis (shown as a box), similar
+to a path diagram. We also display the amount of variance explained by
+each RDA axis within their respect
 
@@
 than individual RDA axes, we present variation partitioning results
-alongside the path diagrams illustrating the details of the RDA results
-(@fig-rda-var-template).
-
-```{r}
-#| label: fig-rda-var-template
-#| fig-cap: Template for visualization of RDA and variation partitioning results.
-Paths between RDA axes and predictor/response variables will be labelled with axis loadings, caluclated as the correlation between the variable and the RDA axis.
-If loadings are non-significant, NS will be shown on the path.
-
-knitr::include_graphics(here::here("outputs", "template.drawio.png"))
-```
-
+alongside the path diagrams illustrating the details of the RDA results.
 RDA and variation partitioning calculations were done in R [@R-base]
@@
 Overall in BC, the intra-annual patterns of productivity as represented
-by the DHIs variables were weakly related to forest structure
+by the DHI variables were weakly related to forest structure
 (@fig-rda-var). Less than 16% of the overall variation in the DHIs was
@@
 axis as representing overall productivity. The overall productivity axis
-had similar loadings (between 0.24-0.27) for all modelled predictors and
-canopy cover. Canopy height had a smaller loading (0.1), while
-structural complexity was negatively correlated with the overall
-productivity axis. RDA2 explained much less of the variation in the
-DHIs. RDA2 had a large positive loading on the minimum DHI (0.55), with
-negative loadings on the cumulative (-0.19) and variation (-0.39) DHIs.
-Due to the minimum and variation having the most influence on RDA2, we
-identified it as a seasonality axis. Canopy cover and structural
-complexity had the strongest (albeit negative; -0.17 and -0.08,
-respectively) loadings on the seasonality axis (@fig-rda-var A).
+had similar loadings (between 0.24-0.27) for all modelled forest
+structural attributes and canopy cover. Canopy height had a smaller
+loading (0.1), while structural complexity was negatively correlated
+with the overall productivity axis. RDA2 explained much less of the
+variation in the DHIs. RDA2 had a large positive loading on the minimum
+DHI (0.55), with negative loadings on the cumulative (-0.19) and
+variation (-0.39) DHIs. Due to the minimum and variation having the most
+influence on RDA2, we identified it as a seasonality axis. Canopy cover
+and structural complexity had the strongest (albeit negative; -0.17 and
+-0.08, respectively) loadings on the seasonality axis (@fig-rda-var A).
 Variation partitioning highlighted that the majority of the explained
-variation in the DHIs was due to the primary attributes (canopy cover,
-canopy height, and structural complexity; 9% variation explained). The
-overlap between primary and modelled attributes was 3.5%, and the
-modelled attributes explained 3.2% of the variation on their own
-(@fig-rda-var B).
+variation in the DHIs was due to the primary forest structural
+attributes (canopy cover, canopy height, and structural complexity; 9%
+variation explained). The overlap between primary and modelled
+attributes was 3.5%, and the modelled attributes explained 3.2% of the
+variation on their own (@fig-rda-var B).
 
@@
 Both visualized analyses are across the entire dataset.
-Non-significant paths are labelled as "NS".
-See supplementary information for results from each BEC zone and forest type.
+See supplementary information for results from each Biogeoclimatic Ecosystem Classification zone and forest type.
 
-knitr::include_graphics(here::here("outputs", "drawio", "British Columbia.drawio.png"))
+knitr::include_graphics(here::here("outputs", "British Columbia.drawio.png"))
 ```
 
-Subdividing the sample by forest type led to similar or increased
-amounts of variance explained when compared to the overall dataset
-(@fig-fracts-bar). Subdividing by BEC zones generally led to smaller
-amounts of variation explained than BC or stratifying by forest type,
-which may be due to the length of the environmental gradients included
-in the resulting subsamples (@fig-fracts-bar). The overlap between the
-primary and modelled datasets often explained the majority of the
-variation in the DHIs, except when stratifying by forest types, where
-this is only true for the coniferous subsample. The broadleaf, mixed
-wood, and wetland-treed forest types contained most of the DHIs
-explained variation in the primary structural attributes. The BEC zones
-followed a similar pattern, with most zones' DHI variation principally
-explained by the overlap, with the exception being CWH, which has no
-variation explained by the overlap between primary and modelled
-datasets. The Coastal BEC group had the lowest amount of variation
-explained, with Coastal Douglas-fir having \<1% of variation explained
-by the structural data. Overall, the DHIs are decoupled from the forest
-structural attributes, as shown by the overall variation explained being
-under 30% regardless of data stratification (@fig-fracts-bar).
+By forest type, results indicated similar or increased amounts of
+variance explained when compared to the overall dataset
+(@fig-fracts-bar). By BEC zones results generally indicated smaller
+amounts of variation explained, which may be due to the relatively short
+environmental gradients within the zones. The overlap between the
+primary and modelled datasets was often responsible for the majority of
+the variation explained in the DHIs, except when stratifying by forest
+types, where this was only true for the coniferous forests. For
+broadleaf, mixed wood, and wetland-treed forest types most of the
+explained variation in the DHIs was contributed by the primary forest
+structural attributes. A similar pattern was found by BEC zones, where
+the variation in the DHIs was principally explained by the overlap
+between primary and modelled forest structural attributes, except for
+CWH, where no variation was explained by the overlap. The Coastal BEC
+group had the lowest amount of variation explained, with Coastal
+Douglas-fir having \<1% of variation explained by the structural data.
+Overall, the DHIs were decoupled from the forest structural attributes,
+as shown by the overall variation explained being under 30% regardless
+of data stratification.
 
@@
 
-Stratifying the dataset by forest type and ecosystem led to varying
-loadings between the RDA axes and predictor/response variables
-(@fig-radar). The DHIs' loadings on RDA1 are large (similar to the
-overall dataset), except in colder BEC groups (Montane, North, and
-Alpine), where the minimum DHI loading half as large as the cumulative
-or variation DHI loadings. As such, we do not identify the first axis as
-being consistently associated with overall productivity. Non-coniferous
-forest types (broadleaf, mixed wood and wetland-treed) typically had
-large loadings between RDA1 and structural complexity. The mixed wood
-forest and Coastal BEC group had no significant loadings from modelled
-forest structural attributes for RDA1. Canopy cover generally had the
-highest significant predictor loading on average in RDA1. Examining the
-predictor loadings for RDA1 spatially across the province shows a
-geographic pattern in the loading strengths of primary structural
-attributes (@fig-fcc A). The interior (Northern and Southern BEC groups)
-of the province generally showed high canopy cover loadings, while the
-Boreal zones in the northwest had equal loadings in canopy cover and
-canopy height. Coastal were driven similarly by canopy height and
-structural complexity, with very low or non-significant loadings from
-canopy cover (@fig-fcc A; @fig-radar). Modelled attributes for RDA1
+Stratification by forest type and ecosystem indicated varying loadings
+between the RDA axes and predictor/response variables (@fig-radar). All
+DHI variables had high loadings on RDA1 (similar to the overall
+dataset), with positive cumulative and minimum DHI loadings, and
+negative variation DHI loadings. Colder BEC groups (Montane, North, and
+Alpine) had smaller minimum DHI loadings, which were half as large as
+the other DHI loadings. As such, we identify the first axis as being
+consistently associated with overall productivity. Canopy cover
+generally had the highest significant predictor loading in RDA1, except
+for in non-coniferous forest types (broadleaf, mixed wood and
+wetland-treed). These forest types typically had large negative loadings
+between RDA1 and structural complexity. The mixed wood forest and
+Coastal BEC group had no significant loadings from modelled forest
+structural attributes for RDA1. Examining the predictor loadings for
+RDA1 spatially across the province shows a geographic pattern in the
+loading strengths of primary forest structural attributes (@fig-fcc A).
+The interior (Northern and Southern BEC groups) of the province
+generally showed high canopy cover loadings, while the Boreal zones in
+the northwest had equal loadings in canopy cover and canopy height.
+Coastal zones were driven similarly by canopy height and structural
+complexity, with very low or non-significant loadings from canopy cover
+(@fig-fcc A; @fig-radar). Modelled forest structural attributes for RDA1
 generally show similar loadings across the province (@fig-fcc B).
@@
 Only seven of sixteen BEC zones had a second RDA axis, with none being
-found in Coastal or Alpine BEC zones (@fig-radar; @fig-fcc) . All of the
+found in Coastal or Alpine BEC zones (@fig-radar; @fig-fcc). All of the
 Montane BEC zones had a secondary axis, while some in the Northern and
 Southern interior had a secondary axis. RDA2 often showed the largest
-absolute loadings in the minimum DHI, which was frequently negative
-(@fig-radar; see supplementary material). The secondary axis had smaller
+absolute loadings in the minimum DHI, which was frequently, but not
+always, negative (@fig-radar; see supplementary material). The montane
+group, which had the largest absolute minimum DHI loading, also had the
+largest absolute variation DHI loading. The secondary axis had smaller
 loadings in the variation and cumulative DHIs, with only the Southern
 BEC group having a lower variation DHI loading than cumulative DHI. RDA2
-has lower predictor loadings (maximum of \~0.25) when compared to RDA1
-(maximum of \~0.5), with infrequently significant loadings for canopy
-cover and structural complexity (@fig-radar). Canopy height had much
-larger primary attribute loadings in the Southern group, while the
-Northern group had slightly larger structural complexity loadings
-(@fig-radar; @fig-fcc C) Similar to RDA1, the modelled structural
-attribute loadings were generally similar to one another, although gross
-stem volume was slightly higher in some southern BEC zones (@fig-radar;
-@fig-fcc D).
+has lower absolute predictor loadings (maximum of \~0.25) when compared
+to RDA1 (maximum of \~0.5), with infrequently significant loadings for
+canopy cover and structural complexity (@fig-radar). Among the primary
+forest structural attributes, RDA2 had the highest loadings with canopy
+height in the Southern group, while the Northern group had slightly
+larger structural complexity loadings (@fig-radar; @fig-fcc C). Similar
+to RDA1, the modelled forest structural attribute loadings were
+generally similar to one another (@fig-radar; @fig-fcc D).
 
@@
 theoretical [@knyazikhin1998] and empirical [@atkins2018] relationships
-between forest ecosystem structure and function. We examined these
-relationships across the province of BC, Canada, which has significant
-environmental gradients [@valentine1978soil; @pojar1987] by using
-wall-to-wall lidar-derived forest structural attributes [@matasci2018;
-@matasci2018b] and intra-annual satellite-derived summaries of
-vegetation productivity [@radeloff2019;
-@razenkovaMediumresolutionDynamicHabitatInPress]. We found a weak
-relationship between in forest ecosystem structure and function overall
-across BC (15.7% of variation explained in the DHIs by forest structural
-attributes), as well as within individual BEC zones and forest types
-found in BC (@fig-fracts-bar).
+between forest ecosystem structure and function, however, the direction
+and significance of these relationships have not been consistent
+[@ali2019]. We examined these relationships across the province of BC,
+Canada, which has significant environmental gradients
+[@valentine1978soil; @pojar1987] by using wall-to-wall forest structural
+attributes generated by imputing the lidar-derived attributes with
+Landsat best-available-pixel composites [@matasci2018; @matasci2018b]
+and Landsat-derived summaries of intra-annual vegetation productivity
+[@radeloff2019; @razenkovaMediumresolutionDynamicHabitatInPress]. We
+found a weak relationship between these remote sensing indicators of
+forest ecosystem structure and function overall across BC (15.7% of
+variation explained in the DHIs by forest structural attributes), as
+well as within individual BEC zones and forest types found in BC
+(@fig-fracts-bar).
 
-Ecosystem structure and function are key EBV classes which are capable
-of being measured using satellite remote sensing at global extents and
-with large temporal depth [@pereira2013; @skidmore2021]. Both EBV
-classes have been related to biodiversity metrics [@radeloff2019;
-@bergen2009]. One of the key attributes of the EBV framework is the
-complementarity of the proposed datasets, therefore it is necessary to
-determine if potential EBVs comprise unique or duplicate information
-when compared with one another [@skidmore2021]. This can be especially
-important when both datasets being examined were primarily derived using
-data from the same series of satellites, in our case, the Landsat series
-[@matasci2018; @matasci2018b;
-@razenkovaMediumresolutionDynamicHabitatInPress]. Our findings show weak
-relationships between these two classes, indicating that these datasets
-are very suitable to be used in a fashion complementary to one another,
-especially when examining ecosystem integrity, even with the same data
-provenance [@hansen2021].
+Our findings indicate that remote sensing-derived forest ecosystem
+structure and forest ecosystem function EBVs are well suited to be used
+in a complementary fashion for ecosystem integrity monitoring, despite
+both EBVs being derived from surface reflectance from the Landsat series
+of satellites [@hansen2021]. While we assessed the complementarity of
+ecosystem structure and function, a single pair of EBVs measurable at
+the ecosystem scale, other EBV class combinations at finer scales should
+also be assessed for complementarity. This is especially relevant in
+cases such as ours, where remote sensing datasets such as Landsat can
+generate a large number of EBVs in different classes [@skidmore2021]. In
+addition, Skidmore et al. [@skidmore2021] highlighted that many
+satellite-derived biodiversity products can fall into multiple EBV
+classes simultaneously -- e.g., leaf area index being in the ecosystem
+structure, ecosystem function, and species trait EBV classes --
+depending on the spatial scale of analysis and goal of the monitoring
+program. As such, it is relevant to consider which classes each
+biodiversity product should belong to, for given spatial and temporal
+extents. Other EBVs, such as those found in the community composition
+and species population classes, may also be inherently linked due to
+being based on the same data sources, even if not observed using remote
+sensing platforms.
 
-While we assess the complementarity of ecosystem structure and function,
-a single pair of EBVs easily measurable at similar scales, other EBV
-class combinations should also be assessed for complementarity. This is
-especially relevant in cases such as ours, where remote sensing datasets
-can proliferate to a number of EBVs in different classes
-[@skidmore2021]. Other EBVs, such as those found in the community
-composition and species population classes, will be inherently linked
-due to being based on the same data. In addition, @skidmore2021
-highlighted that many satellite-derived biodiversity products can fall
-into multiple EBV classes simultaneously -- e.g., leaf area index being
-in the ecosystem structure, ecosystem function, and species trait EBV
-classes -- depending on the spatial scale of analysis and goal of the
-monitoring system. As such, it is relevant to consider which classes
-each biodiversity product should belong to, for given extents and
-temporal depths.
-
-Further, we assessed the utility of primary forest structural attributes
-when compared to modelled forest structural attributes to examine
-duplicated information within a single EBV class. We found that standard
-lidar-derived measurements, such as canopy height, canopy cover, and
-structural complexity (as proposed by Valbuena et al., 2020) comprise
-the majority of the explained variation in the DHIs, either
-independently or in their overlap with the modelled attributes
-(@fig-fracts-bar). This could be expected as modelled forest structural
-attributes are generally directly based on the primary forest structural
-attributes [@coops2021], however, the creation of modelled forest
-structural attributes involves modelling expertise and processing time
-overhead [@matasci2018; @matasci2018b; @duncanson2022]. One advantage of
-these interpreted products is potential uptake by a broader set of
-users, and their ability to be more closely linked to management actions
-and monitoring results, as is the case with forest biomass for climate
-change mitigation [@duncanson2022].
+Further, we assessed the information-richness of primary forest
+structural attributes when compared to modelled forest structural
+attributes to support indicator selection within a single EBV class. We
+found that standard lidar-derived measurements, such as canopy height,
+canopy cover, and structural complexity (as proposed by Valbuena et al.
+[@valbuena2020]) contribute to the majority of the explained variation
+in the DHIs, either independently or in their overlap with the modelled
+forest structural attributes (@fig-fracts-bar). This could be expected
+as modelled forest structural attributes are generally directly based on
+the primary forest structural attributes [@coops2021], however, the
+creation of modelled forest structural attributes involves modelling
+expertise and processing time overhead [@matasci2018; @matasci2018b;
+@duncanson2022]. One advantage of these interpreted products is
+potential uptake by a broader set of users, and their ability to be more
+closely linked to management actions and monitoring results, as is the
+case with forest biomass for climate change mitigation [@duncanson2022].
+While either set of forest structural attributes may be most relevant
+for a given monitoring task, we note that the modelled structural
+attributes do not strongly overlap with structural complexity, and as
+such, we generally recommend to use the primary forest structural
+attributes.
 
@@
 Forest types with deciduous trees (namely broadleaf and mixed wood) had
-large, positive structural complexity loadings when compared to the
+large, negative structural complexity loadings when compared to the
 other forest structural attributes (@fig-radar), with mixed wood forests
@@
 providing additional photosynthesis in these forest types
-[@nilsson2005]. Coniferous forests DHIs, on the other hand, show a
-weaker relationship with structural complexity than all other forest
+[@nilsson2005]. In coniferous forests, on the other hand, the DHIs show
+a weaker relationship with structural complexity than all other forest
 structural attributes. Notably, the Alpine, Montane, and Northern BEC
@@
 loadings between the Minimum DHI and RDA1. This decoupling in the DHIs
-is potentially due to the the near certainty of a pixel containing snow
+is potentially due to the near certainty of a pixel containing snow
 cover over the course of a year leading to consistently low minimum DHI
-values across the ecosystem and thus to weaker minimum DHI loadings
+values across the ecosystem, and thus to weaker minimum DHI loadings
 [@razenkova2020; @beck2006]. These ecosystems with DHI decoupling also
 showed relatively strong relationships to canopy cover and weak
-relationships with structural complexity, potentially due to the
-dominance of coniferous forests in these ecosystems.
+relationships with structural complexity. This weak relationship to
+structural complexity may be due to the dominance of coniferous forests
+in these ecosystems, which have been shown to have lower forest
+structural complexity across the globe [@ehbrecht2021], with weaker
+relationships between canopy height and structural complexity being
+found [@atkins2022].
 
 BC shows varying forest structural drivers of the DHIs depending on the
-geographic location [@fig-fcc]. The interior of BC is drier and warmer,
+geographic location (@fig-fcc). The interior of BC is drier and warmer,
 and also has productivity being strongly driven by canopy cover, while
@@
 the DHIs [@perez2016], leading to varying loadings strengths depending
-on the distance to the coastline. Further, the extreme environmental and
-topographic gradients within BC's ecosystems [@pojar1987;
-@valentine1978soil] may be leading to highly variable data within the
-BEC zones we used for this study, and a finer ecoregionalization may
-reveal a different relationship between ecosystem structure and
-function.
+on the distance to the coast. For example, water availability has been
+shown to be a driver of canopy height [@klein2015]. In these wetter
+regions such as the Coastal group, structural variability is more likely
+to be present in the vertical, rather than horizontal direction,
+especially given the relationship between canopy height and structural
+complexity [@atkins2022]. Further, the weakest relationships between
+structure and function were found in the Coastal group of BEC zones
+(@fig-fracts-bar), which may be due to the consistently high amounts of
+canopy cover, and lower amounts of snow found in these zones leading to
+potentially saturated NDVI values throughout the year. These consistent
+forest structural metrics --- such as the high canopy cover values found
+in the Coastal BEC zones --- may also lead to lower amounts of variation
+explained by these metrics, leading to them being less informative in
+the loadings, and other metrics, such as structural complexity having
+higher amounts of variance explained (@fig-radar). These complex
+linkages between climate, forest structure, and forest productivity
+become further complicated by the extreme environmental and topographic
+gradients within BC's ecosystems [@pojar1987; @valentine1978soil], which
+may warrant study into the scale of these interactions, with a finer (or
+non-climate based) ecoregionalization potentially revealing different
+relationships between ecosystem structure and function.
+
+Various relationship directions and strengths have been found between
+forest structure and forest function [@ali2019]. Temperate and boreal
+forests often have mixed or negative structure-function relationships,
+while tropical forests often have positive relationships [@ali2019]. We
+found that forest structure and function were not strongly linked to one
+another, both across the full study area, and when stratified by
+ecosystem and forest type (@fig-radar). Forest type models showed higher
+amounts of variation explained, which may indicate the linkages between
+forest structure and function are stronger at finer spatial scales,
+potentially driven by finer scale landscape patterns such as terrain and
+land cover, rather than climate [@thom2021]. Notably, the BEC zones and
+forest types are delineated at different spatial scales; forest types
+are mapped at a 30 m pixel size [@hermosilla2022; @hermosilla2018], and
+can be found across the entire province, while BEC zones do not span the
+entirety of the province, and are mapped at a much coarser scale
+[@pojar1987].
 
@@
 correlated with one another (@fig-rda-var). While this does imply that
-utilizing a single one of the three DHI axis may be suitable within this
-geographic region, the the DHIs do provide unique information to one
+using a single one of the three DHI axis may be suitable within this
+geographic region, the DHIs do provide unique information to one
 another, which can be especially relevant when stratifying the study
@@
 
+A potential limitation of our study is due to the uncertainty inherent
+to remotely sensed data. Wall-to-wall forest structural information was
+produced by imputing lidar-derived metrics over satellite spectral
+reflectance and geospatial data across the forested landscapes of
+Canada. The accuracy of the forest structure dataset varied between
+forest structural metrics [@matasci2018; @matasci2018b]. Likewise,
+spectral vegetation indices, as discussed, are not direct measurements
+of productivity, but rather are based on surface reflectance of
+vegetation, which can be influenced by many factors such as soil, canopy
+closure, viewing geometry, and atmospheric effects [@myneni1994;
+@yoder1994]. These uncertainties may weaken the relationship between
+these remote sensing indicators of forest structure and function.
+Further, while forest structure and function have both been shown to be
+drivers of biodiversity [@radeloff2019; @atkins2023], there are
+ecological processes that cannot be assessed through remotely sensed
+data [@coops2009exploring], including predation and fecundity, and thus
+*in situ* data are needed to monitor these processes. Finally, our study
+assumes that structure and function are indicators of biodiversity in
+this system, and we draw on published demonstrations of the relevance of
+each of them to species richness and composition, but we recommend
+further evaluations of these indicators with biodiversity observations
+to determine the relative contributions of each to estimating species
+richness and turnover [@coops2018a].
+
 # Conclusion
@@
 complementarity of two potential EBV datasets - forest structure and the
-DHIs. We also separated the forest structure datasets into primary and
-modelled structure variables in order to assess the need to develop more
-complex structure variables, or if data derived directly from lidar
-datasets are suitable. We found that the structure variables were not
-strongly related to the DHIs, indicating that they are suitable to be
-used together as complementary ecosystem-scale EBVs when monitoring
-forest environments. It was also found that variation in the DHIs
-explained by the overlap between primary and modelled structure
-variables was often higher than the variation explained by either
-individually, both in the BC-wide dataset and the stratified datasets.
-These ecosystem structure and function variables are often attainable at
-global scales using satellite remote sensing, and recent advances are
-allowing them both to be generated at medium spatial resolutions. We
-suggest that biodiversity researchers focus on using forest structural
-attributes derived directly from the lidar data, and if needed, use a
-single modelled forest structural attribute, such as total aboveground
-biomass. In addition, we highlight that the usage of intra-annual
-summarizations of productivity provide novel information for
-biodiversity monitoring when used with forest structural attributes, and
-utilizing them as complementary EBVs for a holistic monitoring system
-for forest ecological integrity.
+DHIs. We also separated the forest structure indicator datasets into
+primary and modelled structure metrics in order to assess the need to
+develop more complex structure variables, or if the primary forest
+structural attributes are suitable on their own. We found that the
+structure metrics were not strongly related to the DHIs, indicating that
+they are suitable to be used together as complementary ecosystem-scale
+EBVs when monitoring forest environments. It was also found that
+variation in the DHIs explained by the overlap between primary and
+modelled structure variables was often higher than the variation
+explained by either individually, both in the BC-wide dataset and the
+stratified datasets. These ecosystem structure and function indicators
+are often attainable at global scales using satellite remote sensing,
+and recent advances are allowing them both to be generated at medium
+spatial resolutions. We suggest that biodiversity researchers focus on
+using forest structural attributes derived directly from the lidar data,
+and if needed, use a single modelled forest structural attribute, such
+as total aboveground biomass. In addition, we highlight that the usage
+of intra-annual summarizations of productivity provide novel information
+for biodiversity monitoring when used with forest structural attributes
+as complementary EBVs for a holistic monitoring system for forest
+ecological integrity.
 
@@
 This research was funded by NSERC support of Coops (RGPIN-2018-03851).
-Remote sensing data products utilized in this research are free and open
-and available for download at <https://ca.nfis.org/maps_eng.html>. We
-thank Dr. Michael Wulder and Dr. Joanne White for development and early
+Remote sensing data products used in this research are free and open and
+available for download at <https://ca.nfis.org/maps_eng.html>. We thank
+Dr. Michael A. Wulder and Dr. Joanne C. White for development and early
 access to these National Terrestrial Ecosystem Mapping System (NTEMS)
@@
 
+# Author Contributions
+
+Conceptualization: ERM, MEA, NCC, ACB and SSB. Methodology: ERM, MEA,
+NCC and TH. Software: ERM and MEA. Formal Analysis: ERM, MEA, and NCC.
+Investigation: ERM. Writing - Original Draft: ERM. Writing - Review and
+Editing: ERM, MEA, NCC, TH, ACB and SSB. Visualization: ERM, MEA, TH,
+NCC. Supervision: NCC. All authors approve the submitted version of the
+manuscript.
+
+# Competing Interests
+
+MEA is an Editorial Board Member of *Scientific Reports*. The authors
+declare no other competing interests.
+
+# Data Availability
+
+The 30 m spatial coverages of the NTEMS data are available online at
+https://opendata.nfis.org/mapserver/nfis-change_eng.html. BEC zone
+boundaries were obtained using the *bcmaps* R package. The Dynamic
+Habitat Indices were calculated using freely available Landsat imagery
+in Google Earth Engine. Details on their production can be found in
+Razenkova et al. [@razenkovaMediumresolutionDynamicHabitatInPress]. All
+code used for data manipulation and analysis can be found at
+https://github.com/emuise/code-structProdSem.
+
 \newpage
@@
 # References {.unnumbered}
+
+::: {#refs}
+:::
+
+\newpage
+
+**Legends**
+
+Figure 1: Location of Biogeoclimatic Ecosystem Classification (BEC)
+zones in British Columbia.
+
+Figure 2: A) Axis loadings from redundancy analysis (RDA) of primary and
+modelled forest structure variables on the dynamic habitat indices
+(DHIs). B) Results from variation partitioning of primary and modelled
+forest structure variables on the DHIs. Both visualized analyses are
+across the entire dataset. See supplementary information for results
+from each Biogeoclimatic Ecosystem Classification zone and forest type.
+
+Figure 3: Stacked bar plot proportion of the variation in the Dynamic
+Habitat Indices (DHIs) variation explained by primary, modelled, and the
+overlap between primary and modelled structure variables across the
+stratified and overall datasets.
+
+Figure 4: Parallel coordinate plots of average absolute significant
+loading strength by Biogeoclimatic Ecosystem Classification (BEC) group,
+forest type, and British Columbia. Note the varying y axes.
+
+Figure 5: False colour maps by Biogeoclimatic Ecosystem Classification
+(BEC) zone of axis loadings for the first redundancy analysis (RDA) axis
+(A and B) and second RDA axis (C and D). Colour values are normalized to
+the maximum loading of each variable. A and C show axis loadings for
+canopy height (CH; red), canopy cover (CC; green) and structural
+complexity (SC; blue). B and D show axis loadings for basal area (BA;
+red), total biomass (TB; green) and gross stem volume (GSV; blue). Black
+indicates no significant loadings, while greyscale colours indicate
+similar loadings across the three variables. Annotated colour wheels for
+each set of variables also shown.

### 2023-11-14 scripts/4_vegan.qmd
resubmit following revisions

@@
 # non normalized
-loadings_rgb <- clean_loads %>% 
+loadings_rgb <- clean_loads_abs %>% 
   group_by(axis) %>%

### 2023-11-14 scripts/manuscript/manuscript.qmd
resubmit following revisions

@@
 Biodiversity Variables (EBVs) [@pereira2013], which are standardized,
-comparable variables that represents a minimum set of biological
+comparable variables that represent a minimum set of biological
 information required for monitoring biodiversity change at large spatial
@@
 EBV datasets, it becomes important to assess the interrelationships
-between these datasets and the complementary of the information they
+between these datasets and the complementarity of the information they
 provide to reduce redundancy in future EBV development efforts
@@
 
-The relationship between ecosystem structure and function have been
+The relationships between ecosystem structure and function have been
 examined within a remote sensing context for decades [@huete2002;
@@
 last 35 years were discarded by using the disturbance mask generated for
-the forested ecosystems of Canada [@hermosilla2016 ] . A maximum of 500
+the forested ecosystems of Canada [@hermosilla2016]. A maximum of 500
 sample units were selected from each stratum; in the many strata where
 this was not achieved, all sampling units meeting the above restrictions
-were selected. Sampling was conducted in R version 4.2.2 \[@R-base\]
-using the **sgsR** package [@R-sgsR]. Neighbourhood analyses for the
-land cover classes and coefficient of variations of canopy height and
-cover were calculated in Python version 3.9.
+were selected. Sampling was conducted in R version 4.2.2 [@R-base] using
+the **sgsR** package [@R-sgsR]. Neighbourhood analyses for the land
+cover classes and coefficient of variations of canopy height and cover
+were calculated in Python version 3.9.
 
@@
 the primary and modelled forest structure variables to ecosystem
-function across a broad environment range. RDA has widely been used in
+function across a broad environmental range. RDA has widely been used in
 community ecology where environmental variables of interest are compared
@@
 types, as well as individually by BEC zone and forest type, with results
-aggregated to BEC zone groups (Table 1) . All measured values were
+aggregated to BEC zone groups (Table 1). All measured values were
 natural-log transformed and standardized to Z-scores based on the
@@
 were used to determine which RDA axes were significant, with p-values
-below 0.1 using an F-test. The proportion of variance attributable to
+below 0.1, using an F-test. The proportion of variance attributable to
 each axis was calculated from the eigenvalues generated by the RDA, and
@@
 variance explained when compared to the overall dataset
-(@fig-fracts-bar). By BEC zones results generally indicated smaller
+(@fig-fracts-bar). By BEC zone, results generally indicated smaller
 amounts of variation explained, which may be due to the relatively short
@@
 #| label: fig-fracts-bar 
-#| fig-cap: Stacked bar plot proportion of the variation in the Dynamic Habitat Indices (DHIs) variation explained by primary, modelled, and the overlap between primary and modelled structure variables across the stratified and overall datasets.
+#| fig-cap: Stacked bar plot showing proportion of the variation in the Dynamic Habitat Indices (DHIs) explained by primary, modelled, and the overlap between primary and modelled structure variables across the stratified and overall datasets.
 knitr::include_graphics(here::here("outputs", "fracts_bar_plot.png"))
@@
 consistently associated with overall productivity. Canopy cover
-generally had the highest significant predictor loading in RDA1, except
+generally had the highest significant predictor loading on RDA1, except
 for in non-coniferous forest types (broadleaf, mixed wood and
@@
 #| label: fig-radar
-#| fig-cap: Parallel coordinate plots of average absolute significant loading strength by Biogeoclimatic Ecosystem Classification (BEC) group, forest type, and British Columbia.
-Note the varying y axes.
+#| fig-cap: Parallel coordinate plots of average significant loading strength by Biogeoclimatic Ecosystem Classification (BEC) group, forest type, and British Columbia. Note the varying y axes.
 
@@
 absolute loadings in the minimum DHI, which was frequently, but not
-always, negative (@fig-radar; see supplementary material). The montane
+always, negative (@fig-radar; see supplementary material). The Montane
 group, which had the largest absolute minimum DHI loading, also had the
@@
 A and C show axis loadings for canopy height (CH; red), canopy cover (CC; green) and structural complexity (SC; blue).
-B and D show axis loadings for basal area (BA; red), total biomass (TB; green) and gross stem volume (GSV; blue).
+B and D show axis loadings for basal area (BA; red), total aboveground biomass (AB; green) and gross stem volume (GSV; blue).
 Black indicates no significant loadings, while greyscale colours indicate similar loadings across the three variables.
@@
 and also has productivity being strongly driven by canopy cover, while
-the Coastal zones are the only BEC zone with strong structural
-complexity loadings. The climate of these regions could be influencing
-the DHIs [@perez2016], leading to varying loadings strengths depending
-on the distance to the coast. For example, water availability has been
-shown to be a driver of canopy height [@klein2015]. In these wetter
-regions such as the Coastal group, structural variability is more likely
-to be present in the vertical, rather than horizontal direction,
-especially given the relationship between canopy height and structural
-complexity [@atkins2022]. Further, the weakest relationships between
-structure and function were found in the Coastal group of BEC zones
-(@fig-fracts-bar), which may be due to the consistently high amounts of
-canopy cover, and lower amounts of snow found in these zones leading to
-potentially saturated NDVI values throughout the year. These consistent
-forest structural metrics --- such as the high canopy cover values found
-in the Coastal BEC zones --- may also lead to lower amounts of variation
-explained by these metrics, leading to them being less informative in
-the loadings, and other metrics, such as structural complexity having
-higher amounts of variance explained (@fig-radar). These complex
-linkages between climate, forest structure, and forest productivity
-become further complicated by the extreme environmental and topographic
-gradients within BC's ecosystems [@pojar1987; @valentine1978soil], which
-may warrant study into the scale of these interactions, with a finer (or
-non-climate based) ecoregionalization potentially revealing different
-relationships between ecosystem structure and function.
+the Coastal BEC zone group has very low canopy cover loadings and high
+structural complexity loadings (@fig-radar). The climate of these
+regions could be influencing the DHIs [@perez2016], leading to varying
+loadings strengths depending on the distance to the coast. For example,
+water availability has been shown to be a driver of canopy height
+[@klein2015]. In these wetter regions such as the Coastal group,
+structural variability is more likely to be present in the vertical,
+rather than horizontal direction, especially given the relationship
+between canopy height and structural complexity [@atkins2022]. Further,
+the weakest relationships between structure and function were found in
+the Coastal group of BEC zones (@fig-fracts-bar), which may be due to
+the consistently high amounts of canopy cover, and lower amounts of snow
+found in these zones leading to potentially saturated NDVI values
+throughout the year. These consistent forest structural metrics --- such
+as the high canopy cover values found in the Coastal BEC zones --- may
+also lead to lower amounts of variation explained by these metrics,
+leading to them being less informative in the loadings, and other
+metrics, such as structural complexity having higher amounts of variance
+explained (@fig-radar). These complex linkages between climate, forest
+structure, and forest productivity become further complicated by the
+extreme environmental and topographic gradients within BC's ecosystems
+[@pojar1987; @valentine1978soil], which may warrant study into the scale
+of these interactions, with a finer (or non-climate based)
+ecoregionalization potentially revealing different relationships between
+ecosystem structure and function.
 
@@
 modelled structure variables was often higher than the variation
-explained by either individually, both in the BC-wide dataset and the
-stratified datasets. These ecosystem structure and function indicators
-are often attainable at global scales using satellite remote sensing,
-and recent advances are allowing them both to be generated at medium
-spatial resolutions. We suggest that biodiversity researchers focus on
-using forest structural attributes derived directly from the lidar data,
-and if needed, use a single modelled forest structural attribute, such
-as total aboveground biomass. In addition, we highlight that the usage
-of intra-annual summarizations of productivity provide novel information
+explained by either individually. The exception to this was found in
+non-coniferous forest types, Coastal BEC zones, and the BC-wide
+analysis. These ecosystem structure and function indicators are often
+attainable at global scales using satellite remote sensing, and recent
+advances are allowing them both to be generated at medium spatial
+resolutions. We suggest that biodiversity researchers focus on using
+forest structural attributes derived directly from the lidar data, and
+if needed, use a single modelled forest structural attribute, such as
+total aboveground biomass. In addition, we highlight that the usage of
+intra-annual summarizations of productivity provide novel information
 for biodiversity monitoring when used with forest structural attributes
@@
 
-Figure 3: Stacked bar plot proportion of the variation in the Dynamic
-Habitat Indices (DHIs) variation explained by primary, modelled, and the
-overlap between primary and modelled structure variables across the
+Figure 3: Stacked bar plot showing the proportion of the variation in
+the Dynamic Habitat Indices (DHIs) explained by primary, modelled, and
+the overlap between primary and modelled structure variables across the
 stratified and overall datasets.
 
-Figure 4: Parallel coordinate plots of average absolute significant
-loading strength by Biogeoclimatic Ecosystem Classification (BEC) group,
-forest type, and British Columbia. Note the varying y axes.
+Figure 4: Parallel coordinate plots of average significant loading
+strength by Biogeoclimatic Ecosystem Classification (BEC) group, forest
+type, and British Columbia. Note the varying y axes.
 
@@
 complexity (SC; blue). B and D show axis loadings for basal area (BA;
-red), total biomass (TB; green) and gross stem volume (GSV; blue). Black
-indicates no significant loadings, while greyscale colours indicate
-similar loadings across the three variables. Annotated colour wheels for
-each set of variables also shown.
+red), total aboveground biomass (AB; green) and gross stem volume (GSV;
+blue). Black indicates no significant loadings, while greyscale colours
+indicate similar loadings across the three variables. Annotated colour
+wheels for each set of variables also shown.

### 2024-01-10 scripts/4_vegan_clean.qmd
began revisions round 2, cleaned 4_vegan to be easier to work with (less defunct extra code)

--- 2023-07-24 scripts/4_vegan.qmd
@@
 library(patchwork)
-library(ggradar)
+
 source(here::here("scripts", "get_keys.R"))
@@
 ```{r data-prep}
+zones <- keys$bec %>% pull(zone)
+groups <- keys$bec %>% pull(group) %>% unique()
+
 bec_sample_loc <- here::here("data", "samples", "vlce_extracted.shp")
@@
   mutate(across(basal_area:VarDHI,  .fns = ~ normal(.x))) 
-
+```
+
+```{r bec-nest}
+to_nest <- bec_normal %>%
+  select(-percentage_first_returns_above_mean,
+         -loreys_height,
+         -elev_stddev,
+         -elev_mean)
+
+bec_nested <- to_nest %>%
+  group_by(bec) %>%
+  nest() %>%
+  ungroup() %>%
+  rename(strata = bec)
+
+class_nested <- to_nest %>%
+  group_by(category) %>%
+  nest() %>%
+  ungroup() %>%
+  rename(strata = category)
+
+set.seed(69420)
+
+strata_nested <- bind_rows(bec_nested, class_nested) %>%
+  mutate(data = map(data, slice_sample, n = 500))
+# this slice sample is the ANALYTICAL difference between 4_vegan and 
+# 4_vegan_clean, other than the actual cleanup of the .rmd file
+
+global <- strata_nested %>%
+  filter(strata %in% zones) %>%
+  unnest(data)
 ```
@@
 ```{r global-model}
-to_rda <- bec_normal %>%
-  select(-elev_mean,
-         -percentage_first_returns_above_mean,
-         -elev_stddev,
-         -loreys_height)
-
-forest_types <- bec_normal %>%
+forest_types <- global %>%
   pull(category) %>%
@@
 
-# to_rda <- bec_normal %>% 
-#   group_by(category) %>% 
-#   slice_sample(n = 3000)  %>% 
-#   filter(category == "Mixed Wood") %>% 
-#   ungroup()
-
-dhi_tab <- to_rda %>%
+dhi_tab <- global %>%
   select(ends_with("DHI"))
 
-primary <- to_rda %>%
+primary <- global %>%
   select(elev_p95,
          elev_cv,
-         # elev_stddev,
          percentage_first_returns_above_2m)
 
-modelled <- to_rda %>%
+modelled <- global %>%
   select(basal_area,
@@
 rda_out <-
-  rda(dhi_tab ~ elev_p95 + percentage_first_returns_above_2m + elev_cv + # simple/raw/primary
-      basal_area + gross_stem_volume + total_biomass, # modelled
-      data = to_rda)
+  rda(
+    dhi_tab ~ elev_p95 + percentage_first_returns_above_2m + elev_cv + # simple/raw/primary
+      basal_area + gross_stem_volume + total_biomass,
+    # modelled
+    data = global
+  )
 
@@
 
-# scores(rda_out, choices = c(1:3))$species
-# scores(rda_out, choices = c(1:3))$biplot
-# not actually the loadings, need to get those from correlations
-
 rcorr <- summary(rda_out)$sites %>%
@@
   select(starts_with("RDA")) %>%
-  bind_cols(to_rda) %>%
-  select(-bec, -category) %>% 
+  bind_cols(global) %>%
+  select(-strata,-category) %>%
   as.matrix() %>%
@@
   filter(axis != "RDA3") %>%
-  mutate(sig = cut(axis_p, 
+  mutate(sig = cut(
+    axis_p,
     breaks = c(0, 0.001, 0.01, 0.05, .10, 1),
     labels = c("***", "**", "*", ".", "NS"),
-                   right = F))
-
-for_radar_loads <- all_data_loadings %>%
-  filter(!endsWith(var, "DHI")) %>%
-  mutate(axis_loading = ifelse(sig == "NS", 0, abs(axis_loading))) %>%
-  select(var, axis, axis_loading) %>%
-  pivot_wider(names_from = var, values_from = axis_loading) %>%
-  mutate(cluster = paste(axis, "- British Columbia")) %>%
-  select(-axis) %>%
-  relocate(cluster)
+    right = F
+  ))
 
@@
 
-Metrics derived solely from the point cloud explain 10.6% of the variation in the DHIs, while metrics modelled based on the point cloud explain 3.4% of the variation in the DHIs.
-85.9% of the variation is unexplained by structural attributes.
-
 ```{r r2-global}
@@
 ```
-
-This model shows very little explanatory power. (13.9%)
 
@@
 
-The model is statistically significant (P \< 0.001). All of the axes are significant, although this may be related to the large sample size.
-
-```{r bec-nest}
-to_nest <- bec_normal %>%
-  select(-percentage_first_returns_above_mean,
-         -loreys_height,
-         -elev_stddev,
-         -elev_mean)
-
-bec_nested <- to_nest %>%
-  group_by(bec) %>%
-  nest() %>%
-  ungroup() %>%
-  rename(strata = bec)
-
-class_nested <- to_nest %>%
-  group_by(category) %>%
-  nest() %>%
-  ungroup() %>%
-  rename(strata = category)
-
-# # vlce sampling regime to get the samples up
-# switched away from this to a single double stratified sample
-# vlce_samples <- read_sf(here::here("data", "samples", "vlce_extracted.shp"))
-# 
-# vlce_df <- vlce_samples %>%
-#   st_drop_geometry() %>%
-#   select(-category_1, -category_2) %>%
-#   rename(category = strata) %>%
-#   relocate(bec) %>%
-#   drop_na() %>%
-#   ungroup()
-# 
-# names(vlce_df) <- names_manual
-# 
-# vlce_normal <- vlce_df %>%
-#   mutate(across(basal_area:VarDHI,  .fns = ~ normal(.x))) 
-# 
-# to_nest <- vlce_normal %>%
-#   select(-percentage_first_returns_above_mean,
-#          -loreys_height,
-#          -elev_stddev,
-#          -elev_mean)
-# 
-# 
-# 
-# 
-# class_nested <- to_nest %>%
-#   group_by(category) %>%
-#   slice_sample(n = 3000) %>%
-#   nest() %>% 
-#   ungroup() %>%
-#   rename(strata = category)
-
-strata_nested <- bind_rows(bec_nested, class_nested)
+```{r}
+# linear relationships * homoschedasticity check
+
+bec_normal %>%
+  pivot_longer(cols = ends_with("DHI"),
+               names_to = "dhi",
+               values_to = "dhi_val") %>%
+  pivot_longer(
+    cols = basal_area:total_biomass,
+    names_to = "struct",
+    values_to = "struct_val"
+  ) %>%
+  ggplot(aes(x = struct_val, y = dhi_val)) +
+  geom_hex() +
+  facet_grid(rows = vars(dhi),
+             cols = vars(struct),
+             scales = "free") +
+  labs(x = "Structure Value (Normalized)",
+       y = "DHI Value (Normalized)")
 ```
@@
 ```{r fracts-out}
-
 fracts_out <- outs %>%
@@
 
+Fracts Bar Plot (Figure)
+
 ```{r}
@@
        height = 6, width = 8)
-```
-
-```{r raincloud-plot}
-library(ggdist)
-fracts_out %>%
-  mutate(total = primary + modelled + overlap) %>%
-  #left_join(keys$bec, by = c("bec" = "zone")) %>%
-  pivot_longer(-strata) %>%
-  mutate(name = str_to_title(name)) %>%
-  mutate(value = ifelse(value < 0, 0, value),
-         class = ifelse(strata %in% forest_types, 
-                        "Forest Types", 
-                        "BEC Zones")) %>%
-  ggplot(aes(x = value, y = name)) +
-  stat_halfeye(
-    # adjust bandwidth
-    adjust = .5,
-    # move to the right
-    justification = -0.2,
-    # remove the slub interval
-    .width = 0,
-    point_colour = NA
-    ) +
-  geom_boxplot(
-    width = 0.12,
-    # removing outliers
-    outlier.color = NA,
-    alpha = 0.5
-  ) +
-  stat_dots(
-    # ploting on left side
-    side = "left",
-    # adjusting position
-    justification = 1.1,
-    # adjust grouping (binning) of observations
-    binwidth = 0.0025,
-    #position = "stack",
-    # aes(fill = group, col = group),
-    alpha = 0.5,
-    aes(shape = class)
-  ) +
-  labs(y = NULL,
-       x = "% Variance Explained",
-       fill = "BEC Group",
-       col = "BEC Group") +
-  scale_x_continuous(labels = scales::label_percent()) +
-  theme_bw() +
-  theme(panel.grid = element_blank(),
-        legend.position = c(0.8, .2))
-
-```
-
-```{r path-setup}
-path_base <- tibble(x1 = c(0,
-             0,
-             0,
-             0,
-             1),
-       y1 = c(1,
-              -1,
-              1,
-              -1,
-              0),
-       x2 = c(1.85,
-              1.85,
-              .85,
-              .85,
-              1.85),
-       y2 = c(.15,
-              -.15,
-              .15,
-              -.15,
-              0)) %>%
-  ggplot(aes(x = x1, y = y1)) +
-  geom_segment(aes(xend = x2, yend = y2), arrow = arrow()) +
-  geom_point(data = tibble(x1 = c(.85, .85), 
-                           y1 = c(-.5, .5)),
-             size = 60,
-             shape = 21,
-             col = "grey") +
-  # geom_text(aes(x = 0, y = -1.15), label = "Modelled\nAttributes") +
-  # geom_text(aes(x = 0, y = 1.15), label = "primary\nAttributes") +
-  # geom_text(aes(x = .65, y = 0), label = "Overlap") +
-  # geom_text(aes(x = 2, y = 0), label = "Dynamic\nHabitat\nIndices") +
-  theme_void() +
-  scale_x_continuous(limits = c(-0.1, 2.1))
-
-ggsave(here::here("outputs", "path_base.png"), path_base)
-
-path_r2_plot <- tibble(type = c("primary", "modelled", "overlap"),
-       x1 = c(0.85, 0.85, 0.85),
-       y1 = c(0.75, -0.75, 0))
-```
-
-```{r path-global}
-var_oi <- c("VarDHI", "MinDHI", "CumDHI")
-#var_oi <- c("VarDHI", "MinDHI")
-
-var_out <- varpart(dhi_tab %>%
-                     select(any_of(var_oi)),
-                   primary, modelled)
-
-# plot(var_out, Xnames = c("Primary", "Modelled"))
-
-path_r2 <- var_out$part$indfract %>%
-      mutate(type = c("primary", "modelled", "overlap", "residual")) %>%
-      select(type, Adj.R.squared) %>%
-  tibble()
-
-path_r2_clean <- path_r2_plot %>%
-  left_join(path_r2) %>%
-  mutate(r2_label = paste0(round(Adj.R.squared * 100, 2), "%"))
-
-path_base +
-  geom_text(data = path_r2_clean, aes(label = r2_label))
-
-ggsave(here::here("outputs", "path_venn.png"),
-        height = 3.5, width = 10)
 ```
@@
 
-```{r cum-min-hexplot}
-bec_df %>% 
-  select(bec, CumDHI, MinDHI) %>%
-  left_join(keys$bec, by = c("bec" = "zone")) %>%
-  ggplot(aes(x = CumDHI, y = MinDHI)) +
-  geom_hex() +
-  facet_wrap(~bec)
-```
-
 ```{r eig-anova}
@@
 axis_varexp %>%
-  filter(anova_sig != "NS") %>%
-  mutate(category = ifelse(strata %in% forest_types, "Forest Types", "BEC Zones")) %>%
-  # left_join(keys$bec, by = c("bec" = "zone")) %>%
-  ggplot(aes(x = axis, y = var_exp, col = category, shape = category)) +
-  geom_jitter(height = 0, width = 0.05) +
-  theme(legend.position = "bottom") +
-  labs(x = NULL,
-       y = "% of Variance Explained",
-       col = NULL,
-       shape = NULL) +
-  scale_y_continuous(labels = scales::label_percent())
-# no secondary axis in alpine & coast.
-
-axis_varexp %>%
   group_by(strata) %>%
@@
 
-```{r varexp-coniferous}
+```{r varexp}
 rda_varexp <- outs %>%
@@
           desc(varexp))
-
-source(here::here("scripts", "get_bec_proportion_vlce.R"))
-
-prop_con <- bec_normal %>% 
-  select(bec, category) %>% 
-  group_by(bec) %>% 
-  summarize(per_sample_for = mean(category == "Coniferous")) %>% 
-  left_join(per_cons %>% 
-              select(-value)) %>% 
-  mutate(for_dif = per_sample_for - per_con) %>%
-  arrange(desc(abs(for_dif))) %>%
-  select(-for_dif) %>%
-  pivot_longer(starts_with("per"))
-
-con_var_plot <- prop_con %>%
-  filter(name == "per_con") %>%
-  left_join(rda_varexp, by = c("bec" = "strata")) %>% 
-  left_join(keys$bec, by = c("bec" = "zone")) %>%
-  mutate(name = ifelse(name == "per_con", 
-                       "Global Percent Coniferous", 
-                       "Sample Percent Coniferous")) %>%
-  ggplot(aes(x = value, y = varexp)) +
-  geom_point(aes(col = group)) +
-  geom_smooth(method = "lm") +
-  scale_x_continuous(labels = scales::label_percent()) +
-  scale_y_continuous(labels = scales::label_percent()) +
-  labs(x = "Percent Coniferous",
-       y = "Proportion of variance explained",
-       col = NULL) +
-  theme(legend.position = c(0.1, 0.85))
-
-ggsave(here::here("outputs", "con_var.png"), plot = con_var_plot)
 ```
@@
 # when a second axis is present, it has a weak loading on the cumulative DHI, as compared to the other DHIs
-# while the majoirty of the variability is contained within all three DHIs together, the second axis, when present, is more linked to the minmimum and variation DHI.
-
-sig_loads %>%
-  select(strata, var, axis, axis_loading, loading_sig) %>%
-  filter(endsWith(var, "DHI"),
-         loading_sig != "NS") %>%
-  ggplot(aes(x = var, y = abs(axis_loading), group = strata)) +
-  geom_point(aes(col = strata)) +
-  geom_line(aes(col = strata), alpha = 0.5, lty = "dashed") +
-  geom_hline(yintercept = 0, col = "red", alpha = 0.2) +
-  facet_wrap(~axis) +
-  theme_bw() +
-  theme(panel.grid = element_blank()) +
-  labs(x = NULL,
-       y = "Axis Loading")
-
-sig_loads %>%
-  select(strata, var, axis, axis_loading) %>%
-  filter(!endsWith(var, "DHI")) %>%
-  left_join(keys$continuous, by = c("var" = "variable")) %>%
-  ggplot(aes(x = var_long, y = abs(axis_loading), group = strata)) +
-  geom_point(aes(col = strata)) +
-  geom_line(aes(col = strata), alpha = 0.5, lty = "dashed") +
-  geom_hline(yintercept = 0, col = "red", alpha = 0.2) +
-  facet_wrap(~axis) +
-  theme(axis.text.x = element_text(angle = 45,
-                                   hjust = 1)) +
-  labs(x = NULL,
-       y = "Axis Loading (Absolute Value)")
-
-sig_loads %>%
-  filter(!endsWith(var, "DHI"),
-         loading_sig != "NS") %>%
-  group_by(strata, axis) %>%
-  mutate(loading_rank = data.table::frankv(abs(axis_loading), 
-                                           order = -1)) %>%
-  ggplot(aes(x = fct_reorder(var, loading_rank),
-             y = loading_rank, col = strata)) +
-  geom_jitter(height = 0) +
-  facet_wrap(~axis) +
-  theme(axis.text.x = element_text(angle = 45, 
-                                   hjust = 1)) +
-  labs(x = NULL,
-       y = "Axis Loading Rank")
-
-# need to remove the correlations for variables that arent actually involved
-# percentage_first_returns_above_mean, elev_mean, elev_stddev
-out_loadings %>% count(var)
+# while the majority of the variability is contained within all three DHIs together, the second axis, when present, is more linked to the minmimum and variation DHI.
 
@@
 
-##### run to here for analysis to be done
-
-radar chart based on groups, not hierarchical clustering
-
-```{r radar}
-gg_color_hue <- function(n) {
-  hues = seq(15, 375, length = n + 1)
-  hcl(h = hues, l = 65, c = 100)[1:n]
-}
-
-n = 8
-cols = c4a(palette = "rainbow", n = 8)
-
-radar_loads <- clean_loads %>%
-  pivot_wider(names_from = var, 
-              values_from = axis_loading) %>%
-  left_join(keys$bec, by = c("strata" = "zone")) %>%
-  mutate(group = ifelse(is.na(group), "Forest Types", group),
-         group = ifelse(strata == "British Columbia", "British Columbia", group),
-         group = ifelse(axis == "RDA2", "RDA2", group)) %>%
-  group_by(group) %>%
-  summarize(across(CumDHI:elev_p95, \(x) mean(x, na.rm = T))) %>%
-  mutate(group = fct_relevel(group, "RDA2", after = Inf),
-         group = fct_relevel(group, "Forest Types", after = 1))
-
-radar_labels <- radar_loads %>%
-  select(-group, -ends_with("DHI")) %>%
-  colnames() %>%
-  tibble(variable = .) %>%
-  left_join(keys$continuous) %>%
-  pull(var_long)
-
-in_radar <- radar_loads %>%
-  select(!ends_with("DHI")) %>%
-  replace(is.na(.), 0) %>%
-  ggradar(axis.labels = radar_labels,
-          legend.title = "Cluster",
-          values.radar = c("", "", 1),
-          base.size = 1,
-          background.circle.transparency = 0,
-          group.point.size = 2,
-          group.colours = cols,
-          axis.label.size = 3,
-          plot.extent.x.sf = 1.5,
-          group.line.width = 1) +
-  theme(legend.position = "none")
-
-out_radar <- radar_loads %>%
-  select(group, ends_with("DHI")) %>%
-  replace(is.na(.), 0) %>%
-
-  ggradar(legend.title = "Cluster",
-          values.radar = c("", "", 1),
-          base.size = 1,
-          background.circle.transparency = 0,
-          group.point.size = 2,
-          group.colours = cols,
-          axis.label.size = 3,
-          plot.extent.x.sf = 1.5,
-          group.line.width = 1) +
-  theme(legend.position = "none")
-
-boxplots <- clean_loads %>% 
-  left_join(keys$continuous, by = c("var" = "variable")) %>%
-  mutate(var_long = ifelse(is.na(var_long), var, var_long)) %>%
-  select(!(divide_by:unit)) %>%
-  left_join(keys$bec, by = c("strata" = "zone")) %>%
-  mutate(group = ifelse(is.na(group), "Forest Types", group),
-         group = ifelse(axis == "RDA2", "RDA2", group),
-         group = ifelse(strata == "British Columbia", "British Columbia", group),
-         inout = ifelse(endsWith(var, "DHI"), "Response", "Predictor")) %>%
-  mutate(group = fct_relevel(group, "RDA2", after = Inf),
-         group = fct_relevel(group, "Forest Types", after = 1)) %>%
-  ggplot(aes(x = abs(axis_loading), y = fct_rev(fct_reorder(var_long, abs(axis_loading))))) +
-  geom_boxplot(outlier.colour = "#00000000", width = .5) +
-  geom_jitter(height = 0.1, aes(col = group, shape = group)) +
-  facet_grid(cols = vars(axis),
-             rows = vars(inout),
-             scales = "free_y",
-             space = "free_y") +
-  labs(x = "Axis Loading (absolute)",
-       y = NULL,
-       col = NULL,
-       shape = NULL) +
-  theme(panel.spacing.y = unit(0, "lines"),
-        legend.position = "right") +
-  scale_shape_manual(values = c(17, 15, 16, 16, 16, 16, 16, 16)) +
-  scale_colour_manual(values = cols)
-
-radar_boxplot <- ((in_radar / out_radar) | boxplots)  +
-  plot_annotation(tag_levels = "A") +
-  plot_layout(guides = "collect") &
-  theme(plot.tag = element_text(size = 12))
-
-ggsave(here::here("outputs", "radar_boxplot.png"),
-       plot = radar_boxplot,
-       device = "png",
-       height = 6,
-       width = 8)
-```
+# run to here for analysis to be done
 
@@
 }
-```
-
-```{r complexity-iqr}
+
+file.copy(here::here("outputs", "drawio", "British Columbia.drawio"), here::here("outputs", "British Columbia.drawio"), overwrite = T)
+```
+
+Climate Table (Table 2)
+
+```{r bec-clim-tab}
+climate_folder <- here::here("data", "climate")
+clim_vars <- climate_folder %>%
+  list.files(pattern = "csv$") %>%
+  tools::file_path_sans_ext()
+
+clim_vars <- clim_vars[!str_detect(clim_vars, "bec")]
+
+bec_clim <- climate_folder %>%
+  list.files(pattern = "bec.csv$",
+             full.names = T) %>%
+  read_csv() %>%
+  mutate(var = clim_vars %>% rep(16) %>% sort) %>%
+  select(bec = ZONE, mean = MEAN, var) %>%
+  pivot_wider(values_from = mean,
+              names_from = var)
+
+bec_tab <- keys$bec %>%
+  left_join(bec_clim, by = c("zone" = "bec")) %>%
+  select(-order) %>%
+  arrange(group, zone) %>%
+  relocate(group) %>%
+  mutate(across(ppt:tmin, \(x) round(x, digits = 1))) %>%
+  left_join(strata_nested %>% unnest(data) %>% count(strata),
+            by = c("zone" = "strata"))
+
+write_csv(bec_tab, here::here("data", "climate.csv"))
+
+knitr::kable(bec_tab)
+```
+
+# rgb map
+
+```{r}
+bec_dissolve <- read_sf(here::here("data", "shapefiles", "bec_dissolve.shp")) %>%
+  janitor::clean_names() %>%
+  select(zone)
+
+# normalized to max
+loadings_rgb <- clean_loads_abs %>% 
+  group_by(var) %>%
+  mutate(axis_loading = axis_loading / max(axis_loading)) %>%
+  group_by(axis) %>%
+  pivot_wider(names_from = var,
+              values_from = axis_loading) %>%
+  ungroup() %>%
+  replace(is.na(.), 0) %>%
+  #mutate(across(elev_cv:elev_p95, \(x) ifelse(x > 0.5, 0.5, x))) %>%
+  mutate(dhi_rgb = rgb(CumDHI, VarDHI, MinDHI),
+         pri_rgb = rgb(elev_p95, percentage_first_returns_above_2m, elev_cv, maxColorValue = 1),
+         mod_rgb = rgb(basal_area, total_biomass, gross_stem_volume, maxColorValue = 1)) %>%
+  #select(strata, axis, percentage_first_returns_above_2m, elev_p95, elev_cv, ends_with("rgb")) %>%
+  pivot_longer(ends_with("rgb"))
+
+rgb_loads_split <- loadings_rgb %>%
+  filter(strata %in% zones) %>%
+  group_by(axis, name) %>%
+  group_split()
+```
+
+```{r}
+make_rgb_map <- function(tibble) {
+
+  input <- bec_dissolve %>%
+    left_join(tibble, by = c("zone" = "strata")) %>%
+    mutate(value = ifelse(is.na(value), "#000000", value))
+
+  rgb_cols = input %>%
+    st_drop_geometry() %>%
+    pull(value) %>%
+    unique() %>%
+    sort()
+
+  input %>%
+    ggplot(aes(x = zone, y = 2, fill = value)) +
+    geom_col() +
+    scale_fill_manual(values = rgb_cols)
+
+   input %>%
+    ggplot() +
+    geom_sf(aes(geometry = geometry, fill = value, col = value), linewidth = 0.1) +
+    scale_fill_manual(values = rgb_cols) +
+    scale_color_manual(values = rgb_cols) +
+    theme_void() +
+    theme(legend.position = "none")
+}
+
+rgb_maps <- map(rgb_loads_split, make_rgb_map)
+
+cowplot_rgb <- cowplot::plot_grid(rgb_maps[[3]], rgb_maps[[2]],
+                                  rgb_maps[[6]], rgb_maps[[5]],
+                                  labels = c("A - Primary RDA1", "B - Modelled RDA1",
+                                             "C - Primary RDA2", "D - Modelled RDA2"),
+                                  vjust = 1.25)
+
+ggsave(here::here("outputs", "rgbplots_cow.png"), cowplot_rgb, height = 7, width = 8, bg = "white")
+
+```
+
+# flip cum to positive
+
+```{r}
+flip_cum <- function(data) {
+  if(data %>% filter(var == "CumDHI") %>% nrow() == 0) {
+    return(data)
+  }
+  flip <- data %>%
+    filter(var == "CumDHI") %>%
+    pull(axis_loading) %>%
+    sign()
+
+  data %>% mutate(axis_loading = axis_loading * flip)
+}
+
+clean_loads_cum_pos <- clean_loads_raw %>%
+  group_by(strata, axis) %>%
+  group_split() %>%
+  map_dfr(., .f = flip_cum)
+
+clean_loads_cum_pos %>%
+  filter(strata %in% zones) %>%
+  count(var, axis, sign = sign(axis_loading)) %>%
+  pivot_wider(names_from = sign, values_from = n) %>%
+  arrange(axis)
+
+clean_loads_joined <- clean_loads_cum_pos %>%
+  left_join(keys$bec %>%
+              select(strata = zone, group)) %>%
+  left_join(keys$continuous %>%
+              select(var = variable, var_long)) %>%
+  mutate(
+    group = ifelse(is.na(group), strata, group),
+    var_long = ifelse(is.na(var_long), var, var_long),
+    pos_neg = ifelse(axis_loading > 0, "Positive", "Negative")
+  )
+
+```
+
+Parcoord Noflip (Figure)
+
+```{r}
+data_parcoord <- clean_loads_joined %>%
+  # mutate(axis_loading = abs(axis_loading)) %>%
+  group_by(group, var_long, axis) %>%
+  summarize(axis_loading = mean(axis_loading)) %>%
+  ungroup() %>%
+  complete(group, var_long, axis, fill = list(axis_loading = NA)) %>%
+  mutate(inout = ifelse(endsWith(var_long, "DHI"), "Response", "Predictor"), 
+         group = fct_relevel(group, "British Columbia", 
+                             "Broadleaf", 
+                             "Coniferous", 
+                             "Mixed Wood", 
+                             "Wetland-Treed", 
+                             after = 0),
+         var_long = ifelse(endsWith(var_long, "DHI"), var_long, str_replace_all(var_long, " ", "\n")),
+         var_long = fct_relevel(var_long, 
+                                c("Structural\nComplexity", "Canopy\nHeight", "Canopy\nCover",
+                                "Basal\nArea", "Aboveground\nBiomass", "Gross\nStem\nVolume",
+                                "Cumulative DHI", "Variation DHI", "Minimum DHI")))
+
+# nojitter/lines
+parcoord_noflip <- data_parcoord %>%
+  ggplot(aes(x = var_long, y = axis_loading)) +
+  #geom_boxplot(col = "#00000025", outlier.alpha = 0) +
+
+  geom_line(aes(col = group, group = group), lty = "dashed") +
+  geom_point(aes(col = group, group = group, shape = group, fill = group)) +
+  geom_hline(yintercept = 0, lty = "dotted", col = "grey", linewidth = 0.5) +
+  facet_wrap(~paste(axis, inout),
+             scale = "free") +
+  #theme_bw() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
+  labs(x = NULL,
+       y = "Axis Loading",
+       col = NULL,
+       shape = NULL,
+       fill = NULL) +
+  # coord_flip() +
+  scale_colour_discrete_c4a_cat(palette = "bold") +
+  scale_fill_discrete_c4a_cat(palette = "bold") +
+  scale_shape_manual(values = c(21, 22, 22, 22, 22, 25, 25, 25, 25, 25))
+
+parcoord_noflip
+
+ggsave(here::here("outputs", "parcoord_noflip.png"), plot = parcoord_noflip,
+       height = 6, width = 8)
+
+# ggsave(here::here("outputs", "parcoord_noflip.svg"), plot = parcoord_noflip,
+#        height = 6, width = 8)
+```
+
+# meg scatterplots
+
+```{r}
+scatter <- bec_normal %>%
+  mutate(label = case_when(bec %in% c("CWH", "MS") ~ bec,
+                           T ~ "Other"))
+
+ggplot(data = scatter, aes(col = label), alpha = 0.25) +
+  geom_point(aes(x = percentage_first_returns_above_2m, y = CumDHI)) +
+  geom_density(aes(x = percentage_first_returns_above_2m)) +
+  geom_density(aes(y = CumDHI)) +
+  scale_fill_manual(values = c("grey", "red", "blue")) +
+  scale_colour_manual(values = c("grey", "red", "blue"))
+
+ggplot() +
+  geom_point(data = bec_df, aes(x = percentage_first_returns_above_2m, y = CumDHI), alpha = 0.5) +
+  geom_point(data = bec_df %>%
+               filter(bec == "CWH"), aes(x = percentage_first_returns_above_2m, y = CumDHI), col = "orange") +
+  geom_point(data = bec_df %>%
+               filter(bec == "MS"), aes(x = percentage_first_returns_above_2m, y = CumDHI), col = "red") +
+  geom_histogram(data = bec_df, aes(x = percentage_first_returns_above_2m)) +
+  geom_histogram(data = bec_df, aes(y = CumDHI))
+```
+
+```{r}
+fracts_out
+
+strata_nested %>% 
+  unnest(cols = data) %>%
+  select(-bec, -category) %>%
+  left_join(keys$bec, by = c("strata" = "zone")) %>%
+  mutate(group = ifelse(is.na(group), "Forest Type", group)) %>%
+  ggplot() +
+  geom_boxplot(aes(x = fct_reorder(.f = strata, .x = MinDHI, .fun = mean), y = MinDHI, fill = group))
+```
+
+```{r}
+strata_nested %>% 
+  unnest(cols = data) %>%
+  select(-bec, -category) %>%
+  group_by(strata) %>%
+  summarize(across(basal_area:VarDHI, .fns = sd)) %>%
+  left_join(keys$bec, by = c("strata" = "zone")) %>%
+  mutate(group = ifelse(is.na(group), "Forest Type", group)) %>%
+  select(strata, group, basal_area:VarDHI) %>%
+  pivot_longer(-c(strata, group)) %>%
+  left_join(fracts_out) %>%
+  mutate(varexp = primary + modelled + overlap) %>%
+  ggplot(aes(x = value, y = varexp, col = group)) +
+  geom_point() +
+  facet_wrap(~name, strip.position = "bottom") +
+  labs(x = NULL,
+       y = "Variation Explained",
+       title = "x axis is sd of variable") +
+  theme(strip.placement = "outside",
+        strip.background = element_blank())
+```
+
+Supplemental information Hexplot template
+
+```{r}
+bec_df %>%
+  left_join(fracts_out %>%
+              mutate(all = (primary + modelled + overlap) * 100) %>%
+              select(strata, all), by = c("bec" = "strata")) %>%
+  mutate(facet = paste0(bec, " (", round(all, 2), "%)")) %>%
+  ggplot(aes(x = percentage_first_returns_above_2m / 100, y = CumDHI)) +
+  geom_hex(aes(fill = after_stat(log(count)))) +
+  labs(x = "Canopy Cover (%)") +
+  facet_wrap(~fct_reorder(facet, all)) +
+  theme(legend.position = "bottom")
+
+ggsave(here::here("outputs", "supplement_hex.png"),
+       height = 6, width = 8)
+```
+
+IQR sorted plot of canopy cover
+
+```{r iqr-plot}
 strata_nested %>% 
   unnest(data) %>% 
-  ggplot(aes(x = fct_reorder(strata, elev_cv, IQR), y = elev_cv)) + 
+  ggplot(aes(x = fct_reorder(strata, percentage_first_returns_above_2m, IQR), y = percentage_first_returns_above_2m)) + 
   geom_boxplot() + 
@@
 ```
+
+Study Area Map (Figure 1)
 
@@
 ```
-
-```{r bec-clim-tab}
-climate_folder <- here::here("data", "climate")
-clim_vars <- climate_folder %>%
-  list.files(pattern = "csv$") %>%
-  tools::file_path_sans_ext()
-
-clim_vars <- clim_vars[!str_detect(clim_vars, "bec")]
-
-bec_clim <- climate_folder %>%
-  list.files(pattern = "bec.csv$",
-             full.names = T) %>%
-  read_csv() %>%
-  mutate(var = clim_vars %>% rep(16) %>% sort) %>%
-  select(bec = ZONE, mean = MEAN, var) %>%
-  pivot_wider(values_from = mean,
-              names_from = var)
-
-bec_tab <- keys$bec %>%
-  left_join(bec_clim, by = c("zone" = "bec")) %>%
-  select(-order) %>%
-  arrange(group, zone) %>%
-  relocate(group) %>%
-  mutate(across(ppt:tmin, \(x) round(x, digits = 1)))
-
-write_csv(bec_tab, here::here("data", "climate.csv"))
-
-knitr::kable(bec_tab)
-```
-
-```{r}
-bec_normal %>%
-  select(-elev_stddev,-loreys_height,-percentage_first_returns_above_mean,-elev_mean) %>%
-  left_join(keys$bec, by = c("bec" = "zone")) %>%
-  pivot_longer(basal_area:VarDHI) %>%
-  ggplot(aes(x = value, y = bec, fill = group)) +
-  geom_boxplot() +
-  facet_wrap( ~ name)
-```
-
-# new radar loads
-
-```{r}
-clean_loads %>%
-  pivot_wider(names_from = var, 
-              values_from = axis_loading) %>%
-  left_join(keys$bec, by = c("strata" = "zone")) %>%
-  mutate(group = ifelse(is.na(group), strata, group),
-         group = ifelse(strata == "British Columbia", "British Columbia", group)) %>%
-  group_by(group, axis) %>%
-  summarize(across(CumDHI:elev_p95, \(x) mean(x, na.rm = T))) %>%
-  pivot_longer(CumDHI:elev_p95) %>%
-  mutate(value = ifelse(is.na(value), 0, value)) %>%
-  left_join(keys$continuous, by = c("name" = "variable")) %>%
-  mutate(var_long = ifelse(is.na(var_long), name, var_long),
-         order = endsWith(var_long, "DHI")) %>%
-  ggplot(aes(x = fct_reorder(var_long, value, .desc = T), y = value)) +
-  geom_col() +
-  #coord_polar() +
-  coord_flip() +
-  facet_grid(cols = vars(group), rows = vars(axis)) +
-  labs(x = "Variable",
-       y = "Axis Loading (absolute)")
-
-clean_loads %>%
-  left_join(keys$bec, by = c("strata" = "zone")) %>%
-  mutate(group = ifelse(is.na(group), strata, group),
-         group = ifelse(strata == "British Columbia", "British Columbia", group)) %>%
-  ggplot(aes(x = var, y = axis_loading, col = group)) +
-  geom_line()
-```
-
-```{r}
-clean_loads %>%
-  left_join(keys$bec, by = c("strata" = "zone")) %>%
-  mutate(group = ifelse(is.na(group), strata, group),
-         group = ifelse(strata == "British Columbia", "British Columbia", group)) %>%
-  pivot_wider(names_from = var, values_from = axis_loading) %>%
-  select(-zone_nm, -order) %>%
-  replace(is.na(.), 0) %>%
-  GGally::ggparcoord(columns = 4:12,
-                     groupColumn = "group",
-                     scale = "globalminmax") +
-  facet_wrap(~axis) +
-  coord_flip()
-```
-
-# parallel coordinate plots
-
-```{r}
-data <- clean_loads_abs %>%
-  left_join(keys$bec %>%
-              select(strata = zone, group)) %>%
-  left_join(keys$continuous %>%
-              select(var = variable, var_long)) %>%
-  mutate(group = ifelse(is.na(group), strata, group),
-         var_long = ifelse(is.na(var_long), var, var_long)) %>%
-  group_by(group, var_long, axis) %>%
-  summarize(axis_loading = mean(axis_loading)) %>%
-  ungroup() %>%
-  complete(group, var_long, axis, fill = list(axis_loading = NA)) %>%
-  mutate(inout = ifelse(endsWith(var_long, "DHI"), "Response", "Predictor"), 
-         group = fct_relevel(group, "British Columbia", 
-                             "Broadleaf", 
-                             "Coniferous", 
-                             "Mixed Wood", 
-                             "Wetland-Treed", 
-                             after = 0),
-         var_long = fct_relevel(var_long, 
-                                c("Canopy Height", "Canopy Cover", "Structural Complexity",
-                                "Basal Area", "Aboveground Biomass", "Gross Stem Volume",
-                                "Cumulative DHI", "Variation DHI", "Minimum DHI")))
-
-# nojitter/lines
-parcoord_noflip <- data %>%
-  ggplot(aes(x = var_long, y = axis_loading)) +
-  #geom_boxplot(col = "#00000025", outlier.alpha = 0) +
-
-  geom_line(aes(col = group, group = group), lty = "dashed") +
-  geom_point(aes(col = group, group = group, shape = group, fill = group)) +
-  # geom_hline(yintercept = 0, lty = "dashed", col = "red") +
-  facet_wrap(~paste(axis, inout),
-             scale = "free") +
-  #theme_bw() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
-  labs(x = NULL,
-       y = "Axis Loading (absolute)",
-       col = NULL,
-       shape = NULL,
-       fill = NULL) +
-  # coord_flip() +
-  scale_colour_discrete_c4a_cat(palette = "bold") +
-  scale_fill_discrete_c4a_cat(palette = "bold") +
-  scale_shape_manual(values = c(21, 22, 22, 22, 22, 25, 25, 25, 25, 25))
-
-parcoord_noflip
-
-ggsave(here::here("outputs", "parcoord_noflip.png"), plot = parcoord_noflip,
-       height = 6, width = 8)
-
-ggsave(here::here("outputs", "parcoord_noflip.svg"), plot = parcoord_noflip,
-       height = 6, width = 8)
-```
-
-# rgb map
-
-```{r}
-bec_dissolve <- read_sf(here::here("data", "shapefiles", "bec_dissolve.shp")) %>%
-  janitor::clean_names() %>%
-  select(zone)
-
-# non normalized
-loadings_rgb <- clean_loads %>% 
-  group_by(axis) %>%
-  pivot_wider(names_from = var,
-              values_from = axis_loading) %>%
-  ungroup() %>%
-  replace(is.na(.), 0) %>%
-  mutate(across(elev_cv:elev_p95, \(x) ifelse(x > 0.5, 0.5, x))) %>%
-  mutate(dhi_rgb = rgb(CumDHI, VarDHI, MinDHI),
-         pri_rgb = rgb(elev_p95, percentage_first_returns_above_2m, elev_cv, maxColorValue = 1),
-         mod_rgb = rgb(basal_area, total_biomass, gross_stem_volume, maxColorValue = 1)) %>%
-  select(strata, axis, percentage_first_returns_above_2m, elev_p95, elev_cv, ends_with("rgb")) %>%
-  pivot_longer(ends_with("rgb"))
-
-# normalized to max
-loadings_rgb <- clean_loads %>% 
-  group_by(var) %>%
-  mutate(axis_loading = axis_loading / max(axis_loading)) %>%
-  group_by(axis) %>%
-  pivot_wider(names_from = var,
-              values_from = axis_loading) %>%
-  ungroup() %>%
-  replace(is.na(.), 0) %>%
-  #mutate(across(elev_cv:elev_p95, \(x) ifelse(x > 0.5, 0.5, x))) %>%
-  mutate(dhi_rgb = rgb(CumDHI, VarDHI, MinDHI),
-         pri_rgb = rgb(elev_p95, percentage_first_returns_above_2m, elev_cv, maxColorValue = 1),
-         mod_rgb = rgb(basal_area, total_biomass, gross_stem_volume, maxColorValue = 1)) %>%
-  #select(strata, axis, percentage_first_returns_above_2m, elev_p95, elev_cv, ends_with("rgb")) %>%
-  pivot_longer(ends_with("rgb"))
-
-becs <- bec_normal %>%
-  pull(bec) %>%
-  unique()
-
-rgb_loads_split <- loadings_rgb %>%
-  filter(strata %in% becs) %>%
-  group_by(axis, name) %>%
-  group_split()
-```
-
-```{r}
-make_rgb_map <- function(tibble) {
-
-  input <- bec_dissolve %>%
-    left_join(tibble, by = c("zone" = "strata")) %>%
-    mutate(value = ifelse(is.na(value), "#000000", value))
-
-  rgb_cols = input %>%
-    st_drop_geometry() %>%
-    pull(value) %>%
-    unique() %>%
-    sort()
-
-  input %>%
-    ggplot(aes(x = zone, y = 2, fill = value)) +
-    geom_col() +
-    scale_fill_manual(values = rgb_cols)
-
-   input %>%
-    ggplot() +
-    geom_sf(aes(geometry = geometry, fill = value, col = value), linewidth = 0.1) +
-    scale_fill_manual(values = rgb_cols) +
-    scale_color_manual(values = rgb_cols) +
-    theme_void() +
-    theme(legend.position = "none")
-}
-
-rgb_maps <- map(rgb_loads_split, make_rgb_map)
-
-labels = c("DHI", "B - Modelled RDA1", "A - Primary RDA1", "DHI", "D - Modelled RDA2", "C - Primary RDA2")
-
-# for (i in 1:length(rgb_maps)) {
-#   print(i)
-#   gg <- rgb_maps[[i]] + labs(title = labels[[i]])
-#   
-#   ggsave(here::here("outputs", "black_plots", paste(i, ".png")),
-#          plot = gg,
-#          height = 3,
-#          width = 3)
-# }
-
-ggsave(here::here("outputs", "test_black.png"), rgb_maps[[6]])
-
-patchwork_rgbplots <- (rgb_maps[[3]] + rgb_maps[[2]]) / 
-                      (rgb_maps[[6]] + rgb_maps[[5]]) &
-  plot_annotation(tag_levels = "A")
-
-cowplot_rgb <- cowplot::plot_grid(rgb_maps[[3]], rgb_maps[[2]],
-                                  rgb_maps[[6]], rgb_maps[[5]],
-                                  labels = c("A - Primary RDA1", "B - Modelled RDA1",
-                                             "C - Primary RDA2", "D - Modelled RDA2"),
-                                  vjust = 1.25)
-
-ggsave(here::here("outputs", "rgbplots_cow.png"), cowplot_rgb, height = 7, width = 8, bg = "white")
-
-# ggsave(here::here("outputs", "rgbplots_norm.png"), patchwork_rgbplots,
-#        height = 6,
-#        width = 8)
-# 
-# print("png saved")
-
-# ggsave(here::here("outputs", "rgbplots_norm.pdf"), patchwork_rgbplots,
-#        height = 6,
-#        width = 8)
-# 
-# print("pdf saved")
-```

### 2024-01-10 scripts/manuscript/manuscript.qmd
began revisions round 2, cleaned 4_vegan to be easier to work with (less defunct extra code)

@@
 nocite: |
-  @matasci2018, @matasci2018b
+  @R-ggplot2, @sf2023
 ---
@@
 structural attributes and modelled forest structure to explaining
-ecosystem function. Third, we examine the global applicability of these
-results by evaluating the consistency of the observed forest
-structure-function relationships across a broad range of environment
-gradients and forest types.
+ecosystem function. Third, we examine the potential geographic
+extensibility of these results by evaluating the consistency of the
+observed forest structure-function relationships across a broad range of
+environment gradients and forest types.
 
@@
 #| label: fig-study-map
-#| fig-cap: Location of Biogeoclimatic Ecosystem Classification (BEC) zones in British Columbia.
+#| fig-cap: Location of Biogeoclimatic Ecosystem Classification (BEC) zones in British Columbia. Map was generated in R version 4.3.1 using the ggplot2 version 2.1.0 and sf version 1.0-14 packages.
 
@@
     tmax = "Average Monthly Max Temperature (°C)",
-    tmin = "Average Monthly Min Temperature (°C)"
+    tmin = "Average Monthly Min Temperature (°C)",
+    n = "n"
   ) %>%
@@
 [@matasci2018; @matasci2018b]. Accuracy metrics for imputed forest
-structural attributes range from an RMSE of 24.5% (structural
+structural attributes range from an RMSE of 29.7% (structural
 complexity) to 82.3% (gross stem volume) and *R^2^* ranging from 0.125
@@
 the forested ecosystems of Canada [@hermosilla2016]. A maximum of 500
-sample units were selected from each stratum; in the many strata where
-this was not achieved, all sampling units meeting the above restrictions
-were selected. Sampling was conducted in R version 4.2.2 [@R-base] using
-the **sgsR** package [@R-sgsR]. Neighbourhood analyses for the land
-cover classes and coefficient of variations of canopy height and cover
-were calculated in Python version 3.9.
+sample units were selected from each stratum; in the strata where this
+was not achieved, all sampling units meeting the above restrictions were
+selected. Sampling was conducted in R version 4.2.2 [@R-base] using the
+**sgsR** package [@R-sgsR]. Neighbourhood analyses for the land cover
+classes and coefficient of variations of canopy height and cover were
+calculated in Python version 3.9.
 
@@
 Annotated colour wheels for each set of variables also shown.
+Maps were generated in R version 4.3.1 using the ggplot2 version 2.1.0 and sf version 1.0-14 packages.
 
@@
 also lead to lower amounts of variation explained by these metrics,
-leading to them being less informative in the loadings, and other
-metrics, such as structural complexity having higher amounts of variance
-explained (@fig-radar). These complex linkages between climate, forest
+leading to them having smaller loadings, and other metrics with higher
+variation in their values, such as structural complexity, having larger
+loadings (@fig-radar). These complex linkages between climate, forest
 structure, and forest productivity become further complicated by the

### 2024-01-24 scripts/4_vegan_clean.qmd
third submission

@@
 
-bec_normal <- bec_df %>%
-  mutate(across(basal_area:VarDHI,  .fns = ~ normal(.x))) 
-```
-
-```{r bec-nest}
-to_nest <- bec_normal %>%
+set.seed(29640)
+
+raw_sliced <- bec_df %>%
   select(-percentage_first_returns_above_mean,
@@
          -elev_stddev,
-         -elev_mean)
-
-bec_nested <- to_nest %>%
+         -elev_mean) %>%
+  group_by(bec) %>%
+  slice_sample(n = 500) %>%
+  ungroup()
+
+to_nest_normal <- raw_sliced %>% 
+  mutate(across(basal_area:VarDHI,  .fns = ~ normal(.x)))
+```
+
+```{r bec-nest}
+# already slice_sampled
+bec_nested <- to_nest_normal %>%
   group_by(bec) %>%
@@
 
-class_nested <- to_nest %>%
+# sample down to 500 here
+class_nested <- to_nest_normal %>%
   group_by(category) %>%
+  slice_sample(n = 500) %>%
   nest() %>%
@@
 
-set.seed(69420)
-
-strata_nested <- bind_rows(bec_nested, class_nested) %>%
-  mutate(data = map(data, slice_sample, n = 500))
+strata_nested <- bind_rows(bec_nested, class_nested)
 # this slice sample is the ANALYTICAL difference between 4_vegan and 
@@
 
-global <- strata_nested %>%
-  filter(strata %in% zones) %>%
-  unnest(data)
+global <- to_nest_normal
 ```
@@
   bind_cols(global) %>%
-  select(-strata,-category) %>%
+  select(-bec,-category) %>%
   as.matrix() %>%
@@
   relocate(axis)
-```
-
-```{r}
-# linear relationships * homoschedasticity check
-
-bec_normal %>%
-  pivot_longer(cols = ends_with("DHI"),
-               names_to = "dhi",
-               values_to = "dhi_val") %>%
-  pivot_longer(
-    cols = basal_area:total_biomass,
-    names_to = "struct",
-    values_to = "struct_val"
-  ) %>%
-  ggplot(aes(x = struct_val, y = dhi_val)) +
-  geom_hex() +
-  facet_grid(rows = vars(dhi),
-             cols = vars(struct),
-             scales = "free") +
-  labs(x = "Structure Value (Normalized)",
-       y = "DHI Value (Normalized)")
 ```
@@
 
+```{r}
+egg <- anova.cca(rda_out, step = 1000)
+
+outs %>%
+  mutate(anova_sig = map(.x = rda, .f = \(x) {
+    anova.cca(rda, step = 1000)
+  })
+```
+
 ```{r varexp}
@@
 # run to here for analysis to be done
-
-# drawio
-
-```{r drawio}
-
-# make dir
-drawio_dir <- here::here("outputs", "drawio")
-
-dir.create(drawio_dir, showWarnings = F)
-
-# templates
-one_rda <-
-  read_lines(here::here(drawio_dir, "_rda_path-text_replace-RDA1.drawio"))
-
-two_rda <-
-  read_lines(here::here(drawio_dir, "_rda_path-text_replace-RDA2.drawio"))
-
-# list to iterate
-stratum <- clean_loads_raw %>%
-  pull(strata) %>%
-  unique()
-
-# make sure everything gets change in rda fig
-possibles <- clean_loads_raw %>%
-  select(var, axis) %>%
-  complete() %>%
-  mutate(possible = paste(var, axis, sep = "-")) %>%
-  pull(possible) %>%
-  unique()
-
-# fractions dfs
-fract_all <- path_r2 %>% 
-  pivot_wider(names_from = type, values_from = Adj.R.squared) %>% 
-  select(-residual) %>%
-  mutate(strata = "British Columbia") %>% 
-  relocate(strata)
-
-fracts_all <- fracts_out %>%
-  bind_rows(fract_all)
-
-# variance explained by axis
-
-all_varexp <- rda_eig %>%
-  filter(name != "RDA3") %>%
-  mutate(strata = "British Columbia") %>%
-  rename(axis = name) %>%
-  select(-value)
-
-varexp_all <- axis_varexp %>%
-  filter(anova_sig != "NS") %>%
-  select(strata, axis, prop, var_exp) %>%
-  bind_rows(all_varexp)
-
-for (s in stratum) {
-  print(s)
-  strata_tp <- clean_loads_raw %>%
-    filter(strata == s) %>%
-    mutate(textreplace = paste(var, axis, sep = "-"),
-           axis_loading = as.character(round(axis_loading, digits = 2)))
-
-  if (strata_tp %>% pull(axis) %>% unique() %>% length() == 1) {
-    strata_drawio <- one_rda
-  } else {
-    strata_drawio <- two_rda
-  }
-
-  textreplace <- strata_tp %>%
-    pull(textreplace)
-
-  replaceval <- strata_tp %>%
-    pull(axis_loading)
-
-  for (i in 1:length(textreplace)) {
-    text <- textreplace[[i]]
-    val <- replaceval[[i]]
-
-    strata_drawio <- str_replace(strata_drawio, text, val)
-  }
-
-  strata_drawio <- str_replace(strata_drawio, "title_strata", s)
-
-  #strata_drawio <- strata_drawio[!str_detect(strata_drawio, "RDA")]
-  # filter through all possibilities remaining, replace them with NS
-
-  for (p in possibles) {
-    strata_drawio <- str_replace(strata_drawio, p, "NS")
-  }
-
-  strata_fract <- fracts_all %>%
-    filter(strata == s) %>%
-    pivot_longer(primary:overlap) %>%
-    mutate(value = (value * 100) %>%
-             round(1) %>%
-             paste0(., "%"))
-
-  fract_names <- strata_fract %>%
-    pull(name)
-
-  fract_value <- strata_fract %>%
-    pull(value)
-
-  for (i in 1:length(fract_names)) {
-    strata_drawio <-
-      str_replace(strata_drawio, fract_names[[i]], fract_value[[i]])
-  }
-
-  strata_varexp <- varexp_all %>%
-    filter(strata == s) %>%
-    mutate(var_exp = (var_exp * 100) %>%
-             round(1) %>%
-             paste0(., "%"))
-
-  axis_names <- strata_varexp %>%
-    pull(axis) %>%
-    paste("var_exp", ., sep = "-")
-
-  axis_vals <- strata_varexp %>%
-    pull(var_exp)
-
-  for (i in 1:length(axis_names)) {
-    strata_drawio <-
-      str_replace(strata_drawio, axis_names[[i]], axis_vals[[i]])
-  }
-
-  write_lines(strata_drawio, here::here(drawio_dir, paste0(s, ".drawio")))
-}
-
-file.copy(here::here("outputs", "drawio", "British Columbia.drawio"), here::here("outputs", "British Columbia.drawio"), overwrite = T)
-```
-
-Climate Table (Table 2)
-
-```{r bec-clim-tab}
-climate_folder <- here::here("data", "climate")
-clim_vars <- climate_folder %>%
-  list.files(pattern = "csv$") %>%
-  tools::file_path_sans_ext()
-
-clim_vars <- clim_vars[!str_detect(clim_vars, "bec")]
-
-bec_clim <- climate_folder %>%
-  list.files(pattern = "bec.csv$",
-             full.names = T) %>%
-  read_csv() %>%
-  mutate(var = clim_vars %>% rep(16) %>% sort) %>%
-  select(bec = ZONE, mean = MEAN, var) %>%
-  pivot_wider(values_from = mean,
-              names_from = var)
-
-bec_tab <- keys$bec %>%
-  left_join(bec_clim, by = c("zone" = "bec")) %>%
-  select(-order) %>%
-  arrange(group, zone) %>%
-  relocate(group) %>%
-  mutate(across(ppt:tmin, \(x) round(x, digits = 1))) %>%
-  left_join(strata_nested %>% unnest(data) %>% count(strata),
-            by = c("zone" = "strata"))
-
-write_csv(bec_tab, here::here("data", "climate.csv"))
-
-knitr::kable(bec_tab)
-```
-
-# rgb map
-
-```{r}
-bec_dissolve <- read_sf(here::here("data", "shapefiles", "bec_dissolve.shp")) %>%
-  janitor::clean_names() %>%
-  select(zone)
-
-# normalized to max
-loadings_rgb <- clean_loads_abs %>% 
-  group_by(var) %>%
-  mutate(axis_loading = axis_loading / max(axis_loading)) %>%
-  group_by(axis) %>%
-  pivot_wider(names_from = var,
-              values_from = axis_loading) %>%
-  ungroup() %>%
-  replace(is.na(.), 0) %>%
-  #mutate(across(elev_cv:elev_p95, \(x) ifelse(x > 0.5, 0.5, x))) %>%
-  mutate(dhi_rgb = rgb(CumDHI, VarDHI, MinDHI),
-         pri_rgb = rgb(elev_p95, percentage_first_returns_above_2m, elev_cv, maxColorValue = 1),
-         mod_rgb = rgb(basal_area, total_biomass, gross_stem_volume, maxColorValue = 1)) %>%
-  #select(strata, axis, percentage_first_returns_above_2m, elev_p95, elev_cv, ends_with("rgb")) %>%
-  pivot_longer(ends_with("rgb"))
-
-rgb_loads_split <- loadings_rgb %>%
-  filter(strata %in% zones) %>%
-  group_by(axis, name) %>%
-  group_split()
-```
-
-```{r}
-make_rgb_map <- function(tibble) {
-
-  input <- bec_dissolve %>%
-    left_join(tibble, by = c("zone" = "strata")) %>%
-    mutate(value = ifelse(is.na(value), "#000000", value))
-
-  rgb_cols = input %>%
-    st_drop_geometry() %>%
-    pull(value) %>%
-    unique() %>%
-    sort()
-
-  input %>%
-    ggplot(aes(x = zone, y = 2, fill = value)) +
-    geom_col() +
-    scale_fill_manual(values = rgb_cols)
-
-   input %>%
-    ggplot() +
-    geom_sf(aes(geometry = geometry, fill = value, col = value), linewidth = 0.1) +
-    scale_fill_manual(values = rgb_cols) +
-    scale_color_manual(values = rgb_cols) +
-    theme_void() +
-    theme(legend.position = "none")
-}
-
-rgb_maps <- map(rgb_loads_split, make_rgb_map)
-
-cowplot_rgb <- cowplot::plot_grid(rgb_maps[[3]], rgb_maps[[2]],
-                                  rgb_maps[[6]], rgb_maps[[5]],
-                                  labels = c("A - Primary RDA1", "B - Modelled RDA1",
-                                             "C - Primary RDA2", "D - Modelled RDA2"),
-                                  vjust = 1.25)
-
-ggsave(here::here("outputs", "rgbplots_cow.png"), cowplot_rgb, height = 7, width = 8, bg = "white")
-
-```
 
@@
 
+# drawio
+
+```{r drawio}
+
+# make dir
+drawio_dir <- here::here("outputs", "drawio")
+
+dir.create(drawio_dir, showWarnings = F)
+
+# templates
+one_rda <-
+  read_lines(here::here(drawio_dir, "_rda_path-text_replace-RDA1.drawio"))
+
+two_rda <-
+  read_lines(here::here(drawio_dir, "_rda_path-text_replace-RDA2.drawio"))
+
+# list to iterate
+stratum <- clean_loads_cum_pos %>%
+  pull(strata) %>%
+  unique()
+
+# make sure everything gets change in rda fig
+possibles <- clean_loads_cum_pos %>%
+  select(var, axis) %>%
+  complete() %>%
+  mutate(possible = paste(var, axis, sep = "-")) %>%
+  pull(possible) %>%
+  unique()
+
+# fractions dfs
+fract_all <- path_r2 %>% 
+  pivot_wider(names_from = type, values_from = Adj.R.squared) %>% 
+  select(-residual) %>%
+  mutate(strata = "British Columbia") %>% 
+  relocate(strata)
+
+fracts_all <- fracts_out %>%
+  bind_rows(fract_all)
+
+# variance explained by axis
+
+all_varexp <- rda_eig %>%
+  filter(name != "RDA3") %>%
+  mutate(strata = "British Columbia") %>%
+  rename(axis = name) %>%
+  select(-value)
+
+varexp_all <- axis_varexp %>%
+  filter(anova_sig != "NS") %>%
+  select(strata, axis, prop, var_exp) %>%
+  bind_rows(all_varexp)
+
+for (s in stratum) {
+  print(s)
+  strata_tp <- clean_loads_cum_pos %>%
+    filter(strata == s) %>%
+    mutate(textreplace = paste(var, axis, sep = "-"),
+           axis_loading = as.character(round(axis_loading, digits = 2)))
+
+  if (strata_tp %>% pull(axis) %>% unique() %>% length() == 1) {
+    strata_drawio <- one_rda
+  } else {
+    strata_drawio <- two_rda
+  }
+
+  textreplace <- strata_tp %>%
+    pull(textreplace)
+
+  replaceval <- strata_tp %>%
+    pull(axis_loading)
+
+  for (i in 1:length(textreplace)) {
+    text <- textreplace[[i]]
+    val <- replaceval[[i]]
+
+    strata_drawio <- str_replace(strata_drawio, text, val)
+  }
+
+  strata_drawio <- str_replace(strata_drawio, "title_strata", s)
+
+  #strata_drawio <- strata_drawio[!str_detect(strata_drawio, "RDA")]
+  # filter through all possibilities remaining, replace them with NS
+
+  for (p in possibles) {
+    strata_drawio <- str_replace(strata_drawio, p, "NS")
+  }
+
+  strata_fract <- fracts_all %>%
+    filter(strata == s) %>%
+    pivot_longer(primary:overlap) %>%
+    mutate(value = (value * 100) %>%
+             round(1) %>%
+             paste0(., "%"))
+
+  fract_names <- strata_fract %>%
+    pull(name)
+
+  fract_value <- strata_fract %>%
+    pull(value)
+
+  for (i in 1:length(fract_names)) {
+    strata_drawio <-
+      str_replace(strata_drawio, fract_names[[i]], fract_value[[i]])
+  }
+
+  strata_varexp <- varexp_all %>%
+    filter(strata == s) %>%
+    mutate(var_exp = (var_exp * 100) %>%
+             round(1) %>%
+             paste0(., "%"))
+
+  axis_names <- strata_varexp %>%
+    pull(axis) %>%
+    paste("var_exp", ., sep = "-")
+
+  axis_vals <- strata_varexp %>%
+    pull(var_exp)
+
+  for (i in 1:length(axis_names)) {
+    strata_drawio <-
+      str_replace(strata_drawio, axis_names[[i]], axis_vals[[i]])
+  }
+
+  write_lines(strata_drawio, here::here(drawio_dir, paste0(s, ".drawio")))
+}
+
+# generate and run batch script to export all files to png, then copy the main
+# one to the main folder
+
+in_drawio <- list.files(here::here("outputs", "drawio"), full.names = T) %>%
+  str_subset("_rda", negate = T)
+
+out_drawio <- in_drawio %>%
+  str_replace("\\.drawio", ".png")
+
+bat_loc <- here::here("scripts", "export_drawio.bat")
+
+glue::glue('"C:/Program Files/draw.io/draw.io.exe" -x "{in_drawio}" -o "{out_drawio}"') %>%
+  write_lines(bat_loc)
+
+system2(bat_loc)
+
+file.copy(here::here("outputs", "drawio", "British Columbia.png"), here::here("outputs", "British Columbia.drawio.png"), overwrite = T)
+```
+
+Climate Table (Table 2)
+
+```{r bec-clim-tab}
+climate_folder <- here::here("data", "climate")
+clim_vars <- climate_folder %>%
+  list.files(pattern = "csv$") %>%
+  tools::file_path_sans_ext()
+
+clim_vars <- clim_vars[!str_detect(clim_vars, "bec")]
+
+bec_clim <- climate_folder %>%
+  list.files(pattern = "bec.csv$",
+             full.names = T) %>%
+  read_csv() %>%
+  mutate(var = clim_vars %>% rep(16) %>% sort) %>%
+  select(bec = ZONE, mean = MEAN, var) %>%
+  pivot_wider(values_from = mean,
+              names_from = var)
+
+bec_tab <- keys$bec %>%
+  left_join(bec_clim, by = c("zone" = "bec")) %>%
+  select(-order) %>%
+  arrange(group, zone) %>%
+  relocate(group) %>%
+  mutate(across(ppt:tmin, \(x) round(x, digits = 1))) %>%
+  left_join(strata_nested %>% unnest(data) %>% count(strata),
+            by = c("zone" = "strata"))
+
+write_csv(bec_tab, here::here("data", "climate.csv"))
+
+knitr::kable(bec_tab)
+```
+
+# rgb map
+
+```{r}
+bec_dissolve <- read_sf(here::here("data", "shapefiles", "bec_dissolve.shp")) %>%
+  janitor::clean_names() %>%
+  select(zone)
+
+# normalized to max
+loadings_rgb <- clean_loads_abs %>% 
+  group_by(var) %>%
+  mutate(axis_loading = axis_loading / max(axis_loading)) %>%
+  group_by(axis) %>%
+  pivot_wider(names_from = var,
+              values_from = axis_loading) %>%
+  ungroup() %>%
+  replace(is.na(.), 0) %>%
+  #mutate(across(elev_cv:elev_p95, \(x) ifelse(x > 0.5, 0.5, x))) %>%
+  mutate(dhi_rgb = rgb(CumDHI, VarDHI, MinDHI),
+         pri_rgb = rgb(elev_p95, percentage_first_returns_above_2m, elev_cv, maxColorValue = 1),
+         mod_rgb = rgb(basal_area, total_biomass, gross_stem_volume, maxColorValue = 1)) %>%
+  #select(strata, axis, percentage_first_returns_above_2m, elev_p95, elev_cv, ends_with("rgb")) %>%
+  pivot_longer(ends_with("rgb"))
+
+rgb_loads_split <- loadings_rgb %>%
+  filter(strata %in% zones) %>%
+  group_by(axis, name) %>%
+  group_split()
+```
+
+```{r}
+make_rgb_map <- function(tibble) {
+
+  input <- bec_dissolve %>%
+    left_join(tibble, by = c("zone" = "strata")) %>%
+    mutate(value = ifelse(is.na(value), "#000000", value))
+
+  rgb_cols = input %>%
+    st_drop_geometry() %>%
+    pull(value) %>%
+    unique() %>%
+    sort()
+
+  input %>%
+    ggplot(aes(x = zone, y = 2, fill = value)) +
+    geom_col() +
+    scale_fill_manual(values = rgb_cols)
+
+   input %>%
+    ggplot() +
+    geom_sf(aes(geometry = geometry, fill = value, col = value), linewidth = 0.1) +
+    scale_fill_manual(values = rgb_cols) +
+    scale_color_manual(values = rgb_cols) +
+    theme_void() +
+    theme(legend.position = "none")
+}
+
+rgb_maps <- map(rgb_loads_split, make_rgb_map)
+
+cowplot_rgb <- cowplot::plot_grid(rgb_maps[[3]], rgb_maps[[2]],
+                                  rgb_maps[[6]], rgb_maps[[5]],
+                                  labels = c("A - Primary RDA1", "B - Modelled RDA1",
+                                             "C - Primary RDA2", "D - Modelled RDA2"),
+                                  vjust = 1.25)
+
+ggsave(here::here("outputs", "rgbplots_cow.png"), cowplot_rgb, height = 7, width = 8, bg = "white")
+
+```
+
 # meg scatterplots
@@
 ```{r}
-scatter <- bec_normal %>%
+scatter <- raw_sliced %>%
   mutate(label = case_when(bec %in% c("CWH", "MS") ~ bec,
@@
         strip.background = element_blank())
-```
-
-Supplemental information Hexplot template
-
-```{r}
-bec_df %>%
-  left_join(fracts_out %>%
-              mutate(all = (primary + modelled + overlap) * 100) %>%
-              select(strata, all), by = c("bec" = "strata")) %>%
-  mutate(facet = paste0(bec, " (", round(all, 2), "%)")) %>%
-  ggplot(aes(x = percentage_first_returns_above_2m / 100, y = CumDHI)) +
-  geom_hex(aes(fill = after_stat(log(count)))) +
-  labs(x = "Canopy Cover (%)") +
-  facet_wrap(~fct_reorder(facet, all)) +
-  theme(legend.position = "bottom")
-
-ggsave(here::here("outputs", "supplement_hex.png"),
-       height = 6, width = 8)
 ```
@@
 ```
+
+Supplemental information Hexplot template
+
+```{r}
+bec_df %>%
+  left_join(fracts_out %>%
+              mutate(all = (primary + modelled + overlap) * 100) %>%
+              select(strata, all), by = c("bec" = "strata")) %>%
+  mutate(facet = paste0(bec, " (", round(all, 2), "%)")) %>%
+  ggplot(aes(x = percentage_first_returns_above_2m / 100, y = CumDHI)) +
+  geom_hex(aes(fill = after_stat(log(count)))) +
+  labs(x = "Canopy Cover (%)") +
+  facet_wrap(~fct_reorder(facet, all)) +
+  theme(legend.position = "bottom")
+
+ggsave(here::here("outputs", "supplement_hex.png"),
+       height = 6, width = 8)
+```
+
+Matrix of plots as requested by reviewer #4
+
+Density hexplots of forest structural attributes and the Dynamic Habitat Indices. Note the log scale.
+
+```{r}
+(hex_matrix_plot <- raw_sliced %>%
+  pivot_longer(cols = basal_area:total_biomass, names_to = "struct", values_to = "struct_vals") %>%
+  pivot_longer(ends_with("DHI"), names_to = "DHI", values_to = "dhi_vals") %>%
+  left_join(keys$continuous, by = c("struct" = "variable")) %>%
+  mutate(struct_vals = struct_vals / divide_by,
+         unit = ifelse(struct == "elev_cv", "", glue::glue("({unit})")),
+         label = glue::glue("{var_long} \n{unit}")) %>%
+  ggplot(aes(x = struct_vals, y = dhi_vals)) +
+  geom_hex(aes(fill = after_stat(log10(count)))) +
+  facet_grid(rows = vars(DHI),
+             cols = vars(label),
+             scales = "free",
+             switch = "both") +
+  labs(x = NULL,
+       y = NULL) +
+  scale_fill_continuous_c4a_seq(palette = "viridis") +
+  theme(legend.position = "bottom",
+        strip.placement = "outside",
+        strip.background = element_blank())
+
+)
+
+ggsave(here::here("outputs", "hex_matrix_plot.png"),
+       height = 6, width = 10)
+```
+
+```{r}
+# linear relationships * homoschedasticity check
+
+to_nest_normal %>%
+  pivot_longer(cols = ends_with("DHI"),
+               names_to = "dhi",
+               values_to = "dhi_val") %>%
+  pivot_longer(
+    cols = basal_area:total_biomass,
+    names_to = "struct",
+    values_to = "struct_val"
+  ) %>%
+  left_join(keys$continuous, by = c("struct" = "variable")) %>%
+  mutate(struct_vals = struct_val / divide_by,
+         unit = ifelse(struct == "elev_cv", "", glue::glue("({unit})")),
+         label = glue::glue("{var_long} \n{unit}")) %>%
+  ggplot(aes(x = struct_val, y = dhi_val)) +
+  geom_hex(aes(fill = after_stat(log10(count)))) +
+  facet_grid(rows = vars(dhi),
+             cols = vars(var_long),
+             scales = "free",
+             switch = "both") +
+  labs(x = "Structure Values (Normalized)",
+       y = "DHI Values (Normalized)") +
+  scale_fill_continuous_c4a_seq(palette = "viridis") +
+  theme(legend.position = "bottom",
+        strip.placement = "outside",
+        strip.background = element_blank())
+```

### 2024-01-24 scripts/manuscript/manuscript.qmd
third submission

@@
 productivity (represented by the Dynamic Habitat Indices, DHIs), with
-the structural information only explaining 16% of the variation in the
-DHIs. Complex metrics of forest structure, such as above ground biomass,
+the structural information only explaining 15.7% of the variation in the
+DHIs. Complex metrics of forest structure, such as aboveground biomass,
 did not contribute additional information over simpler height-based
@@
 productivity in the region [@shi2017]. Likewise, relationships between
-productivity and biodiversity also contain uncertainty, recognizing that
-biodiversity is governed by a variety of factors of which productivity
-is but one [@skidmore2021]. Despite these uncertainties, remotely sensed
-vegetation indices have enabled the assessment of biodiversity patterns
-at single time points [@bonn2004; @nagendra2010; @phillips2008], or
-through time when using time series data [@berry2007; @radeloff2019;
-@coops2019]
+productivity indicators and biodiversity also contain uncertainty,
+recognizing that biodiversity is governed by a variety of factors of
+which productivity is but one [@skidmore2021]. Despite these
+uncertainties, remotely sensed vegetation indices have enabled the
+assessment of biodiversity patterns at single time points [@bonn2004;
+@nagendra2010; @phillips2008], or through time when using time series
+data [@berry2007; @radeloff2019; @coops2019]
 
@@
   ) %>%
-  tab_header(title = "Biogeoclimatic Ecosystem Classification (BEC) Zones, their aggregated biomes, and their monthly average climate values for precipitation, maximum temperature, and minimum temperature.
+  tab_header(title = "Biogeoclimatic Ecosystem Classification (BEC) Zones, their aggregated biomes, monthly average climate values for precipitation, maximum temperature, and minimum temperature, and the number of samples taken from each BEC zone.
 Climate data from Wang et al.
@@
 
-### National Terrestrial Ecosystem Monitoring System data products
+### Forest Structure
+
+The forest structural attributes (canopy height, canopy cover,
+structural complexity \[coefficient of variation of height returns\],
+basal area, aboveground biomass, and gross stem volume) were retrieved
+for 2015 from the 30-m spatial resolution dataset generated by Matasci
+et al. [@matasci2018; @matasci2018b]. Their method extends structural
+attributes derived from a set of lidar acquisitions and field plots
+across Canada. These data were generated for the treed pixels of Canada
+using the BAP Landsat surface reflectance composite [@hermosilla2016]
+and auxiliary data (i.e., topography, geography) using a k-Nearest
+Neighbour imputation [@matasci2018; @matasci2018b]. Accuracy metrics for
+imputed forest structural attributes range from an RMSE of 29.7%
+(structural complexity) to 82.3% (gross stem volume) and *R^2^* ranging
+from 0.125 (structural complexity) to 0.712 (gross stem volume)
+[@matasci2018; @matasci2018b].
+
+### Dynamic Habitat Indices
+
+The DHIs for the terrestrial area of BC were calculated following the
+procedures described in Razenkova et al
+@razenkovaMediumresolutionDynamicHabitatInPress. Although DHIs are
+typically derived from high-temporal, coarse spatial resolution imagery
+such as MODIS, we used Landsat surface reflectance data due to its finer
+spatial resolution, which corresponds to the 30 m ALS derived forest
+structure data. To account for the lower temporal resolution of Landsat,
+we generated a synthetic year of monthly data from a ten-year timespan
+of Landsat imagery. Google Earth Engine (GEE) [@gorelick2017] was used
+to obtain all valid Landsat pixels for the study area over a ten-year
+period centered on 2015. Selected pixels were filtered for shadows,
+clouds, and cloud shadows using the QA band derived from the fmask
+algorithm [@zhu2012], available on GEE. The NDVI was then calculated for
+each pixel in each image and summarized into a synthetic year of monthly
+data by calculating the median of each month's NDVI values, disregarding
+the year of the image. The calculations for the Cumulative DHI, Minimum
+DHI, and Variation DHI can be found in Table 2, alongside a summary of
+the main datasets analyzed in this study.
+
+### Ancillary Data
 
@@
 Hermosilla et al. @hermosilla2016 using the Composite2Change approach.
-The composites were developed for each year from 1984-2019 by selecting
-the best available imagery from the available growing season Landsat
-observations, removing pixels with clouds, cloud shadows, and haze. The
-methodology uses the scoring method from White et al. @white2014 to
-select the best pixel for a given year from Landsat-5 Thematic Mapper,
-Landsat-7 Enhanced Thematic Mapper Plus, and Landsat-8 Operational Land
-Imager imagery. A spectral trend analysis was conducted over these
-initial BAP composites using the Normalized Burn Ratio (a vegetation
-index) on each pixel, to remove unscreened noise, detect changes, and
-temporally interpolate data gaps. This resulted in a gap-free
-surface-reflectance composite across Canada [@hermosilla2015]. We use
-data from 2015 for all National Terrestrial Ecosystem Monitoring System
-products, except forest disturbances, which comprise the period
+The composites were developed for each year from 1984 to 2019 by
+selecting the best available imagery from the available growing season
+Landsat observations, removing pixels with clouds, cloud shadows, and
+haze. The methodology uses the scoring method from White et al.
+@white2014 to select the best pixel for a given year from Landsat-5
+Thematic Mapper, Landsat-7 Enhanced Thematic Mapper Plus, and Landsat-8
+Operational Land Imager imagery. A spectral trend analysis was conducted
+over these initial BAP composites using the Normalized Burn Ratio (a
+vegetation index) on each pixel, to remove unscreened noise, detect
+changes, and temporally interpolate data gaps. This resulted in a
+gap-free surface-reflectance composite across Canada [@hermosilla2015].
+The forest structure and land cover datasets are based on BAP data from
+2015, while the forest disturbance mask covers the period from
 1985-2020.
@@
 
-#### Forest Structure
-
-The forest structural attributes (canopy height, canopy cover,
-structural complexity \[coefficient of variation of height returns\],
-basal area, aboveground biomass, and gross stem volume) were derived
-from the 30-m spatial resolution dataset generated by Matasci et al.
-[@matasci2018; @matasci2018b]. In brief, the method extends structural
-attributes derived from a set of lidar acquisitions and field plots
-across Canada. These data were generated for the treed pixels of Canada
-using the BAP Landsat surface reflectance composite and auxiliary data
-(i.e., topography, geography) using a k-Nearest Neighbour imputation
-[@matasci2018; @matasci2018b]. Accuracy metrics for imputed forest
-structural attributes range from an RMSE of 29.7% (structural
-complexity) to 82.3% (gross stem volume) and *R^2^* ranging from 0.125
-(structural complexity) to 0.712 (gross stem volume) [@matasci2018;
-@matasci2018b].
-
-### Dynamic Habitat Indices
-
-The DHIs were calculated following Razenkova et al.
-@razenkovaMediumresolutionDynamicHabitatInPress for the terrestrial area
-of BC. In brief, we used Google Earth Engine [@gorelick2017] to obtain
-valid Landsat pixels for the study area across a ten-year time span,
-centered on 2015 (2011-2020), filtering out pixels containing shadows,
-clouds, and cloud shadows using the QA band derived from the fmask
-algorithm [@zhu2012], and then calculated the NDVI index. These
-estimates were composited into a synthetic year of monthly data by
-calculating the monthly median NDVI value. Finally, images of the sum,
-minimum, and coefficient of variation across this synthetic year of NDVI
-values were then calculated to represent the cumulative, minimum, and
-variation DHIs, respectively.
-
-A summary of the datasets used in this study is presented in Table 2.
-
 ## Sampling
@@
 were restricted to forested pixels that were surrounded by the same
-forest type to reduce the chance of sampling mixed pixels. We also
-required that sample units were selected from a homogeneous area, to
-reduce uncertainty associated with the input datasets [@shang2020];
-homogeneous areas were defined as 3x3-pixel windows with a coefficient
-of variation for canopy cover and canopy height below 0.5. A minimum
-sampling distance of 1-km was implemented to reduce the effects of
-spatial autocorrelation. Sample units that had been disturbed in the
-last 35 years were discarded by using the disturbance mask generated for
-the forested ecosystems of Canada [@hermosilla2016]. A maximum of 500
-sample units were selected from each stratum; in the strata where this
-was not achieved, all sampling units meeting the above restrictions were
-selected. Sampling was conducted in R version 4.2.2 [@R-base] using the
-**sgsR** package [@R-sgsR]. Neighbourhood analyses for the land cover
-classes and coefficient of variations of canopy height and cover were
-calculated in Python version 3.9.
+forest type to reduce the chance of sampling mixed pixels. Sample units
+were selected from homogeneous areas to reduce uncertainty associated
+with the input datasets [@shang2020]. Homogeneous areas were defined as
+3x3-pixel windows with a coefficient of variation for canopy cover and
+canopy height below 0.5. A minimum sampling distance of 1 km was
+implemented to reduce the effects of spatial autocorrelation. Sample
+units that were disturbed in the last 35 years were excluded by using
+the disturbance mask generated for the forested ecosystems of Canada
+[@hermosilla2016]. A maximum of 500 sample units were selected from each
+stratum. In the strata where this was not achieved, all sampling units
+meeting the above restrictions were selected (see Table 1). The sampling
+was conducted in R version 4.2.2 [@R-base] using the **sgsR** package
+[@R-sgsR]. Neighbourhood analyses for the land cover classes and
+coefficient of variations of canopy height and cover were calculated in
+Python version 3.9.
 
@@
 add paths to the Venn diagram to indicate how the structural attributes
-contribute to the variation in the DHIs. Because the variation
-partitioning results summarize the full multivariate solution, rather
-than individual RDA axes, we present variation partitioning results
-alongside the path diagrams illustrating the details of the RDA results.
-RDA and variation partitioning calculations were done in R [@R-base]
-version 4.2.2 using the **vegan** package [@R-vegan]. The code
-associated with the processing and analysis is available at
+contribute the variation in the DHIs. Because the variation partitioning
+results summarize the full multivariate solution, rather than individual
+RDA axes, we present variation partitioning results alongside the path
+diagrams illustrating the details of the RDA results. RDA and variation
+partitioning calculations were done in R [@R-base] version 4.2.2 using
+the **vegan** package [@R-vegan]. The code associated with the
+processing and analysis is available at
 https://github.com/emuise/code-structProdSem.
+
+We present a simplified flow diagram of the data provenance to result
+pipeline in @fig-flow.
+
+```{r}
+#| label: fig-flow
+#| fig-cap: Flow diagram of the data to indicators to results pipeline implemented in this study.
+
+knitr::include_graphics(here::here("outputs", "p2_flow.drawio.png"))
+```
 
@@
 axis as representing overall productivity. The overall productivity axis
-had similar loadings (between 0.24-0.27) for all modelled forest
+had similar loadings (between 0.26-0.28) for all modelled forest
 structural attributes and canopy cover. Canopy height had a smaller
-loading (0.1), while structural complexity was negatively correlated
+loading (0.13), while structural complexity was negatively correlated
 with the overall productivity axis. RDA2 explained much less of the
-variation in the DHIs. RDA2 had a large positive loading on the minimum
-DHI (0.55), with negative loadings on the cumulative (-0.19) and
-variation (-0.39) DHIs. Due to the minimum and variation having the most
+variation in the DHIs. RDA2 had a large negative loading on the minimum
+DHI (-0.47), with smaller, positive loadings on the cumulative (0.26)
+and variation (0.29) DHIs. Due to the minimum DHI having the largest
 influence on RDA2, we identified it as a seasonality axis. Canopy cover
-and structural complexity had the strongest (albeit negative; -0.17 and
--0.08, respectively) loadings on the seasonality axis (@fig-rda-var A).
+and structural complexity had the strongest (0.18 and 0.08,
+respectively) loadings on the seasonality axis (@fig-rda-var A).
 Variation partitioning highlighted that the majority of the explained
 variation in the DHIs was due to the primary forest structural
-attributes (canopy cover, canopy height, and structural complexity; 9%
+attributes (canopy cover, canopy height, and structural complexity; 7.7%
 variation explained). The overlap between primary and modelled
-attributes was 3.5%, and the modelled attributes explained 3.2% of the
+attributes was 4.2%, and the modelled attributes explained 3.4% of the
 variation on their own (@fig-rda-var B).
@@
 (@fig-fracts-bar). By BEC zone, results generally indicated smaller
-amounts of variation explained, which may be due to the relatively short
+amounts of variation explained, which may be due to the relatively small
 environmental gradients within the zones. The overlap between the
@@
 the variation explained in the DHIs, except when stratifying by forest
-types, where this was only true for the coniferous forests. For
-broadleaf, mixed wood, and wetland-treed forest types most of the
-explained variation in the DHIs was contributed by the primary forest
-structural attributes. A similar pattern was found by BEC zones, where
-the variation in the DHIs was principally explained by the overlap
-between primary and modelled forest structural attributes, except for
-CWH, where no variation was explained by the overlap. The Coastal BEC
-group had the lowest amount of variation explained, with Coastal
-Douglas-fir having \<1% of variation explained by the structural data.
-Overall, the DHIs were decoupled from the forest structural attributes,
-as shown by the overall variation explained being under 30% regardless
-of data stratification.
+types, where this was only true for coniferous forests. For broadleaf,
+mixed wood, and wetland-treed forest types most of the explained
+variation in the DHIs was contributed by the primary forest structural
+attributes. A similar pattern was found by BEC zones, where the
+variation in the DHIs was principally explained by the overlap between
+primary and modelled forest structural attributes, except for CWH, where
+the variation was entirely explained by the primary forest structural
+attributes. The Coastal BEC group had the lowest amount of variation
+explained, with Coastal Douglas-fir having \<1% of variation explained
+by the structural data. Overall, the DHIs were decoupled from the forest
+structural attributes, as shown by the overall variation explained being
+under 32% regardless of data stratification.
 
@@
 the other DHI loadings. As such, we identify the first axis as being
-consistently associated with overall productivity. Canopy cover
-generally had the highest significant predictor loading on RDA1, except
-for in non-coniferous forest types (broadleaf, mixed wood and
-wetland-treed). These forest types typically had large negative loadings
-between RDA1 and structural complexity. The mixed wood forest and
-Coastal BEC group had no significant loadings from modelled forest
-structural attributes for RDA1. Examining the predictor loadings for
-RDA1 spatially across the province shows a geographic pattern in the
-loading strengths of primary forest structural attributes (@fig-fcc A).
-The interior (Northern and Southern BEC groups) of the province
-generally showed high canopy cover loadings, while the Boreal zones in
-the northwest had equal loadings in canopy cover and canopy height.
-Coastal zones were driven similarly by canopy height and structural
-complexity, with very low or non-significant loadings from canopy cover
-(@fig-fcc A; @fig-radar). Modelled forest structural attributes for RDA1
-generally show similar loadings across the province (@fig-fcc B).
+consistently associated with overall productivity. The predictor
+loadings for canopy cover and the modelled forest structural attributes
+were generally similar within the same stratification/BEC groups. The
+mixed wood forest and Coastal BEC group had no significant loadings from
+modelled forest structural attributes for RDA1. Examining the predictor
+loadings for RDA1 spatially across the province shows a geographic
+pattern in the loading strengths of primary forest structural attributes
+(@fig-fcc A). The interior (Northern and Southern BEC groups) of the
+province generally showed high canopy cover loadings, while the Boreal
+zones (Boreal White and Black Spruce and Sub-Boreal Pine – Spruce) in
+the northwest had large canopy height loadings. Coastal zones were
+driven similarly by canopy height and structural complexity, with very
+low or non-significant loadings from canopy cover (@fig-fcc A;
+@fig-radar).
 
@@
 
-Only seven of sixteen BEC zones had a second RDA axis, with none being
-found in Coastal or Alpine BEC zones (@fig-radar; @fig-fcc). All of the
-Montane BEC zones had a secondary axis, while some in the Northern and
-Southern interior had a secondary axis. RDA2 often showed the largest
-absolute loadings in the minimum DHI, which was frequently, but not
-always, negative (@fig-radar; see supplementary material). The Montane
-group, which had the largest absolute minimum DHI loading, also had the
-largest absolute variation DHI loading. The secondary axis had smaller
-loadings in the variation and cumulative DHIs, with only the Southern
-BEC group having a lower variation DHI loading than cumulative DHI. RDA2
-has lower absolute predictor loadings (maximum of \~0.25) when compared
-to RDA1 (maximum of \~0.5), with infrequently significant loadings for
-canopy cover and structural complexity (@fig-radar). Among the primary
-forest structural attributes, RDA2 had the highest loadings with canopy
-height in the Southern group, while the Northern group had slightly
-larger structural complexity loadings (@fig-radar; @fig-fcc C). Similar
-to RDA1, the modelled forest structural attribute loadings were
-generally similar to one another (@fig-radar; @fig-fcc D).
+Only four of sixteen BEC zones had a second RDA axis, with none being
+found in Coastal or Alpine BEC zones (@fig-radar; @fig-fcc). Two Montane
+BEC zones had a secondary axis, while the Northern and Southern interior
+groups each had one zone with a secondary axis. RDA2 often showed the
+largest absolute loadings in the minimum DHI, which was frequently, but
+not always, negative (@fig-radar; see supplementary material). The
+Montane group had the largest absolute minimum and variation DHI
+loadings, which were notably the opposite sign of all other groups with
+a secondary axis. RDA2 had lower absolute predictor loadings (maximum of
+\~0.25) when compared to RDA1 (maximum of \~0.5), with infrequently
+significant loadings for canopy cover and structural complexity
+(@fig-radar). There was no consistent loading pattern found in RDA2 for
+the primary predictor variables (@fig-radar; @fig-fcc C). Similar to
+RDA1, the modelled forest structural attribute loadings were generally
+similar to one another (@fig-radar; @fig-fcc D).
 
@@
 
-Relationships between forest ecosystem structure and function are
-increasingly being examined due to advances in data collection methods
-[@ali2019; @radeloff2019; @ali2016]. Prior studies have shown both
-theoretical [@knyazikhin1998] and empirical [@atkins2018] relationships
-between forest ecosystem structure and function, however, the direction
-and significance of these relationships have not been consistent
-[@ali2019]. We examined these relationships across the province of BC,
-Canada, which has significant environmental gradients
-[@valentine1978soil; @pojar1987] by using wall-to-wall forest structural
-attributes generated by imputing the lidar-derived attributes with
-Landsat best-available-pixel composites [@matasci2018; @matasci2018b]
-and Landsat-derived summaries of intra-annual vegetation productivity
-[@radeloff2019; @razenkovaMediumresolutionDynamicHabitatInPress]. We
-found a weak relationship between these remote sensing indicators of
-forest ecosystem structure and function overall across BC (15.7% of
-variation explained in the DHIs by forest structural attributes), as
-well as within individual BEC zones and forest types found in BC
-(@fig-fracts-bar).
+Relationships between forest ecosystem structure and function are being
+examined more frequently due to advances in data collection methods,
+such as remote sensing [@ali2019; @radeloff2019; @ali2016]. This paper
+assesses the complementarity of two potential ecosystem-scale EBV
+indicators for ecosystem monitoring across BC, Canada. Redundancy
+analysis was used to determine the information overlap of wall-to-wall
+maps of forest structural attributes and the DHIs, which were both
+generated from Landsat imagery. This analysis revealed a weak
+relationship between the two sets of remote sensing derived indicators,
+with only 15.7% of the variation in the DHIs being explained by the
+forest structural attributes (@fig-rda-var). This relationship was also
+assessed across the different forest classes and ecosystems of British
+Columbia, and similar results were found (@fig-fracts-bar). The small
+amount of overlap between the two datasets suggests that they are
+suitable for use in ecosystem monitoring programs, despite both being
+derived from Landsat imagery.
+
+We also evaluated whether primary forest structural attributes provide
+more information than modelled forest structural attributes for
+selecting indicators within a single EBV class. It was found that
+standard lidar-derived measurements, such as canopy height, canopy
+cover, and structural complexity (as proposed by Valbuena et al.
+[@valbuena2020], contribute the majority of the explained variation in
+the DHIs, either independently or in their overlap with the modelled
+forest structural attributes. This could be expected as modelled forest
+structural attributes are generally directly based on the primary forest
+structural attributes [@coops2021], however, the creation of modelled
+forest structural attributes involves modelling expertise and processing
+time overhead [@matasci2018; @matasci2018b; @duncanson2022]. One
+advantage of interpreted products is their potential uptake by a broader
+set of users and their ability to be closely linked with management
+actions and monitoring results. This is exemplified by the use of forest
+biomass monitoring for climate change mitigation [@duncanson2022]. While
+either set of forest structural attributes may be relevant for a given
+monitoring task, it is important to note that the primary structural
+attributes contain the majority of the variation across both classes of
+structural attributes (@fig-fracts-bar). Additionally, the modelled
+structural attributes are generally not related to structural
+complexity, which has often been found to be strongly related to
+biodiversity metrics [@ehbrecht2021]. Therefore, we generally recommend
+using the primary forest structural attributes.
 
 Our findings indicate that remote sensing-derived forest ecosystem
-structure and forest ecosystem function EBVs are well suited to be used
-in a complementary fashion for ecosystem integrity monitoring, despite
-both EBVs being derived from surface reflectance from the Landsat series
-of satellites [@hansen2021]. While we assessed the complementarity of
+structure and forest ecosystem function EBVs can be applied in a
+complementary fashion for ecosystem monitoring [@hansen2021], despite
+both EBVs using surface reflectance values from the Landsat series of
+satellites in either their calculation or imputation [@radeloff2019;
+@matasci2018; @matasci2018b]. While we assessed the complementarity of
 ecosystem structure and function, a single pair of EBVs measurable at
 the ecosystem scale, other EBV class combinations at finer scales should
-also be assessed for complementarity. This is especially relevant in
-cases such as ours, where remote sensing datasets such as Landsat can
-generate a large number of EBVs in different classes [@skidmore2021]. In
-addition, Skidmore et al. [@skidmore2021] highlighted that many
-satellite-derived biodiversity products can fall into multiple EBV
-classes simultaneously -- e.g., leaf area index being in the ecosystem
-structure, ecosystem function, and species trait EBV classes --
-depending on the spatial scale of analysis and goal of the monitoring
-program. As such, it is relevant to consider which classes each
-biodiversity product should belong to, for given spatial and temporal
-extents. Other EBVs, such as those found in the community composition
-and species population classes, may also be inherently linked due to
-being based on the same data sources, even if not observed using remote
-sensing platforms.
-
-Further, we assessed the information-richness of primary forest
-structural attributes when compared to modelled forest structural
-attributes to support indicator selection within a single EBV class. We
-found that standard lidar-derived measurements, such as canopy height,
-canopy cover, and structural complexity (as proposed by Valbuena et al.
-[@valbuena2020]) contribute to the majority of the explained variation
-in the DHIs, either independently or in their overlap with the modelled
-forest structural attributes (@fig-fracts-bar). This could be expected
-as modelled forest structural attributes are generally directly based on
-the primary forest structural attributes [@coops2021], however, the
-creation of modelled forest structural attributes involves modelling
-expertise and processing time overhead [@matasci2018; @matasci2018b;
-@duncanson2022]. One advantage of these interpreted products is
-potential uptake by a broader set of users, and their ability to be more
-closely linked to management actions and monitoring results, as is the
-case with forest biomass for climate change mitigation [@duncanson2022].
-While either set of forest structural attributes may be most relevant
-for a given monitoring task, we note that the modelled structural
-attributes do not strongly overlap with structural complexity, and as
-such, we generally recommend to use the primary forest structural
-attributes.
+also be assessed for complementarity. This is especially relevant when
+remote sensing datasets such as Landsat can generate a large number of
+EBVs in different classes [@skidmore2021]. In addition, Skidmore et al.
+[@skidmore2021] highlighted that many satellite-derived biodiversity
+products can inform upon multiple EBV classes simultaneously -- e.g.,
+leaf area index being recognized as a ecosystem structure, ecosystem
+function, and species trait EBV indicator -- depending on the spatial
+scale of analysis and goal of the monitoring program. As such, it is
+relevant to consider which EBV classes biodiversity products can inform,
+for given spatial and temporal extents. Other EBVs, such as those found
+in the community composition and species population classes may also be
+informed by these products. Further, their reliance on species level
+data also means that they could be inherently correlated with one
+another.
 
@@
 large, negative structural complexity loadings when compared to the
-other forest structural attributes (@fig-radar), with mixed wood forests
+other forest structural attributes (@fig-radar), with forest types
 having no modelled structural attributes associated with the DHIs. This
@@
 relationships between canopy height and structural complexity being
-found [@atkins2022].
+found than deciduous forests [@atkins2022].
 
@@
 ecoregionalization potentially revealing different relationships between
-ecosystem structure and function.
+ecosystem structure and function. Further, when examining these
+relationships through remote-sensing derived indicators, we recognize
+that they may change over time and space in response to disturbances,
+both anthropogenic and natural. For example, modifying the biomass of an
+area through forest clearing, fires, or insect disturbances may disrupt
+the underlying assumptions of the DHIs, as abrupt changes in the middle
+of the year may lead to low Minimum DHI values, high Variation DHI
+values, and lower than normal Cumulative DHI values.
 
@@
 while tropical forests often have positive relationships [@ali2019]. We
-found that forest structure and function were not strongly linked to one
-another, both across the full study area, and when stratified by
-ecosystem and forest type (@fig-radar). Forest type models showed higher
-amounts of variation explained, which may indicate the linkages between
-forest structure and function are stronger at finer spatial scales,
-potentially driven by finer scale landscape patterns such as terrain and
-land cover, rather than climate [@thom2021]. Notably, the BEC zones and
-forest types are delineated at different spatial scales; forest types
-are mapped at a 30 m pixel size [@hermosilla2022; @hermosilla2018], and
-can be found across the entire province, while BEC zones do not span the
-entirety of the province, and are mapped at a much coarser scale
-[@pojar1987].
+found that indicators of forest structure and function were not strongly
+linked to one another, both across the full study area and when
+stratified by ecosystem and forest type (@fig-radar). The forest type
+models showed higher amounts of variation explained, which may indicate
+the linkages between forest structure and function are stronger at finer
+spatial scales, potentially driven by finer scale landscape patterns
+such as terrain and land cover, rather than climate [@thom2021].
+Notably, the BEC zones and forest types are delineated at different
+spatial scales; forest types are mapped at a 30 m pixel size
+[@hermosilla2022; @hermosilla2018], and can be found across the entire
+province, while BEC zones do not span the entirety of the province, and
+are mapped at a much coarser scale [@pojar1987].
 
@@
 Figure 1: Location of Biogeoclimatic Ecosystem Classification (BEC)
-zones in British Columbia.
-
-Figure 2: A) Axis loadings from redundancy analysis (RDA) of primary and
+zones in British Columbia. Map was generated in R version 4.3.1 using
+the ggplot2 version 2.1.0 and sf version 1.0-14 packages.
+
+Figure 2: Flow diagram of the data to indicators to results pipeline
+implemented in this study.
+
+Figure 3: A) Axis loadings from redundancy analysis (RDA) of primary and
 modelled forest structure variables on the dynamic habitat indices
@@
 
-Figure 3: Stacked bar plot showing the proportion of the variation in
+Figure 4: Stacked bar plot showing the proportion of the variation in
 the Dynamic Habitat Indices (DHIs) explained by primary, modelled, and
@@
 
-Figure 4: Parallel coordinate plots of average significant loading
+Figure 5: Parallel coordinate plots of average significant loading
 strength by Biogeoclimatic Ecosystem Classification (BEC) group, forest
@@
 
-Figure 5: False colour maps by Biogeoclimatic Ecosystem Classification
+Figure 6: False colour maps by Biogeoclimatic Ecosystem Classification
 (BEC) zone of axis loadings for the first redundancy analysis (RDA) axis
@@
 indicate similar loadings across the three variables. Annotated colour
-wheels for each set of variables also shown.
+wheels for each set of variables also shown. Maps were generated in R
+version 4.3.1 using the ggplot2 version 2.1.0 and sf version 1.0-14
+packages.
